### 2023-05-30 Manuscript_Nilsen_Nater.qmd
Version 1 of ms - draft

---
title: "An open population age structured distance sampling model"
format: docx
author: Erlend B. Nilsen & Chloé Nater
editor: visual
---

## Introduction

Estimating abundance and demographic rates for wildlife populations is a key element of population ecology, demography, eco-evolution research, conservation and wildlife management (refs). Tremendous progress during the last few decades both in terms of new technologies that allow new types of field data being collected (e.g. eDNA ref; camera traps; refs) and the statistical machinery available to interpret the ecological and demographic signals in the data (refs). Nevertheless, large scale and long term data sets are still relatively rare (ref), due to high costs of field data collection.

A cost efficient and well documented approach suitable is the use of line transect surveys. However, because the detectability is often less than 1 and variable between years and habitats,

-   Importance of abundance trends and demography

-   Short about methods -- detection probability -- marked vs unmarked animals

-   Approaches to estimate abundance and demography from unmarked animals

-   Distance sampling -- such methods are still lacking/not well utilized

-   Integrated pop. Model;

Here, we will present a novel integrated population model based on distance sampling line transect survey data and demographic data from marked birds.

-   Estimate parameters with and without augmentation with demographic data -- compare estimates and precision
-   Estimate latent variable (juv summer survival) -- compare to literature values
-   Temporal variation in S and R -- via random effects models
-   Simulate - robustness

## Methods

### Field data

Two different sources of data were used in the open population demographic distance sampling (ODDS) model: i) data from a line transect survey program targeting willow ptarmigan operated as part of the natural resources management authorities, and ii) data from an individual based research project based on radio collared willow ptarmigan. The temporal coverage of the two data sets partially overlap, with the line transect survey spanning 2007-2019 (and ongoing) and the individual based data spanning 2015-2019 (and ongoing).

Line transect survey data were collected in August each year, prior to the annual autumn harvest season. Line transects are surveyed by educated volunteers, using pointing dogs to located the birds. When located, the geographical coordinate, perpendicular distance from the sampling line, the number of birds as well as the age (juvenile and adult) and sex of the birds are recorded. Additional information about the surveys are found in Nilsen *et al.* (2020) and Bowler *et al.* (2020). \[Add sample sizes\].

The individual longitudinal study based on radio collared willow ptarmigan was conducted in 2015-2020. Each winter (in February-March), willow ptarmigan were located at night using snowmobiles and large hand nets with prolonged handles, as described in Israelsen et al. (2020). To prevent birds from flying off before the field personnel were close enough to capture them, a high-powered head lamp was used to dazzle the birds. After capture, birds were placed in an opaque bag to reduce stress. Before releasing the birds, they were fitted with a uniquely numbered leg ring (\~ 2.4g) and a Holohil RI-2BM or Holohil RI-2DM radio transmitter (\~ 14.1g). The radio transmitters had an expected battery lifetime of 24 months (RI-2BM) or 30 months (RI-2DM), and included a mortality circuit that was activated if a bird had been immobile for 12 hours. We monitored the birds throughout the year, by triangulation from the ground at least once a month for 10 months of the year (February -- November) by qualified field personnel. If a mortality signal was heard from the transmitter, we recovered it as soon as possible to determine cause of death. A number of birds dispersed out of the main study areas and was thus out of signal range for field personnel on the ground. To avoid loss of data, we conducted aerial triangulation using a helicopter or airplane three times a year (May, September and November) in the years 2016-2020.

### The open population demographic distance sampling model

Distance sampling models have been applied to a wide range of contexts, and has in recent years been extended to allow a direct modelling of the population process by formulating the state process as a open population process (ref). Here, we extend this framework by directly modelling the underlying state dynamics in terms of annual survival and recruitment rates.

In line with standard age- or stage structured modelling of wildlife populations, we first defined the life cycle for the focal species -- the willow ptarmigan. Based on previous models for willow ptarmigan (ref), and based on the age-structure of the data, the model included two age classes: juveniles and adults (Figure 1).

**Figure 1.** Life cycle graph

We assumed that the population size and structured was surveyed each year at the end of the summer season in August, equivalent to the schedule of the real survey data. Ptarmigan breed annually, and the clutch hatch around last week of June. By the time of survey, they are around 1.5 months of age during the annual line transect surveys in August. Ptarmigan reach maturity at the age of one (ref). The life cycle is therefore described in term of i) annual survival probability (S~t~) which describe the probability to survive from August in year *t* to August in year *t+1*. Here, we assume no age-or sex-specificity in the survival rates, acknowledging that this is a simplifying assumption that could be alleviated with additional auxiliary data (Sandercock *et al.* 2011; Israelsen *et al.* 2020). Moreover, following the standard for post-breeding census models year specific recruitment (R~t~) will be given by the product of annual survival (S~t~) and fecundity (F~t~):

(equation 1)

Based on the life cycle and the timing of the annual surveys, the transition from August year *t* to year *t+1* is given by the year-specific transition matrix - defined as:

### 2023-06-01 Manuscript_Nilsen_Nater.qmd
Embedded figures in manuscript
For now, I've just added all figures that may be relevant. Some may be moved to supplementary. Figures from simulated data tests still to follow. 

Note that I have not properly set up the referencing yet. We should also look into automatic numbering of figures.

@@
 
-### Field data
+### The open population demographic distance sampling model
+
+Distance sampling models have been applied to a wide range of contexts, and has in recent years been extended to allow a direct modelling of the population process by formulating the state process as a open population process (ref).
+Here, we extend this framework by directly modelling the underlying state dynamics in terms of annual survival and recruitment rates.
+
+In line with standard age- or stage structured modelling of wildlife populations, we first defined the life cycle for the focal species -- the willow ptarmigan.
+Based on previous models for willow ptarmigan (ref), and based on the age-structure of the data, the model included two age classes: juveniles and adults (Figure 1).
+
+![**Figure 1.** Simplified graphical representation of the ptarmigan life cycle with two age classes and the data sources included in the integrated distance sampling model.
+Juv\[t\] = juveniles in year t.
+Ad\[t\] = adults in year t.
+R\[t\] = recruitment rate in year t.
+S\[t\] = survival probability from year t to t+1.](Figures/RypeIDSM_structure.png)
+
+We assumed that the population size and structured was surveyed each year at the end of the summer season in August, equivalent to the schedule of the real survey data.
+Ptarmigan breed annually, and the clutch hatch around last week of June.
+By the time of survey, they are around 1.5 months of age during the annual line transect surveys in August.
+Ptarmigan reach maturity at the age of one (ref).
+The life cycle is therefore described in term of i) annual survival probability (S~t~) which describe the probability to survive from August in year *t* to August in year *t+1*.
+Here, we assume no age-or sex-specificity in the survival rates, acknowledging that this is a simplifying assumption that could be alleviated with additional auxiliary data (Sandercock *et al.* 2011; Israelsen *et al.* 2020).
+Moreover, following the standard for post-breeding census models year specific recruitment (R~t~) will be given by the product of annual survival (S~t~) and fecundity (F~t~).
+
+Based on the life cycle and the timing of the annual surveys, the transition from August year *t* to year *t+1* is given by the year-specific transition matrix - defined as:
+
+### Field data (case study)
 
@@
 Additional information about the surveys are found in Nilsen *et al.* (2020) and Bowler *et al.* (2020).
-\[Add sample sizes\].
+\[Add sample sizes and additional information on data processing/storage post collection, i.e.
+how do volunteers report data, storage in hønsefuglportalen, quality control, pushes to LivingNorway and GBIF\].
 
@@
 
-### The open population demographic distance sampling model
+\[Figure 2: Map of Lierne study area?\]
 
-Distance sampling models have been applied to a wide range of contexts, and has in recent years been extended to allow a direct modelling of the population process by formulating the state process as a open population process (ref).
-Here, we extend this framework by directly modelling the underlying state dynamics in terms of annual survival and recruitment rates.
+## Results
 
-In line with standard age- or stage structured modelling of wildlife populations, we first defined the life cycle for the focal species -- the willow ptarmigan.
-Based on previous models for willow ptarmigan (ref), and based on the age-structure of the data, the model included two age classes: juveniles and adults (Figure 1).
+### Model performance on simulated datasets
 
-**Figure 1.** Life cycle graph
+\[One or several figures of estimates vs. true values?\]
 
-We assumed that the population size and structured was surveyed each year at the end of the summer season in August, equivalent to the schedule of the real survey data.
-Ptarmigan breed annually, and the clutch hatch around last week of June.
-By the time of survey, they are around 1.5 months of age during the annual line transect surveys in August.
-Ptarmigan reach maturity at the age of one (ref).
-The life cycle is therefore described in term of i) annual survival probability (S~t~) which describe the probability to survive from August in year *t* to August in year *t+1*.
-Here, we assume no age-or sex-specificity in the survival rates, acknowledging that this is a simplifying assumption that could be alleviated with additional auxiliary data (Sandercock *et al.* 2011; Israelsen *et al.* 2020).
-Moreover, following the standard for post-breeding census models year specific recruitment (R~t~) will be given by the product of annual survival (S~t~) and fecundity (F~t~):
+### Case study on willow ptarmigans in Lierne
 
-(equation 1)
+#### Population density
 
-Based on the life cycle and the timing of the annual surveys, the transition from August year *t* to year *t+1* is given by the year-specific transition matrix - defined as:
+![**Figure X.** Estimated density of willow ptarmigans in Lierne from 2007 to 2021. Solid line represents the posterior median, ribbon marks 95% credible interval.](Figures/TimeSeries_popDens.png)
+
+#### Vital rates
+
+![**Figure X.** Posterior densities of A) seasonal survival, B) average annual survival, and C) recruitment rate.
+For the latter, the yellow distribution is for the intercept, representing a baseline recruitment rate when rodent occupancy is low.
+The turquoise distributions are for year-specific estimates or recruitment rate, with darker colors indicating later years.](Figures/PostDens_Mu_S_tR.png)
+
+![**Figure X.** Estimated recruitment rate of willow ptarmigans in Lierne from 2007 to 2021.
+Solid line represents the posterior median, ribbon marks the 95% credible interval.](Figures/TimeSeries_rRep.png)
+
+![**Figure X.** Predicted effect of rodent occupancy on recruitment rate. Solid line represent the posterior median, ribbon marks the 95% credible interval.](Figures/Rep_betaR.R.png)
+
+#### Detection parameters
+
+![**Figure X.** Predicted average detection probabilities for distance sampling surveys in Lierne from 2007 to 2021.
+Solid line represents the posterior median, ribbon marks the 95% credible interval.](Figures/TimeSeries_pDetect.png)
+
+## Discussion
+
+### Conclusion
+
+### Author contributions
+
+### Acknowledgements
+
+### Data and code availability

### 2023-06-02 Manuscript_Nilsen_Nater.qmd
HTML render

@@
 ---
-title: "An open population age structured distance sampling model"
-format: docx
-author: Erlend B. Nilsen & Chloé Nater
+title: "An integrated distance sampling approach for modelling age-structured populations"
+format: html
+author: Erlend B. Nilsen & Chloé R. Nater
 editor: visual
+bibliography: rypeIDSM_MS1.bib
 ---
@@
 
-### The open population demographic distance sampling model
+### An integrated age-structured distance sampling model
+
+Distance sampling models have been used for estimating animal abundance in a wide range of contexts and for a variety of taxa (+ref).
+In recent years, the framework has been extended and also applied for estimating changes in abundance over time in open populations by embedding a hidden state model representing population dynamics [@moore2011; @sollmann2015].
 

### 2023-06-02 Manuscript_Nilsen_Nater.qmd

@@
 
--   Distance sampling -- such methods are still lacking/not well utilized
-
--   Integrated pop. Model;
+Distance sampling (DS) models have been used for estimating animal abundance in a wide range of contexts and for a variety of taxa [@buckland2015].
+In recent years, they have been extended and also applied for estimating changes in abundance over time in open populations by embedding a hidden state model representing population dynamics [@moore2011; @sollmann2015].
+These latter frameworks may allow to accurately quantify population changes but provide little information on the drivers of these changes, i.e.
+the underlying vital rates.
+Concurrent with the development of more sophisticated DS models, another group of models has emerged and gained popularity, not least for their ability to disentangle demographic processes underlying population dynamic: integrated population models [IPMs, @schaub2021].
+Through joint analysis of multiple datasets, IPMs allow simultaneous estimation of population size and composition, as well as all vital rates that form part of an underlying typically age- or stage-structured population model.
+Since both DS models and IPMs estimate population size/density, a combination of the two frameworks has the potential to provide good estimates of both population- and demographic parameters and maximizing knowledge gained through transect surveys by augmenting them with other available data [e.g.
+@schmidt2020].
+In this study [...]
 
@@
 ### An integrated age-structured distance sampling model
-
-Distance sampling models have been used for estimating animal abundance in a wide range of contexts and for a variety of taxa (+ref).
-In recent years, the framework has been extended and also applied for estimating changes in abundance over time in open populations by embedding a hidden state model representing population dynamics [@moore2011; @sollmann2015].
 

### 2023-06-02 testRun_LierneVest.R
Modified model run for Lierne West

@@ -1,0 +1,213 @@
+
+
+library(tidyverse)
+
+# SETUP #
+#-------#
+
+## Source all functions in "R" folder
+sourceDir <- function(path, trace = TRUE, ...) {
+  for (nm in list.files(path, pattern = "[.][RrSsQq]$")) {
+    if(trace) cat(nm,":")
+    source(file.path(path, nm), ...)
+    if(trace) cat("\n")
+  }
+}
+sourceDir('R')
+
+
+## Set switches 
+
+# (Re-)downloading data
+# downloadData <- FALSE
+downloadData <- TRUE
+
+# Aggregation to area level
+areaAggregation <- FALSE
+
+# Recruitment per adult or per adult female
+R_perF <- FALSE
+
+# Drop observations of juveniles with no adults present
+R_parent_drop0 <- TRUE
+
+# Aggregation level for reproduction data
+# NOTE: if this is not defined, will default to group level
+sumR.Level <- "line" # Summing at the line level
+
+# Random effects shared across areas
+shareRE <- TRUE
+
+# Time variation in survival
+survVarT <- FALSE
+
+# Rodent covariate on reproduction
+fitRodentCov <- TRUE
+
+# DOWNLOAD/FETCH DATA #
+#---------------------#
+
+if(downloadData){
+  #Rype_arkiv <- downloadLN(datasets = "Fjellstyrene", versions = 1.6, save = TRUE)
+  Rype_arkiv <- downloadLN(datasets = c("Fjellstyrene", "Statskog", "FeFo"), versions = c(1.7, 1.8, 1.12), save = TRUE)
+}else{
+  stop("downloadData = FALSE not supported yet. There is an issue with encoding when using LivingNorwayR::initializeDwCArchive() that needs to be resolved first.")
+  #Rype_arkiv <- initializeDwCArchive("data/Rype_arkiv.zip")
+}
+
+
+# WRANGLE LINE TRANSECT DATA #
+#----------------------------#
+
+## Set localities/areas and time period of interest
+localities <- listLocations()
+areas <- listAreas()
+minYear <- 2007
+maxYear <- 2021
+
+## List duplicate transects to remove
+duplTransects <- listDuplTransects()
+
+## Extract transect and observational data from DwC archive
+LT_data <- wrangleData_LineTrans(DwC_archive_list = Rype_arkiv, 
+                                 duplTransects = duplTransects,
+                                 localities = localities,
+                                 #areas = areas,
[…]

### 2023-06-02 Manuscript_Nilsen_Nater.qmd
Added text to methods section

@@
 @schmidt2020].
-In this study [...]
+In this study \[...\]
 
@@
 
-### An integrated age-structured distance sampling model
+### An integrated distance sampling model
 
-Distance sampling models have been applied to a wide range of contexts, and has in recent years been extended to allow a direct modelling of the population process by formulating the state process as a open population process (ref).
-Here, we extend this framework by directly modelling the underlying state dynamics in terms of annual survival and recruitment rates.
+Our open population integrated distance sampling model consists of two major components: a latent structured population model and a set of likelihoods for data originating from distance sampling surveys and auxiliary survival monitoring (radio-telemetry in the example case).
 
-In line with standard age- or stage structured modelling of wildlife populations, we first defined the life cycle for the focal species -- the willow ptarmigan.
-Based on previous models for willow ptarmigan (ref), and based on the age-structure of the data, the model included two age classes: juveniles and adults (Figure 1).
+#### Age-structured population model
 
-![**Figure 1.** Simplified graphical representation of the ptarmigan life cycle with two age classes and the data sources included in the integrated distance sampling model.
+The population model follows a post-breeding census and includes two age classes: juveniles (young of the year) and adults (\> 1 year of age, @fig-model).
+This structure is inspired by earlier models for our focal species, the willow ptarmigan (+ref), and is commonly used for populations of passerine and game birds (+ref).
+In the context of our willow ptarmigan case study (see below), the census falls into late summer and coincides with the annual distance-sampling survey in August.
+
+Both juveniles and adults survive from year $t$ census to year $t+1$ census with survival probability $S_t$.
+As ptarmigan can reproduce already at 1-year old, all survivors then produce offspring in late June which recruit into the population as juveniles just prior to the census in year $t+1$ according to a recruitment rate $R_{t+1}$.
+The changes in densities (numbers) of juveniles and adults in the population, $D_{juv}$ and $D_{ad}$, can thus be expressed as
+
+$$
+\begin{aligned}
+  D_{juv, t+1} & = D_{ad,t+1} * R_{t+1} \\
+  D_{ad, t+1} & = S_t * (D_{juv,t} + D_{ad, t})  
+\end{aligned}
+$$ or, alternatively, in matrix notation as
+
+$$
+\left[
+\begin{array}{} 
+D_{juv,t+1} \\
+D_{ad,t+1}
+\end{array}
+\right]
+= 
+\left[
+\begin{array}{cc} 
+S_t*R_{t+1} & S_t*R_{t+1}\\
+S_t & S_t
+\end{array}
+\right]
+\left[\begin{array}{} 
+D_{juv,t} \\
+D_{ad,t}
+\end{array}\right]
+$$
+
+Note that recruitment rate $R$ is defined as juveniles/adult (not juveniles/female).
+We also make the simplifying assumption that there is no age- or sex-dependence of vital rates, but this assumption could be relaxed by including additional auxiliary data (Sandercock *et al.* 2011; Israelsen *et al.* 2020).
+
+![Simplified graphical representation of the ptarmigan life cycle with two age classes and the data sources included in the integrated distance sampling model.
 Juv\[t\] = juveniles in year t.
@@
 R\[t\] = recruitment rate in year t.
-S\[t\] = survival probability from year t to t+1.](Figures/RypeIDSM_structure.png)
+S\[t\] = survival probability from year t to t+1.](Figures/RypeIDSM_structure.png){#fig-model}
 
-We assumed that the population size and structured was surveyed each year at the end of the summer season in August, equivalent to the schedule of the real survey data.
-Ptarmigan breed annually, and the clutch hatch around last week of June.
-By the time of survey, they are around 1.5 months of age during the annual line transect surveys in August.
-Ptarmigan reach maturity at the age of one (ref).
-The life cycle is therefore described in term of i) annual survival probability (S~t~) which describe the probability to survive from August in year *t* to August in year *t+1*.
-Here, we assume no age-or sex-specificity in the survival rates, acknowledging that this is a simplifying assumption that could be alleviated with additional auxiliary data (Sandercock *et al.* 2011; Israelsen *et al.* 2020).
-Moreover, following the standard for post-breeding census models year specific recruitment (R~t~) will be given by the product of annual survival (S~t~) and fecundity (F~t~).
+#### Likelihoods for distance sampling data
 
-Based on the life cycle and the timing of the annual surveys, the transition from August year *t* to year *t+1* is given by the year-specific transition matrix - defined as:
+The implementation of the modelling framework we present makes three assumptions about the distance sampling survey: 1) the survey consists of line transects, 2) animals may be detected alone or in groups, and 3) juveniles and adults can be distinguished during surveys.
+These assumptions are inspired by our willow ptarmigan case study (details below).
+Our model includes three likelihoods for different components of the age-structured distance sampling data.
+
+#### Likelihood for radio-telemetry data
 
@@
 When located, the geographical coordinate, perpendicular distance from the sampling line, the number of birds as well as the age (juvenile and adult) and sex of the birds are recorded.
-Additional information about the surveys are found in Nilsen *et al.* (2020) and Bowler *et al.* (2020).
+\[Some more information on the surveys, e.g.
+why are individuals in groups, how are sexes and age classes distinguished, etc.\] Additional information about the surveys are found in Nilsen *et al.* (2020) and Bowler *et al.* (2020).
 \[Add sample sizes and additional information on data processing/storage post collection, i.e.

### 2023-06-05 Manuscript_Nilsen_Nater.qmd
Completed draft of model description

@@
 The population model follows a post-breeding census and includes two age classes: juveniles (young of the year) and adults (\> 1 year of age, @fig-model).
-This structure is inspired by earlier models for our focal species, the willow ptarmigan (+ref), and is commonly used for populations of passerine and game birds (+ref).
+This structure is inspired by earlier models for our focal species, the willow ptarmigan (+ref), and is commonly used for populations of passerine and game birds [@williams2002; @schaub2021].
 In the context of our willow ptarmigan case study (see below), the census falls into late summer and coincides with the annual distance-sampling survey in August.
@@
 \end{aligned}
-$$ or, alternatively, in matrix notation as
+$$
+
+or, alternatively, in matrix notation as
 
@@
 Our model includes three likelihoods for different components of the age-structured distance sampling data. 
+First is the likelihood for the perpendicular detection distances from line transect, $y$, which are linked to distance-dependent detection probability $p$ through a half-normal detection function:
+
+$$
+\begin{equation}
+p = exp(-\frac{y^2}{2\sigma^2})
+\end{equation}
+$$
+where $\sigma$ is the half-normal detection parameter.
+We assumed $\sigma$ to vary among years (index $t$) but not transect lines or animal group size.
+Following @moore2011, the resulting $\sigma_t$ can be used to calculate effective strip width ($esw_{t}$) and, consequently, average detection probability per line transect with a truncation distance $W$ according to:
+
+$$
+\begin{aligned}
+  esw_t & = \sqrt{\frac{\pi * \sigma_{t}^2}{2}}  \\
+  \hat{p_t} & = esw_t/W
+\end{aligned}
+$$ 
+The average detection probability $\hat{p_t}$ is an integral part of the second data likelihood which relates the observed number of animals in each age class $a$, $obsN_{a,j,t}$ ($j$ = transect) to the corresponding true number per transect, $N_{a,j,t}$:
+
+$$
+\begin{equation}
+  obsN_{a,j,t} \sim Poisson(\hat{p_t}*N_{a,j,t})
+\end{equation}
+$$
+$N_{juv,j,t}$ and $N_{ad,j,t}$ are then linked back to the population model by converting them to densities through multiplication with $2L_{j,t}W$ (where $L_{j,t}$ is length of transect $j$ in year $t$).
+
+The third data likelihood focuses on the counts of adults ($obsAd_{j,t}$) and juveniles ($obsJuv_{j,t}$) observed during the distance sampling surveys and links them to time-specific recruitment rate:
+
+$$
+\begin{equation}
+  obsJuv_{j,t} \sim Poisson(R_t*obsAd_{j,t})
+\end{equation}
+$$
 
@@
 
+The final likelihood is for the auxiliary telemetry data.
+It is set up under the assumption of perfect detection, and hence known fates, of animals bearing transmitters and links the numbers of animals released at the start of season $k$ of year $t$ to the number of survivors at the end of the same season:
+
+$$
+\begin{equation}
+  survivors_{k,t} \sim Binomial(released_{k,t}, Sk_t)
+\end{equation}
+$$
+Here, $Sk_t$ is the seasonal survival rate, and annual survival rate, $S_t$ is calculated as $S1_t * S2_t$.
+
+#### Hierarchical models with time-variation in parameters
+
+Vital rates (survival probabilities $S$, recruitment rates $R$), detection parameters (half-normal detection parameters $\sigma$), and initial population densities can all be modelled as time-dependent in our framework.
+For both the tests with simulated data and the case study described below, we implemented log-normally distributed random year effects on all parameters except survival, which was set to be constant.
+In the case study, we additionally included an effect of rodent occupancy (see details below) on log recruitment rates, resulting in the following model:
+
+$$
+\begin{equation}
+  log(R_t) = log(\mu_R) + \beta*RodentOcc_t + \epsilon_t
+\end{equation}
+$$
+where $\mu_R$ is the baseline recruitment rate, $\beta$ the slope of the effect of rodent occupancy, and $\epsilon_t$ the normally distributed random effects.
+
+### Model testing with simulated data
+
+TBA
+
 ### Field data (case study)
 
-Two different sources of data were used in the open population demographic distance sampling (ODDS) model: i) data from a line transect survey program targeting willow ptarmigan operated as part of the natural resources management authorities, and ii) data from an individual based research project based on radio collared willow ptarmigan.
+\[Add some basic information about willow ptarmigan in Norway, plus Lierne study area\]
+
+The two datasets used for the case study were i) data from a line transect survey program targeting willow ptarmigan operated as part of the natural resources management authorities, and ii) data from an individual based research project based on radio collared willow ptarmigan.
 The temporal coverage of the two data sets partially overlap, with the line transect survey spanning 2007-2019 (and ongoing) and the individual based data spanning 2015-2019 (and ongoing).
@@
 \[Figure 2: Map of Lierne study area?\]
+
+### Bayesian implementation
+
+We implemented the model in a Bayesian framework using NIMBLE [@valpine2017] in R version 4.2.2 [@R2022].
+The likelihood for line transect observation distances was set up using a custom half-normal distribution developed by Michael Scroggie as part of the "nimbleDistance" package (https://github.com/scrogster/nimbleDistance).
+We used non-informative uniform priors (with biologically reasonable boundaries where possible) for all parameters.
+For each model fit, we ran x MCMC chains with NIMBLE's standard samples for xxxk iterations.
+xxxk thereof were discarded as burn-in prior to thining with a factor x, leaving us with xxxx posterior samples per chain (total of xxxxx samples per run).
 

### 2023-06-06 Analysis_RealData.R
Added and included the new map
Unfortunately, something seems to be wrong with the shapefile. 
I get an error: Shape contains invalid polygons.
Additionally, there are six instances of the following warning: bounding box has potentially an invalid value range for longlat data.

The new map object is also substantially smaller than the previous one: 357 elements totaling 2.2 MB as opposed to 428 elements totaling 16.8 MB previously.

@@ -177,7 +177,7 @@ plotTimeSeries(mcmc.out = IDSM.out.tidy,
 #---------------------#
 
 ## Make map of Norwegian municipalities ("fylke")
-NorwayMunic.map <- setupMap_NorwayMunic(shp.path = "data/Kommuner_2018_WGS84/Kommuner_2018_WGS84.shp",
+NorwayMunic.map <- setupMap_NorwayMunic(shp.path = "data/NorgeKommuner_clipped_kystkontor/NorgeKommuner_clipped_kystkontor.shp",
                                         d_trans = LT_data$d_trans,
                                         areas = areas, areaAggregation = TRUE)
 

### 2023-06-06 R/setupMap_NorwayMunic.R

@@ -28,12 +28,12 @@ setupMap_NorwayMunic <- function(shp.path, d_trans,
     dplyr::filter(!(locality %in% c("Harodalen", "Savngovann"))) %>%
     dplyr::select(verbatimLocality, municipality_adj) %>%
     dplyr::distinct() %>%
-    dplyr::rename(KOMMUNENAV = municipality_adj,
+    dplyr::rename(navn = municipality_adj,
                   Area = verbatimLocality)
   
   ## Add area names to map data
   matchData <- as_tibble(mapNM@data) %>%
-    dplyr::left_join(., areaMunic, by = "KOMMUNENAV") 
+    dplyr::left_join(., areaMunic, by = "navn") 
 
   mapNM@data <- as.data.frame(matchData)
   

### 2023-06-06 Manuscript_Nater_etal.qmd
Updated files and deleted MS1 content

--- 2023-06-02 Manuscript_Nilsen_Nater.qmd
@@
 ---
-title: "An integrated distance sampling approach for modelling age-structured populations"
+title: "Large-scale spatiotemporal variation in vital rates of an alpine bird"
 format: html
-author: Erlend B. Nilsen & Chloé R. Nater
+author: Chloé R. Nater, James Martin, Erlend B. Nilsen
 editor: visual
-bibliography: rypeIDSM_MS1.bib
+bibliography: rypeIDSM_MS2.bib
 ---
@@
 
-Estimating abundance and demographic rates for wildlife populations is a key element of population ecology, demography, eco-evolution research, conservation and wildlife management (refs).
-Tremendous progress during the last few decades both in terms of new technologies that allow new types of field data being collected (e.g.
-eDNA ref; camera traps; refs) and the statistical machinery available to interpret the ecological and demographic signals in the data (refs).
-Nevertheless, large scale and long term data sets are still relatively rare (ref), due to high costs of field data collection.
-
-A cost efficient and well documented approach suitable is the use of line transect surveys. However, because the detectability is often less than 1 and variable between years and habitats,
-
--   Importance of abundance trends and demography
-
--   Short about methods -- detection probability -- marked vs unmarked animals
-
--   Approaches to estimate abundance and demography from unmarked animals
-
-Distance sampling (DS) models have been used for estimating animal abundance in a wide range of contexts and for a variety of taxa [@buckland2015].
-In recent years, they have been extended and also applied for estimating changes in abundance over time in open populations by embedding a hidden state model representing population dynamics [@moore2011; @sollmann2015].
-These latter frameworks may allow to accurately quantify population changes but provide little information on the drivers of these changes, i.e.
-the underlying vital rates.
-Concurrent with the development of more sophisticated DS models, another group of models has emerged and gained popularity, not least for their ability to disentangle demographic processes underlying population dynamic: integrated population models [IPMs, @schaub2021].
-Through joint analysis of multiple datasets, IPMs allow simultaneous estimation of population size and composition, as well as all vital rates that form part of an underlying typically age- or stage-structured population model.
-Since both DS models and IPMs estimate population size/density, a combination of the two frameworks has the potential to provide good estimates of both population- and demographic parameters and maximizing knowledge gained through transect surveys by augmenting them with other available data [e.g.
-@schmidt2020].
-In this study [...]
-
-Here, we will present a novel integrated population model based on distance sampling line transect survey data and demographic data from marked birds.
-
--   Estimate parameters with and without augmentation with demographic data -- compare estimates and precision
--   Estimate latent variable (juv summer survival) -- compare to literature values
--   Temporal variation in S and R -- via random effects models
--   Simulate - robustness
-
 ## Methods
 
-### An integrated age-structured distance sampling model
-
-Distance sampling models have been applied to a wide range of contexts, and has in recent years been extended to allow a direct modelling of the population process by formulating the state process as a open population process (ref).
-Here, we extend this framework by directly modelling the underlying state dynamics in terms of annual survival and recruitment rates.
-
-In line with standard age- or stage structured modelling of wildlife populations, we first defined the life cycle for the focal species -- the willow ptarmigan.
-Based on previous models for willow ptarmigan (ref), and based on the age-structure of the data, the model included two age classes: juveniles and adults (Figure 1).
-
-![**Figure 1.** Simplified graphical representation of the ptarmigan life cycle with two age classes and the data sources included in the integrated distance sampling model.
-Juv\[t\] = juveniles in year t.
-Ad\[t\] = adults in year t.
-R\[t\] = recruitment rate in year t.
-S\[t\] = survival probability from year t to t+1.](Figures/RypeIDSM_structure.png)
-
-We assumed that the population size and structured was surveyed each year at the end of the summer season in August, equivalent to the schedule of the real survey data.
-Ptarmigan breed annually, and the clutch hatch around last week of June.
-By the time of survey, they are around 1.5 months of age during the annual line transect surveys in August.
-Ptarmigan reach maturity at the age of one (ref).
-The life cycle is therefore described in term of i) annual survival probability (S~t~) which describe the probability to survive from August in year *t* to August in year *t+1*.
-Here, we assume no age-or sex-specificity in the survival rates, acknowledging that this is a simplifying assumption that could be alleviated with additional auxiliary data (Sandercock *et al.* 2011; Israelsen *et al.* 2020).
-Moreover, following the standard for post-breeding census models year specific recruitment (R~t~) will be given by the product of annual survival (S~t~) and fecundity (F~t~).
-
-Based on the life cycle and the timing of the annual surveys, the transition from August year *t* to year *t+1* is given by the year-specific transition matrix - defined as:
-
-### Field data (case study)
-
-Two different sources of data were used in the open population demographic distance sampling (ODDS) model: i) data from a line transect survey program targeting willow ptarmigan operated as part of the natural resources management authorities, and ii) data from an individual based research project based on radio collared willow ptarmigan.
-The temporal coverage of the two data sets partially overlap, with the line transect survey spanning 2007-2019 (and ongoing) and the individual based data spanning 2015-2019 (and ongoing).
-
-Line transect survey data were collected in August each year, prior to the annual autumn harvest season.
-Line transects are surveyed by educated volunteers, using pointing dogs to located the birds.
-When located, the geographical coordinate, perpendicular distance from the sampling line, the number of birds as well as the age (juvenile and adult) and sex of the birds are recorded.
-Additional information about the surveys are found in Nilsen *et al.* (2020) and Bowler *et al.* (2020).
-\[Add sample sizes and additional information on data processing/storage post collection, i.e.
-how do volunteers report data, storage in hønsefuglportalen, quality control, pushes to LivingNorway and GBIF\].
-
-The individual longitudinal study based on radio collared willow ptarmigan was conducted in 2015-2020.
-Each winter (in February-March), willow ptarmigan were located at night using snowmobiles and large hand nets with prolonged handles, as described in Israelsen et al.
-(2020).
-To prevent birds from flying off before the field personnel were close enough to capture them, a high-powered head lamp was used to dazzle the birds.
-After capture, birds were placed in an opaque bag to reduce stress.
-Before releasing the birds, they were fitted with a uniquely numbered leg ring (\~ 2.4g) and a Holohil RI-2BM or Holohil RI-2DM radio transmitter (\~ 14.1g).
-The radio transmitters had an expected battery lifetime of 24 months (RI-2BM) or 30 months (RI-2DM), and included a mortality circuit that was activated if a bird had been immobile for 12 hours.
-We monitored the birds throughout the year, by triangulation from the ground at least once a month for 10 months of the year (February -- November) by qualified field personnel.
-If a mortality signal was heard from the transmitter, we recovered it as soon as possible to determine cause of death.
-A number of birds dispersed out of the main study areas and was thus out of signal range for field personnel on the ground.
-To avoid loss of data, we conducted aerial triangulation using a helicopter or airplane three times a year (May, September and November) in the years 2016-2020.
-
-\[Figure 2: Map of Lierne study area?\]
-
 ## Results
-
-### Model performance on simulated datasets
-
-\[One or several figures of estimates vs. true values?\]
-
-### Case study on willow ptarmigans in Lierne
-
-#### Population density
-
-![**Figure X.** Estimated density of willow ptarmigans in Lierne from 2007 to 2021. Solid line represents the posterior median, ribbon marks 95% credible interval.](Figures/TimeSeries_popDens.png)
-
-#### Vital rates
-
-![**Figure X.** Posterior densities of A) seasonal survival, B) average annual survival, and C) recruitment rate.
-For the latter, the yellow distribution is for the intercept, representing a baseline recruitment rate when rodent occupancy is low.
-The turquoise distributions are for year-specific estimates or recruitment rate, with darker colors indicating later years.](Figures/PostDens_Mu_S_tR.png)
-
-![**Figure X.** Estimated recruitment rate of willow ptarmigans in Lierne from 2007 to 2021.
-Solid line represents the posterior median, ribbon marks the 95% credible interval.](Figures/TimeSeries_rRep.png)
-
-![**Figure X.** Predicted effect of rodent occupancy on recruitment rate. Solid line represent the posterior median, ribbon marks the 95% credible interval.](Figures/Rep_betaR.R.png)
-
-#### Detection parameters
-
-![**Figure X.** Predicted average detection probabilities for distance sampling surveys in Lierne from 2007 to 2021.
-Solid line represents the posterior median, ribbon marks the 95% credible interval.](Figures/TimeSeries_pDetect.png)
 

### 2023-06-06 Manuscript_Nater_etal.qmd
Added some worknote content

@@
 format: html
-author: Chloé R. Nater, James Martin, Erlend B. Nilsen
+author: Chloé R. Nater, James A. Martin, Erlend B. Nilsen
 editor: visual
@@
 
-## Introduction
+## About this worknote
+The purpose of this worknote is to introduce and showcase preliminary results from application of the open-population integrated distance-sampling model (IDSM) for willow ptarmigan to data from all of Norway.
+Eventually, this model extension and resulting insights into large-scale spatiotemporal variation in density, population growth rate, and vital rates will be the subject of a second scientific paper.
+This paper will be written and submitted in the second half of 2023, and this worknote serves as a temporary substitute for it in the reporting of project results to Miljødirektoratet in June 2022.
+As such, the worknote contains a general description of the extension of the model to analyse data from multiple areas simultaneously, visualizations of some preliminary results, and an introduction to a reproducible workflow that allows to easily and robustly re-run analyses as new data becomes available in future years.
 
-## Methods
+## IDSM extension to multiple areas
 
-## Results
+Distance-sampling data on willow ptarmigan is collected on line transects not only in western Lierne, but across most of its Norwegian range (https://honsefugl.nina.no/).
+Taking the line transect datasets from Fjellstyrene, Statskog, and FeFo together, there is data for willow ptarmigan in 41 reporting areas (for a list, see https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp/blob/main/R/listAreas.R).
+The model developed as part of this project and presented in @nilsen2023 was designed with the potential of extension to run on data from multiple areas simultaneously in mind, and this extension involves two steps in practice.
 
-## Discussion
+The first step is the inclusion of an additional dimension, area, to enable using data collected from multiple areas (@fig-model).
+The area-dimension is added across all model components with the exception of the known-fate telemetry data (and seasonal decomposition of survival estimated from it), as that is only available for the Lierne area.
 
-### Conclusion
+![Simplified graphical representation of the ptarmigan life cycle with two age classes and the data sources included in the integrated distance sampling model.
+The pink "x"'s indicate the added dimension for area.
+Juv\[x,t\] = juveniles in year t.
+Ad\[x,t\] = adults in year t.
+R\[x,t\] = recruitment rate in year t.
+S\[x,t\] = survival probability from year t to t+1.](Figures/RypeIDSM_multiArea_structure.png){#fig-model}
 
-### Author contributions
+The second step is making adjustments to capitalize on the integration of data from multiple areas to make maximal use of available information. 
+There are several not mutually exclusive options for approaching this, and one of them is the use of "hyper-parameters" for sharing information across areas [e.g. @horswill2019; @horswill2021].
+As an illustrative example, take the area-specific average annual survival probability, $\mu^S_x$.
+Instead of assigning independent non-informative priors to each $\mu^S_x$, we now assume that the variation in average survival across space can be described by a common (logit) normal distribution such that:
 
-### Acknowledgements
+$$
+\begin{equation}
+\mu^S_x \sim Normal(logit(h.\mu^S), h.\sigma^S)
+\end{equation}
+$$
+where $h.\mu^S$ is the "hyper mean" and $h.\sigma^S$ is the "hyper standard deviation". 
+Non-informative priors are then assigned to the hyper-parameters instead of the area-specific realizations:
 
-### Data and code availability
+$$
+\begin{align}
+h.\mu^S & \sim Uniform(0,1) \\
+h.\sigma^S & \sim Uniform(0, 5)
+\end{align}
+$$
+
+For the multi-area implementation of the IDSM, we used the equivalent of the above approach for parameters representing average survival probabilities, baseline recruitment rates, rodent effect on recruitment, and average distance sampling detection.
+Initial population density, on the other hand, was estimated independently for each area. 
+Random effects standard deviations were assumed to be the same for all areas, effectively resulting in space-for-time subtitution for random effect estimation.
+
+Another approach to maximising information gain when jointly analysing data from different areas is explicit modelling of spatial and spatio-temporal variation.
+While this is commonly done for modern species distribution models [e.g.
+@pacifici2017; @guelat2018], it has rarely been implemented in demographic models.
+This is not least due to the fact that few demographic models span large spatial extents.
+Since our IDSM and the willow ptarmigan data do actually do that though, extending the framework to account for spatial and spatiotemporal correlations may be feasible.
+Per now, we have not yet attempted this, but are planning to do so in the near future.
+
+## Preliminary results
+
+### Population density
+
+::: {#fig-dens layout-ncol=2}
+
+![Median density estimates](Figures/density_2018-2020_Median.png){#fig-densA}
+
+![Uncertainty in density estimates](Figures/density_2018-2020_SD.png){#fig-densB}
+
+Median (a) and standard deviation (b) of posterior estimates of average ptarmigan density in the three most recent years (2018-2020) across 41 reporting areas in Norway.
+:::
+
+### Population growth rate
+
+::: {#fig-lam layout-ncol=2}
+
+![Median population growth rate estimates](Figures/lambda_2018-2020_Median.png){#fig-lamA}
+
+![Uncertainty in population growth rate estimates](Figures/lambda_2018-2020_SD.png){#fig-lamB}
+
+Median (a) and standard deviation (b) of posterior estimates of average annual population growth rate over the three most recent years (2018-2020) across 41 reporting areas in Norway.
+:::
+
+### Survival probability
+
+::: {#fig-pSurv layout-ncol=2}
+
+![Median survival estimates](Figures/pSurv_Median.png){#fig-pSurvA}
+
+![Uncertainty in survival estimates](Figures/pSurv_SD.png){#fig-pSurvB}
+
+Median (a) and standard deviation (b) of posterior estimates of average annual survival probability across 41 reporting areas in Norway.
+:::
+
+### Recruitment rate
+
+::: {#fig-rRep layout-ncol=2}
+
+![Median recruitment estimates](Figures/rRep_Median.png){#fig-rRepA}
+
+![Uncertainty in recruitment estimates](Figures/rRep_SD.png){#fig-rRepB}
+
+Median (a) and standard deviation (b) of posterior estimates of average annual recruitment rate across 41 reporting areas in Norway.
+:::
+
+### Effect of rodent occupancy
+
+::: {#fig-betaR layout-ncol=2}
+
+![Median rodent effect estimates](Figures/betaR_Median.png){#fig-betaRA}
+
+![Uncertainty in rodent effect estimates](Figures/betaR_SD.png){#fig-betaRB}
+
+Median (a) and standard deviation (b) of posterior estimates of effects of rodent occupancy on recruitment rate across 41 reporting areas in Norway.
+:::
+
+## Reproducible workflow with "targets"
+
+## Data and code availability

### 2023-06-07 Manuscript_Nater_etal.qmd
Completed worknote draft

@@
 ## About this worknote
+
 The purpose of this worknote is to introduce and showcase preliminary results from application of the open-population integrated distance-sampling model (IDSM) for willow ptarmigan to data from all of Norway.
@@
 Distance-sampling data on willow ptarmigan is collected on line transects not only in western Lierne, but across most of its Norwegian range (https://honsefugl.nina.no/).
-Taking the line transect datasets from Fjellstyrene, Statskog, and FeFo together, there is data for willow ptarmigan in 41 reporting areas (for a list, see https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp/blob/main/R/listAreas.R).
+Taking the line transect datasets from Fjellstyrene, Statskog, and FeFo together, there is data for willow ptarmigan in 41 reporting areas (for a list, see [here]( https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp/blob/main/R/listAreas.R)).
 The model developed as part of this project and presented in @nilsen2023 was designed with the potential of extension to run on data from multiple areas simultaneously in mind, and this extension involves two steps in practice.
@@
 The second step is making adjustments to capitalize on the integration of data from multiple areas to make maximal use of available information.
-There are several not mutually exclusive options for approaching this, and one of them is the use of "hyper-parameters" for sharing information across areas [e.g. @horswill2019; @horswill2021].
+There are several not mutually exclusive options for approaching this, and one of them is the use of "hyper-parameters" for sharing information across areas [e.g.
+@horswill2019; @horswill2021].
 As an illustrative example, take the area-specific average annual survival probability, $\mu^S_x$.
@@
 \end{equation}
-$$
-where $h.\mu^S$ is the "hyper mean" and $h.\sigma^S$ is the "hyper standard deviation". 
-Non-informative priors are then assigned to the hyper-parameters instead of the area-specific realizations:
+$$ where $h.\mu^S$ is the "hyper mean" and $h.\sigma^S$ is the "hyper standard deviation". Non-informative priors are then assigned to the hyper-parameters instead of the area-specific realizations:
 
@@
 
+In the following, we present some preliminary results from the model fit to data from all 41 reporting areas simultaneously.
+Results are presented in the format of pairs of maps.
+For each quantity/result, we present one map of the posterior median and one map of the standard deviation of the posterior distribution, where the latter serves as a measurement of uncertainty for the former.
+
 ### Population density
 
-::: {#fig-dens layout-ncol=2}
+@fig-densA shows that population density is generally lowest in the Northern parts of Norway and highest in central Norway, close to the Swedish borders.
+Higher median density estimates also seem to be associated with higher estimation uncertainty (@fig-densB) and therefore have to be interpreted with care.
 
+::: {#fig-dens layout-ncol="2"}
 ![Median density estimates](Figures/density_2018-2020_Median.png){#fig-densA}
@@
 Median (a) and standard deviation (b) of posterior estimates of average ptarmigan density in the three most recent years (2018-2020) across 41 reporting areas in Norway.
+Darker colors indicate higher median values and higher uncertainty.
 :::
@@
 
-::: {#fig-lam layout-ncol=2}
+In most areas, populations of willow ptarmigan have been increasing over the period 2018-2020, potentially indicating a positive effect of more restricted hunting in recent years (@fig-lam).
+Some areas -- predominantly in central Norway -- also had declining populations, but many of those declines followed upon a period of increase between the start of data colection in 2007 and sometime between 2016 and 2018.
 
+::: {#fig-lam layout-ncol="2"}
 ![Median population growth rate estimates](Figures/lambda_2018-2020_Median.png){#fig-lamA}
@@
 Median (a) and standard deviation (b) of posterior estimates of average annual population growth rate over the three most recent years (2018-2020) across 41 reporting areas in Norway.
+In a), pinkish colors indicate declining populations while greenish colors indicate growing populations (white = stable populations).
+In b) darker colors indicate higher median values and higher uncertainty.
 :::
@@
 
-::: {#fig-pSurv layout-ncol=2}
+Annual survival probabilities ranged from aroung 35% to 50% across reporting areas in Norway, with the highest values occurring in the far north and in the mountains in the south (@fig-pSurv).
+
+::: {#fig-pSurv layout-ncol="2"}
 
@@
 Median (a) and standard deviation (b) of posterior estimates of average annual survival probability across 41 reporting areas in Norway.
+Darker colors indicate higher median values and higher uncertainty.
 :::
@@
 
-::: {#fig-rRep layout-ncol=2}
+Annual recruitment rates showed a spatial pattern (@fig-rRep) that was inverse relative to that of annual survival (@fig-pSurv), with lower recruitment rates co-occurring with higher survival rates and vice-versa.
+This is consistent with basic life history theory [@stearns1992] and with previous studies finding relatively slower life histories in alpine / high altitude areas [e.g.
+@sandercock2005; @bears2009; @alice2016].
 
+::: {#fig-rRep layout-ncol="2"}
 ![Median recruitment estimates](Figures/rRep_Median.png){#fig-rRepA}
@@
 
-Median (a) and standard deviation (b) of posterior estimates of average annual recruitment rate across 41 reporting areas in Norway.
+Median (a) and standard deviation (b) of posterior estimates of average annual recruitment rate across 41 reporting areas in Norway. Darker colors indicate higher median values and higher uncertainty.
 :::
@@
 
-::: {#fig-betaR layout-ncol=2}
+The model predicted substantial variation in the effect of the rodent occupancy covariate on recruitment (@fig-betaR).
+Areas in the north and south west generally had recruitment increasing with rodent occupancy while the remaining areas were a mosaic of positive and negative effects, albeit with some spatial clustering.
 
+::: {#fig-betaR layout-ncol="2"}
 ![Median rodent effect estimates](Figures/betaR_Median.png){#fig-betaRA}
@@
 Median (a) and standard deviation (b) of posterior estimates of effects of rodent occupancy on recruitment rate across 41 reporting areas in Norway.
+In a), blueish colors indicate negative effects while reddish colors indicate positive effects (yellow = no effect).
+In b) darker colors indicate higher median values and higher uncertainty.
 :::
@@
 
+Reproducibility and ease of repeating analyses was a key focus when developing the multi-area IDSM.To that end, we set up the workflow as a "targets pipeline", implemented throught the R packages "targets" [@landau2021].
+A visual representation of the pipeline is displayed in @fig-workflow.
+
+![Graphical representation of the "targets pipeline" for the multi-area modelling setup.
+Upward facing triangles are functions, downward facing triangles are general options/arguments, circles are objects and outputs (="targets") created as part of the workflow.
+](Figures/Workflow_layout.png){#fig-workflow width="424"}
+
 ## Data and code availability
+
+The raw data from the line transect surveys is deposited on GBIF and can be accessed freely via the Living Norway Data Portal (https://data.livingnorway.no/).
+The work presented above is based on versions 1.7, 1.8, and 1.12 for the datasets from Fjellstyrene (key b49a2978-0e30-4748-a99f-9301d17ae119), Statskog (key 6a948a1c-7e23-4d99-b1c1-ec578d0d3159), and FeFo (key c47f13c1-7427-45a0-9f12-237aad351040), respectively.
+
+The auxiliary radio-telemetry data and all code, including the "targets pipeline" can be found in the project's repository on GitHub: [https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp](https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp).
+
+## References

### 2023-08-02 Manuscript_Nilsen_Nater.qmd
Some text updates

@@
 
-Estimating abundance and demographic rates for wildlife populations is a key element of population ecology, demography, eco-evolution research, conservation and wildlife management (refs).
-Tremendous progress during the last few decades both in terms of new technologies that allow new types of field data being collected (e.g.
-eDNA ref; camera traps; refs) and the statistical machinery available to interpret the ecological and demographic signals in the data (refs).
+Estimating abundance and demographic rates for wildlife populations is an integral part of basic and applied ecology [@Skalski2005].
+Over the last few decades, a tremendous progress have been made towards this end.
+This progress is partly driven the development and application of new field data collection methods and apporaches, such as citizen science data (ref), camera trap data (ref) and the collection of environmental DNA data (ref).
+In addition, developments of new analysis methods allow us to estimate abundance and demographic rates it situations where it was not feasible before (ref).
+combined, these advance put us in a much better position to estimate state variables needed for population management (ref) but also estimate indexes that are relevant for large scale policy applications, e.g.
+such as Essential Biodiversity Variables (ref).
+
 Nevertheless, large scale and long term data sets are still relatively rare (ref), due to high costs of field data collection.
@@
 \end{equation}
-$$
-where $\sigma$ is the half-normal detection parameter.
+$$ where $\sigma$ is the half-normal detection parameter.
 We assumed $\sigma$ to vary among years (index $t$) but not transect lines or animal group size.
@@
 \end{aligned}
-$$ 
-The average detection probability $\hat{p_t}$ is an integral part of the second data likelihood which relates the observed number of animals in each age class $a$, $obsN_{a,j,t}$ ($j$ = transect) to the corresponding true number per transect, $N_{a,j,t}$:
+$$ The average detection probability $\hat{p_t}$ is an integral part of the second data likelihood which relates the observed number of animals in each age class $a$, $obsN_{a,j,t}$ ($j$ = transect) to the corresponding true number per transect, $N_{a,j,t}$:
 
@@
 \end{equation}
-$$
-$N_{juv,j,t}$ and $N_{ad,j,t}$ are then linked back to the population model by converting them to densities through multiplication with $2L_{j,t}W$ (where $L_{j,t}$ is length of transect $j$ in year $t$).
+$$ $N_{juv,j,t}$ and $N_{ad,j,t}$ are then linked back to the population model by converting them to densities through multiplication with $2L_{j,t}W$ (where $L_{j,t}$ is length of transect $j$ in year $t$).
 
@@
 \end{equation}
-$$
-Here, $Sk_t$ is the seasonal survival rate, and annual survival rate, $S_t$ is calculated as $S1_t * S2_t$.
+$$ Here, $Sk_t$ is the seasonal survival rate, and annual survival rate, $S_t$ is calculated as $S1_t * S2_t$.
 
@@
 \end{equation}
-$$
-where $\mu_R$ is the baseline recruitment rate, $\beta$ the slope of the effect of rodent occupancy, and $\epsilon_t$ the normally distributed random effects.
+$$ where $\mu_R$ is the baseline recruitment rate, $\beta$ the slope of the effect of rodent occupancy, and $\epsilon_t$ the normally distributed random effects.
 

### 2023-11-30 R/writeModelCode.R
Generalised area loops

@@ -41,11 +41,12 @@ writeModelCode <- function(survVarT){
     # POPULATION MODEL #
     ####################
     
-    #-----------------------------------------#
-    # Initial population size/density (t = 1) #
-    #-----------------------------------------#
-    
     for(x in 1:N_areas){
+      
+      #-----------------------------------------#
+      # Initial population size/density (t = 1) #
+      #-----------------------------------------#
+      
       for (j in 1:N_sites[x]){
         
         N_exp[x, 1, j, 1] ~ dpois(Density[x, 1, j, 1]*L[x, j, 1]*W*2) 
@@ -54,14 +55,13 @@ writeModelCode <- function(survVarT){
         Density[x, 1, j, 1] <- exp(log(Mu.D1[x]) + eps.D1[x, j])*ratio.JA1[x]
         Density[x, 2, j, 1] <- exp(log(Mu.D1[x]) + eps.D1[x, j])*(1-ratio.JA1[x])
       }
-    }
-    
-    
-    #-------------------------------#
-    # Population dynamics for t > 1 #
-    #-------------------------------#
-    
-    for(x in 1:N_areas){
+      
+      
+      
+      #-------------------------------#
+      # Population dynamics for t > 1 #
+      #-------------------------------#
+      
       for(j in 1:N_sites[x]){
         for(t in 2:N_years){
           
@@ -79,27 +79,26 @@ writeModelCode <- function(survVarT){
           N_exp[x, 1:N_ageC, j, t] <- Density[x, 1:N_ageC, j, t]*L[x, j, t]*W*2
         }
       }
-    }
-    
-    #--------------------#
-    # Derived parameters #
-    #--------------------#
-    
-    ## Area- and year-specific total densities
-    for(x in 1:N_areas){
+      
+      #--------------------#
+      # Derived parameters #
+      #--------------------#
+      
+      ## Area- and year-specific total densities
       for (t in 1:N_years){
         N_tot_exp[x, t] <- sum(N_exp[x, 1, 1:N_sites[x], t] + N_exp[x, 2, 1:N_sites[x], t])
       }
-    }
+    } # x
     
     
     ####################
     # DATA LIKELIHOODS #
     ####################
     
-    ## Age-specific line transect counts
[…]

### 2023-11-30 Analysis_RealData.R
Updated functions and masterscripts to work with new RE setup

@@ -30,14 +30,11 @@ R_parent_drop0 <- TRUE
 # NOTE: if this is not defined, will default to group level
 sumR.Level <- "line" # Summing at the line level
 
-# Random effects shared across areas
-shareRE <- TRUE
-
 # Time variation in survival
-survVarT <- FALSE
+survVarT <- TRUE
 
 # Rodent covariate on reproduction
-fitRodentCov <- FALSE
+fitRodentCov <- TRUE
 
 # DOWNLOAD/FETCH DATA #
 #---------------------#
@@ -118,7 +115,6 @@ modelCode <- writeModelCode(survVarT = survVarT)
 ## Setup for model using nimbleDistance::dHN
 model_setup <- setupModel(modelCode = modelCode,
                           R_perF = R_perF,
-                          shareRE = shareRE, 
                           survVarT = survVarT, 
                           fitRodentCov = fitRodentCov,
                           nim.data = input_data$nim.data,
@@ -136,27 +132,31 @@ IDSM.out <- nimbleMCMC(code = model_setup$modelCode,
                        inits = model_setup$initVals, 
                        monitors = model_setup$modelParams,
                        nchains = model_setup$mcmcParams$nchains, 
-                       niter = model_setup$mcmcParams$niter, 
+                       #niter = model_setup$mcmcParams$niter, 
+                       niter = 500,
                        nburnin = model_setup$mcmcParams$nburn, 
                        thin = model_setup$mcmcParams$nthin, 
                        samplesAsCodaMCMC = TRUE, 
                        setSeed = 0)
 #Sys.time() - t.start
 
-saveRDS(IDSM.out, file = 'rypeIDSM_dHN_multiArea_realData_Lierne.rds')
+saveRDS(IDSM.out, file = "rypeIDSM_dHN_multiArea_realData_Lierne.rds")
 
 
 # TIDY UP POSTERIOR SAMPLES #
 #---------------------------#
 
-IDSM.out.tidy <- tidySamples(IDSM.out = IDSM.out, save = TRUE)
+IDSM.out.tidy <- tidySamples(IDSM.out = IDSM.out, 
+                             save = TRUE,
+                             fileName = "rypeIDSM_dHN_multiArea_realData_allAreas_tidy.rds")
 
 
 # OPTIONAL: MCMC TRACE PLOTS #
 #----------------------------#
 
 plotMCMCTraces(mcmc.out = IDSM.out.tidy,
[…]

### 2023-11-30 Analysis_RealData_LierneVest.R

@@ -33,9 +33,6 @@ R_parent_drop0 <- TRUE
 # NOTE: if this is not defined, will default to group level
 sumR.Level <- "line" # Summing at the line level
 
-# Random effects shared across areas
-shareRE <- TRUE
-
 # Time variation in survival
 survVarT <- FALSE
 
@@ -115,23 +112,18 @@ input_data <- prepareInputData(d_trans = LT_data$d_trans,
 # MODEL SETUP #
 #-------------#
 
-# Original version (zeroes-trick)
-# model_setup <- setupModel(modelCode.path = "NIMBLE Code/RypeIDSM.R",
-#                           customDist = FALSE,
-#                           nim.data = input_data$nim.data,
-#                           nim.constants = input_data$nim.constants,
-#                           testRun = FALSE, initVals.seed = 0)
-  
-# Updated version (nimbleDistance::dHN)
-model_setup <- setupModel(modelCode.path = "NIMBLE Code/RypeIDSM_multiArea_dHN.R",
-                          customDist = TRUE,
+## Write model code
+modelCode <- writeModelCode(survVarT = survVarT)
+
+## Setup for model using nimbleDistance::dHN
+model_setup <- setupModel(modelCode = modelCode,
                           R_perF = R_perF,
-                          shareRE = shareRE, 
                           survVarT = survVarT, 
                           fitRodentCov = fitRodentCov,
                           nim.data = input_data$nim.data,
                           nim.constants = input_data$nim.constants,
-                          testRun = TRUE, nchains = 4,
+                          testRun = TRUE, 
+                          nchains = 4,
                           initVals.seed = 0)
 
 # MODEL (TEST) RUN #
@@ -150,20 +142,21 @@ IDSM.out <- nimbleMCMC(code = model_setup$modelCode,
                        setSeed = 0)
 Sys.time() - t.start
 
-saveRDS(IDSM.out, file = 'rypeIDSM_dHN_multiArea_realData_Lierne.rds')
+saveRDS(IDSM.out, file = "rypeIDSM_dHN_multiArea_realData_Lierne.rds")
 
 
 # TIDY UP POSTERIOR SAMPLES #
 #---------------------------#
 
-IDSM.out.tidy <- tidySamples(IDSM.out = IDSM.out, save = TRUE)
[…]

### 2023-11-30 R/plotMCMCTraces.R

@@ -2,6 +2,7 @@
 #'
 #' @param mcmc.out an mcmc list containing posterior samples from a model run.
 #' @param fitRodentCov logical. If TRUE, rodent covariate on reproduction is included.
+#' @param survVarT logical. If TRUE, temporal and residual variation in survival are included. 
 #' @param VitalRates logical. If TRUE (default), plots traces and posterior densities for vital rate parameters.
 #' @param DetectParams logical. If TRUE (default), plots traces and posterior densities for detection parameters.
 #' @param PopSizes logical. If TRUE (default), plots traces and posterior densities for population sizes.
@@ -12,19 +13,24 @@
 #'
 #' @examples
 #' 
-plotMCMCTraces <- function(mcmc.out, fitRodentCov, VitalRates = TRUE, DetectParams = TRUE, PopSizes = TRUE, Densities = TRUE){
+plotMCMCTraces <- function(mcmc.out, fitRodentCov, survVarT, VitalRates = TRUE, DetectParams = TRUE, PopSizes = TRUE, Densities = TRUE){
   
   ## Make parameter lists
-  mVR_params <- c("Mu.R", "h.Mu.R", "h.sigma.R", "sigmaT.R",
+  mVR_params <- c("Mu.R", "h.Mu.R", "h.sigma.R", "sigmaT.R", "sigmaR.R",
                   "Mu.D1", "sigma.D", "ratio.JA1",
                   "Mu.S", "Mu.S1", "h.Mu.S", "h.sigma.S")
+  
+  if(survVarT){
+    mVR_params <- c(mVR_params, c("sigmaT.S", "sigmaR.S"))
+  }
+  
   if(fitRodentCov){
     mVR_params <- c(mVR_params, "betaR.R", "h.Mu.betaR.R", "h.sigma.betaR.R")
   }
   
   tVR_params <- c("R_year", "S")
   
-  mDet_params <- c("mu.dd", "sigmaT.dd")
+  mDet_params <- c("mu.dd", "sigmaT.dd", "sigmaR.dd")
   
   tDet_params <- c("esw", "p", "sigma")
   

### 2023-11-30 R/setupModel.R

@@ -6,9 +6,11 @@
 #' @param nim.constants list of input objects representing constants
 #' @param R_perF logical. If TRUE, treats recruitment rate as juvenile per adult female.
 #' If FALSE, treats recruitment rate as juvenile per adult (sum of both sexes).
-#' @param shareRE logical. If TRUE, temporal random effects are shared across locations.
 #' @param survVarT logical. If TRUE, survival is simulated including annual variation.
 #' @param fitRodentCov logical. If TRUE, rodent covariate on reproduction is included.
+#' @param addDummyDim logical. If TRUE (default) adds a dummy "area" dimension when 
+#' simulating initial values for a single area implementation. This is necessary 
+#' for the multi-area setup/model to run with data from only one area. 
 #' @param niter integer. Number of MCMC iterations (default = 25000)
 #' @param nthin integer. Thinning factor (default = 5)
 #' @param nburn integer. Number of iterations to discard as burn-in (default = 5000)
@@ -25,7 +27,7 @@
 
 setupModel <- function(modelCode, customDist,
                        nim.data, nim.constants,
-                       R_perF, shareRE, survVarT, fitRodentCov, addDummyDim = TRUE,
+                       R_perF, survVarT, fitRodentCov, addDummyDim = TRUE,
                        niter = 100000, nthin = 20, nburn = 40000, nchains = 3,
                        testRun = FALSE, initVals.seed){
   
@@ -34,8 +36,8 @@ setupModel <- function(modelCode, customDist,
   
   ## Set parameters to monitor
   params <- c("esw", "p", #"D",
-              "R_year", "Mu.R", "h.Mu.R", "h.sigma.R", "sigmaT.R",
-              "sigma", "mu.dd", "sigmaT.dd",
+              "R_year", "Mu.R", "h.Mu.R", "h.sigma.R", "sigmaT.R", "sigmaR.R",
+              "sigma", "mu.dd", "sigmaT.dd", "sigmaR.dd",
               "Density", "N_exp", "N_tot_exp",
               "Mu.D1", "sigma.D",
               "S", "Mu.S", "h.Mu.S", "h.sigma.S",
@@ -43,7 +45,7 @@ setupModel <- function(modelCode, customDist,
               "ratio.JA1")
   
   if(survVarT){
-    params <- c(params, "sigmaT.S", "epsT.S1.prop")
[…]

### 2023-11-30 R/simulateInits.R

@@ -4,7 +4,6 @@
 #' @param nim.constants list of input objects representing constants
 #' @param R_perF logical. If TRUE, treats recruitment rate as juvenile per adult female.
 #' If FALSE, treats recruitment rate as juvenile per adult (sum of both sexes).
-#' @param shareRE logical. If TRUE, temporal random effects are shared across locations.
 #' @param survVarT logical. If TRUE, survival is simulated including annual variation.
 #' @param fitRodentCov logical. If TRUE, initial values are generated for rodent 
 #' covariate (effect) and covariate is included in data simulation.
@@ -14,7 +13,7 @@
 #'
 #' @examples
 
-simulateInits <- function(nim.data, nim.constants, R_perF, shareRE, survVarT, fitRodentCov){
+simulateInits <- function(nim.data, nim.constants, R_perF, survVarT, fitRodentCov){
   
   # Limits and constants #
   #----------------------#
@@ -63,28 +62,32 @@ simulateInits <- function(nim.data, nim.constants, R_perF, shareRE, survVarT, fi
   mu.S <- EnvStats::rnormTrunc(N_areas, qlogis(h.Mu.S), sd = h.sigma.S, max = qlogis(Mu.S1))
 
   sigmaT.S <- runif(1, 0, 0.05)
+  sigmaR.S <- runif(1, 0, 0.05)
   
   Mu.S <- rep(NA, N_areas)
   S <-  matrix(NA, nrow = N_areas, ncol = N_years)
   S1 <- S2 <- rep(NA, N_years)
   
-  epsT.S1.prop <- runif(1, 0.3, 0.8)
+  eps.S1.prop <- runif(1, 0.3, 0.8)
   
   if(survVarT){
-    epsT.S <- matrix(0, nrow = N_areas, ncol = N_years)
-    #epsT.S <- matrix(rnorm(N_areas*N_years, 0, sigmaS.R), nrow = N_areas, ncol = N_years)
+    epsT.S <- rep(0, N_years)
+    #epsT.S <- rnorm(N_years, 0, sigmaT.S)
[…]

### 2023-11-30 R/writeModelCode.R
Added previously missing RE

@@ -158,7 +158,7 @@ writeModelCode <- function(survVarT){
       for(t in 1:N_years){
         
         # Detection decay
-        log(sigma[x, t]) <- mu.dd[x] + epsR.dd[x, t]
+        log(sigma[x, t]) <- mu.dd[x] + epsT.dd[t] + epsR.dd[x, t]
         sigma2[x, t] <- sigma[x, t] * sigma[x, t]
         
         # Effective strip width

### 2023-11-30 Analysis_RealData.R
Adjustments to mapping function
This still runs with the archived version of the maptools package, but it will need updating to work with "live" packages.

@@ -1,4 +1,6 @@
 library(tidyverse)
+library(sf)
+library(terra)
 
 # SETUP #
 #-------#

### 2023-11-30 R/plotMaps.R

@@ -211,11 +211,11 @@ plotMaps <- function(mcmc.out, mapNM,
   pdf("Plots/AreaMaps/Avg_rRep_Map.pdf", width = 5, height = 6)
   
   print(
-    tm_shape(mapNM.rRep) + tm_polygons("Median", palette = "plasma", style = "cont", colorNA = "grey80")
+    tmap::tm_shape(mapNM.rRep) + tmap::tm_polygons("Median", palette = "plasma", style = "cont", colorNA = "grey80")
   )
   
   print(
-    tm_shape(mapNM.rRep) + tm_polygons("SD", palette = rev("BuGn"), style = "cont", colorNA = "grey80")
+    tmap::tm_shape(mapNM.rRep) + tmap::tm_polygons("SD", palette = rev("BuGn"), style = "cont", colorNA = "grey80")
   )
   
   dev.off()
@@ -224,11 +224,11 @@ plotMaps <- function(mcmc.out, mapNM,
   pdf("Plots/AreaMaps/Avg_pSurv_Map.pdf", width = 5, height = 6)
   
   print(
-    tm_shape(mapNM.pSurv) + tm_polygons("Median", palette = "plasma", style = "cont", colorNA = "grey80")
+    tmap::tm_shape(mapNM.pSurv) + tmap::tm_polygons("Median", palette = "plasma", style = "cont", colorNA = "grey80")
   )
   
   print(
-    tm_shape(mapNM.pSurv) + tm_polygons("SD", palette = rev("BuGn"), style = "cont", colorNA = "grey80")
+    tmap::tm_shape(mapNM.pSurv) + tmap::tm_polygons("SD", palette = rev("BuGn"), style = "cont", colorNA = "grey80")
   )
   
   dev.off()
@@ -238,11 +238,11 @@ plotMaps <- function(mcmc.out, mapNM,
     pdf("Plots/AreaMaps/betaR_Map.pdf", width = 5, height = 6)
     
     print(
-      tm_shape(mapNM.betaR) + tm_polygons("Median", palette = colorspace::divergingx_hcl(10, palette = "Zissou1"), style = "cont", colorNA = "grey80")
+      tmap::tm_shape(mapNM.betaR) + tmap::tm_polygons("Median", palette = colorspace::divergingx_hcl(10, palette = "Zissou1"), style = "cont", colorNA = "grey80")
     )
     
     print(
-      tm_shape(mapNM.betaR) + tm_polygons("SD", palette = rev("BuGn"), style = "cont", colorNA = "grey80")
[…]

### 2023-11-30 Manuscript_Nilsen_Nater.qmd
Edits to text (.qmd) and bib-file

@@
 Over the last few decades, a tremendous progress have been made towards this end.
-This progress is partly driven the development and application of new field data collection methods and apporaches, such as citizen science data (ref), camera trap data (ref) and the collection of environmental DNA data (ref).
-In addition, developments of new analysis methods allow us to estimate abundance and demographic rates it situations where it was not feasible before (ref).
-combined, these advance put us in a much better position to estimate state variables needed for population management (ref) but also estimate indexes that are relevant for large scale policy applications, e.g.
+This progress is partly driven the development and application of new field data collection methods and approaches, such as citizen science data (ref), camera trap data (ref) and the collection of environmental DNA data (ref).
+In addition, developments of novel statistical methods allow researchers to estimate abundance and demographic rates it situations where it was not feasible before (ref).
+Combined, these advance put us in a much better position to estimate state variables needed for population management (ref) or to estimate indexes that are relevant for large scale policy applications, e.g.
 such as Essential Biodiversity Variables (ref).
@@
 @schmidt2020].
-In this study \[...\]
 
-Here, we will present a novel integrated population model based on distance sampling line transect survey data and demographic data from marked birds.
+In this study, we will present a new IPM, where we integrate data from line transect distance sampling survey data and survival data from marked birds.
 
@@
 Both juveniles and adults survive from year $t$ census to year $t+1$ census with survival probability $S_t$.
-As ptarmigan can reproduce already at 1-year old, all survivors then produce offspring in late June which recruit into the population as juveniles just prior to the census in year $t+1$ according to a recruitment rate $R_{t+1}$.
+As ptarmigan can reproduce already as 1-year old, all survivors then produce offspring in late June which recruit into the population as juveniles just prior to the census in year $t+1$ according to a recruitment rate $R_{t+1}$.
 The changes in densities (numbers) of juveniles and adults in the population, $D_{juv}$ and $D_{ad}$, can thus be expressed as
@@
 Note that recruitment rate $R$ is defined as juveniles/adult (not juveniles/female).
-We also make the simplifying assumption that there is no age- or sex-dependence of vital rates, but this assumption could be relaxed by including additional auxiliary data (Sandercock *et al.* 2011; Israelsen *et al.* 2020).
+We also make the simplifying assumption that there is no age- or sex-dependence of vital rates, but this assumption could be relaxed by including additional auxiliary data [@Israelsen2020; @Sandercock2011].
 
@@
 $$ where $\sigma$ is the half-normal detection parameter.
-We assumed $\sigma$ to vary among years (index $t$) but not transect lines or animal group size.
+We assumed $\sigma$ to vary among years (index $t$) but not between transect lines or animal group size.
 Following @moore2011, the resulting $\sigma_t$ can be used to calculate effective strip width ($esw_{t}$) and, consequently, average detection probability per line transect with a truncation distance $W$ according to:
@@
 \end{equation}
-$$ $N_{juv,j,t}$ and $N_{ad,j,t}$ are then linked back to the population model by converting them to densities through multiplication with $2L_{j,t}W$ (where $L_{j,t}$ is length of transect $j$ in year $t$).
+$$ $N_{juv,j,t}$ and $N_{ad,j,t}$ are then linked back to the population model by converting them to densities through multiplication with $2L_{j,t}W$ (where $L_{j,t}$ is length of transect $j$ in year $t$, and $W$ is the truncation distance).
 
-The third data likelihood focuses on the counts of adults ($obsAd_{j,t}$) and juveniles ($obsJuv_{j,t}$) observed during the distance sampling surveys and links them to time-specific recruitment rate:
+The third data likelihood focuses on the counts of adults ($obsAd_{j,t}$) and juveniles ($obsJuv_{j,t}$) observed during the distance sampling surveys and links them to year-specific recruitment rate:
 
@@
 \end{equation}
-$$ Here, $Sk_t$ is the seasonal survival rate, and annual survival rate, $S_t$ is calculated as $S1_t * S2_t$.
+$$ Here, $Sk_t$ is the seasonal survival probability, and annual survival probability, $S_t$ is calculated as $S1_t * S2_t$.
 
@@
 
-### Field data (case study)
+### Case study
 
-\[Add some basic information about willow ptarmigan in Norway, plus Lierne study area\]
+The willow ptarmigan has a circumpolar distribution (Fuglei et al., 2020), and lives year-round in heterogeneous alpine and artic ecosystems.
+In Norway, there has been a long-term decline in the willow ptarmigan abundance spanning more than a century (ref), but the last few decades the trend in abundance has fluctuated both in time and space.
+In Scandinavia, willow ptarmigan is a valued game species (ref), and there has been several long-term research projects devoted to understanding how they respond to environmental variation and harvest management [@Israelsen2020; @Sandercock2011].
+A key insight from across several study areas is the the annual recruitment rate (i.e.
+$R_t$ in our model, as outlined above) is highly variable, and is affected both by spring conditions [@Eriksen2023] and the abundance of small rodents as alternative prey for shared predators (i.e.
+the Alternative Prey Hypothesis ref).
+Adult survival show less inter-annual fluctuations [@Israelsen2020], although spatial (and potentially temporal) variation due to e.g.
+harvest management is evident when comparing across studies [@Israelsen2020].
 
-The two datasets used for the case study were i) data from a line transect survey program targeting willow ptarmigan operated as part of the natural resources management authorities, and ii) data from an individual based research project based on radio collared willow ptarmigan.
+Our case study was based on an ongoing long-term research project on willow ptarmigan in Lierne municipality in Central Norway (approximately 62.4 degrees north and 13.2 degrees east).
+The study area is located in a sub-alpine ecosystem, and the landscape is a mosaic of open heath and shrub vegetation (dominated by Ericacea, willow shrub *Salix spp*.
+and dwarf birch *Betula nana*), interspersed with bogs and forest patches (mainly birch *Betula spp*.).
+The climate is strongly seasonal, with snow typically covering the ground from October/November through April/May.
+
+From this study system, two datasets were used for the case study:
+
+1.  Data from a line transect survey program targeting willow ptarmigan operated as part of the natural resources management authorities
+
+2.
+Data from an individual based research project based on radio collared willow ptarmigan.
 The temporal coverage of the two data sets partially overlap, with the line transect survey spanning 2007-2019 (and ongoing) and the individual based data spanning 2015-2019 (and ongoing).
@@
 
-\[One or several figures of estimates vs. true values?\]
+In general, the model was able to estimate both parameters of the detection model (Figure X) of the simulated data with little bias.
+
+![](Figures/SimCheck_Detects_Temps.png)
+
+**Figure X** Model performance on simulated data. Upper panel depicts model parameters, and lower panels show derived estimated of effective strip width and mean detection probability
+
+![](Figures/SimCheck_VRs_Temps.png)
+
+**Figure X** Vital rates estimated based on simulated data
 
@@
 
-#### Population density
+```{r}
+#| echo: FALSE
+#| message: FALSE
+#| warning: FALSE
+
+library(tidyverse)
+library(tidybayes)
+
+## Posterior samples added to folder "Posterior_samples". 
+## .rds are included in gitignore
+
+test <- readRDS("Posterior_samples/rypeIDSM_dHN_multiArea_realData_Lierne_tidy.rds")
+
+test1 <- test %>% spread_draws(Mu.R[])
+mean_MuR <- mean(test1$Mu.R)
+ci_MuR <- quantile(test1$Mu.R, probs = c(0.025, 0.975))
+
+```
+
+Having evaluated the overall performance of our model on simulated data, we used data from our case study in Lierne as a case study to estimate abundance, vital rates and detection probabilities from a real-world data set.
+As is evident from Figure X, the population density increased markedly across the study period, from \< 10 ptarmigan / $km^{2}$ in 2007 to \> 35 ptarmigan / $km^{2}$ in in 2021.
+The increase was most distinct from 2016 and onward.
+Across years, mean survival probability for August - January ($S_1$) was estimated at X (95% C.I = ), survival probability for February - July ($S_1$) was estimated as X (95% C.I = x), and annual survival probability $S$ is given by the product of $S_1$ and $S_2$ was estimated at X.
+Note that we assumed that $S$ was constant across all years in the study period from 2007 - 2021.
+Recruitment ($R_t$) was allowed to vary across years (see model specification), and did as expected show large inter-annual variability.
+While the mean (baseline) recruitment $\mu_R$ was estimated as `r round(mean_MuR, 1)` (95% C.I.
+= `r as.numeric(round(ci_MuR[1], 1))` - `r as.numeric(round(ci_MuR[2], 1))` ).
 

### 2023-12-04 Analysis_RealData_LierneVest.R
Completed run of Lierne model for manuscript

@@ -153,14 +153,14 @@ IDSM.out <- nimbleMCMC(code = model_setup$modelCode,
                        setSeed = mySeed)
 Sys.time() - t.start
 
-saveRDS(IDSM.out, file = 'rypeIDSM_dHN_multiArea_realData_Lierne_rodentEffZ.rds')
+saveRDS(IDSM.out, file = 'rypeIDSM_dHN_multiArea_realData_Lierne.rds')
 
 
 # TIDY UP POSTERIOR SAMPLES #
 #---------------------------#
 
 IDSM.out.tidy <- tidySamples(IDSM.out = IDSM.out, save = FALSE)
-saveRDS(IDSM.out.tidy, file = 'rypeIDSM_dHN_multiArea_realData_Lierne_rodentEffZ_tidy.rds')
+saveRDS(IDSM.out.tidy, file = 'rypeIDSM_dHN_multiArea_realData_Lierne_tidy.rds')
 
 
 # OPTIONAL: MCMC TRACE PLOTS #

### 2023-12-04 Analysis_RealData.R
Moved area-specific posterior summaries into separate function
These now get used by >1 function downstream, so makes sense to calculate them as a separate step in workflow.

@@ -96,7 +96,7 @@ d_rodent <- wrangleData_Rodent(duplTransects = duplTransects,
 input_data <- prepareInputData(d_trans = LT_data$d_trans, 
                                d_obs = LT_data$d_obs,
                                d_cmr = d_cmr,
-                               d_rodent = d_rodent,
+                               d_rodent = d_rodent$rodentAvg,
                                #localities = localities, 
                                areas = areas,
                                areaAggregation = TRUE,
@@ -151,6 +151,21 @@ saveRDS(IDSM.out, file = "rypeIDSM_dHN_multiArea_realData_Lierne.rds")
 IDSM.out.tidy <- tidySamples(IDSM.out = IDSM.out, 
                              save = TRUE,
                              fileName = "rypeIDSM_dHN_multiArea_realData_allAreas_tidy.rds")
+
+
+
+# MAKE POSTERIOR SUMMARIES PER AREA #
+#-----------------------------------#
+
+PostSum.list <- summarisePost_areas(mcmc.out = IDSM.out.tidy, 
+                                    N_areas = input_data$nim.constant$N_areas, 
+                                    area_names = input_data$nim.constant$area_names, 
+                                    N_sites = input_data$nim.constant$N_sites, 
+                                    min_years = input_data$nim.constant$min_years, 
+                                    max_years = input_data$nim.constant$max_years, 
+                                    minYear = minYear, maxYear = maxYear,
+                                    fitRodentCov = fitRodentCov,
+                                    save = TRUE)
 
 
 # OPTIONAL: MCMC TRACE PLOTS #
@@ -211,27 +226,17 @@ NorwayMunic.map <- setupMap_NorwayMunic(shp.path = "data/Kommuner_2018_WGS84/Kom
                                         areas = areas, areaAggregation = TRUE)
 
 ## Plot population growth rate, density, and vital rates on map
-plotMaps(mcmc.out = IDSM.out.tidy, 
+plotMaps(PostSum.list = PostSum.list, 
          mapNM = NorwayMunic.map,
-         N_areas = input_data$nim.constant$N_areas, 
-         area_names = input_data$nim.constant$area_names, 
-         N_sites = input_data$nim.constant$N_sites, 
-         min_years = input_data$nim.constant$min_years, 
-         max_years = input_data$nim.constant$max_years, 
          minYear = minYear, maxYear = maxYear,
[…]

### 2023-12-04 R/summarisePost_areas.R

@@ -1,0 +1,201 @@
+#' Summarises posteriors for average vital rates, population densities, and 
+#' population growth rates per area
+#'
+#' @param mcmc.out an mcmc list containing posterior samples from a fitted model.
+#' @param N_areas integer. Number of areas in the analysis. 
+#' @param area_names character vector (length = N_areas) containing names of all
+#' areas in the analysis. 
+#' @param N_sites integer vector (length = N_areas) specifying the number of 
+#' sites per area. 
+#' @param min_years integer vector (length = N_areas) specifying the time index 
+#' for the first year with data for each area. 
+#' @param max_years integer vector (length = N_areas) specifying the time index 
+#' for the first year with data for each area. 
+#' @param minYear integer. First year included in analysis. 
+#' @param maxYear integer. Last year included in analysis.
+#' @param fitRodentCov logical. If TRUE, makes posterir summaries for rodent 
+#' effect in addition to other parameters. 
+#' @param save logical. If TRUE (default), saves the posterior summaries in an 
+#' RDS file ()
+#' 
+#' @return a list containing 5 dataframes with summarised posteriors for several
+#' parameters by area: average recruitment (rRep.sum), average survival 
+#' (pSurv.sum), rodent effect slope (betaR.sum, = NA when fitRodentCov = FALSE), 
+#' average population density (popDens.sum), and population growth rate 
+#' (lambda.sum). 
+#' @export
+#'
+#' @examples
+#' 
+summarisePost_areas <- function(mcmc.out, 
+                                N_areas, area_names, N_sites, 
+                                min_years, max_years, minYear, maxYear,
+                                fitRodentCov, save = TRUE){
+  
+  
+  ## Convert posterior samples to matrix format
+  out.mat <- as.matrix(mcmc.out)
+  
+  ## Set year index ranges for plotting shared data collection period only
+  minYearIdx_shared <- max(min_years)
+  maxYearIdx_shared <- min(max_years)
+  
+  ## Determine area-specific year range
+  area_yearIdxs <- (1:(maxYear-minYear+1))
+  area_years <- area_yearIdxs + (minYear - 1)
+  
+  ## Calculate area- and year-specific population growth rates
+  for(i in 1:N_areas){
[…]

### 2023-12-04 R/plotMaps.R
Updated map plotting function

@@ -1,195 +1,66 @@
+#' Plot posterior summaries for vital rate and population parameters on map
+#'
+#' @param PostSum.list a list containing 5 dataframes with summarised posteriors for several
+#' parameters by area: average recruitment (rRep.sum), average survival 
+#' (pSurv.sum), rodent effect slope (betaR.sum, = NA when fitRodentCov = FALSE), 
+#' average population density (popDens.sum), and population growth rate 
+#' (lambda.sum). 
+#' @param mapNM sf / dataframe object containing map of Norwegian municipalities. 
+#' @param minYear integer. First year in the analysis.
+#' @param maxYear integer. Last year in the analysis.
+#' @param fitRodentCov logical. If TRUE, makes plot for rodent effect slope in 
+#' addition to other plots. 
+#'
+#' @return a vector of pdf plot names. The plots can be found in Plots/AreaMaps.
+#' @export
+#'
+#' @examples
 
-plotMaps <- function(mcmc.out, mapNM,
-                     N_areas, area_names, N_sites, 
-                     min_years, max_years, minYear, maxYear,
+plotMaps <- function(PostSum.list, mapNM,
+                     minYear, maxYear,
                      fitRodentCov){
   
-  ## Convert posterior samples to matrix format
-  out.mat <- as.matrix(mcmc.out)
-  
-  ## Set year index ranges for plotting shared data collection period only
-  minYearIdx_shared <- max(min_years)
-  maxYearIdx_shared <- min(max_years)
-  
-  ## Determine area-specific year range
-  area_yearIdxs <- (1:(maxYear-minYear+1))
-  area_years <- area_yearIdxs + (minYear - 1)
-  
-  ## Calculate area- and year-specific population growth rates
-  for(i in 1:N_areas){
-    for(t in 1:length(area_years)){
-      
-      # Summarize annual average population densities (per area, averaged over lines)
-      popDens_juv_t0 <- out.mat[, paste0("Density[",  i, ", 1, ", 1:N_sites[i], ", ", area_yearIdxs[t], "]")]
-      popDens_ad_t0 <- out.mat[, paste0("Density[",  i, ", 2, ", 1:N_sites[i], ", ", area_yearIdxs[t], "]")]
-      popDens_mean_t0 <- rowMeans(popDens_juv_t0 + popDens_ad_t0)
-      
[…]

### 2023-12-04 R/plotLatitude.R
Completed latitude plotting function

@@ -1,190 +1,151 @@
-plotLatitude <- function(mcmc.out, 
-                         N_areas, area_names, area_coord, N_sites, 
-                         min_years, max_years, minYear, maxYear,
+#' Plot posterior summaries for vital rate and population parameters as a function of latitude
+#'
+#' @param PostSum.list a list containing 5 dataframes with summarised posteriors for several
+#' parameters by area: average recruitment (rRep.sum), average survival 
+#' (pSurv.sum), rodent effect slope (betaR.sum, = NA when fitRodentCov = FALSE), 
+#' average population density (popDens.sum), and population growth rate 
+#' (lambda.sum). 
+#' @param area_coord a tibble containing coordinate data for each area in analysis.
+#' @param minYear integer. First year in the analysis.
+#' @param maxYear integer. Last year in the analysis.
+#' @param fitRodentCov logical. If TRUE, makes plot for rodent effect slope in 
+#' addition to other plots. 
+#'
+#' @return a vector of pdf plot names. The plots can be found in Plots/Latitude.
+#' @export
+#'
+#' @examples
+
+plotLatitude <- function(PostSum.list, 
+                         area_coord,
+                         minYear, maxYear,
                          fitRodentCov){
   
-  
-  ## Convert posterior samples to matrix format
-  out.mat <- as.matrix(mcmc.out)
-  
-  ## Set year index ranges for plotting shared data collection period only
-  minYearIdx_shared <- max(min_years)
-  maxYearIdx_shared <- min(max_years)
-  
-  ## Determine area-specific year range
-  area_yearIdxs <- (1:(maxYear-minYear+1))
-  area_years <- area_yearIdxs + (minYear - 1)
-  
-  ## Calculate area- and year-specific population growth rates
-  for(i in 1:N_areas){
-    for(t in 1:length(area_years)){
-      
-      # Summarize annual average population densities (per area, averaged over lines)
-      popDens_juv_t0 <- out.mat[, paste0("Density[",  i, ", 1, ", 1:N_sites[i], ", ", area_yearIdxs[t], "]")]
-      popDens_ad_t0 <- out.mat[, paste0("Density[",  i, ", 2, ", 1:N_sites[i], ", ", area_yearIdxs[t], "]")]
[…]

### 2023-12-04 R/plotPosteriorDens_VR.R
Completed roxygen skeletons

@@ -10,7 +10,8 @@
 #' to time-average survival.
 #' @param fitRodentCov logical. If TRUE, interprets recruitment intercept as
 #' baseline instead of average.
-#' @return
+#' 
+#' @return a vector of pdf plot names. The plots can be found in Plots/VR_PostDens.
 #' @export
 #'
 #' @examples

### 2023-12-04 R/setupMap_NorwayMunic.R

@@ -1,3 +1,16 @@
+#' Set up map object containing Norwegian municipalities
+#'
+#' @param shp.path character string. Path to where the shapefile for Norwegian
+#' municipalities is stored. 
+#' @param d_trans tibble containing information on transects (events). Output of
+#' wrangleData_LineTrans(). 
+#' @param areas character vector containing names of all areas in analysis. 
+#' @param areaAggregation 
+#'
+#' @return sf / dataframe object containing map of Norwegian municipalities. 
+#' @export
+#'
+#' @examples
 
 setupMap_NorwayMunic <- function(shp.path, d_trans,
                                  areas, areaAggregation = TRUE){

### 2023-12-04 tidySummary.R
Added summary function based on tidy_draws

@@ -1,0 +1,116 @@
+
+library(tidybayes)
+library(tidyverse)
+
+
+
+#' Calculate summary stats for selected variables
+#'
+#' @param out.mcmc 
+#'
+#' @return a tibble containg summary stats for selected variables
+#' @export
+#'
+#' @examples
+
+tidySummary <- function(out.mcmc){
+
+
+################################################################################
+### Summarize muR
+muR_tidy <- out.mcmc %>% tidybayes::spread_draws(Mu.R[area], 
+                                                 sep = "[,]", regex = FALSE) %>% 
+  group_by() %>%
+  summarise(Mean = mean(Mu.R), 
+            std = sd(Mu.R), 
+            Median = median(Mu.R), 
+            lower = quantile(Mu.R, probs = 0.025), 
+            upper = quantile(Mu.R, probs = 0.975)) %>%
+            mutate(model_parameter = "Mu_R")
+
+################################################################################
+### Summarize mu.S
+
+muS_tidy <- out.mcmc %>% tidybayes::spread_draws(Mu.S[area], 
+                                                 sep = "[,]", regex = FALSE) %>% 
+  group_by() %>%
+  summarise(Mean = mean(Mu.S), 
+            std = sd(Mu.S), 
+            Median = median(Mu.S), 
+            lower = quantile(Mu.S, probs = 0.025), 
+            upper = quantile(Mu.S, probs = 0.975)) %>%
+            mutate(model_parameter = "Mu_S")
+
+################################################################################
+### Summarize mu.S1
+
+muS1_tidy <- out.mcmc %>% tidybayes::spread_draws(Mu.S1, 
+                                                 sep = "[,]", regex = FALSE) %>% 
+  group_by() %>%
+  summarise(Mean = mean(Mu.S1), 
+            std = sd(Mu.S1), 
+            Median = median(Mu.S1), 
+            lower = quantile(Mu.S1, probs = 0.025), 
+            upper = quantile(Mu.S1, probs = 0.975)) %>%
+  mutate(model_parameter = "Mu_S1")
+
+################################################################################
+### Summarize mu.S2
+
+muS2_tidy <- out.mcmc %>% tidybayes::spread_draws(Mu.S[], Mu.S1, 
+                                                 sep = "[,]", regex = FALSE) %>% 
+  mutate(Mu.S2 = Mu.S/Mu.S1) %>%
+  group_by() %>%
+  summarise(Mean = mean(Mu.S2), 
[…]

### 2023-12-04 Manuscript_Nilsen_Nater.qmd
Updates to ms and .bib file

@@
 
-## Introduction
+## Abstract
+
+*Estimation of abundance and demographic rates for populations of wild vertebrate species is a challenging but fundamental issue for both management and research into ecology and evolution.
+One approach that has been used extensively to estimate abundance of wild living populations is Distance Sampling (DS) methods based on line transect survey data.
+Historically, DS models were only available as open population models, and did not allow for the direct estimation of changes in abundance through time.
+The advent of open population formulations based on the DS framework greatly extended the scope of the models, but so far models that estimate both temporal dynamics in abundance as well as the underlying demographic rates has not been implemented.
+Here, we present an integrated distance sampling apporach, that utilize age-structured survey data and auxiliary data from marked individuals to jointly estimate population dynamics and the temporal variation in the demographic rates (recruitment rate and survival probability) that determine the temporal transition.
+The underlying process model is based on a two-stage transition matrix (*$A_t$*), with matrix entries given by survival probabilities (*$S$*) and time-specific recruitment rates (*$R_t$*).
+This framework allow us to make full use of the available data, and to effectively integrate the two data sources in an integrated modelling framework.
+Moreover, demographic rates often respond to environmental variation, and our approach allow us to directly estimate the effect of such environmental covariates on demographic rate variation.
+We first fit the model to simulated data, to assess it's ability to recover the underlying population dynamics when the true values are known.
+Then, we use data from a study of willow ptarmigan (Lagopus lagopus) in Norway as a case study.*
+
+## 1. Introduction
 
@@
 Combined, these advance put us in a much better position to estimate state variables needed for population management (ref) or to estimate indexes that are relevant for large scale policy applications, e.g.
-such as Essential Biodiversity Variables (ref).
+such as Essential Biodiversity Variables [@Kissling2018].
 
-Nevertheless, large scale and long term data sets are still relatively rare (ref), due to high costs of field data collection.
+Until recently, joint estimation of population dynamics and demography has relied mostly on data from marked individuals through application of open-population capture-mark-recapture models (ref).
+While such methods can provide valuable information for both ecological research and management, they are also in many cases costly to implement over large areas.
+On the other hand, data from monitoring of abundance trends over time typically are based on data from unmarked animals (ref).
+A common approach for such surveys is to structure data collection around the distance sampling (DS) method.
+DS have been used for estimating animal abundance in a wide range of contexts and for a variety of taxa [@buckland2015].
+One reason for their popularity might be that they does not require repeated visits to the same sites to estimate detection probability (ref).
+Rather, it is assumed that detection probability decrease as the objects are located longer away from the transect line (ref).
+This makes them particularily useful for implementation in participatory monitoring program, allowing stakeholders to take part in the data collection process.
 
-A cost efficient and well documented approach suitable is the use of line transect surveys. However, because the detectability is often less than 1 and variable between years and habitats,
+The most common implementation of the DS model has been to use close-population formulations, that does not allow to formulate a process model that project abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates (ref).
+In recent years, DS approahces have been extended and also applied for estimating changes in abundance over time in open populations by embedding a hidden state model representing population dynamics [@moore2011; @sollmann2015].
+Such approaches has greatly extended the application of DS approaches to study ecological dynamics in time and space.
+However, while these latter frameworks may allow to accurately quantify population changes but provide little information on the drivers of these changes, i.e.
+the underlying vital rates.
+In fact, if the data does not contain information about the age (and/or sex-) structure of the surveyed population, there is no stratight forward way to estimate demographic rates from such data.
+In cases where the age (and sex) of the observed individuals can be observed, this information can under some assumptions be used to inform about recruitment rates and survival probabilities (ref).
+For instance, in a integrated population model based on harvest statistics and observations of population age structure, both population abundance and demographic rates was estimated without even without additional data from marked individuals [@Nilsen2018].
 
--   Importance of abundance trends and demography
-
--   Short about methods -- detection probability -- marked vs unmarked animals
-
--   Approaches to estimate abundance and demography from unmarked animals
-
-Distance sampling (DS) models have been used for estimating animal abundance in a wide range of contexts and for a variety of taxa [@buckland2015].
-In recent years, they have been extended and also applied for estimating changes in abundance over time in open populations by embedding a hidden state model representing population dynamics [@moore2011; @sollmann2015].
-These latter frameworks may allow to accurately quantify population changes but provide little information on the drivers of these changes, i.e.
-the underlying vital rates.
 Concurrent with the development of more sophisticated DS models, another group of models has emerged and gained popularity, not least for their ability to disentangle demographic processes underlying population dynamic: integrated population models [IPMs, @schaub2021].
@@
 In this study, we will present a new IPM, where we integrate data from line transect distance sampling survey data and survival data from marked birds.
+The process model is a stage-structured matrix model, that projects population size from one time step to the next based on the underlying survival and recruitment rates.
+This is made possible by the fact that the line transect surveys provide data on population age structure which can be used to inform the estimation of demographic rates.
+In the IPM, we integrate these data with data from marked birds from which we estimate survival probabilities.
+The model is thus formally an IPM, as we jointly estimate parameters of interest based on several data sets (ref).
+In the paper, we first present the model and assess the robustness based on simulated data.
+In addition, we apply the model to real data collected from a willow ptarmigan (*Lagopus lagopus*) study in Central Norway.
+Because demographic rates are often affected by environmental covariates, we also illustrate how such covariates can be included in the modelling framework.
 
--   Estimate parameters with and without augmentation with demographic data -- compare estimates and precision
--   Estimate latent variable (juv summer survival) -- compare to literature values
--   Temporal variation in S and R -- via random effects models
--   Simulate - robustness
+## 2. Methods
 
-## Methods
-
-### An integrated distance sampling model
+### 2.1 An integrated distance sampling model
 
@@
 
-#### Age-structured population model
+#### 2.1.1 Age-structured population model
 
@@
 
-#### Likelihoods for distance sampling data
+#### 2.1.2 Likelihoods for distance sampling data
 
@@
 
-#### Likelihood for radio-telemetry data
+#### 2.1.3 Likelihood for radio-telemetry data
 
@@
 
-#### Hierarchical models with time-variation in parameters
+#### 2.1.3 Hierarchical models with time-variation in parameters
 
@@
 
-### Model testing with simulated data
+### 2.2 Model testing with simulated data
 
@@
 
-### Case study
+### 2.3 Case study
 
@@
 
-\[Figure 2: Map of Lierne study area?\]
-
-### Bayesian implementation
+### 2.4 Bayesian implementation
 
@@
 
-## Results
+## 3. Results
 
-### Model performance on simulated datasets
+### 3.1 Model performance on simulated datasets
 
@@
 
-### Case study on willow ptarmigans in Lierne
+### 3.2 Case study on willow ptarmigans in Lierne
 
@@
 library(tidybayes)
+source("tidySummary.R")
 
@@
 
-test <- readRDS("Posterior_samples/rypeIDSM_dHN_multiArea_realData_Lierne_tidy.rds")
+out.data <- readRDS("Posterior_samples/rypeIDSM_dHN_multiArea_realData_Lierne_tidy.rds")
 
-test1 <- test %>% spread_draws(Mu.R[])
-mean_MuR <- mean(test1$Mu.R)
-ci_MuR <- quantile(test1$Mu.R, probs = c(0.025, 0.975))
+tidy_summary_stats <- tidySummary(out.data)
 
@@
 The increase was most distinct from 2016 and onward.
-Across years, mean survival probability for August - January ($S_1$) was estimated at X (95% C.I = ), survival probability for February - July ($S_1$) was estimated as X (95% C.I = x), and annual survival probability $S$ is given by the product of $S_1$ and $S_2$ was estimated at X.
+Across years, mean survival probability for August - January ($S_1$) was estimated at `r filter(tidy_summary_stats, model_parameter =="Mu_S1")$Mean` (95% C.I = `r filter(tidy_summary_stats, model_parameter =="Mu_S1")$lower` - `r filter(tidy_summary_stats, model_parameter =="Mu_S1")$upper`), survival probability for February - July ($S_2$) was estimated as `r filter(tidy_summary_stats, model_parameter =="Mu_S2")$Mean` (95% C.I = `r filter(tidy_summary_stats, model_parameter =="Mu_S2")$lower` - `r filter(tidy_summary_stats, model_parameter =="Mu_S2")$upper`), and annual survival probability $S$ is given by the product of $S_1$ and $S_2$ was estimated at `r filter(tidy_summary_stats, model_parameter =="Mu_S")$Mean` (95% C.I = `r filter(tidy_summary_stats, model_parameter =="Mu_S")$lower` - `r filter(tidy_summary_stats, model_parameter =="Mu_S")$upper`).
 Note that we assumed that $S$ was constant across all years in the study period from 2007 - 2021.
 Recruitment ($R_t$) was allowed to vary across years (see model specification), and did as expected show large inter-annual variability.
-While the mean (baseline) recruitment $\mu_R$ was estimated as `r round(mean_MuR, 1)` (95% C.I.
-= `r as.numeric(round(ci_MuR[1], 1))` - `r as.numeric(round(ci_MuR[2], 1))` ).
+While the mean (baseline) recruitment $\mu_R$ was estimated as `r filter(tidy_summary_stats, model_parameter =="Mu_R")$Mean` (95% C.I = `r filter(tidy_summary_stats, model_parameter =="Mu_R")$lower` - `r filter(tidy_summary_stats, model_parameter =="Mu_R")$upper`).
 

### 2023-12-04 tidySummary.R
Added additional summaries for rodent covariate effect size

@@ -102,12 +102,25 @@ R_year_tidy <- out.mcmc %>% tidybayes::spread_draws(R_year[area, year], sep = "[
   ungroup() %>%
   mutate(model_parameter = "R_year")
 
+################################################################################
+##### Effect of rodent covariate betaR.R
+
+betaR.R_tidy <- out.mcmc %>% tidybayes::spread_draws(betaR.R[area], 
+                                                     sep = "[,]", regex = FALSE) %>% 
+  group_by() %>%
+  summarise(Mean = mean(betaR.R), 
+            std = sd(betaR.R), 
+            Median = median(betaR.R), 
+            lower = quantile(betaR.R, probs = 0.025), 
+            upper = quantile(betaR.R, probs = 0.975)) %>%
+  mutate(model_parameter = "betaR.R")
+
 
 ################################################################################
 #### Putting together output
 
 pooled_results <- bind_rows(dens_tidy, R_year_tidy, muR_tidy, 
-                            muS_tidy, muS1_tidy, muS2_tidy)
+                            muS_tidy, muS1_tidy, muS2_tidy, betaR.R_tidy)
 
 
 pooled_results

### 2023-12-04 Manuscript_Nilsen_Nater.qmd
Edits to manuscript

@@
 
-### 3.2 Case study on willow ptarmigans in Lierne
+### 3.2 Case study on willow ptarmigans in Central Norway
 
@@
 Note that we assumed that $S$ was constant across all years in the study period from 2007 - 2021.
+
 Recruitment ($R_t$) was allowed to vary across years (see model specification), and did as expected show large inter-annual variability.
 While the mean (baseline) recruitment $\mu_R$ was estimated as `r filter(tidy_summary_stats, model_parameter =="Mu_R")$Mean` (95% C.I = `r filter(tidy_summary_stats, model_parameter =="Mu_R")$lower` - `r filter(tidy_summary_stats, model_parameter =="Mu_R")$upper`).
+The estimated recruitment rate varied (as expected) among years: while the annual recruitment rate was estimated at `r max(filter(tidy_summary_stats, model_parameter =="R_year")$Mean)` in year 2007, it was estimated at `r min(filter(tidy_summary_stats, model_parameter =="R_year")$Mean)` in year 2012.
+Moreover, as expected based on the Alternative Prey Hypothesis (APH), our model estimated a strong link between the annual recruitment and the abundance of small rodents (slope-paramater for the z-standardized rodent occurrence data = `r filter(tidy_summary_stats, model_parameter =="betaR.R")$Mean` ; 95% C.I.
+= `r filter(tidy_summary_stats, model_parameter =="betaR.R")$lower` - `r filter(tidy_summary_stats, model_parameter =="betaR.R")$upper`).
 
 ![**Figure X.** Estimated density of willow ptarmigans in Lierne from 2007 to 2021. Solid line represents the posterior median, ribbon marks 95% credible interval.](Figures/TimeSeries_popDens.png)
-
-#### Vital rates
 

### 2023-12-04 R/setupModel.R
Added missing monitor for sigma.h.dd

@@ -37,7 +37,7 @@ setupModel <- function(modelCode, customDist,
   ## Set parameters to monitor
   params <- c("esw", "p", #"D",
               "R_year", "Mu.R", "h.Mu.R", "h.sigma.R", "sigmaT.R", "sigmaR.R",
-              "sigma", "mu.dd", "sigmaT.dd", "sigmaR.dd",
+              "sigma", "mu.dd", "h.sigma.dd", "sigmaT.dd", "sigmaR.dd",
               "Density", "N_exp", "N_tot_exp",
               "Mu.D1", "sigma.D",
               "S", "Mu.S", "h.Mu.S", "h.sigma.S",

### 2023-12-04 Analysis_RealData.R
Wrote function to perform and plot variance decomposition

@@ -217,6 +217,16 @@ if(fitRodentCov){
 }
 
 
+# OPTIONAL: CALCULATE AND PLOT VARIANCE DECOMPOSITION #
+#-----------------------------------------------------#
+
+VarDecomp <- plotVarDecomposition(mcmc.out = IDSM.out.tidy, 
+                                  N_areas = input_data$nim.constants$N_areas, 
+                                  N_years = input_data$nim.constants$N_years, 
+                                  fitRodentCov = fitRodentCov, 
+                                  RodentOcc_data = input_data$nim.data$RodentOcc)
+  
+
 # OPTIONAL: MAP PLOTS #
 #---------------------#
 

### 2023-12-04 R/plotVarDecomposition.R

@@ -1,0 +1,137 @@
+#' Decompose variance by component and plot proportions
+#'
+#' @param mcmc.out n mcmc list containing posterior samples from a model run.
+#' @param N_areas integer. Number of areas in the analysis.
+#' @param N_years integer. Number of years in the analysis.
+#' @param fitRodentCov logical. Indicates whether (TRUE) or not (FALSE) variance
+#' due to effect of a rodent covariate should be considered.
+#' @param RodentOcc_data numeric vector containing the rodent occupancy data
+#' used for fitting the model. Optional argument that needs to be provided if
+#' fitRodenCov = TRUE.
+#'
+#' @return a list containing a dataframe of posterior distributions for proportions
+#' of variance explained by different components, as well as the path to the 
+#' plot. 
+#' @export
+#'
+#' @examples
+
+plotVarDecomposition <- function(mcmc.out, N_areas, N_years, fitRodentCov, RodentOcc_data = NULL){
+  
+  ## Check rodent data is provided if needed
+  if(fitRodentCov & missing("RodentOcc_data")){
+    stop("RodentOcc_data has to be provided if fitRodentCov = TRUE.")
+  }
+  
+  ## Convert posterior samples into matrix
+  sam.mat <- as.matrix(mcmc.out)
+  nsamples <- nrow(sam.mat)
+  
+  ## Extract posterior distributions for random variance components
+  # Random year variation
+  varR_year <- sam.mat[, "sigmaT.R"]^2
+  varS_year <- sam.mat[, "sigmaT.S"]^2
+  varDet_year <- sam.mat[, "sigmaT.dd"]^2
+  
+  # Random area variation
+  varR_area <- sam.mat[, "h.sigma.R"]^2
+  varS_area <- sam.mat[, "h.sigma.S"]^2
+  varDet_area <- sam.mat[, "h.sigma.dd"]^2
+  
+  # Residual variation
+  varR_res <- sam.mat[, "sigmaR.R"]^2
+  varS_res <- sam.mat[, "sigmaR.S"]^2
+  varDet_res <- sam.mat[, "sigmaR.dd"]^2
+  
+  
+  ## Calculate posterior distribution for variance due to area-specific rodent effect
+  if(fitRodentCov){
+    
+    varR_rodent <- rep(NA, nsamples)
+    
+    for(i in 1:nsamples){
+      
+      rodent_Eff <- c()
+      
+      for(x in 1:N_areas){
+        
[…]

### 2023-12-05 Manuscript_Nater_etal.qmd
Set up manuscript structure with bullet points and text to recycle

@@
 
-## About this worknote
+# Abstract
 
-The purpose of this worknote is to introduce and showcase preliminary results from application of the open-population integrated distance-sampling model (IDSM) for willow ptarmigan to data from all of Norway.
-Eventually, this model extension and resulting insights into large-scale spatiotemporal variation in density, population growth rate, and vital rates will be the subject of a second scientific paper.
-This paper will be written and submitted in the second half of 2023, and this worknote serves as a temporary substitute for it in the reporting of project results to Miljødirektoratet in June 2022.
-As such, the worknote contains a general description of the extension of the model to analyse data from multiple areas simultaneously, visualizations of some preliminary results, and an introduction to a reproducible workflow that allows to easily and robustly re-run analyses as new data becomes available in future years.
+# Introduction
 
-## IDSM extension to multiple areas
+# Methods
+
+## Study species
+
+-   Willow ptarmigan distribution, life history, and (historical) abundance
+
+## Data collection, management, and preparation
+
+### Line transect sampling
+
+-   Setup of Hønsefugltakseringen monitoring programme (volunteer training, line transect sampling, spatio-temporal coverage, etc.)
+-   Data & metadata recording via app
+-   Quality control of data
+-   Open access publishing of data (GBIF, LivingNorway)
+-   Pre-modelling data wrangling
 
 Distance-sampling data on willow ptarmigan is collected on line transects not only in western Lierne, but across most of its Norwegian range (https://honsefugl.nina.no/).
-Taking the line transect datasets from Fjellstyrene, Statskog, and FeFo together, there is data for willow ptarmigan in 41 reporting areas (for a list, see [here]( https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp/blob/main/R/listAreas.R)).
+Taking the line transect datasets from Fjellstyrene, Statskog, and FeFo together, there is data for willow ptarmigan in 41 reporting areas (for a list, see [here](https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp/blob/main/R/listAreas.R)).
+
+### Radio-telemetry study in Lierne
+
+-   Description of setup, data collection, temporal coverage, etc.
+-   Pre-modelling data wrangling
+
+## National-scale integrated model
+
+### Integrated distance sampling model (IDSM) for willow ptarmigan
+
+-   Introduction of integrated Distance sampling model (IDSM) from MS1
+
+### Multi-area model extension
+
+-   Addition of spatial dimension
+-   Hyper-parameter distributions for spatial variation
+-   Temporal variation (shared year RE and "residual" variation)
+
 The model developed as part of this project and presented in @nilsen2023 was designed with the potential of extension to run on data from multiple areas simultaneously in mind, and this extension involves two steps in practice.
@@
 Another approach to maximising information gain when jointly analysing data from different areas is explicit modelling of spatial and spatio-temporal variation.
-While this is commonly done for modern species distribution models [e.g.
-@pacifici2017; @guelat2018], it has rarely been implemented in demographic models.
-This is not least due to the fact that few demographic models span large spatial extents.
-Since our IDSM and the willow ptarmigan data do actually do that though, extending the framework to account for spatial and spatiotemporal correlations may be feasible.
-Per now, we have not yet attempted this, but are planning to do so in the near future.
 
-## Preliminary results
+### Model implementation
 
-In the following, we present some preliminary results from the model fit to data from all 41 reporting areas simultaneously.
-Results are presented in the format of pairs of maps.
-For each quantity/result, we present one map of the posterior median and one map of the standard deviation of the posterior distribution, where the latter serves as a measurement of uncertainty for the former.
+-   Bayesian, NIMBLE
+-   Priors
+-   Default samplers
+-   Chains, iterations, burn-in, thinning, posterior samples
 
-### Population density
+## Variance decomposition
+
+-   How variance decomposition was done
+
+## Reproducible workflow with "targets"
+
+Reproducibility and ease of repeating analyses was a key focus when developing the multi-area IDSM.To that end, we set up the workflow as a "targets pipeline", implemented throught the R packages "targets" [@landau2021].
+A visual representation of the pipeline is displayed in @fig-workflow.
+
+![Graphical representation of the "targets pipeline" for the multi-area modelling setup.
+Upward facing triangles are functions, downward facing triangles are general options/arguments, circles are objects and outputs (="targets") created as part of the workflow.](Figures/Workflow_layout.png){#fig-workflow width="424"}
+
+# Results
+
+## Population density
 
@@
 
-### Population growth rate
+## Population growth rate
 
@@
 
-### Survival probability
+## Survival probability
 
@@
 ::: {#fig-pSurv layout-ncol="2"}
-
 ![Median survival estimates](Figures/pSurv_Median.png){#fig-pSurvA}
@@
 
-### Recruitment rate
+## Recruitment rate
 
@@
 
-### Effect of rodent occupancy
+## Effects of rodent occupancy
 
@@
 
-## Reproducible workflow with "targets"
+## Detection parameters
 
-Reproducibility and ease of repeating analyses was a key focus when developing the multi-area IDSM.To that end, we set up the workflow as a "targets pipeline", implemented throught the R packages "targets" [@landau2021].
-A visual representation of the pipeline is displayed in @fig-workflow.
+\[TBA\]
 
-![Graphical representation of the "targets pipeline" for the multi-area modelling setup.
-Upward facing triangles are functions, downward facing triangles are general options/arguments, circles are objects and outputs (="targets") created as part of the workflow.
-](Figures/Workflow_layout.png){#fig-workflow width="424"}
+## Variance decomposition
 
-## Data and code availability
+\[TBA\]
+
+# Discussion
+
+\[TBA\]
+
+Another approach to maximising information gain when jointly analysing data from different areas is explicit modelling of spatial and spatio-temporal variation.
+While this is commonly done for modern species distribution models [e.g.
+@pacifici2017; @guelat2018], it has rarely been implemented in demographic models.
+This is not least due to the fact that few demographic models span large spatial extents.
+Since our IDSM and the willow ptarmigan data do actually do that though, extending the framework to account for spatial and spatiotemporal correlations may be feasible.
+Per now, we have not yet attempted this, but are planning to do so in the near future.
+
+# Acknowledgements
+
+\[TBA\]
+
+# Data and code availability
 
@@
 
-The auxiliary radio-telemetry data and all code, including the "targets pipeline" can be found in the project's repository on GitHub: [https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp](https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp).
+The auxiliary radio-telemetry data and all code, including the "targets pipeline" can be found in the project's repository on GitHub: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.
 
-## References
+# References

### 2023-12-06 Manuscript_Nilsen_Nater.qmd
Updated text and bib-file

--- 2023-12-04 Manuscript_Nilsen_Nater.qmd
@@
 The advent of open population formulations based on the DS framework greatly extended the scope of the models, but so far models that estimate both temporal dynamics in abundance as well as the underlying demographic rates has not been implemented.
-Here, we present an integrated distance sampling apporach, that utilize age-structured survey data and auxiliary data from marked individuals to jointly estimate population dynamics and the temporal variation in the demographic rates (recruitment rate and survival probability) that determine the temporal transition.
+Here, we present an integrated distance sampling approach, that utilize age-structured survey data and auxiliary data from marked individuals to jointly estimate population dynamics and the temporal variation in the demographic rates (recruitment rate and survival probability) that determine the temporal transition.
 The underlying process model is based on a two-stage transition matrix (*$A_t$*), with matrix entries given by survival probabilities (*$S$*) and time-specific recruitment rates (*$R_t$*).
@@
 
-Our open population integrated distance sampling model consists of two major components: a latent structured population model and a set of likelihoods for data originating from distance sampling surveys and auxiliary survival monitoring (radio-telemetry in the example case).
+Our open population integrated distance sampling model consists of two major components: a latent structured population model and a set of likelihoods for data originating from distance sampling surveys and auxiliary survival monitoring.
+In the example case, these auxiliary data come from a radio-telemetry study, but in principle other types of capture-recapture data can also be used.
 
@@
 Data from an individual based research project based on radio collared willow ptarmigan.
-The temporal coverage of the two data sets partially overlap, with the line transect survey spanning 2007-2019 (and ongoing) and the individual based data spanning 2015-2019 (and ongoing).
+The temporal coverage of the two data sets partially overlap, with the line transect survey spanning 2007-2021 (and ongoing) and the individual based data spanning 2015-2021 (and ongoing).
 
-Line transect survey data were collected in August each year, prior to the annual autumn harvest season.
+Line transect survey data were collected in August each year, prior to the annual autumn harvest season, as part of the program "Hønsefuglportalen".
+Hønsefuglportalen is a national program for line transect surveys of tetraonid birds, and the effort is directed mainly to willow ptarmigan habitats.
+In our case study, we used data from the western part of Lierne municipality.
 Line transects are surveyed by educated volunteers, using pointing dogs to located the birds.
 When located, the geographical coordinate, perpendicular distance from the sampling line, the number of birds as well as the age (juvenile and adult) and sex of the birds are recorded.
-\[Some more information on the surveys, e.g.
-why are individuals in groups, how are sexes and age classes distinguished, etc.\] Additional information about the surveys are found in Nilsen *et al.* (2020) and Bowler *et al.* (2020).
-\[Add sample sizes and additional information on data processing/storage post collection, i.e.
-how do volunteers report data, storage in hønsefuglportalen, quality control, pushes to LivingNorway and GBIF\].
+As the surveys are conducted in early August, juveniles can be distinguished from adults by their smaller body size.
+Nevertheless, mis-identification can occur, and in addition a proportion of the birds are registered as "unknown" age and/or sex.
+Since 2019, data has been collected through a mobile app tailormade for this project.
+Before 2019, field workers reported their data through a dedicated web portal.
+After data is collected, they go through several steps of quality control: first by local contacts and subsequently by personnel at NINA.
+The data are standardized based on the Darwin-Core standard (ref), and made publicly available as sampling-event data set published through GBIF [@Nilsen2023a].
+For additional description of the data collection procedures, see [@Bowler2020]; [@Kvasnes2018][@Nilsen2023a].
 
 The individual longitudinal study based on radio collared willow ptarmigan was conducted in 2015-2020.
-Each winter (in February-March), willow ptarmigan were located at night using snowmobiles and large hand nets with prolonged handles, as described in Israelsen et al.
-(2020).
+Each winter (in February-March), willow ptarmigan were located at night using snowmobiles and large hand nets with prolonged handles, as described in [@Israelsen2020].
 To prevent birds from flying off before the field personnel were close enough to capture them, a high-powered head lamp was used to dazzle the birds.
@@
 
+Based on simulated data, we assessed the models ability to estimate the underlying parameters. We run a total of 10 simulations, and the results are summarized in
+
 In general, the model was able to estimate both parameters of the detection model (Figure X) of the simulated data with little bias.
 
-![](Figures/SimCheck_Detects_Temps.png)
+![Model performance on simulated data.
+Upper panel depicts model parameters, and lower panels show derived estimated of effective strip width and mean detection probability](Figures/SimCheck_Detects_Temps.png){#@fig-sim-detection}
 
@@
 Having evaluated the overall performance of our model on simulated data, we used data from our case study in Lierne as a case study to estimate abundance, vital rates and detection probabilities from a real-world data set.
-As is evident from Figure X, the population density increased markedly across the study period, from \< 10 ptarmigan / $km^{2}$ in 2007 to \> 35 ptarmigan / $km^{2}$ in in 2021.
+As is evident from Figure X, the population density increased markedly across the study period, from \< 10 ptarmigan / $km^{2}$ in 2007 to \> 35 ptarmigan / $km^{2}$ in in 2021 (@fig-lierne-density).
 The increase was most distinct from 2016 and onward.
-Across years, mean survival probability for August - January ($S_1$) was estimated at `r filter(tidy_summary_stats, model_parameter =="Mu_S1")$Mean` (95% C.I = `r filter(tidy_summary_stats, model_parameter =="Mu_S1")$lower` - `r filter(tidy_summary_stats, model_parameter =="Mu_S1")$upper`), survival probability for February - July ($S_2$) was estimated as `r filter(tidy_summary_stats, model_parameter =="Mu_S2")$Mean` (95% C.I = `r filter(tidy_summary_stats, model_parameter =="Mu_S2")$lower` - `r filter(tidy_summary_stats, model_parameter =="Mu_S2")$upper`), and annual survival probability $S$ is given by the product of $S_1$ and $S_2$ was estimated at `r filter(tidy_summary_stats, model_parameter =="Mu_S")$Mean` (95% C.I = `r filter(tidy_summary_stats, model_parameter =="Mu_S")$lower` - `r filter(tidy_summary_stats, model_parameter =="Mu_S")$upper`).
+Across years, mean survival probability for August - January ($S_1$) was estimated at `r round(filter(tidy_summary_stats, model_parameter =="Mu_S1")$Mean, 2)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_S1")$lower, 2)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_S1")$upper, 2)`), survival probability for February - July ($S_2$) was estimated as `r round(filter(tidy_summary_stats, model_parameter =="Mu_S2")$Mean, 2)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_S2")$lower, 2)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_S2")$upper, 2)`) (@fig-lierne-demo).
+Annual survival probability $S$ is given by the product of $S_1$ and $S_2$ was estimated at `r round(filter(tidy_summary_stats, model_parameter =="Mu_S")$Mean, 2)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_S")$lower, 2)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_S")$upper, 2)`).
 Note that we assumed that $S$ was constant across all years in the study period from 2007 - 2021.
+
 Recruitment ($R_t$) was allowed to vary across years (see model specification), and did as expected show large inter-annual variability.
-While the mean (baseline) recruitment $\mu_R$ was estimated as `r filter(tidy_summary_stats, model_parameter =="Mu_R")$Mean` (95% C.I = `r filter(tidy_summary_stats, model_parameter =="Mu_R")$lower` - `r filter(tidy_summary_stats, model_parameter =="Mu_R")$upper`).
+While the mean (baseline) recruitment $\mu_R$ was estimated as `r round(filter(tidy_summary_stats, model_parameter =="Mu_R")$Mean, 1)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_R")$lower, 1)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_R")$upper, 1)`).
+The estimated recruitment rate varied (as expected) among years: while the annual recruitment rate was estimated at `r round(max(filter(tidy_summary_stats, model_parameter =="R_year")$Mean), 1)` in year 2007, it was estimated at `r round(min(filter(tidy_summary_stats, model_parameter =="R_year")$Mean), 1)` in year 2012.
+Moreover, in contrast to the Alternative Prey Hypothesis (APH), our model did not find a clear link between the annual recruitment and the abundance of small rodents (slope-paramater for the z-standardized rodent occurrence data = `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$Mean, 3)` ; 95% C.I.
+= `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$lower, 3)` - `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$upper, 3)`).
 
-![**Figure X.** Estimated density of willow ptarmigans in Lierne from 2007 to 2021. Solid line represents the posterior median, ribbon marks 95% credible interval.](Figures/TimeSeries_popDens.png)
+![Estimated density of willow ptarmigans in Lierne from 2007 to 2021.
+Solid line represents the posterior median, ribbon marks 95% credible interval.](Figures/TimeSeries_popDens.png){#fig-lierne-density}
 
-#### Vital rates
+![Posterior densities of A) seasonal survival, B) average annual survival, and C) recruitment rate.
+For the latter, the yellow distribution is for the intercept, representing a baseline recruitment rate when rodent occupancy is low.
+The turquoise distributions are for year-specific estimates or recruitment rate, with darker colors indicating later years.](Figures/PostDens_Mu_S_tR.png){#fig-lierne-demo}
 
-![**Figure X.** Posterior densities of A) seasonal survival, B) average annual survival, and C) recruitment rate.
-For the latter, the yellow distribution is for the intercept, representing a baseline recruitment rate when rodent occupancy is low.
-The turquoise distributions are for year-specific estimates or recruitment rate, with darker colors indicating later years.](Figures/PostDens_Mu_S_tR.png)
-
-![**Figure X.** Estimated recruitment rate of willow ptarmigans in Lierne from 2007 to 2021.
-Solid line represents the posterior median, ribbon marks the 95% credible interval.](Figures/TimeSeries_rRep.png)
-
-![**Figure X.** Predicted effect of rodent occupancy on recruitment rate. Solid line represent the posterior median, ribbon marks the 95% credible interval.](Figures/Rep_betaR.R.png)
-
-#### Detection parameters
-
-![**Figure X.** Predicted average detection probabilities for distance sampling surveys in Lierne from 2007 to 2021.
-Solid line represents the posterior median, ribbon marks the 95% credible interval.](Figures/TimeSeries_pDetect.png)
+####
 
 ## Discussion
+
+-   In our process model, the population growth rate ($\lambda$) is determind by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, creating a dependence between the demographic parameters.
+If the age ratio in the data are biased or contain frequent misclassifications, this is likely to affect the relative contribution of survival and recruitment to the growth rate.
+As a test of the phenomena, we also
 

### 2023-12-06 Manuscript_Nater_etal.qmd
Added introduction outline in bullet points

@@
 # Introduction
+
+-   There is a growing need/demand for indicators for biodiversity, species abundance and demographic rates (see e.g. EBVs)
+
+-   On one hand, these indicators need to be representative at large scales
+
+-   This for e.g. international reporting towards biodiversity targets
+
+-   On the other hand, indicators with decent spatial resolution are also needed
+
+-   The reason for that is that that is the relevant scale for local-level managements and planning
+
+-   This goes both for development decisions and species management and conservation (which typically happen at a county or even municipality level)
+
+-   Besides organisational scale, there is also spatial variation not only in biodiversity, but also in population dynamics and life history of (target) species
+
+-   This variation needs to be accounted for to develop successful and sustainable (harvest) management and conservation strategies
+
+-   There is clear need for large-scale and spatially explicit indicators, but their development hinges on the availability of data from large-scale and long-term monitoring programmes
+
+-   Several countries (in Europe) have been working towards setting up such programmes and been successful particularly for monitoring of some species groups, i.e. breeding birds and butterflies
+
+-   Unfortunately, most of these monitoring programmes te limited to presence-absence or very simple count observations, making them suitable for the development of indicators of distribution and population trends, but usually not for indicators of abundance, population dynamics, and demographic rates
+
+-   The primary reason here is that large-scale long-term monitoring programmes are very costly to run, in particular when they need to be carried out by specialists
+
+-   Citizen science, i.e. the collection of ecological data by members of the public, has been growing rapidly and providing data at larger spatial and taxonomic scales than previously possible.
+
+-   However, increased coverage and total sampling effort in citizen science monitoring often comes at the cost of data being presence-only and prone to bias, limiting usefulness in particular for the study of population dynamics and demography.
+
+-   The exception are monitoring programmes that make use of a large number of citizen scientists that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol (+ 1-2 internationally "widely known" programmes?)
+
+-   Norway has one such monitoring programme that focuses on terrestrial gamebird species: Hønsefugltakseringen (needs an English name)
+
+-   Brief intro of monitoring programme: sampling design (mention age- and sex- assigment), coverage, data handling, OA publishing of data
+
+-   Brief intro of previously developed modelling framework: integrated distance sampling model using age-specific line transect data and auxiliary telemetry data to get estimates of population size and structure, as well as key vital rates (survival, recruitment), over time
+
+-   Combined, national-scale monitoring and this modelling framework have potential for large-scale high-resolution modelling of population dynamics and its demographic and environmental drivers
+
+-   In this study, we extended model of Nielsen & Nater 2023 to run not just on a single area but all areas with publicly available line transect data simultaneously
+
+-   Capitalize on "sharing information" across locations and effectively space-for-time substitution to estimate population size, age-structure, survival, and recruitment across 41 reporting districts and 15 years (2007-2021) for willow ptarmigan
+
+-   Shortly about willow ptarmigan (hereafter "ptarmigan"): sentinel species (sensitive to changes in climate and land-use), position in the foodweb (maybe?
+does it matter?), high cultural value (part of people's "nature experience", popular game species), not redlisted but popular hunting making sustainable management crucial
+
+-   Wrap up with a sentence on why large-scale spatially explicit indicators for ptarmigan populations would be good to have (national and international context)?
 

### 2023-12-06 Manuscript_Nater_etal.qmd
Added section headers and bullet points for discussion

@@
 
+## Abundance and vital rates across space and time
+
+-   Biological results (interpretations and comparisons with other case studies / species / systems)
+-   With some focus on what we are able to get insights on thanks to the integration across time and space (i.e.
+better estimates through data sharing, relative importance of different kinds of variation, etc.
+)
+
+## Implications for management (preliminary section title)
+
+-   What results mean for management (of habitat and of harvest)
+
+## Model limitations and outlook
+
+-   Some limitations of "base model" that need mentioning (e.g. age assignment, ignoring potential effects of flock size on detection)
+
+-   Limited usefulness of auxiliary radio-telemetry data as only from one area (big potential for including additional datasets from other small-scale studies, telemetry but also other, e.g.
+nest monitoring?)
+
+-   Outlook: improved modelling of spatio-temporal patterns through correlated REs
+
 Another approach to maximising information gain when jointly analysing data from different areas is explicit modelling of spatial and spatio-temporal variation.
@@
 Per now, we have not yet attempted this, but are planning to do so in the near future.
+(BUT: we do already have some computational limitations now, so likely to need development of more efficient sampling strategies.
+Which is fortunately possible with NIMBLE but there is a chance that another platform, e.g.
+INLA, may be required)
+
+-   Outlook: integration with / coupling to models for harvest to develop tools for (automatic) calculation of e.g. hunting quotas
+
+## Reproducible workflows for a sustainable future
+
+-   Structure / setup of this workflow (from data collection to analytical results)
+-   Why this matters: transparency and credibility, feedback to volunteer data collectors, cost-effective to re-do analyses for annual/regular reporting, easy to re-use and build upon, publicly accessible and well documented to allow more effective resource use in follow-up projects (sustainable use of funding), etc.
+-   Also: useful to link up to different reporting routes and/or more general (multi-indicator) projects
+
+## Conclusion
+
+-   Summarise neatly
+-   Things worth highlighting: inference across large spatial scale but with "useful" resolution for management, advantages of integration across space (we can estimate better and more), integration of small locality-specific datasets with large national-scale monitoring, importance of reproducible workflows
 

### 2023-12-06 Manuscript_Nilsen_Nater.qmd
Edits to ms file

@@
 ---
-title: "An integrated distance sampling approach for modelling age-structured populations"
+title: "An integrated open population distance sampling approach for modelling age-structured populations"
 format: html
@@
 the underlying vital rates.
-In fact, if the data does not contain information about the age (and/or sex-) structure of the surveyed population, there is no stratight forward way to estimate demographic rates from such data.
+In fact, if the data does not contain information about the age (and/or sex-) structure of the surveyed population, there is no straight forward way to estimate demographic rates from such data.
 In cases where the age (and sex) of the observed individuals can be observed, this information can under some assumptions be used to inform about recruitment rates and survival probabilities (ref).
@@
 
-Based on simulated data, we assessed the models ability to estimate the underlying parameters. We run a total of 10 simulations, and the results are summarized in
-
+Based on simulated data, we assessed the models ability to estimate the underlying parameters.
+We run a total of 10 simulations, and the results are summarized in @fig-sim-detection and @fig-sim-demo.
 In general, the model was able to estimate both parameters of the detection model (Figure X) of the simulated data with little bias.
+This was true for all 10 simulations, and for underlying model parameters than span a wide variation in values.
+The full output from each simulation is available in the appendix.
 
 ![Model performance on simulated data.
-Upper panel depicts model parameters, and lower panels show derived estimated of effective strip width and mean detection probability](Figures/SimCheck_Detects_Temps.png){#@fig-sim-detection}
+Upper panel depicts model parameters, and lower panels show derived estimated of effective strip width and mean detection probability.
+Dots represents true value, and density plot represent the posterior samples for each simulation](Figures/SimCheck_Detects_Temps.png){#fig-sim-detection}
 
-**Figure X** Model performance on simulated data. Upper panel depicts model parameters, and lower panels show derived estimated of effective strip width and mean detection probability
-
-![](Figures/SimCheck_VRs_Temps.png)
-
-**Figure X** Vital rates estimated based on simulated data
+![Posterior samples for the vital rates estimated based on simulated data.
+Upper panel show annual and seasonal survival probabilities, hyperparameters for the random effects model for recruitment (mean and sd), and lower panel show annual estimates of the recruitment rate.
+Dots represent true values, and density plots represents the posterior samples from each simulation](Figures/SimCheck_VRs_Temps.png){#fig-sim-demo}
 
@@
 
-####
-
 ## Discussion
 
--   In our process model, the population growth rate ($\lambda$) is determind by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, creating a dependence between the demographic parameters.
+-   In general, the model does a very good job in recreating the underlying parameters when fitted to simulated data where the true values are known.
+For the simulated data, a relative wide range of parameter values were allowed, and we did not detect any particular problems or biases with respect to this variation in the underlying parameters.
+-   The fact that this formulation allow us to separately estimate the underlying demographic rates (survival and recruitment) is a clear improvement compared to previous open population DS models, as this allow us both to obtain estimates of the rates, but also to assess the effect of environmental variation directly on demographic rates and not just on population growth rate ($\lambda$) .
+-   Nevertheless, we did not detect a clear effect of rodent abundance (indexed by the occurrence of rodents on the transect lines) on recruitment rate.
+This might be somewhat surprising, given that such a pattern has been reported repeatedly in the literature (see e.g.
+@Bowler2020).
+However, the data on rodent abundance was heavily zero-inflated, and the annual variation in the index was rather small.
+\[A few sentences about fading rodent cycles, and perhaps a weakened link to ptarmigan dynamics\].
+-   In our process model, the population growth rate ($\lambda$) is determined by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, creating a dependence between the demographic parameters.
 If the age ratio in the data are biased or contain frequent misclassifications, this is likely to affect the relative contribution of survival and recruitment to the growth rate.
-As a test of the phenomena, we also
-
-### Conclusion
+As a test of the phenomena, we also assessed how sensitive the model output was with regard to the treatment of birds classified as "unknown sex and age" by the field personnel (see **Methods**).
+In the version presented in the results section, we made the assumption that these birds were in fact juveniles.
+In an alternative scenario, we discarded all birds classified as unknowns, and rerun the model (Appendix).
+As expected the estimated population density was virtually unaffected by this decision, whereas the effect on the demographic parameters were somewhat more affected.
+Thus biases in the age ratio will affect the demographic rates, but not so much the estimated density.
+Compared to previous studies on ptarmigan, we could have expected a higher survival rate indicating that there is some misclassification in the data with respect to age specification.
+-   The model is flexible, and can be extended with additional auxiliary data, such as separate data on recruitment rates that could improve the precision of the demographic rate estimates.
+It is also possible to extent the model with additional covariates, to gain a better understanding of how species respond to environmental variation through the effects on vital rates.
 
@@
 
+EBN Lead - Drafted the model code; Lead - Wrote first version of ms; Lead; data collection Contributed - Modelling and analyses
+
+CRN Lead - Updated and finalized model code; Lead - Modelling; Contributed - Wrote first version of ms
+
 ### Acknowledgements
 
+We are grateful to field workers that collected the line transect data through the Hønsefuglportalen program.
+Fjellstyrene i Lierne contributed both to the line transect program and the field work related to the marked willow ptarmigan.
+Norwegian Environment Agency funded the projects (grant nrs **TO BE ADDED**).
+\[+ additional people contributing to the various projects\]
+
 ### Data and code availability

### 2023-12-06 Manuscript_Nater_etal.qmd
Added flags for Erlend to start drafting
Added text + references about ptarmigan ecology in introduction

@@
 
--   Shortly about willow ptarmigan (hereafter "ptarmigan"): sentinel species (sensitive to changes in climate and land-use), position in the foodweb (maybe?
-does it matter?), high cultural value (part of people's "nature experience", popular game species), not redlisted but popular hunting making sustainable management crucial
-
--   Wrap up with a sentence on why large-scale spatially explicit indicators for ptarmigan populations would be good to have (national and international context)?
+The willow ptarmigan (*Lagopus lagopus*) has a circumpolar distribution, mainly inhabiting sub-alpine and arctic ecosystems [see e.g.
+@fuglei2020].
+While the species is currently listed as Least Concern (LC) both in the global and Norwegian red list of species, it has undergone rather dramatic declines in abundance in Norway since the turn of the 20th century [@hjeljord2022].
+While the main reason for the long-term decline in abundance remain unresolved, they are considered sentinel species that are sensitive to both climate change and land use management [@Henden2017; @Storch2007].
+Moreover, being one of only a handful of bird species that spend the winter in mountain ecosystems in Scandinavia, they are important components of the ecosystem as prey species for the gyrfalcon [@Alastair2020].
+The willow ptarmigan has a relatively fast pace of life [@sandercock2005; @Steen1996], and can have substantial spatio-temporal variation in demographic rates [@Bowler2020].
+Their population dynamics are characterized by large inter-annual fluctuations in abundance [@hjeljord2022], and previous research has tied these fluctuations to rodent cycle through shared predators [@hagen1952; @Bowler2020].
+This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis and has been a central part of Fennoscandian research on grouse population dynamics for many decades [@hagen1952; @Linden1988: @Steen1985].
+In addition, spring weather conditions and phenology is known to have considerable effects on breeding success and recruitment rates [@Eriksen2023, @Steen1985].
+Across their distributional range, they are iconic species with a high cultural value, partly linked to their popularity as game species.
+The fact that they are valued game species, information about spatio-temporal variation in demographic rates and population dynamics is particularly important in order to design sustainable harvest strategies.
+In addition, as a sentinel species the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is in fact included in both the main national biodiversity (Nature Index for Norway; refs) and ecosystem (Assessment of the Ecological Condition; refs) assessments.
+The model presented here therefore potentially has direct application in real world management and conservation contexts.
 
@@
 -   Pre-modelling data wrangling
+
+\[End - ERLEND drafts\]
 

### 2023-12-18 Manuscript_Nater_etal.qmd
Updated text and references (.bib file)

@@
 
--   Setup of Hønsefugltakseringen monitoring programme (volunteer training, line transect sampling, spatio-temporal coverage, etc.)
--   Data & metadata recording via app
--   Quality control of data
--   Open access publishing of data (GBIF, LivingNorway)
--   Pre-modelling data wrangling
+The line transect survey data were collected through a structured participatory monitoring program called "Hønsefuglportalen" (<https://honsefugl.nina.no/Innsyn/en>).
+In August each year, trained volunteer fieldworkers collect willow ptarmigan observations along predefined line transects.
+To increase the detection probability, each team of fieldworkers use a pointing dog to locate the birds.
+Each survey team typically consist of at least two persons (one dog handler and one person responsible for following the transect line) and one dog.
+Often, more than one dog is used for a survey, but only one dog should be used at a time.
+The transect lines vary in length, but is typically between x and y km (range: xx, mean: yy).
+When an observation is made, the exact location of observation is noted, along with perpendicular distance to the transect line, and the age- and sex of the observed birds.
+An observation typically include 1 - 12 birds (mean = ?), representing one brood (female and or male with young-of-the-year chicks).
+When the surveys are conducted in August, the chicks of the year are able to fly but can be distinguished from older birds as they are still of smaller body size.
+Since 2019, most of the data has been collected using a mobile app tailored made for the monitoring program, where the field workers can register and get access to the transect lines allocated to them by the local organizers.
+Before 2019, data were collected on a dedicated field form, and entered manually in a web portal.
+After field data has been collected, several steps of quality control takes place before, both by local stakeholders and personnel from NINA.
+Surveys are carried out on both public and private land.
+After an initial embargo period, all data from public land are published and made freely available as a sampling-event data set on GBIF (refs).
+This data set contain both meta data about the transect surveys (survey date, line transect length and location, study area ID and other information) as well as data pertaining to the observations (species, number of birds of different categories (adult males, adult females, juveniles, and birds of unknown category), perpendicular distance to transect line, exact location, and time of observation).
+For this study, we used only data from public land that is freely available on GBIF; thus the workflow we describe should be fully reproducible by others.
+Formally, the data is publsihed as three distinct data sets, based on the three main public land owners (Statskog, FeFo and Fjellstyrene, respectively).
 
-Distance-sampling data on willow ptarmigan is collected on line transects not only in western Lierne, but across most of its Norwegian range (https://honsefugl.nina.no/).
-Taking the line transect datasets from Fjellstyrene, Statskog, and FeFo together, there is data for willow ptarmigan in 41 reporting areas (for a list, see [here](https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp/blob/main/R/listAreas.R)).
+The program is not designed as a national monitoring, but rather a collection of local and regional survey programs.
+All involved survey areas use a common field protocol and data collection model.
+In addition, the local study designs are reviewed by staff at NINA, and common recommendations for study design is provided.
+However, both a local scale and in particular at a regional scale, the design is unbalanced due to the fact that participation in voluntary by stakeholders.
+In general, the effort is higher in South-Eastern and Central Norway, medium in Northern Norway, and in general low in Western and Southern Norway.
+In total, we used data from Xx different survey areas, across xx counties and xx municipalities.
 
@@
 
--   Description of setup, data collection, temporal coverage, etc.
--   Pre-modelling data wrangling
-
-\[End - ERLEND drafts\]
+In our model, we integrated data from an ongoing field study of marked birds in Lierne municipality in Central Norway.
+These data were used to inform the survival component of the model (se below).
+The data is based on a sample of radio marked birds.
+The birds were captured in winter (late February or early March each year in the study period), and fitted with VHF collars.
+The marked birds were monitored on a regular basis until they either i) died, ii) battery stopped working, or iii) we lost contact with the bird for other reasons.
+For most of the year, the birds were monitored at least monthly.
+However, during the breeding and chick-rearing season (May to July) the birds were monitored more often, and during december and January we obtained few observations due to challenging field work conditions.
+A proportion of the birds were harvest annually in the regular recreational harvest, and birds that were harvested were reported as shot to the field personnel.
+In addition, as the collars had mortality switch, we were also able to locate and retrieve a high proportion of birds that died for natural causes.
+We used data from the years 2015 - 2021, and the total sample size across these years was xx birds.
+The study is described in detail in @Israelsen2020 and in @Arnekleiv2022.
 

### 2023-12-18 Manuscript_Nilsen_Nater.qmd
Worked over Erlend's draft

@@
 
-Estimating abundance and demographic rates for wildlife populations is an integral part of basic and applied ecology [@Skalski2005].
-Over the last few decades, a tremendous progress have been made towards this end.
+Estimating abundance and demographic rates for wildlife populations is an integral part of basic and applied ecology [@Skalski2005, @williams2002].
+Over the last few decades, tremendous progress has been made towards this end.
 This progress is partly driven the development and application of new field data collection methods and approaches, such as citizen science data (ref), camera trap data (ref) and the collection of environmental DNA data (ref).
-In addition, developments of novel statistical methods allow researchers to estimate abundance and demographic rates it situations where it was not feasible before (ref).
-Combined, these advance put us in a much better position to estimate state variables needed for population management (ref) or to estimate indexes that are relevant for large scale policy applications, e.g.
-such as Essential Biodiversity Variables [@Kissling2018].
+In addition, developments of novel statistical methods alongside decreases in computational costs now allow researchers to estimate abundance and demographic rates in situations where it was not feasible before (ref).
+Combined, these advances put us in a much better position for estimating quantities needed for population management [@williams2002] and indices relevant for large scale policy applications, e.g.
+Essential Biodiversity Variables [@Kissling2018].
 
-Until recently, joint estimation of population dynamics and demography has relied mostly on data from marked individuals through application of open-population capture-mark-recapture models (ref).
-While such methods can provide valuable information for both ecological research and management, they are also in many cases costly to implement over large areas.
-On the other hand, data from monitoring of abundance trends over time typically are based on data from unmarked animals (ref).
+Until recently, joint estimation of population dynamics and demography has relied mostly on data from marked individuals through application of open-population capture-mark-recapture models [@schaub2021].
+While such methods can provide valuable information for both ecological research and management, collecting the necessary data is typically costly and logistically challenging to implement over large areas.
+Monitoring focusing on abundance trends over larger areas, on the other hand, are typically based on data from unmarked animals (ref).
 A common approach for such surveys is to structure data collection around the distance sampling (DS) method.
-DS have been used for estimating animal abundance in a wide range of contexts and for a variety of taxa [@buckland2015].
-One reason for their popularity might be that they does not require repeated visits to the same sites to estimate detection probability (ref).
-Rather, it is assumed that detection probability decrease as the objects are located longer away from the transect line (ref).
-This makes them particularily useful for implementation in participatory monitoring program, allowing stakeholders to take part in the data collection process.
+DS has been used for estimating animal abundance in a wide range of contexts and for a variety of taxa [@buckland2015].
+One reason for the method's popularity is that it does not require repeated visits to the same sites for estimating detection probability (ref).
+Rather, it is assumed that detection probability decreases the further away from the transect line animals are located (ref).
+This makes DS particularly useful for implementation in participatory monitoring programs, allowing stakeholders to take part in the data collection process.
+<!-- #It does not become evident why the fact that detection is assumed to decrease with distance makes a survey design more suitable for participatory monitoring.
+May need some more explanation.
+-->
 
-The most common implementation of the DS model has been to use close-population formulations, that does not allow to formulate a process model that project abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates (ref).
-In recent years, DS approahces have been extended and also applied for estimating changes in abundance over time in open populations by embedding a hidden state model representing population dynamics [@moore2011; @sollmann2015].
-Such approaches has greatly extended the application of DS approaches to study ecological dynamics in time and space.
-However, while these latter frameworks may allow to accurately quantify population changes but provide little information on the drivers of these changes, i.e.
+The most common implementations of DS models have long used close-population formulations, and thus do not allow formulating a process model that projects abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates (ref).
+In recent years, DS approaches have been extended in many ways, including applications that estimate changes in abundance over time in open populations via a hidden state model representing population dynamics [@moore2011; @sollmann2015].
+This has greatly extended the potential of DS approaches for studying ecological dynamics across time and space.
+However, while these latter frameworks may allow to accurately quantify population changes, they typically provide little information on the drivers of these changes, i.e.
 the underlying vital rates.
-In fact, if the data does not contain information about the age (and/or sex-) structure of the surveyed population, there is no straight forward way to estimate demographic rates from such data.
-In cases where the age (and sex) of the observed individuals can be observed, this information can under some assumptions be used to inform about recruitment rates and survival probabilities (ref).
-For instance, in a integrated population model based on harvest statistics and observations of population age structure, both population abundance and demographic rates was estimated without even without additional data from marked individuals [@Nilsen2018].
+In fact, if the data does not contain information about the age (and/or sex-) structure of the surveyed population, there is no straightforward way to estimate demographic rates from such data.
+On the contrary, if age (and sex) of detected individuals can be determined, this information can be used to provide information on recruitment rates and survival probabilities (ref).
+@Nilsen2018, for example, used a model based on harvest statistics and observations of population age structure to estimate population abundance and demographic rates without the need for any additional data from marked individuals.
+<!-- #I have removed the mention of integrated population models from this sentence since those only get specifically introduced in the next paragraph -->
 
-Concurrent with the development of more sophisticated DS models, another group of models has emerged and gained popularity, not least for their ability to disentangle demographic processes underlying population dynamic: integrated population models [IPMs, @schaub2021].
-Through joint analysis of multiple datasets, IPMs allow simultaneous estimation of population size and composition, as well as all vital rates that form part of an underlying typically age- or stage-structured population model.
-Since both DS models and IPMs estimate population size/density, a combination of the two frameworks has the potential to provide good estimates of both population- and demographic parameters and maximizing knowledge gained through transect surveys by augmenting them with other available data [e.g.
+Concurrent with the development of more sophisticated DS models, another group of models has emerged and rapidly gained popularity, not least for their ability to disentangle demographic processes underlying population dynamic: integrated population models [IPMs, @schaub2021].
+Through joint analysis of multiple datasets, IPMs allow simultaneous estimation of population size and composition, as well as all vital rates that form part of an underlying age- or stage-structured population model.
+Since both DS models and IPMs estimate population size/density, a combination of the two frameworks has the potential to provide good estimates of both population- and demographic parameters by maximizing knowledge gained from transect surveys by augmenting them with other available data [e.g.
 @schmidt2020].
 
-In this study, we will present a new IPM, where we integrate data from line transect distance sampling survey data and survival data from marked birds.
-The process model is a stage-structured matrix model, that projects population size from one time step to the next based on the underlying survival and recruitment rates.
-This is made possible by the fact that the line transect surveys provide data on population age structure which can be used to inform the estimation of demographic rates.
+In this study, we present a new IPM (IDSM, integrated distance sampling model) which integrates data from line transect distance sampling survey data and survival data from marked birds.
+The model's core is a stage-structured matrix population model that projects population size from one time step to the next based on underlying survival and recruitment rates.
+<!-- # I suggest removing the following two sentences as they mostly re-state information we have already provided: "This is made possible by the fact that the line transect surveys provide data on population age structure which can be used to inform the estimation of demographic rates.
 In the IPM, we integrate these data with data from marked birds from which we estimate survival probabilities.
-The model is thus formally an IPM, as we jointly estimate parameters of interest based on several data sets (ref).
-In the paper, we first present the model and assess the robustness based on simulated data.
-In addition, we apply the model to real data collected from a willow ptarmigan (*Lagopus lagopus*) study in Central Norway.
-Because demographic rates are often affected by environmental covariates, we also illustrate how such covariates can be included in the modelling framework.
+The model is thus formally an IPM, as we jointly estimate parameters of interest based on several data sets (ref)." --> Below, we first present the model and assess its robustness and performance through application to simulated data.
+We then apply the model to real data collected from a willow ptarmigan (*Lagopus lagopus*) study in Central Norway.
+Because demographic rates are often affected by environmental covariates (e.g.
+rodent abundance in the case of willow ptarmigans), we also illustrate how such covariates can be included in the modelling framework.
 
@@
 
-Our open population integrated distance sampling model consists of two major components: a latent structured population model and a set of likelihoods for data originating from distance sampling surveys and auxiliary survival monitoring.
+Our open population integrated distance sampling model (IDSM) consists of two major components: a latent structured population model and a set of likelihoods for data originating from distance sampling surveys and auxiliary survival monitoring.
 In the example case, these auxiliary data come from a radio-telemetry study, but in principle other types of capture-recapture data can also be used.
@@
 
-TBA
+We assessed the model's overall performance and ability to estimate abundance, demographic rates, and detection parameters without bias by testing it on simulated data.
+We generated a total of 10 simulated datasets in five steps.
+First, we simulated 15-year time-series of survival and recruitment rates from biologically plausible values for averages of and -- in the case of recruitment -- among-year variation in demographic rates (survival was held constant across years).
+Second, we used the yearly demographic rate and realistic initial population densities to simulate stochastic population dynamics in 50 distinct sites.
+Third, we simulated the grouping of individuals in each site by first determining the expected number of groups in a site (based on the average group size of 5.6 from our ptarmigan data) and then distributing individuals among groups via multinomial trials.
+Fourth, we assigned a distance from transect line to each group and simulated the line transect survey in all 50 sites across 15 years.
+Finally, we simulated 5-year time-series of radio-telemetry data (= survival from one year to the next) for a subset of individuals (30 on average) using the simulated survival probabilities for each relevant year.
+We then fit the IDSM to each of the 10 simulated datasets three times, using distinct seeds for both simulating initial values and initiating and running the MCMC.
+Model implementation for simulated data tests was largely identical to that for real data and is described in detail below.
 
@@
 The willow ptarmigan has a circumpolar distribution (Fuglei et al., 2020), and lives year-round in heterogeneous alpine and artic ecosystems.
-In Norway, there has been a long-term decline in the willow ptarmigan abundance spanning more than a century (ref), but the last few decades the trend in abundance has fluctuated both in time and space.
-In Scandinavia, willow ptarmigan is a valued game species (ref), and there has been several long-term research projects devoted to understanding how they respond to environmental variation and harvest management [@Israelsen2020; @Sandercock2011].
+In Norway, there has been a long-term decline in the willow ptarmigan abundance across more than a century (ref), but in the last few decades abundance trends has fluctuated both in time and space.
+In Scandinavia, willow ptarmigan is a valued game species (ref), and there have been several long-term research projects devoted to understanding how they respond to environmental variation and harvest management [@Israelsen2020; @Sandercock2011].
 A key insight from across several study areas is the the annual recruitment rate (i.e.
-$R_t$ in our model, as outlined above) is highly variable, and is affected both by spring conditions [@Eriksen2023] and the abundance of small rodents as alternative prey for shared predators (i.e.
+$R_t$ in our model, as outlined above) is highly variable, and is affected both by spring conditions [@Eriksen2023] and the abundance of small rodents, which constitute alternative prey for common predators (i.e.
 the Alternative Prey Hypothesis ref).
@@
 Our case study was based on an ongoing long-term research project on willow ptarmigan in Lierne municipality in Central Norway (approximately 62.4 degrees north and 13.2 degrees east).
-The study area is located in a sub-alpine ecosystem, and the landscape is a mosaic of open heath and shrub vegetation (dominated by Ericacea, willow shrub *Salix spp*.
-and dwarf birch *Betula nana*), interspersed with bogs and forest patches (mainly birch *Betula spp*.).
+The study area is located in a sub-alpine ecosystem, and the landscape is a mosaic of open heath and shrub vegetation (dominated by Ericacea, willow shrub *Salix spp*., and dwarf birch *Betula nana*), interspersed with bogs and forest patches (mainly birch *Betula spp*.).
 The climate is strongly seasonal, with snow typically covering the ground from October/November through April/May.
@@
 
-1.  Data from a line transect survey program targeting willow ptarmigan operated as part of the natural resources management authorities
+1.  Data from a line transect survey program targeting willow ptarmigan operated under the natural resources management authorities (2007-2021, ongoing)
 
-2.
-Data from an individual based research project based on radio collared willow ptarmigan.
-The temporal coverage of the two data sets partially overlap, with the line transect survey spanning 2007-2021 (and ongoing) and the individual based data spanning 2015-2021 (and ongoing).
+2.  Data from an individual-based monitoring programme based on radio collared willow ptarmigan (2015-2021, ongoing)
 
 Line transect survey data were collected in August each year, prior to the annual autumn harvest season, as part of the program "Hønsefuglportalen".
-Hønsefuglportalen is a national program for line transect surveys of tetraonid birds, and the effort is directed mainly to willow ptarmigan habitats.
+Hønsefuglportalen is a national program for line transect surveys of tetraonid birds, and the effort is directed mainly towards willow ptarmigan habitats.
 In our case study, we used data from the western part of Lierne municipality.
-Line transects are surveyed by educated volunteers, using pointing dogs to located the birds.
-When located, the geographical coordinate, perpendicular distance from the sampling line, the number of birds as well as the age (juvenile and adult) and sex of the birds are recorded.
+Line transects are surveyed by trained volunteers that use pointing dogs to locate the birds.
+When located, the geographical coordinate, perpendicular distance from the sampling line, the number of birds in the group, as well as the age (juvenile or adult) and sex of the birds are recorded.
 As the surveys are conducted in early August, juveniles can be distinguished from adults by their smaller body size.
-Nevertheless, mis-identification can occur, and in addition a proportion of the birds are registered as "unknown" age and/or sex.
-Since 2019, data has been collected through a mobile app tailormade for this project.
+<!-- maybe also mention how males and females are distinguished?
+Also, is it just body size for the juveniles or also plumage?
+--> Nevertheless, mis-identification can occur, and in addition a proportion of the birds are registered as "unknown" age and/or sex.
+Since 2019, data has been collected through a mobile app tailormade for this project <!-- Add name + maybe link to download in app store?
+--> .
 Before 2019, field workers reported their data through a dedicated web portal.
-After data is collected, they go through several steps of quality control: first by local contacts and subsequently by personnel at NINA.
-The data are standardized based on the Darwin-Core standard (ref), and made publicly available as sampling-event data set published through GBIF [@Nilsen2023a].
+After data are collected and reported, they undergo several steps of quality control: first by local contacts and subsequently by personnel at the Norwegian Institute for Nature Research (NINA).
+The data are then standardized based on the Darwin-Core standard (ref), and made publicly available as a sampling-event data set published through GBIF [@Nilsen2023a].
 For additional description of the data collection procedures, see [@Bowler2020]; [@Kvasnes2018][@Nilsen2023a].
@@
 The individual longitudinal study based on radio collared willow ptarmigan was conducted in 2015-2020.
-Each winter (in February-March), willow ptarmigan were located at night using snowmobiles and large hand nets with prolonged handles, as described in [@Israelsen2020].
+<!-- Above you stated that radio telemetry data collection is until 2021 and ongoing.
+Until what year was it?
+--> Each winter (in February-March), willow ptarmigan were located at night using snowmobiles and large hand nets with prolonged handles, as described in [@Israelsen2020].
 To prevent birds from flying off before the field personnel were close enough to capture them, a high-powered head lamp was used to dazzle the birds.
 After capture, birds were placed in an opaque bag to reduce stress.
-Before releasing the birds, they were fitted with a uniquely numbered leg ring (\~ 2.4g) and a Holohil RI-2BM or Holohil RI-2DM radio transmitter (\~ 14.1g).
+They were then fitted with a uniquely numbered leg ring (\~ 2.4g) and a Holohil RI-2BM or Holohil RI-2DM radio transmitter (\~ 14.1g) and subsequently released.
 The radio transmitters had an expected battery lifetime of 24 months (RI-2BM) or 30 months (RI-2DM), and included a mortality circuit that was activated if a bird had been immobile for 12 hours.
-We monitored the birds throughout the year, by triangulation from the ground at least once a month for 10 months of the year (February -- November) by qualified field personnel.
-If a mortality signal was heard from the transmitter, we recovered it as soon as possible to determine cause of death.
+We monitored the birds throughout most of the year by triangulation from the ground at least once a month for 10 months of the year (February -- November) by qualified field personnel.
+If a mortality signal was emitted from a transmitter, we attempted to recover it as soon as possible to determine cause of death.
 A number of birds dispersed out of the main study areas and was thus out of signal range for field personnel on the ground.
@@
 
-### 2.4 Bayesian implementation
+### 2.4 Bayesian model implementation
 
-We implemented the model in a Bayesian framework using NIMBLE [@valpine2017] in R version 4.2.2 [@R2022].
+We implemented the model in a Bayesian framework using NIMBLE version 1.0.1 [@valpine2017] in R version 4.3.1 [@R2023].
 The likelihood for line transect observation distances was set up using a custom half-normal distribution developed by Michael Scroggie as part of the "nimbleDistance" package (https://github.com/scrogster/nimbleDistance).
 We used non-informative uniform priors (with biologically reasonable boundaries where possible) for all parameters.
-For each model fit, we ran x MCMC chains with NIMBLE's standard samples for xxxk iterations.
-xxxk thereof were discarded as burn-in prior to thining with a factor x, leaving us with xxxx posterior samples per chain (total of xxxxx samples per run).
+We assumed constant survival and time-varying recruitment rate in models fit to both simulated and real data.
+
+For the model fits to simulated and real data we ran 3 and 4 MCMC chains with NIMBLE's standard samples for 500k and 100k iterations, respectively.
+300k and 40k and thereof were discarded as burn-in prior to thinning with factors 5 and 20 , leaving us with 40k and 3k posterior samples per chain (total of 120k and 12k samples per run), repectively.
 
@@
 
-Based on simulated data, we assessed the models ability to estimate the underlying parameters.
-We run a total of 10 simulations, and the results are summarized in @fig-sim-detection and @fig-sim-demo.
-In general, the model was able to estimate both parameters of the detection model (Figure X) of the simulated data with little bias.
-This was true for all 10 simulations, and for underlying model parameters than span a wide variation in values.
-The full output from each simulation is available in the appendix.
+Posterior distributions for parameters estimated in three model fits to each of 10 simulated data sets are shown in @fig-sim-detection and @fig-sim-demo.
+Overall, the IDSM was able to corretly estimate both detection parameters and demographic rate parameters from all 10 simulated datasets without any systematic bias.
+The replicate runs for each dataset resulted in very similar posterior distributions, demonstrating that the models converged towards the same posterior distributions irrespective of starting values.
+Estimates of population sizes / densities were also adequate, and are presented in the Supplementary Materials (see folder SimCheck_byDataSet).
 
-![Model performance on simulated data.
+![Detection parameters estimated from 10 distinct sets of simulated data (= colors).
 Upper panel depicts model parameters, and lower panels show derived estimated of effective strip width and mean detection probability.
-Dots represents true value, and density plot represent the posterior samples for each simulation](Figures/SimCheck_Detects_Temps.png){#fig-sim-detection}
+Dots represents true values (black for global values, colored for dataset-specific values), and density plot represent the posterior distributions for each model run.
+](Figures/SimCheck_Detects_Temps.png){#fig-sim-detection}
 
-![Posterior samples for the vital rates estimated based on simulated data.
+![Demographic rate parameters estimated from 10 distinct sets of simulated data (= colors).
 Upper panel show annual and seasonal survival probabilities, hyperparameters for the random effects model for recruitment (mean and sd), and lower panel show annual estimates of the recruitment rate.
-Dots represent true values, and density plots represents the posterior samples from each simulation](Figures/SimCheck_VRs_Temps.png){#fig-sim-demo}
+Dots represent true values (black for global values, colored for dataset-specific values), and density curves represent the posterior distributions from each model run.
+](Figures/SimCheck_VRs_Temps.png){#fig-sim-demo}
 
@@
 Having evaluated the overall performance of our model on simulated data, we used data from our case study in Lierne as a case study to estimate abundance, vital rates and detection probabilities from a real-world data set.
-As is evident from Figure X, the population density increased markedly across the study period, from \< 10 ptarmigan / $km^{2}$ in 2007 to \> 35 ptarmigan / $km^{2}$ in in 2021 (@fig-lierne-density).
+Ptarmigan population density increased markedly across the study period, from \< 10 ptarmigan / $km^{2}$ in 2007 to \> 35 ptarmigan / $km^{2}$ in in 2021 (@fig-lierne-density).
 The increase was most distinct from 2016 and onward.
-Across years, mean survival probability for August - January ($S_1$) was estimated at `r round(filter(tidy_summary_stats, model_parameter =="Mu_S1")$Mean, 2)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_S1")$lower, 2)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_S1")$upper, 2)`), survival probability for February - July ($S_2$) was estimated as `r round(filter(tidy_summary_stats, model_parameter =="Mu_S2")$Mean, 2)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_S2")$lower, 2)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_S2")$upper, 2)`) (@fig-lierne-demo).
-Annual survival probability $S$ is given by the product of $S_1$ and $S_2$ was estimated at `r round(filter(tidy_summary_stats, model_parameter =="Mu_S")$Mean, 2)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_S")$lower, 2)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_S")$upper, 2)`).
-Note that we assumed that $S$ was constant across all years in the study period from 2007 - 2021.
-
-Recruitment ($R_t$) was allowed to vary across years (see model specification), and did as expected show large inter-annual variability.
-While the mean (baseline) recruitment $\mu_R$ was estimated as `r round(filter(tidy_summary_stats, model_parameter =="Mu_R")$Mean, 1)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_R")$lower, 1)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_R")$upper, 1)`).
-The estimated recruitment rate varied (as expected) among years: while the annual recruitment rate was estimated at `r round(max(filter(tidy_summary_stats, model_parameter =="R_year")$Mean), 1)` in year 2007, it was estimated at `r round(min(filter(tidy_summary_stats, model_parameter =="R_year")$Mean), 1)` in year 2012.
-Moreover, in contrast to the Alternative Prey Hypothesis (APH), our model did not find a clear link between the annual recruitment and the abundance of small rodents (slope-paramater for the z-standardized rodent occurrence data = `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$Mean, 3)` ; 95% C.I.
-= `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$lower, 3)` - `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$upper, 3)`).
 
@@
 Solid line represents the posterior median, ribbon marks 95% credible interval.](Figures/TimeSeries_popDens.png){#fig-lierne-density}
+
+Average survival probability for August - January ($S_1$) was estimated at `r round(filter(tidy_summary_stats, model_parameter =="Mu_S1")$Mean, 2)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_S1")$lower, 2)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_S1")$upper, 2)`) while average survival probability for February - July ($S_2$) was estimated as `r round(filter(tidy_summary_stats, model_parameter =="Mu_S2")$Mean, 2)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_S2")$lower, 2)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_S2")$upper, 2)`) (@fig-lierne-demo A).
+Annual survival probability $S$, given by the product of $S_1$ and $S_2$, was estimated at `r round(filter(tidy_summary_stats, model_parameter =="Mu_S")$Mean, 2)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_S")$lower, 2)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_S")$upper, 2)`), @fig-lierne-demo A).
+
+Recruitment ($R_t$) was allowed to vary across years (see model specification), and estimates displayed large inter-annual variability (@fig-lierne-demo C, Figure S1).
+While the mean (baseline) recruitment $\mu_R$ was estimated as `r round(filter(tidy_summary_stats, model_parameter =="Mu_R")$Mean, 1)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="Mu_R")$lower, 1)` - `r round(filter(tidy_summary_stats, model_parameter =="Mu_R")$upper, 1)`) the yearly recruitment rates ranged from `r round(min(filter(tidy_summary_stats, model_parameter =="R_year")$Mean), 1)` in year 2012 to `r round(max(filter(tidy_summary_stats, model_parameter =="R_year")$Mean), 1)` in year 2007.
+
+Given the available data, the IDSM was not able to estimate a clear effect of small rodent abundance on ptarmigan recruitment (slope-paramater for the z-standardized rodent occurrence data = `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$Mean, 3)` ; 95% C.I.
+= `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$lower, 3)` - `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$upper, 3)`).
 
@@
 For the simulated data, a relative wide range of parameter values were allowed, and we did not detect any particular problems or biases with respect to this variation in the underlying parameters.
+-   Both simulated data tests and real data application have shown that mixing is suboptimal for the average recruitment rate parameter (Mu.R).
+Future work should perhaps look into optimizing sampling?
+(For biological interpretation that is not a big problem as the yearly recruitment rate estimates are fine...
+but have to keep in mind when interpreting) <!-- I leave it to you Erlend to decide where, whether, and how much you want to go into this one -->
 -   The fact that this formulation allow us to separately estimate the underlying demographic rates (survival and recruitment) is a clear improvement compared to previous open population DS models, as this allow us both to obtain estimates of the rates, but also to assess the effect of environmental variation directly on demographic rates and not just on population growth rate ($\lambda$) .
@@
 \[A few sentences about fading rodent cycles, and perhaps a weakened link to ptarmigan dynamics\].
+\[Also: a time-series of 15 years is still relatively short, so it is to be expected that we have low statistical power to detect temporal covariate effects --\> more years of data, or space-for-time substitution, will help)
 -   In our process model, the population growth rate ($\lambda$) is determined by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, creating a dependence between the demographic parameters.
@@
 -   The model is flexible, and can be extended with additional auxiliary data, such as separate data on recruitment rates that could improve the precision of the demographic rate estimates.
+Another one might be data that allows to distinguish between true and apparent survival<!-- or is that accounted for in the telemetry data, i.e.
+are the individuals that left hte study area censored?
+-->.
 It is also possible to extent the model with additional covariates, to gain a better understanding of how species respond to environmental variation through the effects on vital rates.
+\[See my point above on limited statistical power with 15 years.
+Maybe could be good to link to the space-for-time substitution idea here and "tease" the extension to multi-area since the monitoring programme, at least for line transects, covers many more areas).
 
@@
 
-EBN Lead - Drafted the model code; Lead - Wrote first version of ms; Lead; data collection Contributed - Modelling and analyses
+EBN Lead - Drafted the model code; Lead - Wrote first version of ms; Lead - Data collection; Contributed - Modelling and analyses
 
-CRN Lead - Updated and finalized model code; Lead - Modelling; Contributed - Wrote first version of ms
+CRN Lead - Updated, further developed, and finalized model code; Lead - Modelling and analyses; Contributed - Wrote first version of ms
 
@@
 
+\[Should we thank James for discussions etc.?\]
+
 ### Data and code availability

### 2023-12-20 Manuscript_Nater_etal.qmd
Wrote a first (short?) draft for the introduction

@@
 
--   There is a growing need/demand for indicators for biodiversity, species abundance and demographic rates (see e.g. EBVs)
+There is growing demand for biodiversity indicators from international unions, national government, local management bodies, and corporate and industry actors.
+Such indicators ideally represent a wide range of biodiversity's states and functions (e.g.
++ ref on basic EBV idea), yet the development of suitable indicators for certain qualities, such as species abundance and demography, has been more difficult than for others (+ref).
+This is at least partially due to challenging requirements regarding spatial scales of useful biodiversity indicators.
+On one hand, indicators need to be representative at large geographic scales, for example, in the context of countries' reporting towards biodiversity targets (+ ref).
+On the other hand, indicators also ideally have good spatial resolution as the scales relevant for local-level management and planning are often much smaller (+ ref).
+This latter requirement is particularly crucial for (infrastructure) development strategies and for species management and conservation, both of which tend to require knowledge on species abundance and population dynamics (i.e.
+demographic rates) that is relevant for their county- or municipality-level decision making (+ refs).
+Another reason why abundance and population indicators ideally come with good spatial resolution is that there can be substantial amounts of variation in population dynamics and life history of species across space (+ref).
+This is variation that needs to be accounted for in order to develop successful and sustainable strategies for area use, (harvest) management, and species and biodiversity conservation [@williams2002].
 
--   On one hand, these indicators need to be representative at large scales
+While large-scale, spatially-explicit indicators for species abundance and populations are clearly needed, development and practical implementation are greatly limited due to the reliance of such indicators on the availability of data from large-scale, long-term monitoring programmes (+ref).
+Consequently, many countries have been working on setting up, maintaining, and improving such monitoring programmes over the last decades.
+Many now well-established programmes focus on breeding birds and butterflies, and examples include the North American Breeding Bird Survey (https://www.pwrc.usgs.gov/bbs/), UK Butterfly Monitoring Scheme (https://ukbms.org/), Swiss Biodiversity Monitoring (https://www.biodiversitymonitoring.ch/), and many more.
+<!-- is there any other one we want to name here specifically?
+-->
 
--   This for e.g. international reporting towards biodiversity targets
+There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly and requires trained specialists, thus limiting the amount of data that can be collected, while participatory monitoring, i.e.
+the collection of ecological data by members of the public (also called citizen science, +ref), allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias (+ref).
+Consequently, many large-scale monitoring programmes are limited to presence-absence or very simple count observations, making them suitable for the development of indicators of distribution and population trends, but usually not for indicators of abundance, population dynamics, and demographic rates (+ ref).
+The exception here are monitoring programmes that succeed in making use of a large number of citizen scientists that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol.
+<!-- can we think of 1-2 internationally well-known examples?
+--> In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", https://honsefugl.nina.no/Innsyn/en).
+It is a line transect survey programme carried out annually in > xxxx locations across the country (xxxx > transects) by trained volunteers using pointing dogs.
+The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata (+ref to protocol?) and established routines for quality control and annual releases of publicly available data via GBIF.
+As such, it is particularly well suited to become part of a workflow for producing and updating abundance and population indicators on an annual basis.
 
--   On the other hand, indicators with decent spatial resolution are also needed
+The line transect data from "Hønsefuglportalen" has been used previously for estimating abundance trends of willow ptarmigan (*Lagopus lagopus*) across Norway [e.g.
+@bowler2020; @nilsen2020] but large-scale estimation of demographic rates underlying these trends has thus far remained as untapped potential of the dataset.
+@nilsen2023 recently developed a novel integrated distance sampling model (IDSM) which successfully uses age information in line transect data alongside radio-telemetry data to estimate not just abundance, but also survival and recruitment across years.
+In this study, we adapt the model of @nilsen2023 to run not just on a single site but on all areas with publicly available line transect data from "Hønsefuglportalen" simultaneously.
+Unlike several previous studies applying integrated models for population dynamics to multiple (sub-) populations separately and comparing results [e.g.
+@robinson2014; @nater2023], we opt for an approach explicitly integrating across space, thus allowing for sharing of information across locations and -- in effect -- space-for-time substitution [e.g.
+@horswill2019; @morrison2022].
+We then apply the resulting multi-area IDSM to "Hønsefuglportalen" data on willow ptarmigan to estimate population size, age-structure, survival, recruitment, and impacts of small rodent abundance across  41 reporting districts and 15 years (2007-2021) for willow ptarmigan.
+We further embed the modelling workflow in a reproducible, semi-automated pipeline that will greatly facilitate the repeated calculation of abundance and population indicators at different spatial scales as new data is added every year.
 
--   The reason for that is that that is the relevant scale for local-level managements and planning
+# Methods
 
--   This goes both for development decisions and species management and conservation (which typically happen at a county or even municipality level)
-
--   Besides organisational scale, there is also spatial variation not only in biodiversity, but also in population dynamics and life history of (target) species
-
--   This variation needs to be accounted for to develop successful and sustainable (harvest) management and conservation strategies
-
--   There is clear need for large-scale and spatially explicit indicators, but their development hinges on the availability of data from large-scale and long-term monitoring programmes
-
--   Several countries (in Europe) have been working towards setting up such programmes and been successful particularly for monitoring of some species groups, i.e. breeding birds and butterflies
-
--   Unfortunately, most of these monitoring programmes te limited to presence-absence or very simple count observations, making them suitable for the development of indicators of distribution and population trends, but usually not for indicators of abundance, population dynamics, and demographic rates
-
--   The primary reason here is that large-scale long-term monitoring programmes are very costly to run, in particular when they need to be carried out by specialists
-
--   Citizen science, i.e. the collection of ecological data by members of the public, has been growing rapidly and providing data at larger spatial and taxonomic scales than previously possible.
-
--   However, increased coverage and total sampling effort in citizen science monitoring often comes at the cost of data being presence-only and prone to bias, limiting usefulness in particular for the study of population dynamics and demography.
-
--   The exception are monitoring programmes that make use of a large number of citizen scientists that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol (+ 1-2 internationally "widely known" programmes?)
-
--   Norway has one such monitoring programme that focuses on terrestrial gamebird species: Hønsefugltakseringen (needs an English name)
-
--   Brief intro of monitoring programme: sampling design (mention age- and sex- assigment), coverage, data handling, OA publishing of data
-
--   Brief intro of previously developed modelling framework: integrated distance sampling model using age-specific line transect data and auxiliary telemetry data to get estimates of population size and structure, as well as key vital rates (survival, recruitment), over time
-
--   Combined, national-scale monitoring and this modelling framework have potential for large-scale high-resolution modelling of population dynamics and its demographic and environmental drivers
-
--   In this study, we extended model of Nielsen & Nater 2023 to run not just on a single area but all areas with publicly available line transect data simultaneously
-
--   Capitalize on "sharing information" across locations and effectively space-for-time substitution to estimate population size, age-structure, survival, and recruitment across 41 reporting districts and 15 years (2007-2021) for willow ptarmigan
+## Study species
 
@@
 
-# Methods
-
-## Study species
-
--   Willow ptarmigan distribution, life history, and (historical) abundance
-
 ## Data collection, management, and preparation
@@
 The transect lines vary in length, but is typically between x and y km (range: xx, mean: yy).
-When an observation is made, the exact location of observation is noted, along with perpendicular distance to the transect line, and the age- and sex of the observed birds.
-An observation typically include 1 - 12 birds (mean = ?), representing one brood (female and or male with young-of-the-year chicks).
+When an observation is made, the exact location of observation is noted, along with perpendicular distance to the transect line, and the age and sex of the observed birds.
+An observation typically include 1 - 12 birds (mean = 5.6), representing one brood (female and or male with young-of-the-year chicks).
 When the surveys are conducted in August, the chicks of the year are able to fly but can be distinguished from older birds as they are still of smaller body size.
@@
 In general, the effort is higher in South-Eastern and Central Norway, medium in Northern Norway, and in general low in Western and Southern Norway.
-In total, we used data from Xx different survey areas, across xx counties and xx municipalities.
+In total, we used data from 41 different survey areas, across xx counties and xx municipalities.
 

### 2023-12-20 Manuscript_Nater_etal.qmd
Added numbers and references to existing text

@@
 
+TBA
+
 # Introduction
+
+<!-- @JAMES & ERLEND: First draft is relatively "compact". Input on extending it and/or addition of relevant references appreciated -->
 
@@
 Consequently, many countries have been working on setting up, maintaining, and improving such monitoring programmes over the last decades.
-Many now well-established programmes focus on breeding birds and butterflies, and examples include the North American Breeding Bird Survey (https://www.pwrc.usgs.gov/bbs/), UK Butterfly Monitoring Scheme (https://ukbms.org/), Swiss Biodiversity Monitoring (https://www.biodiversitymonitoring.ch/), and many more.
+Many now well-established programmes focus on breeding birds and butterflies, and examples include the North American Breeding Bird Survey (<https://www.pwrc.usgs.gov/bbs/>), UK Butterfly Monitoring Scheme (<https://ukbms.org/>), Swiss Biodiversity Monitoring (<https://www.biodiversitymonitoring.ch/>), and many more.
 <!-- is there any other one we want to name here specifically?
@@
 <!-- can we think of 1-2 internationally well-known examples?
---> In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", https://honsefugl.nina.no/Innsyn/en).
-It is a line transect survey programme carried out annually in > xxxx locations across the country (xxxx > transects) by trained volunteers using pointing dogs.
+Check out MAPS (Monitoring of Avian Productivity and Survivorship) and CES (Constant Effort Sites) --> In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", <https://honsefugl.nina.no/Innsyn/en>).
+It is a line transect survey programme carried out annually in >120 localities across the country (>2000 transects) by trained volunteers using pointing dogs.
 The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata (+ref to protocol?) and established routines for quality control and annual releases of publicly available data via GBIF.
@@
 @horswill2019; @morrison2022].
-We then apply the resulting multi-area IDSM to "Hønsefuglportalen" data on willow ptarmigan to estimate population size, age-structure, survival, recruitment, and impacts of small rodent abundance across  41 reporting districts and 15 years (2007-2021) for willow ptarmigan.
+We then apply the resulting multi-area IDSM to "Hønsefuglportalen" data on willow ptarmigan to estimate population size, age-structure, survival, recruitment, and impacts of small rodent abundance across 41 reporting districts and 15 years (2007-2021) for this popular small-game species.
 We further embed the modelling workflow in a reproducible, semi-automated pipeline that will greatly facilitate the repeated calculation of abundance and population indicators at different spatial scales as new data is added every year.
@@
 
-The willow ptarmigan (*Lagopus lagopus*) has a circumpolar distribution, mainly inhabiting sub-alpine and arctic ecosystems [see e.g.
+The willow ptarmigan (*Lagopus lagopus*) is a tetraoid bird with a circumpolar distribution, mainly inhabiting sub-alpine and arctic ecosystems [see e.g.
 @fuglei2020].
-While the species is currently listed as Least Concern (LC) both in the global and Norwegian red list of species, it has undergone rather dramatic declines in abundance in Norway since the turn of the 20th century [@hjeljord2022].
-While the main reason for the long-term decline in abundance remain unresolved, they are considered sentinel species that are sensitive to both climate change and land use management [@Henden2017; @Storch2007].
-Moreover, being one of only a handful of bird species that spend the winter in mountain ecosystems in Scandinavia, they are important components of the ecosystem as prey species for the gyrfalcon [@Alastair2020].
-The willow ptarmigan has a relatively fast pace of life [@sandercock2005; @Steen1996], and can have substantial spatio-temporal variation in demographic rates [@Bowler2020].
+While the species is currently listed as Least Concern (LC) both in the global and Norwegian Red List of Species, it has undergone rather dramatic declines in abundance in Norway since the turn of the 20th century [@hjeljord2022].
+While the main reason for the long-term decline in abundance remain unresolved, willow ptarmigan are considered sentinel species that are sensitive to both climate change and land use changes [@Henden2017; @Storch2007].
+Moreover, being one of only a handful of bird species that spend the winter in mountain ecosystems in Scandinavia, they are important components of the ecosystem as prey species for the gyrfalcon [@Alastair2020] <!-- do we have other "important" species to add here.
+Arctic fox?
+Some other endangered beasts perhaps?
+-->.
+The willow ptarmigan has a relatively fast pace of life [@sandercock2005; @Steen1996], and can display substantial spatio-temporal variation in demographic rates [@Bowler2020].
 Their population dynamics are characterized by large inter-annual fluctuations in abundance [@hjeljord2022], and previous research has tied these fluctuations to rodent cycle through shared predators [@hagen1952; @Bowler2020].
@@
 Across their distributional range, they are iconic species with a high cultural value, partly linked to their popularity as game species.
-The fact that they are valued game species, information about spatio-temporal variation in demographic rates and population dynamics is particularly important in order to design sustainable harvest strategies.
-In addition, as a sentinel species the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is in fact included in both the main national biodiversity (Nature Index for Norway; refs) and ecosystem (Assessment of the Ecological Condition; refs) assessments.
-The model presented here therefore potentially has direct application in real world management and conservation contexts.
+The latter means that information about spatio-temporal variation in demographic rates and population dynamics is particularly important in order to design sustainable harvest strategies.
+In addition, as a sentinel species the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is included in both the main national biodiversity [Nature Index for Norway; @jakobsson2020; <https://www.naturindeks.no/Indicators/lirype>] and ecosystem [Assessment of the Ecological Condition; @framstad2022] assessments.
 
@@
 The line transect survey data were collected through a structured participatory monitoring program called "Hønsefuglportalen" (<https://honsefugl.nina.no/Innsyn/en>).
-In August each year, trained volunteer fieldworkers collect willow ptarmigan observations along predefined line transects.
+In August each year, trained volunteer fieldworkers collect observations of willow ptarmigan and other grouse species (rock ptarmigan *Lagopus muta*, black grouse *Lyrurus tetrix*, capercaillie *Tetrao urogallus*) along predefined line transects.
 To increase the detection probability, each team of fieldworkers use a pointing dog to locate the birds.
@@
 Often, more than one dog is used for a survey, but only one dog should be used at a time.
-The transect lines vary in length, but is typically between x and y km (range: xx, mean: yy).
-When an observation is made, the exact location of observation is noted, along with perpendicular distance to the transect line, and the age and sex of the observed birds.
-An observation typically include 1 - 12 birds (mean = 5.6), representing one brood (female and or male with young-of-the-year chicks).
+The transect lines vary in length, but are typically between 1 and 8 km (range: 0.3-16.2 km, median: 3 km).
+When birds are observed, the exact location of observation is noted, along with its perpendicular distance from the transect line, as well as the age and sex of the birds.
+An observation typically includes 1 - 12 birds (mean = 5.6), with groups > 1 typically representing one brood (female and or male with young-of-the-year chicks).
 When the surveys are conducted in August, the chicks of the year are able to fly but can be distinguished from older birds as they are still of smaller body size.
 Since 2019, most of the data has been collected using a mobile app tailored made for the monitoring program, where the field workers can register and get access to the transect lines allocated to them by the local organizers.
-Before 2019, data were collected on a dedicated field form, and entered manually in a web portal.
-After field data has been collected, several steps of quality control takes place before, both by local stakeholders and personnel from NINA.
+Prior to 2019, data were collected on a dedicated fieldwork form, and entered manually in a web portal afterwards.
+After field data has been registered, it undergoes several steps of quality control carried out by local stakeholders and personnel from the Norwegian Institute for Nature Research (NINA).
 Surveys are carried out on both public and private land.
 After an initial embargo period, all data from public land are published and made freely available as a sampling-event data set on GBIF (refs).
-This data set contain both meta data about the transect surveys (survey date, line transect length and location, study area ID and other information) as well as data pertaining to the observations (species, number of birds of different categories (adult males, adult females, juveniles, and birds of unknown category), perpendicular distance to transect line, exact location, and time of observation).
-For this study, we used only data from public land that is freely available on GBIF; thus the workflow we describe should be fully reproducible by others.
-Formally, the data is publsihed as three distinct data sets, based on the three main public land owners (Statskog, FeFo and Fjellstyrene, respectively).
+The published datasets contain both metadata about the transect surveys (survey date, line transect length and location, study area ID, etc.) as well as bird observation data (species, number of birds of different categories (adult males, adult females, juveniles, and birds of unknown category), perpendicular distance to transect line, exact location, and time of observation).
+Formally, the data from public land is published as three distinct data sets, one for each of the main public land administrators (Statskog, FeFo and Fjellstyrene, respectively).
+<!-- I changed "owners" to "administrators", just because to me "public" implies it is not owned by anyone.
+But if this is wrong, please change it back -->
 
-The program is not designed as a national monitoring, but rather a collection of local and regional survey programs.
+Notably, the program is not designed as a centralized national monitoring programme, but rather a collection of local and regional survey programs.
 All involved survey areas use a common field protocol and data collection model.
-In addition, the local study designs are reviewed by staff at NINA, and common recommendations for study design is provided.
-However, both a local scale and in particular at a regional scale, the design is unbalanced due to the fact that participation in voluntary by stakeholders.
-In general, the effort is higher in South-Eastern and Central Norway, medium in Northern Norway, and in general low in Western and Southern Norway.
-In total, we used data from 41 different survey areas, across xx counties and xx municipalities.
+In addition, the local study designs are reviewed by staff at NINA, and common recommendations for study design are provided.
+However, due to the fact that participation by stakeholders is voluntary, the spatial distribution of transects and sampling effort is not homogeneous.
+In general, the effort is higher in South-Eastern and Central Norway, medium in Northern Norway, and low in Western and Southern Norway.
+
+In this study we used all publicly available data for the period 2007-2021, which included a total of 2225 transects in 41 different survey areas spanning 9 counties and 50 municipalities.
+Transects on which no willow ptarmigan were ever observed (i.e.
+species absence likely due to low habitat suitability) were dropped, resulting in 2077 transects used in the analyses.
 
@@
 
-In our model, we integrated data from an ongoing field study of marked birds in Lierne municipality in Central Norway.
-These data were used to inform the survival component of the model (se below).
-The data is based on a sample of radio marked birds.
-The birds were captured in winter (late February or early March each year in the study period), and fitted with VHF collars.
-The marked birds were monitored on a regular basis until they either i) died, ii) battery stopped working, or iii) we lost contact with the bird for other reasons.
-For most of the year, the birds were monitored at least monthly.
-However, during the breeding and chick-rearing season (May to July) the birds were monitored more often, and during december and January we obtained few observations due to challenging field work conditions.
+The model of @nilsen2023 integrated line transect data with radio-telemetry data from from an ongoing field study of marked willow ptarmigans in Lierne municipality in Central Norway.
+From 2015 to 2019, around 50 birds were captured in winter (late February or early March) each year and fitted with VHF collars.
+The marked birds were then monitored on a regular basis until they either i) died, ii) their transmitter's battery stopped working, or iii) we lost contact with the bird for other reasons.
+For most of the year, the birds were monitored <!-- how were they actually monitored?
+Can we have a tiny bit more detail here?
+--> at least once a month.
+During the breeding and chick-rearing season (May to July) birds were monitored more often, and during December and January we obtained few observations due to challenging field work conditions.
 A proportion of the birds were harvest annually in the regular recreational harvest, and birds that were harvested were reported as shot to the field personnel.
-In addition, as the collars had mortality switch, we were also able to locate and retrieve a high proportion of birds that died for natural causes.
-We used data from the years 2015 - 2021, and the total sample size across these years was xx birds.
-The study is described in detail in @Israelsen2020 and in @Arnekleiv2022.
+In addition, as the collars had mortality switch, we were also able to locate and retrieve a high proportion of birds that died for natural causes, resulting in a known-fate mark-recapture dataset.
+The radio-telemetry study is described in detail in @Israelsen2020 and in @Arnekleiv2022.
+
+In this study we used data from years 2015 - 2020, and the total sample size across these years was 139 birds for the Aug-Jan period and 258 birds for the Feb-Jul period.
 

### 2023-12-20 Manuscript_Nater_etal.qmd
Added flags/comments for co-authors

@@
 
-\[TBA\]
+TBA
 
@@
 )
+<!-- @ERLEND & @JAMES, once the "new" results are in: thoughts on what we see that is interesting, cool, expected, unexpected etc.
+Can we compare to other large scale studies, either on other grouse-like creatures or harvested species more generally?
+-->
 
-## Implications for management (preliminary section title)
+## Implications for management 
+<!-- preliminary section title) -->
 
@@
 nest monitoring?)
+<!-- @ERLEND: Any thoughts on what the status here is?
+Might there be other data available from beyond Lierne?
+Are there exciting future prospects e.g.
+linked to automated monitoring with cameras at a larger scale?
+-->
 
@@
 INLA, may be required)
+<!-- @ERLEND: related to this, I want to discuss the potential of including the not publicly available data.
+Is there an overview somewhere over what the coverage (in terms of municipalities) would be if we had access to all the Hønsefuglportalen data?
+This might matter not just for increasing sample size, but perhaps also making CAR modelling (spatial structure through neighborhood relations) possible -->
+
+<!-- @MYSELF: Have a closer look at chapter 19 in IPM book to see what they did there and also what other work they reference -->
 
 -   Outlook: integration with / coupling to models for harvest to develop tools for (automatic) calculation of e.g. hunting quotas
+<!-- @JAMES: This one is for you.
+Some ruminations on how and why it would be great to formally model harvest, integration with hunting data, maybe eventually develop a tool to calculate quotas (I know this has been developed for other game species), etc.
+-->
 
@@
 -   Also: useful to link up to different reporting routes and/or more general (multi-indicator) projects
+<!-- @EVERYONE: Do we know of some good references for this? PRseumably we cannot cite our chapter in the Technical Manual yet :-P -->
 
@@
 
-\[TBA\]
+TBA ...
+Additionally, we would like to thank Bernardo Brandão Niebuhr dos Santos for his help with spatial data visualization. 
+<!-- @ERLEND: you maybe know best all the people/institutions we need to thank here -->
+<!-- @JAMES: should we be acknowledging your fullbright scholarship which made the collab possible? If so, please add details. -->
 
@@
 
-The auxiliary radio-telemetry data and all code, including the "targets pipeline" can be found in the project's repository on GitHub: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.
+The auxiliary radio-telemetry data, rodent occupancy data, and all code including the "targets pipeline" can be found in the project's repository on GitHub: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.
+The results presented in this paper were created using version 2.0 of the code (Zenodo citation TBA).
 

### 2023-12-20 Manuscript_Nilsen_Nater.qmd
minor text edits

@@
 DS has been used for estimating animal abundance in a wide range of contexts and for a variety of taxa [@buckland2015].
-One reason for the method's popularity is that it does not require repeated visits to the same sites for estimating detection probability (ref).
-Rather, it is assumed that detection probability decreases the further away from the transect line animals are located (ref).
+One reason for the method's popularity is that it does not require repeated visits to the same sites for estimating detection probability.
 This makes DS particularly useful for implementation in participatory monitoring programs, allowing stakeholders to take part in the data collection process.
-<!-- #It does not become evident why the fact that detection is assumed to decrease with distance makes a survey design more suitable for participatory monitoring.
-May need some more explanation.
--->
 
-The most common implementations of DS models have long used close-population formulations, and thus do not allow formulating a process model that projects abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates (ref).
+The most common implementations of DS models have long used close-population formulations, and thus do not allow formulating a process model that projects abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates @buckland2015.
 In recent years, DS approaches have been extended in many ways, including applications that estimate changes in abundance over time in open populations via a hidden state model representing population dynamics [@moore2011; @sollmann2015].
@@
 In fact, if the data does not contain information about the age (and/or sex-) structure of the surveyed population, there is no straightforward way to estimate demographic rates from such data.
-On the contrary, if age (and sex) of detected individuals can be determined, this information can be used to provide information on recruitment rates and survival probabilities (ref).
+On the contrary, if age (and sex) of detected individuals can be determined, this information can be used to provide information on recruitment rates and survival probabilities.
 @Nilsen2018, for example, used a model based on harvest statistics and observations of population age structure to estimate population abundance and demographic rates without the need for any additional data from marked individuals.
-<!-- #I have removed the mention of integrated population models from this sentence since those only get specifically introduced in the next paragraph -->
 
@@
 The model's core is a stage-structured matrix population model that projects population size from one time step to the next based on underlying survival and recruitment rates.
-<!-- # I suggest removing the following two sentences as they mostly re-state information we have already provided: "This is made possible by the fact that the line transect surveys provide data on population age structure which can be used to inform the estimation of demographic rates.
-In the IPM, we integrate these data with data from marked birds from which we estimate survival probabilities.
-The model is thus formally an IPM, as we jointly estimate parameters of interest based on several data sets (ref)." --> Below, we first present the model and assess its robustness and performance through application to simulated data.
+Below, we first present the model and assess its robustness and performance through application to simulated data.
 We then apply the model to real data collected from a willow ptarmigan (*Lagopus lagopus*) study in Central Norway.
@@
 Upper panel depicts model parameters, and lower panels show derived estimated of effective strip width and mean detection probability.
-Dots represents true values (black for global values, colored for dataset-specific values), and density plot represent the posterior distributions for each model run.
-](Figures/SimCheck_Detects_Temps.png){#fig-sim-detection}
+Dots represents true values (black for global values, colored for dataset-specific values), and density plot represent the posterior distributions for each model run.](Figures/SimCheck_Detects_Temps.png){#fig-sim-detection}
 
@@
 Upper panel show annual and seasonal survival probabilities, hyperparameters for the random effects model for recruitment (mean and sd), and lower panel show annual estimates of the recruitment rate.
-Dots represent true values (black for global values, colored for dataset-specific values), and density curves represent the posterior distributions from each model run.
-](Figures/SimCheck_VRs_Temps.png){#fig-sim-demo}
+Dots represent true values (black for global values, colored for dataset-specific values), and density curves represent the posterior distributions from each model run.](Figures/SimCheck_VRs_Temps.png){#fig-sim-demo}
 

### 2023-12-26 Manuscript_Nater_etal.qmd
Added code chunk for printing relevant stats for methods section

@@
 ---
-title: "Large-scale spatiotemporal variation in vital rates of an alpine bird"
+title: "Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird"
 format: html
@@
 Check out MAPS (Monitoring of Avian Productivity and Survivorship) and CES (Constant Effort Sites) --> In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", <https://honsefugl.nina.no/Innsyn/en>).
-It is a line transect survey programme carried out annually in >120 localities across the country (>2000 transects) by trained volunteers using pointing dogs.
+It is a line transect survey programme carried out annually in \>120 localities across the country (\>2000 transects) by trained volunteers using pointing dogs.
 The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata (+ref to protocol?) and established routines for quality control and annual releases of publicly available data via GBIF.
@@
 When birds are observed, the exact location of observation is noted, along with its perpendicular distance from the transect line, as well as the age and sex of the birds.
-An observation typically includes 1 - 12 birds (mean = 5.6), with groups > 1 typically representing one brood (female and or male with young-of-the-year chicks).
+An observation typically includes 1 - 12 birds (mean = 5.6), with groups \> 1 typically representing one brood (female and or male with young-of-the-year chicks).
 When the surveys are conducted in August, the chicks of the year are able to fly but can be distinguished from older birds as they are still of smaller body size.
@@
 
-In this study we used all publicly available data for the period 2007-2021, which included a total of 2225 transects in 41 different survey areas spanning 9 counties and 50 municipalities.
+In this study we used all publicly available data for the period 2007-2021, which included a total of 2225 transects in 41 different reporting districts spanning 9 counties and 50 municipalities.
 Transects on which no willow ptarmigan were ever observed (i.e.
@@
 
--   Introduction of integrated Distance sampling model (IDSM) from MS1
+@nilsen2023 recently developed an integrated distance sampling model (IDSM) which jointly analyses line transect and radio-telemetry data and applied it to willow ptarmigan in the Western part of Lierne municipality in Norway.
+The model consists of a deterministic population model with two age classes (juveniles and adults) and four data likelihoods: 1) likelihood for observation distances from transect lines for estimating detection probability; 2) likelihood for age-specific counts on transect surveys for estimating numbers of juveniles and adults present; 3) likelihood for juvenile to adult ratios observed at the locality level to provide estimate recruitment rate (as juveniles/adult); and 4) likelihood for known-fate telemetry data to estimate seasonal and annual survival.
+Below, we describe our new extension of this model to include data from several as opposed to just one.
+For more detailed information on the single-site model, including tests of model performance, see @nilsen2023.
 
 ### Multi-area model extension
+
+For applying the ptarmigan IDSM across all 41 reporting districts we included an area index in all model parameters (@fig-model) and enabled sharing of information among areas by explicitly modelling spatial variation alongside shared temporal and residual variation in vital rates and detection parameters.
+
+The spatially-explicit formulation of the two age-class population model can be written as:
+
+$$
+\begin{aligned}
+  D_{juv, x, t+1} & = D_{ad, x, t+1} * R_{x, t+1} \\
+  D_{ad, x, t+1} & = S_{x, t} * (D_{juv, x, t} + D_{ad, x, t})  
+\end{aligned}
+$$
+
+Here, $D_{juv, x, t}$ and $D_{ad, x, t}$ are the densities of juvenile and adult ptarmigan in area $x$ in year $t$, respectively.
+Both juveniles and adult survive from year $t$ to $t+1$ with an area- ($x$) and year- ($t$) specific survival probability $S_{x, t}$, and survivors produce the next generation of juveniles according to an area- and year-specific recruitment rate ($R_{x,t}$).
 
@@
 -   Temporal variation (shared year RE and "residual" variation)
-
-The model developed as part of this project and presented in @nilsen2023 was designed with the potential of extension to run on data from multiple areas simultaneously in mind, and this extension involves two steps in practice.
 
@@
 # Results
+
+```{r, echo = FALSE}
+# Posterior summaries
+post.sum <- readRDS("Results/PosteriorSummaries_byArea.rds")
+
+# Printing function
+printPostSum <- function(table, row = NULL, area = NULL){
+
+  if(!missing(area)){
+    row <- which(table$Area == area)
+  }
+
+  print.out <- list(
+    area.name = table$Area[row],
+    post.sum = paste0(
+      round(table$Median[row], digits = 3), 
+      " [",
+      round(table$lCI[row], digits = 3),
+      ", ",
+      round(table$uCI[row], digits = 3),
+      "]"
+    )
+  )
+  return(print.out)
+}
+
+# Detection stats
+#Det.hyper.mean
+#Det.sd.spatial
+#Det.sd.temporal
+#Det.sd.residual
+#Det.min
+#Det.max
+
+# Survival stats
+#S.hyper.mean
+#S.sd.spatial
+#S.sd.temporal
+#S.sd.residual
+S.min <- printPostSum(table = post.sum$pSurv.sum, 
+                      row = which(post.sum$pSurv.sum$Median == min(post.sum$pSurv.sum$Median)))
+S.max <- printPostSum(table = post.sum$pSurv.sum,
+                      row = which(post.sum$pSurv.sum$Median == max(post.sum$pSurv.sum$Median)))
+
+# Recruitment stats
+#R.hyper.mean
+#R.sd.spatial
+#R.sd.temporal
+#R.sd.residual
+R.min <- printPostSum(table = post.sum$rRep.sum, 
+                      row = which(post.sum$rRep.sum$Median == min(post.sum$rRep.sum$Median)))
+R.max <- printPostSum(table = post.sum$rRep.sum,
+                      row = which(post.sum$rRep.sum$Median == max(post.sum$rRep.sum$Median)))
+
+# Rodent effects stats
+#betaR.hyper.mean
+#betaR.sd.spatial
+#betaR.sd.temporal
+#betaR.sd.residual
+betaR.min <- printPostSum(table = post.sum$betaR.sum, 
+                      row = which(post.sum$betaR.sum$Median == min(post.sum$betaR.sum$Median)))
+betaR.max <- printPostSum(table = post.sum$betaR.sum,
+                      row = which(post.sum$betaR.sum$Median == max(post.sum$betaR.sum$Median)))
+n.pos.betaR <- length(which(post.sum$betaR.sum$Median > 0))
+n.neg.betaR <- length(which(post.sum$betaR.sum$Median < 0))
+
+# Population density stats
+popDens.sum <- subset(post.sum$popDens.sum, SummaryPeriod == "2018-2021")
+dens.min <- printPostSum(table = popDens.sum, 
+                      row = which(popDens.sum$Median == min(popDens.sum$Median)))
+dens.max <- printPostSum(table = popDens.sum,
+                      row = which(popDens.sum$Median == max(popDens.sum$Median)))
+
+# Population growth rate stats
+lambda.sum <- subset(post.sum$lambda.sum, SummaryPeriod == "2018-2021")
+lambda.min <- printPostSum(table = lambda.sum, 
+                      row = which(lambda.sum$Median == min(lambda.sum$Median)))
+lambda.max <- printPostSum(table = lambda.sum,
+                      row = which(lambda.sum$Median == max(lambda.sum$Median)))
+n.pos.lambda <- length(which(lambda.sum$Median > 1))
+n.neg.lambda <- length(which(lambda.sum$Median < 1))
+
+```
 
@@
 
-TBA
-
 # Discussion
@@
 better estimates through data sharing, relative importance of different kinds of variation, etc.
-)
-<!-- @ERLEND & @JAMES, once the "new" results are in: thoughts on what we see that is interesting, cool, expected, unexpected etc.
+) <!-- @ERLEND & @JAMES, once the "new" results are in: thoughts on what we see that is interesting, cool, expected, unexpected etc.
 Can we compare to other large scale studies, either on other grouse-like creatures or harvested species more generally?
@@
 ## Implications for management
+
 <!-- preliminary section title) -->
@@
 -   Limited usefulness of auxiliary radio-telemetry data as only from one area (big potential for including additional datasets from other small-scale studies, telemetry but also other, e.g.
-nest monitoring?)
-<!-- @ERLEND: Any thoughts on what the status here is?
+nest monitoring?) <!-- @ERLEND: Any thoughts on what the status here is?
 Might there be other data available from beyond Lierne?
@@
 Which is fortunately possible with NIMBLE but there is a chance that another platform, e.g.
-INLA, may be required)
-<!-- @ERLEND: related to this, I want to discuss the potential of including the not publicly available data.
+INLA, may be required) <!-- @ERLEND: related to this, I want to discuss the potential of including the not publicly available data.
 Is there an overview somewhere over what the coverage (in terms of municipalities) would be if we had access to all the Hønsefuglportalen data?
@@
 
--   Outlook: integration with / coupling to models for harvest to develop tools for (automatic) calculation of e.g. hunting quotas
-<!-- @JAMES: This one is for you.
+-   Outlook: integration with / coupling to models for harvest to develop tools for (automatic) calculation of e.g.
+hunting quotas <!-- @JAMES: This one is for you.
 Some ruminations on how and why it would be great to formally model harvest, integration with hunting data, maybe eventually develop a tool to calculate quotas (I know this has been developed for other game species), etc.
@@
 -   Why this matters: transparency and credibility, feedback to volunteer data collectors, cost-effective to re-do analyses for annual/regular reporting, easy to re-use and build upon, publicly accessible and well documented to allow more effective resource use in follow-up projects (sustainable use of funding), etc.
--   Also: useful to link up to different reporting routes and/or more general (multi-indicator) projects
-<!-- @EVERYONE: Do we know of some good references for this? PRseumably we cannot cite our chapter in the Technical Manual yet :-P -->
+-   Also: useful to link up to different reporting routes and/or more general (multi-indicator) projects <!-- @EVERYONE: Do we know of some good references for this?
+PRseumably we cannot cite our chapter in the Technical Manual yet :-P -->
 
@@
 Additionally, we would like to thank Bernardo Brandão Niebuhr dos Santos for his help with spatial data visualization.
-<!-- @ERLEND: you maybe know best all the people/institutions we need to thank here -->
-<!-- @JAMES: should we be acknowledging your fullbright scholarship which made the collab possible? If so, please add details. -->
+<!-- @ERLEND: you maybe know best all the people/institutions we need to thank here --> <!-- @JAMES: should we be acknowledging your fullbright scholarship which made the collab possible?
+If so, please add details.
+-->
 

### 2023-12-26 R/summarisePost_areas.R
Added stats for detection parameters

@@ -100,7 +100,7 @@ summarisePost_areas <- function(mcmc.out,
   ## Summarize posteriors for relevant parameters
   
   # Prepare matrices for storage of results
-  rRep.sum <- pSurv.sum <- betaR.sum <- data.frame()
+  rRep.sum <- pSurv.sum <- betaR.sum <- detect.sum <- data.frame()
   popDens.sum <- lambda.sum <- data.frame()
   
   for(i in 1:N_areas){
@@ -142,6 +142,16 @@ summarisePost_areas <- function(mcmc.out,
       betaR.sum <- NA
     }
     
+    # Summarize detection decay parameter
+    detect_name <- paste0("mu.dd[",  i, "]")
+    detect_add <- data.frame(Area = area_names[i],
+                           Median = median(exp(out.mat[, detect_name])),
+                           lCI = unname(quantile(exp(out.mat[, detect_name]), probs = 0.025)),
+                           uCI = unname(quantile(exp(out.mat[, detect_name]), probs = 0.975)),
+                           Mean = mean(exp(out.mat[, detect_name])),
+                           SD = sd(exp(out.mat[, detect_name]))
+    )
+    detec.sum <- rbind(detect.sum, detect_add)
     
     # Summarize average population densities
     Dens_names <- c(paste0("densAvg_tot[",  i, "]"), paste0("densAvg_shared[",  i, "]"))

### 2023-12-26 R/summarisePost_areas.R
Added stats for hyper-parameters

@@ -194,12 +194,29 @@ summarisePost_areas <- function(mcmc.out,
     lambda.sum <- rbind(lambda.sum, lambda_tot_add, lambda_shared_add)
   }
   
+  ## Summarize hyper-parameters
+  hParams.list <- c("h.Mu.R", "h.sigma.R", "sigmaT.R", "sigmaR.R",
+                    "h.Mu.betaR.R", "h.sigma.betaR.R",
+                    "h.Mu.S","h.sigma.S", "sigmaT.S", "sigmaR.S",
+                    "h.mu.dd", "h.sigma.dd", "sigmaT.dd", "sigmaR.dd")
+  hParams.sum <- data.frame()
+  for(i in 1:length(hParams.list)){
+    hParams_add <- data.frame(Parameter = hParams.list[i],
+                              Median = median(out.mat[, hParams.list[i]]),
+                              lCI = unname(quantile(out.mat[, hParams.list[i]], probs = 0.025)),
+                              uCI = unname(quantile(out.mat[, hParams.list[i]], probs = 0.975)),
+                              Mean = mean(out.mat[, hParams.list[i]]),
+                              SD = sd(out.mat[, hParams.list[i]]))
+    hParams.sum <- rbind(hParams.sum, hParams_add)
+  }
+  
   ## Collect summary data into a list 
   PostSum.list <- list(rRep.sum = rRep.sum,
                        pSurv.sum = pSurv.sum,
                        betaR.sum = betaR.sum,
                        popDens.sum = popDens.sum,
-                       lambda.sum = lambda.sum)
+                       lambda.sum = lambda.sum,
+                       hParams.sum = hParams.sum)
   
   ## Save (optional) and return
   if(save){

### 2023-12-26 R/summarisePost_areas.R
Added CV as a summary metric

@@ -112,7 +112,8 @@ summarisePost_areas <- function(mcmc.out,
                            lCI = unname(quantile(out.mat[, rRep_name], probs = 0.025)),
                            uCI = unname(quantile(out.mat[, rRep_name], probs = 0.975)),
                            Mean = mean(out.mat[, rRep_name]),
-                           SD = sd(out.mat[, rRep_name])
+                           SD = sd(out.mat[, rRep_name]),
+                           CV = sd(out.mat[, rRep_name]) / mean(out.mat[, rRep_name])
     )
     rRep.sum <- rbind(rRep.sum, rRep_add)
     
@@ -123,7 +124,8 @@ summarisePost_areas <- function(mcmc.out,
                             lCI = unname(quantile(out.mat[, pSurv_name], probs = 0.025)),
                             uCI = unname(quantile(out.mat[, pSurv_name], probs = 0.975)),
                             Mean = mean(out.mat[, pSurv_name]),
-                            SD = sd(out.mat[, pSurv_name])
+                            SD = sd(out.mat[, pSurv_name]),
+                            CV = sd(out.mat[, pSurv_name])/mean(out.mat[, pSurv_name])
     )
     pSurv.sum <- rbind(pSurv.sum, pSurv_add)
     
@@ -135,8 +137,10 @@ summarisePost_areas <- function(mcmc.out,
                               lCI = unname(quantile(out.mat[, betaR_name], probs = 0.025)),
                               uCI = unname(quantile(out.mat[, betaR_name], probs = 0.975)),
                               Mean = mean(out.mat[, betaR_name]),
-                              SD = sd(out.mat[, betaR_name])
+                              SD = sd(out.mat[, betaR_name]),
+                              CV = sd(out.mat[, betaR_name]) / mean(out.mat[, betaR_name])
       )
+      
       betaR.sum <- rbind(betaR.sum, betaR_add)
     } else{
       betaR.sum <- NA
@@ -149,7 +153,8 @@ summarisePost_areas <- function(mcmc.out,
                            lCI = unname(quantile(exp(out.mat[, detect_name]), probs = 0.025)),
                            uCI = unname(quantile(exp(out.mat[, detect_name]), probs = 0.975)),
                            Mean = mean(exp(out.mat[, detect_name])),
-                           SD = sd(exp(out.mat[, detect_name]))
+                           SD = sd(exp(out.mat[, detect_name])),
[…]

### 2023-12-26 R/summarisePost_areas.R
Small fix to return detection parameter summaries

@@ -156,7 +156,7 @@ summarisePost_areas <- function(mcmc.out,
                            SD = sd(exp(out.mat[, detect_name])),
                            CV = sd(exp(out.mat[, detect_name])) / mean(exp(out.mat[, detect_name])) 
     )
-    detec.sum <- rbind(detect.sum, detect_add)
+    detect.sum <- rbind(detect.sum, detect_add)
     
     # Summarize average population densities
     Dens_names <- c(paste0("densAvg_tot[",  i, "]"), paste0("densAvg_shared[",  i, "]"))
@@ -225,6 +225,7 @@ summarisePost_areas <- function(mcmc.out,
   PostSum.list <- list(rRep.sum = rRep.sum,
                        pSurv.sum = pSurv.sum,
                        betaR.sum = betaR.sum,
+                       detect.sum = detect.sum,
                        popDens.sum = popDens.sum,
                        lambda.sum = lambda.sum,
                        hParams.sum = hParams.sum)

### 2023-12-26 Analysis_RealData.R
Added saving of summary data

@@ -273,11 +273,12 @@ if(fitRodentCov){
 # OPTIONAL: CALCULATE AND PLOT VARIANCE DECOMPOSITION #
 #-----------------------------------------------------#
 
-VarDecomp <- plotVarDecomposition(mcmc.out = IDSM.out.tidy, 
-                                  N_areas = input_data$nim.constants$N_areas, 
-                                  N_years = input_data$nim.constants$N_years, 
-                                  fitRodentCov = fitRodentCov, 
-                                  RodentOcc_data = input_data$nim.data$RodentOcc)
+plotVarDecomposition(mcmc.out = IDSM.out.tidy, 
+                     N_areas = input_data$nim.constants$N_areas, 
+                     N_years = input_data$nim.constants$N_years, 
+                     fitRodentCov = fitRodentCov, 
+                     RodentOcc_data = input_data$nim.data$RodentOcc,
+                     saveResults = TRUE)
   
 
 # OPTIONAL: MAP PLOTS #

### 2023-12-26 R/plotVarDecomposition.R

@@ -8,15 +8,16 @@
 #' @param RodentOcc_data numeric vector containing the rodent occupancy data
 #' used for fitting the model. Optional argument that needs to be provided if
 #' fitRodenCov = TRUE.
+#' @param saveResults logical. If TRUE (default), saves posterior summaries as
+#' an RDS file into working directory. 
 #'
-#' @return a list containing a dataframe of posterior distributions for proportions
-#' of variance explained by different components, as well as the path to the 
-#' plot. 
+#' @return a vector containinf the path to the plot, as well as to the saved
+#' posterior summaries of variance decomposition (if saveResults = TRUE).  
 #' @export
 #'
 #' @examples
 
-plotVarDecomposition <- function(mcmc.out, N_areas, N_years, fitRodentCov, RodentOcc_data = NULL){
+plotVarDecomposition <- function(mcmc.out, N_areas, N_years, fitRodentCov, RodentOcc_data = NULL, saveResults = TRUE){
   
   ## Check rodent data is provided if needed
   if(fitRodentCov & missing("RodentOcc_data")){
@@ -114,6 +115,19 @@ plotVarDecomposition <- function(mcmc.out, N_areas, N_years, fitRodentCov, Roden
     propVar.all <- subset(propVar.all, VarComponent != "RodentOcc")
   }
   
+  # Summarised
+  propVar.sum <- propVar.all %>%
+    dplyr::group_by(Parameter, VarComponent) %>%
+    dplyr::summarise(median = median(PropVar),
+                     lCI = quantile(PropVar, probs = 0.025),
+                     uCI = quantile(PropVar, probs = 0.975),
+                     .groups = "keep")
+  
+  ## Save posterior data
+  if(saveResults){
+    saveRDS(propVar.sum, "PosteriorSummaries_VarDecomp.rds")
+  }
+  
   ## Make plotting directory if it does not exist yet
   ifelse(!dir.exists("Plots/VarDecomposition"), dir.create("Plots/VarDecomposition"), FALSE)
   
@@ -132,8 +146,12 @@ plotVarDecomposition <- function(mcmc.out, N_areas, N_years, fitRodentCov, Roden
   )
   dev.off()
   
-  ## Return data and plot path
-  return(list(propVar.post = propVar.all, 
-              plotPath = "Plots/VarDecomposition/VarDecomposition_overall.pdf"))
+  ## Return data and plot paths
+  if(saveResults){
[…]

### 2023-12-26 R/plotMaps.R
Added detection param map & CV map for all quantities

@@ -22,6 +22,11 @@ plotMaps <- function(PostSum.list, mapNM,
   
   ## Add estimates to map objects
   
+  # Average detection parameters
+  mapNM.detect <- mapNM
+  mapNM.detect <- mapNM.detect %>%
+    dplyr::left_join(., PostSum.list$detect.sum, by = "Area") 
+  
   # Average reproductive rates
   mapNM.rRep <- mapNM
   mapNM.rRep <- mapNM.rRep %>%
@@ -64,6 +69,23 @@ plotMaps <- function(PostSum.list, mapNM,
   
   ## Plot maps and print to pdf
   
+  # Average detection parameters
+  pdf("Plots/AreaMaps/Avg_detect_Map.pdf", width = 5, height = 6)
+  
+  print(
+    tmap::tm_shape(mapNM.detect) + tmap::tm_polygons("Median", palette = "plasma", style = "cont", colorNA = "grey80")
+  )
+  
+  print(
+    tmap::tm_shape(mapNM.detect) + tmap::tm_polygons("SD", palette = rev("BuGn"), style = "cont", colorNA = "grey80")
+  )
+  
+  print(
+    tmap::tm_shape(mapNM.detect) + tmap::tm_polygons("CV", palette = rev("BuGn"), style = "cont", colorNA = "grey80")
+  )
+  
+  dev.off()
+  
   # Average reproductive rates
   pdf("Plots/AreaMaps/Avg_rRep_Map.pdf", width = 5, height = 6)
   
@@ -73,6 +95,10 @@ plotMaps <- function(PostSum.list, mapNM,
   
   print(
     tmap::tm_shape(mapNM.rRep) + tmap::tm_polygons("SD", palette = rev("BuGn"), style = "cont", colorNA = "grey80")
+  )
+  
+  print(
+    tmap::tm_shape(mapNM.rRep) + tmap::tm_polygons("CV", palette = rev("BuGn"), style = "cont", colorNA = "grey80")
   )
   
   dev.off()
@@ -88,6 +114,10 @@ plotMaps <- function(PostSum.list, mapNM,
     tmap::tm_shape(mapNM.pSurv) + tmap::tm_polygons("SD", palette = rev("BuGn"), style = "cont", colorNA = "grey80")
   )
   
+  print(
+    tmap::tm_shape(mapNM.pSurv) + tmap::tm_polygons("CV", palette = rev("BuGn"), style = "cont", colorNA = "grey80")
+  )
+  
   dev.off()
   
   # Rodent effects
[…]

### 2023-12-26 R/plotLatitude.R
Added hlines for beta and lambda plots

@@ -54,6 +54,7 @@ plotLatitude <- function(PostSum.list,
     
     p_betaR <- ggplot(betaR.sum) + 
       geom_pointrange(aes(x = Latitude, y = Median, ymin = lCI, ymax = uCI, colour = Longitude), size = 0.5, fatten = 4, alpha = 0.5) +
+      geom_hline(aes(yintercept = 0), color = "grey70", linetype = "dotted") + 
       ylab("Estimate") + 
       paletteer::scale_color_paletteer_c("grDevices::Temps") + 
       theme_classic()
@@ -81,12 +82,14 @@ plotLatitude <- function(PostSum.list,
   
   p_lambda1 <- ggplot(subset(lambda.sum, SummaryPeriod == paste0(minYear, "-", maxYear))) + 
     geom_pointrange(aes(x = Latitude, y = Median, ymin = lCI, ymax = uCI, colour = Longitude), size = 0.5, fatten = 4, alpha = 0.5) +
+    geom_hline(aes(yintercept = 1), color = "grey70", linetype = "dotted") + 
     ylab("Estimate") + 
     paletteer::scale_color_paletteer_c("grDevices::Temps") + 
     theme_classic()
   
   p_lambda2 <- ggplot(subset(lambda.sum, SummaryPeriod != paste0(minYear, "-", maxYear))) + 
     geom_pointrange(aes(x = Latitude, y = Median, ymin = lCI, ymax = uCI, colour = Longitude), size = 0.5, fatten = 4, alpha = 0.5) +
+    geom_hline(aes(yintercept = 1), color = "grey70", linetype = "dotted") + 
     ylab("Estimate") + 
     paletteer::scale_color_paletteer_c("grDevices::Temps") + 
     theme_classic()

### 2023-12-26 R/plotLatitude.R
Added upper bound 500) for plotting average density across all years

@@ -66,6 +66,7 @@ plotLatitude <- function(PostSum.list,
   
   p_popDens1 <- ggplot(subset(popDens.sum, SummaryPeriod == paste0(minYear, "-", maxYear))) + 
     geom_pointrange(aes(x = Latitude, y = Median, ymin = lCI, ymax = uCI, colour = Longitude), size = 0.5, fatten = 4, alpha = 0.5) +
+    ylim(0, 500) + 
     ylab("Estimate") + 
     paletteer::scale_color_paletteer_c("grDevices::Temps") + 
     theme_classic()

### 2023-12-26 R/plotLatitude.R
Changed plot order

@@ -103,8 +103,8 @@ plotLatitude <- function(PostSum.list,
   if(fitRodentCov){
     pdf("Plots/Latitude/VitalRateParams_Latitude.pdf", width = 8, height = 6)
     print(
-      ggpubr::ggarrange(p_rRep + ggtitle("Average recruitment rate") + theme(axis.title.x = element_blank()), 
-                        p_pSurv + ggtitle("Average survival rate") + theme(axis.title.x = element_blank()),  
+      ggpubr::ggarrange(p_pSurv + ggtitle("Average survival rate") + theme(axis.title.x = element_blank()),
+                        p_rRep + ggtitle("Average recruitment rate") + theme(axis.title.x = element_blank()), 
                         p_betaR + ggtitle("Rodent effect on recruitment"),
                         heights = c(1, 1, 1.2), ncol = 1, 
                         legend = "right", common.legend = TRUE)

### 2023-12-27 Manuscript_Nater_etal.qmd
Completed model description

@@
 \begin{aligned}
-  D_{juv, x, t+1} & = D_{ad, x, t+1} * R_{x, t+1} \\
-  D_{ad, x, t+1} & = S_{x, t} * (D_{juv, x, t} + D_{ad, x, t})  
+  D_{juv, x, j, t+1} & = D_{ad, x, j, t+1} * R_{x, t+1} \\
+  D_{ad, x, j, t+1} & = S_{x, t} * (D_{juv, x, j, t} + D_{ad, x, j, t})  
 \end{aligned}
@@
 
-Here, $D_{juv, x, t}$ and $D_{ad, x, t}$ are the densities of juvenile and adult ptarmigan in area $x$ in year $t$, respectively.
+Here, $D_{juv, j, x, t}$ and $D_{ad, j, x, t}$ are the densities of juvenile and adult ptarmigan in survey site (= transect) $j$ of area $x$ in year $t$, respectively.
 Both juveniles and adult survive from year $t$ to $t+1$ with an area- ($x$) and year- ($t$) specific survival probability $S_{x, t}$, and survivors produce the next generation of juveniles according to an area- and year-specific recruitment rate ($R_{x,t}$).
-
--   Addition of spatial dimension
--   Hyper-parameter distributions for spatial variation
--   Temporal variation (shared year RE and "residual" variation)
-
-The first step is the inclusion of an additional dimension, area, to enable using data collected from multiple areas (@fig-model).
-The area-dimension is added across all model components with the exception of the known-fate telemetry data (and seasonal decomposition of survival estimated from it), as that is only available for the Lierne area.
 
@@
 R\[x,t\] = recruitment rate in year t.
-S\[x,t\] = survival probability from year t to t+1.](Figures/RypeIDSM_multiArea_structure.png){#fig-model}
+S\[x,t\] = survival probability from year t to t+1.
+Note that the additional site (=transect) dimension, "j", is omitted for the sake of illustration.](Figures/RypeIDSM_multiArea_structure.png){#fig-model}
 
-The second step is making adjustments to capitalize on the integration of data from multiple areas to make maximal use of available information.
-There are several not mutually exclusive options for approaching this, and one of them is the use of "hyper-parameters" for sharing information across areas [e.g.
-@horswill2019; @horswill2021].
-As an illustrative example, take the area-specific average annual survival probability, $\mu^S_x$.
-Instead of assigning independent non-informative priors to each $\mu^S_x$, we now assume that the variation in average survival across space can be described by a common (logit) normal distribution such that:
+The initial densities, $D_{juv, x, j, 1}$ and $D_{ad, x, j, 1}$ are modelled for each site (= transect) as random realizations of log normal distributions with area-specific log means ($\mu^{D1}_x$) and log standard deviations ($\sigma^{D1}_x$).
+Survival ($S_{x,t}$) and recruitment ($R_{x,t}$), on the other hand, are assumed to be the same for all sites $j$ within a given area $x$ and were modelled as:
 
 $$
-\begin{equation}
-\mu^S_x \sim Normal(logit(h.\mu^S), h.\sigma^S)
-\end{equation}
-$$ where $h.\mu^S$ is the "hyper mean" and $h.\sigma^S$ is the "hyper standard deviation". Non-informative priors are then assigned to the hyper-parameters instead of the area-specific realizations:
+\begin{aligned}
+  logit(S_{x, t}) & = logit(\mu^S) + \varepsilon^{X.S}_{x} + \varepsilon^{T.S}_{t} + \varepsilon^{R.S}_{x,t} \\
+    log(R_{x, t}) & = log(\mu^R) + \beta_x*rodentOcc_{x,t} +  \varepsilon^{X.R}_{x} + \varepsilon^{T.R}_{t} + \varepsilon^{R.R}_{x,t}
+\end{aligned}
+$$ The global means ,$\mu$, and the normally distributed spatial random effects, $\varepsilon^X$, represent the equivalent of what is elsewhere referred to as "hyper-parameter distributions" for sharing information on demographic rates across areas [e.g.
+@horswill2019; @horswill2021].
+We also used this same approach for defining the area-specific effects ($\beta_x$) of local yearly rodent occupancy ($rodentOcc_{x,t}$) on recruitment.
+In addition to spatial variation in survival and recruitment, we also included large-scale temporal variation through random year effects that were shared by across all areas ($\varepsilon^T_t$) and otherwise unaccounted for variation through year- and area-specific residual random effects ($\varepsilon^R_{x,t}$).
+Spatial, temporal, and residual random effects were modelled as normally distributed with globally defined (= shared) standard deviations.
 
-$$
-\begin{align}
-h.\mu^S & \sim Uniform(0,1) \\
-h.\sigma^S & \sim Uniform(0, 5)
-\end{align}
-$$
-
-For the multi-area implementation of the IDSM, we used the equivalent of the above approach for parameters representing average survival probabilities, baseline recruitment rates, rodent effect on recruitment, and average distance sampling detection.
-Initial population density, on the other hand, was estimated independently for each area.
-Random effects standard deviations were assumed to be the same for all areas, effectively resulting in space-for-time subtitution for random effect estimation.
-
-Another approach to maximising information gain when jointly analysing data from different areas is explicit modelling of spatial and spatio-temporal variation.
+The three likelihoods for data resulting from the line transect sampling (observation distances, age-specific counts, and juvenile to adult ratios, see above) were also formulated as spatially explicit, with year- and area-specific distance sampling detection parameters modelled in the same way as survival and recruitment (bar the effect of rodent occupancy, @fig-model).
+For the known-fate telemetry data (and the seasonal decomposition of survival estimated from it), on the other hand, we did not add an additional area dimension as this data was only available for the Lierne area.
 

### 2023-12-27 Manuscript_Nater_etal.qmd
Added model implementation section

@@
 
--   Bayesian, NIMBLE
--   Priors
--   Default samplers
--   Chains, iterations, burn-in, thinning, posterior samples
+We implemented our multi-area IDSM in a Bayesian framework using NIMBLE version 1.0.1 [@valpine2017] in R version 4.3.1 [@R2023].
+For the likelihood for line transect observation distances we used a custom half-normal distribution developed by Michael Scroggie in the "nimbleDistance" package (https://github.com/scrogster/nimbleDistance).
+We used non-informative uniform priors for all parameters and including biologically sensible boundaries where possible.
+We simulated complete sets of initial values for all model nodes prior to model running and using pre-defined seeds to ensure reproducibility.
+Using NIMBLE's standard samplers, we then ran 4 MCMC chains of 100k iterations each.
+We discarded the first 40k samples of each chain as burn-in, and thinned the remainder by a factor 20, resulting in a final joint posterior containing a total of 4 x 3k = 120k samples (note that high thinning rates were necessary to constrain memory load of the joint posterior, which included 314568 monitored parameters).
 
-## Variance decomposition
+## Post-hoc variance decomposition
 
@@
 
-Reproducibility and ease of repeating analyses was a key focus when developing the multi-area IDSM.To that end, we set up the workflow as a "targets pipeline", implemented throught the R packages "targets" [@landau2021].
+Reproducibility and ease of repeating analyses was a key focus when developing the multi-area IDSM.To that end, we set up the workflow as a "targets pipeline", implemented throught the R package "targets" [@landau2021].
 A visual representation of the pipeline is displayed in @fig-workflow.

### 2023-12-28 Manuscript_Nater_etal.qmd
Added some references

@@
 There is growing demand for biodiversity indicators from international unions, national government, local management bodies, and corporate and industry actors.
-Such indicators ideally represent a wide range of biodiversity's states and functions (e.g.
-+ ref on basic EBV idea), yet the development of suitable indicators for certain qualities, such as species abundance and demography, has been more difficult than for others (+ref).
+Such indicators ideally represent a wide range of biodiversity's states and functions [e.g.
+Essential Biodiversity Variables, @pereira2013; @jetz2019], yet the development of suitable indicators for certain qualities, such as species abundance and demography, has been more difficult than for others [@schmeller2018; @waldock2022].
 This is at least partially due to challenging requirements regarding spatial scales of useful biodiversity indicators.
-On one hand, indicators need to be representative at large geographic scales, for example, in the context of countries' reporting towards biodiversity targets (+ ref).
-On the other hand, indicators also ideally have good spatial resolution as the scales relevant for local-level management and planning are often much smaller (+ ref).
+On one hand, indicators need to be representative at large geographic scales, for example, in the context of countries' reporting towards biodiversity targets [e.g.
+@feld2009].
+On the other hand, indicators also ideally have good spatial resolution as the scales relevant for local-level management and planning are often much smaller [@stevenson2021].
 This latter requirement is particularly crucial for (infrastructure) development strategies and for species management and conservation, both of which tend to require knowledge on species abundance and population dynamics (i.e.
-demographic rates) that is relevant for their county- or municipality-level decision making (+ refs).
-Another reason why abundance and population indicators ideally come with good spatial resolution is that there can be substantial amounts of variation in population dynamics and life history of species across space (+ref).
+demographic rates) that is relevant for their county- or municipality-level decision making (+ refs?).
+Another reason why abundance and population indicators ideally come with good spatial resolution is that there can be substantial amounts of variation in population dynamics and life history of species across space [e.g.
+@robinson2014; @horswill2019].
 This is variation that needs to be accounted for in order to develop successful and sustainable strategies for area use, (harvest) management, and species and biodiversity conservation [@williams2002].
 
-While large-scale, spatially-explicit indicators for species abundance and populations are clearly needed, development and practical implementation are greatly limited due to the reliance of such indicators on the availability of data from large-scale, long-term monitoring programmes (+ref).
+While large-scale, spatially-explicit indicators for species abundance and populations are clearly needed, development and practical implementation are greatly limited due to the reliance of such indicators on the availability of data from large-scale, long-term monitoring programmes \[@proencca2017\].
 Consequently, many countries have been working on setting up, maintaining, and improving such monitoring programmes over the last decades.
@@
 There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly and requires trained specialists, thus limiting the amount of data that can be collected, while participatory monitoring, i.e.
-the collection of ecological data by members of the public (also called citizen science, +ref), allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias (+ref).
-Consequently, many large-scale monitoring programmes are limited to presence-absence or very simple count observations, making them suitable for the development of indicators of distribution and population trends, but usually not for indicators of abundance, population dynamics, and demographic rates (+ ref).
+the collection of ecological data by members of the public [also called citizen science, @fraisl2022], allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias [@johnston2023].
+Consequently, many large-scale monitoring programmes are often limited to presence-absence or very simple count observations, making them suitable for the development of indicators of distribution and population trends, but usually not for indicators of abundance, population dynamics, and demographic rates (+ example refs).
 The exception here are monitoring programmes that succeed in making use of a large number of citizen scientists that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol.
@@
 
--   How variance decomposition was done
+Following model fitting, we calculated posterior distributions for the proportions of variance in survival probabilities, recruitment rates, and detection decay explained by 1) spatial variation ($var_{area}$), 2) temporal variation ($var_{year}$), 3) residual variation ($var_{residual}$), and 4) variation in rodent occupancy ($var_{rodent}$).
+To obtain the proportion variance explained by each of the component, we divided it by the sum of all other components ($var_{area} + var_{year} + var_{residual} + var_{rodent}$).
+The spatial, temporal, and residual variance components were defined as the square of the estimated corresponding random effects standard deviation from the model while $var_{rodent}$ was calculated as the variance of all area- and year-specific $\beta_x*rodentOcc_{x,t}$ products.
 
@@
 
-Reproducibility and ease of repeating analyses was a key focus when developing the multi-area IDSM.To that end, we set up the workflow as a "targets pipeline", implemented throught the R package "targets" [@landau2021].
-A visual representation of the pipeline is displayed in @fig-workflow.
+Reproducibility and ease of repeating analyses was a key focus when developing the multi-area IDSM.To that end, we set up the workflow as a "targets pipeline", implemented through the R package "targets" [@landau2021].
+The pipeline contains a variety of options for controlling modelling decisions in the workflow such as the year range of data to consider, the level of spatial aggregation (i.e.
+reporting district vs.
+survey locality), whether to model time variation in survival and/or effects of rodent abundance, whether to run MCMC chains sequentially or in parallel, etc.
+A visual representation of the pipeline is also displayed in @fig-workflow and for more details on pipeline implementation and options, we refer the reader to the GitHub repository: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.
 
@@
 ```
+
+All numerical results in the following are presented as median \[95% credible interval unless otherwise indicated.
+Posterior summaries (median, 95% credible interval, mean, standard deviation, coefficient of variation) for all main parameters are also provided in the supplementary file "PosteriorSummaries_byAreas.csv".
 
@@
 
-The auxiliary radio-telemetry data, rodent occupancy data, and all code including the "targets pipeline" can be found in the project's repository on GitHub: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.
+The auxiliary radio-telemetry data, rodent occupancy data, and all code including the "targets pipeline" can be found in the project's repository on GitHub:vv.
 The results presented in this paper were created using version 2.0 of the code (Zenodo citation TBA).

### 2023-12-28 Manuscript_Nater_etal.qmd
Added results (pop. density)

@@
 
-While large-scale, spatially-explicit indicators for species abundance and populations are clearly needed, development and practical implementation are greatly limited due to the reliance of such indicators on the availability of data from large-scale, long-term monitoring programmes \[@proencca2017\].
+While large-scale, spatially-explicit indicators for species abundance and populations are clearly needed, development and practical implementation are greatly limited due to the reliance of such indicators on the availability of data from large-scale, long-term monitoring programmes [@proencca2017].
 Consequently, many countries have been working on setting up, maintaining, and improving such monitoring programmes over the last decades.
@@
 
-All numerical results in the following are presented as median \[95% credible interval unless otherwise indicated.
+All numerical results in the following are presented as median \[95% credible interval\] unless otherwise indicated.
 Posterior summaries (median, 95% credible interval, mean, standard deviation, coefficient of variation) for all main parameters are also provided in the supplementary file "PosteriorSummaries_byAreas.csv".
+Supplementary figures (SF) are provided as .pdf files with captions in "SuppFigures_Captions.txt".
 
@@
 
-@fig-densA shows that population density is generally lowest in the Northern parts of Norway and highest in central Norway, close to the Swedish borders.
-Higher median density estimates also seem to be associated with higher estimation uncertainty (@fig-densB) and therefore have to be interpreted with care.
+Only during the most recent four years (2018-2021) has data been collected regularly for all reporting areas included in the analyses.
+During this period, estimated population densities varied between `r dens.min$post.sum` birds/km^2 in the area "`r dens.min$area.name`" close to the coast in central Norway to `r dens.max$post.sum` birds/km^2 in "`r dens.max$area.name`" further south near the Swedish border.
+In general, population density appeared to be lowest in the Northern parts of Norway and highest in the eastern part of central Norway (@fig-densA).
+Uncertainty in density estimates was relatively consistent, with a few areas (inlcuding the one with the lowest estimated density, "`r dens.min$area.name`") sticking out by having substantially less precise estimates (@fig-densB).
+Populations fluctuated substantially over time in any given area (SF "TimeSeries_popDens1.pdf") and some years seemed to be indicative of relatively high (e.g.
+2011, 2014, 2018) or low (e.g.
+2012, 2015) densities across a large number of areas.
 
 ::: {#fig-dens layout-ncol="2"}
-![Median density estimates](Figures/density_2018-2020_Median.png){#fig-densA}
+![Median density estimates](Figures/Avg_popDens_Map-4.png){#fig-densA}
 
-![Uncertainty in density estimates](Figures/density_2018-2020_SD.png){#fig-densB}
+![Uncertainty in density estimates](Figures/Avg_popDens_Map-6.png){#fig-densB}
 
-Median (a) and standard deviation (b) of posterior estimates of average ptarmigan density in the three most recent years (2018-2020) across 41 reporting areas in Norway.
+Median (a) and coefficient of variation (standard deviation / mean) (b) of posterior estimates of average ptarmigan density in the four most recent years (2018-2021) across 41 reporting areas (summarised at the municipality level) in Norway.
 Darker colors indicate higher median values and higher uncertainty.
@@
 
-The auxiliary radio-telemetry data, rodent occupancy data, and all code including the "targets pipeline" can be found in the project's repository on GitHub:vv.
+The auxiliary radio-telemetry data, rodent occupancy data, and all code including the "targets pipeline" can be found in the project's repository on GitHub: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.
 The results presented in this paper were created using version 2.0 of the code (Zenodo citation TBA).

### 2023-12-28 R/plotLatitude.R
Added additional plot with popDens vs. lambda

@@ -105,6 +105,30 @@ plotLatitude <- function(PostSum.list,
     paletteer::scale_color_paletteer_c("grDevices::Temps") + 
     theme_classic()
   
+  ## Average pop. densities vs. average pop. growth rates
+  dens_lambda.sum <- popDens.sum %>%
+    dplyr::rename(dens.Median = Median,
+                  dens.lCI = lCI,
+                  dens.uCI = uCI) %>%
+    dplyr::select(-Mean, -SD, -CV, -Latitude, -Longitude) %>%
+    dplyr::left_join(lambda.sum, by = c("Area", "SummaryPeriod"))
+  
+  p_lambdaDens1 <- ggplot(subset(dens_lambda.sum, SummaryPeriod == paste0(minYear, "-", maxYear))) + 
+    geom_pointrange(aes(x = dens.Median, y = Median, ymin = lCI, ymax = uCI, colour = Latitude), size = 0.5, fatten = 4, alpha = 0.5) +
+    geom_pointrange(aes(x = dens.Median, y = Median, xmin = dens.lCI, xmax = dens.uCI, colour = Latitude), size = 0, fatten = 4, alpha = 0.5) +
+    xlim(0, 500) + 
+    xlab("Population density") +
+    ylab("Population growth rate") + 
+    paletteer::scale_color_paletteer_c("grDevices::Temps") + 
+    theme_classic()
+  
+  p_lambdaDens2 <- ggplot(subset(dens_lambda.sum, SummaryPeriod != paste0(minYear, "-", maxYear))) + 
+    geom_pointrange(aes(x = dens.Median, y = Median, ymin = lCI, ymax = uCI, colour = Latitude), size = 0.5, fatten = 4, alpha = 0.5) +
+    geom_pointrange(aes(x = dens.Median, y = Median, xmin = dens.lCI, xmax = dens.uCI, colour = Latitude), size = 0, fatten = 0, alpha = 0.5) +
+    xlab("Population density") +
+    ylab("Population growth rate") +  
+    paletteer::scale_color_paletteer_c("grDevices::Temps") + 
+    theme_classic()
   
   ## Make plotting directory if it does not exist yet
   ifelse(!dir.exists("Plots/Latitude"), dir.create("Plots/Latitude"), FALSE)
@@ -161,12 +185,27 @@ plotLatitude <- function(PostSum.list,
   )
   dev.off()
   
[…]

### 2023-12-29 R/plotLatitude.R
Adjusted plotting parameters

@@ -114,20 +114,24 @@ plotLatitude <- function(PostSum.list,
     dplyr::left_join(lambda.sum, by = c("Area", "SummaryPeriod"))
   
   p_lambdaDens1 <- ggplot(subset(dens_lambda.sum, SummaryPeriod == paste0(minYear, "-", maxYear))) + 
-    geom_pointrange(aes(x = dens.Median, y = Median, ymin = lCI, ymax = uCI, colour = Latitude), size = 0.5, fatten = 4, alpha = 0.5) +
-    geom_pointrange(aes(x = dens.Median, y = Median, xmin = dens.lCI, xmax = dens.uCI, colour = Latitude), size = 0, fatten = 4, alpha = 0.5) +
+    geom_pointrange(aes(x = dens.Median, y = Median, ymin = lCI, ymax = uCI, colour = Latitude), size = 0, fatten = 1, linewidth = 0.5, alpha = 0.375) +
+    geom_pointrange(aes(x = dens.Median, y = Median, xmin = dens.lCI, xmax = dens.uCI, colour = Latitude), size = 0, fatten = 1, linewidth = 0.5, alpha = 0.375) +
+    geom_point(aes(x = dens.Median, y = Median, fill = Latitude), shape = 21, color = "black", size = 2, alpha = 0.75) +
+    paletteer::scale_color_paletteer_c("grDevices::Temps") + 
+    paletteer::scale_fill_paletteer_c("grDevices::Temps") + 
     xlim(0, 500) + 
     xlab("Population density") +
     ylab("Population growth rate") + 
-    paletteer::scale_color_paletteer_c("grDevices::Temps") + 
     theme_classic()
   
   p_lambdaDens2 <- ggplot(subset(dens_lambda.sum, SummaryPeriod != paste0(minYear, "-", maxYear))) + 
-    geom_pointrange(aes(x = dens.Median, y = Median, ymin = lCI, ymax = uCI, colour = Latitude), size = 0.5, fatten = 4, alpha = 0.5) +
-    geom_pointrange(aes(x = dens.Median, y = Median, xmin = dens.lCI, xmax = dens.uCI, colour = Latitude), size = 0, fatten = 0, alpha = 0.5) +
+    geom_pointrange(aes(x = dens.Median, y = Median, ymin = lCI, ymax = uCI, colour = Latitude), size = 0, fatten = 1, linewidth = 0.5, alpha = 0.375) +
[…]

### 2023-12-29 R/plotLatitude.R
Added survival vs. recruitment plot

@@ -57,6 +57,31 @@ plotLatitude <- function(PostSum.list,
     ylab("Estimate") + 
     theme_classic()
   
+  ## Average annual survival vs. reproductive rates
+  pSurv_rRep.sum <- pSurv.sum %>%
+    dplyr::rename(surv.Median = Median,
+                  surv.lCI = lCI,
+                  surv.uCI = uCI) %>%
+    dplyr::select(-Mean, -SD, -CV, -Latitude, -Longitude) %>%
+    dplyr::left_join(rRep.sum, by = "Area")
+  
+  p_survRep1 <- ggplot(pSurv_rRep.sum) + 
+    geom_pointrange(aes(x = surv.Median, y = Median, ymin = lCI, ymax = uCI, colour = Latitude), size = 0, fatten = 1, linewidth = 0.5, alpha = 0.375) +
+    geom_pointrange(aes(x = surv.Median, y = Median, xmin = surv.lCI, xmax = surv.uCI, colour = Latitude), size = 0, fatten = 1, linewidth = 0.5, alpha = 0.375) +
+    geom_point(aes(x = surv.Median, y = Median, fill = Latitude), shape = 21, color = "black", size = 2, alpha = 0.75) +
+    paletteer::scale_color_paletteer_c("grDevices::Temps") + 
+    paletteer::scale_fill_paletteer_c("grDevices::Temps") + 
+    xlab("Survival probability") +
+    ylab("Recruitment rate") + 
+    theme_classic()
+  
+  p_survRep2 <- ggplot(pSurv_rRep.sum) + 
+    geom_point(aes(x = surv.Median, y = Median, colour = Latitude), size = 2, alpha = 0.75) +
+    xlab("Survival probability") +
+    ylab("Recruitment rate") + 
+    paletteer::scale_color_paletteer_c("grDevices::Temps") + 
+    theme_classic()
+  
   # Rodent effects 
   if(fitRodentCov){
     betaR.sum <- PostSum.list$betaR.sum %>%
@@ -166,7 +191,15 @@ plotLatitude <- function(PostSum.list,
     dev.off()
   }
 
-  
+  ## Assemble plots (vital rate correlation)
+  pdf("Plots/Latitude/SurvRep_Latitude.pdf", width = 5, height = 4)
+  print(
+    p_survRep1
+  )
+  print(
+    p_survRep2
+  )
+  dev.off()
   
   ## Assemble plots (population-level parameters)
[…]

### 2023-12-29 R/plotLatitude.R
Updated plotting parameters

@@ -32,9 +32,11 @@ plotLatitude <- function(PostSum.list,
     dplyr::left_join(area_coord, by = "Area")
   
   p_detect <- ggplot(detect.sum) + 
-    geom_pointrange(aes(x = Latitude, y = Median, ymin = lCI, ymax = uCI, colour = Longitude), size = 0.5, fatten = 4, alpha = 0.5) +
-    paletteer::scale_color_paletteer_c("grDevices::Temps") + 
+    geom_pointrange(aes(x = Latitude, y = Median, ymin = lCI, ymax = uCI, colour = Longitude), size = 0, fatten = 4, alpha = 0.5) +
+    geom_point(aes(x = Latitude, y = Median, fill = Longitude), shape = 21, color = "black", size = 2, alpha = 0.75) +
     ylab("Estimate") + 
+    paletteer::scale_color_paletteer_c("grDevices::Temps") + 
+    paletteer::scale_fill_paletteer_c("grDevices::Temps") + 
     theme_classic()
   
   # Average reproductive rates
@@ -42,9 +44,11 @@ plotLatitude <- function(PostSum.list,
     dplyr::left_join(area_coord, by = "Area")
   
   p_rRep <- ggplot(rRep.sum) + 
-    geom_pointrange(aes(x = Latitude, y = Median, ymin = lCI, ymax = uCI, colour = Longitude), size = 0.5, fatten = 4, alpha = 0.5) +
-    paletteer::scale_color_paletteer_c("grDevices::Temps") + 
+    geom_pointrange(aes(x = Latitude, y = Median, ymin = lCI, ymax = uCI, colour = Longitude), size = 0, fatten = 4, alpha = 0.5) +
+    geom_point(aes(x = Latitude, y = Median, fill = Longitude), shape = 21, color = "black", size = 2, alpha = 0.75) +
     ylab("Estimate") + 
+    paletteer::scale_color_paletteer_c("grDevices::Temps") + 
+    paletteer::scale_fill_paletteer_c("grDevices::Temps") + 
     theme_classic()
   
   # Average annual survival 
@@ -52,9 +56,11 @@ plotLatitude <- function(PostSum.list,
     dplyr::left_join(area_coord, by = "Area")
   
   p_pSurv <- ggplot(pSurv.sum) + 
-    geom_pointrange(aes(x = Latitude, y = Median, ymin = lCI, ymax = uCI, colour = Longitude), size = 0.5, fatten = 4, alpha = 0.5) +
[…]

### 2023-12-29 R/plotLatitude.R
Added plot array for manuscript

@@ -101,6 +101,15 @@ plotLatitude <- function(PostSum.list,
       paletteer::scale_color_paletteer_c("grDevices::Temps") + 
       paletteer::scale_fill_paletteer_c("grDevices::Temps") + 
       theme_classic()
+    
+    p_betaR_long <- ggplot(betaR.sum) + 
+      geom_hline(aes(yintercept = 0), color = "grey70", linetype = "dotted") + 
+      geom_pointrange(aes(x = Longitude, y = Median, ymin = lCI, ymax = uCI, colour = Latitude), size = 0, fatten = 4, alpha = 0.5) +
+      geom_point(aes(x = Longitude, y = Median, fill = Latitude), shape = 21, color = "black", size = 2, alpha = 0.75) +
+      ylab("Rodent effect on recruitment") + 
+      paletteer::scale_color_paletteer_c("grDevices::Temps") + 
+      paletteer::scale_fill_paletteer_c("grDevices::Temps") + 
+      theme_classic()
   }
   
   # Average population densities
@@ -251,6 +260,20 @@ plotLatitude <- function(PostSum.list,
   )
   dev.off()
   
+  ## Assemble plots (manuscript array)
+  if(fitRodentCov){
+    pdf("Plots/Latitude/AllParams_Latitude.pdf", width = 8, height = 6)
+    print(
+      ggpubr::ggarrange(ggpubr::ggarrange(p_lambdaDens2 + theme(legend.position = "none") + ggtitle("A)"),
+                                          p_survRep1 + theme(legend.position = "none") + ggtitle("B)"),
+                                          nrow = 1), 
+                        p_betaR_long + ggtitle("C)"),  
+                        heights = c(1.5, 1), ncol = 1, 
+                        legend = "right", common.legend = TRUE)
+    )
+    dev.off()
+  }
+
   
   ## Write and return plot paths
   plot.paths <- c("Plots/Latitude/DetectParams_Latitude.pdf",
@@ -260,6 +283,10 @@ plotLatitude <- function(PostSum.list,
                   "Plots/Latitude/PopParams2_Latitude.pdf",
                   "Plots/Latitude/LambdaDens1_Latitude.pdf",
                   "Plots/Latitude/LambdaDens2_Latitude.pdf")
+  
+  if(fitRodentCov){
+    plot.paths <- c(plot.paths, "Plots/Latitude/AllParams_Latitude.pdf")
+  }
   
   return(plot.paths)
   

### 2023-12-29 Manuscript_Nater_etal.qmd
Added text to results for survival and recruitment

@@
 # Survival stats
-#S.hyper.mean
-#S.sd.spatial
-#S.sd.temporal
-#S.sd.residual
+S.hyper.mean <- printPostSum(table = post.sum$hParams.sum, 
+                             row = which(post.sum$hParams.sum$Parameter == "h.Mu.S"))
+S.sd.spatial <- printPostSum(table = post.sum$hParams.sum, 
+                             row = which(post.sum$hParams.sum$Parameter == "h.sigma.S"))
+S.sd.temporal <- printPostSum(table = post.sum$hParams.sum, 
+                             row = which(post.sum$hParams.sum$Parameter == "sigmaT.S"))
+S.sd.residual <- printPostSum(table = post.sum$hParams.sum, 
+                             row = which(post.sum$hParams.sum$Parameter == "sigmaR.S"))
 S.min <- printPostSum(table = post.sum$pSurv.sum, 
@@
 # Recruitment stats
-#R.hyper.mean
-#R.sd.spatial
-#R.sd.temporal
-#R.sd.residual
+R.hyper.mean <- printPostSum(table = post.sum$hParams.sum, 
+                             row = which(post.sum$hParams.sum$Parameter == "h.Mu.R"))
+R.sd.spatial <- printPostSum(table = post.sum$hParams.sum, 
+                             row = which(post.sum$hParams.sum$Parameter == "h.sigma.R"))
+R.sd.temporal <- printPostSum(table = post.sum$hParams.sum, 
+                             row = which(post.sum$hParams.sum$Parameter == "sigmaT.R"))
+R.sd.residual <- printPostSum(table = post.sum$hParams.sum, 
+                             row = which(post.sum$hParams.sum$Parameter == "sigmaR.R"))
 R.min <- printPostSum(table = post.sum$rRep.sum, 
@@
 Only during the most recent four years (2018-2021) has data been collected regularly for all reporting areas included in the analyses.
-During this period, estimated population densities varied between `r dens.min$post.sum` birds/km^2 in the area "`r dens.min$area.name`" close to the coast in central Norway to `r dens.max$post.sum` birds/km^2 in "`r dens.max$area.name`" further south near the Swedish border.
+During this period, estimated population densities varied between `r dens.min$post.sum` birds/km\^2 in the area "`r dens.min$area.name`" close to the coast in central Norway to `r dens.max$post.sum` birds/km\^2 in "`r dens.max$area.name`" further south near the Swedish border.
 In general, population density appeared to be lowest in the Northern parts of Norway and highest in the eastern part of central Norway (@fig-densA).
@@
 
-In most areas, populations of willow ptarmigan have been increasing over the period 2018-2020, potentially indicating a positive effect of more restricted hunting in recent years (@fig-lam).
-Some areas -- predominantly in central Norway -- also had declining populations, but many of those declines followed upon a period of increase between the start of data colection in 2007 and sometime between 2016 and 2018.
+Average population growth rates over the last four years (2018-2021) ranged from moderate declines (`r lambda.min$post.sum` in the "`r lambda.min$area.name`" area) to \> 50% increase (`r lambda.max$post.sum` in the "`r lambda.max$area.name`" area).
+In the majority of reporting areas (`r n.pos.lambda` out of 41), populations of willow ptarmigan have been increasing over the period 2018-2021 (@fig-lam).
+Some areas -- predominantly in central Norway -- also had declining populations, but many of those declines followed upon periods of increase between the start of data collection in 2007 and sometime between 2016 and 2018 (SF "TimeSeries_popDens1.pdf").
 
 ::: {#fig-lam layout-ncol="2"}
-![Median population growth rate estimates](Figures/lambda_2018-2020_Median.png){#fig-lamA}
+![Median population growth rate estimates](Figures/Avg_lambda_Map-4.png){#fig-lamA}
 
-![Uncertainty in population growth rate estimates](Figures/lambda_2018-2020_SD.png){#fig-lamB}
+![Uncertainty in population growth rate estimates](Figures/Avg_lambda_Map-5.png){#fig-lamB}
 
-Median (a) and standard deviation (b) of posterior estimates of average annual population growth rate over the three most recent years (2018-2020) across 41 reporting areas in Norway.
+Median (a) and standard deviation (b) of posterior estimates of average annual population growth rate over the four most recent years (2018-2021) across 41 reporting areas (summarised at the municipality level) in Norway.
 In a), pinkish colors indicate declining populations while greenish colors indicate growing populations (white = stable populations).
-In b) darker colors indicate higher median values and higher uncertainty.
+In b) darker colors indicate higher uncertainty.
 :::
 
-## Survival probability
+The highest population growth rates were estimated for areas with relatively low population densities across latitudes [@fig-paramCorrs A] but we did not find evidence for a strong association between population growth rates and population densities overall.
 
-Annual survival probabilities ranged from aroung 35% to 50% across reporting areas in Norway, with the highest values occurring in the far north and in the mountains in the south (@fig-pSurv).
+![Posterior summaries (points = posterior medians, lines = 95% credible intervals) of area-specific population growth rate vs.
+population density (A), recruitment rate vs.
+survival probability (B) and rodent effect on recruitment along a longitudinal gradient (C).
+Color indicates latitude of the midpoint of each area.](Figures/AllParams_Latitude.png){#fig-paramCorrs}
 
-::: {#fig-pSurv layout-ncol="2"}
-![Median survival estimates](Figures/pSurv_Median.png){#fig-pSurvA}
+## Survival probabilities and recruitment rates
 
-![Uncertainty in survival estimates](Figures/pSurv_SD.png){#fig-pSurvB}
+Annual survival probabilities ranged from `r S.min$post.sum` (area "`r S.min$area.name`") to `r S.max$post.sum` (area "`r S.max$area.name`") across reporting areas in Norway, with the highest values occurring in the far north and in the mountains in the south (@fig-pVRA).
+The global average survival probability across all areas and years ($\mu^S$) was estimated at `r S.hyper.mean`.
+Spatial variation in survival (random effect SD = `r S.sd.spatial$post.sum`) was relatively low compared to temporal (`r S.sd.temporal$post.sum`) and residual (`r S.sd.residual$post.sum`) variation.
 
-Median (a) and standard deviation (b) of posterior estimates of average annual survival probability across 41 reporting areas in Norway.
-Darker colors indicate higher median values and higher uncertainty.
-:::
+Recruitment rates varied between `r R.min$post.sum` (area "`r R.min$area.name`") and `r R.max$post.sum` (area "`r R.max$area.name`") and displayed a spatial pattern opposite to that 
+of annual survival, i.e.
+lower recruitment rates co-occurring with higher survival rates and vice-versa [@fig-pVRB; @fig-paramCorrs B).
+Across all areas and year, average recruitment rate was `r R.hyper.mean`.
+Unlike for survival, the model predicted similar magnitudes of spatial and temporal variation (random effect SDs of `r R.sd.spatial` and `r R.sd.temporal`, respectively) and about twice as much residual variation (`r R.sd.residual`).
 
-## Recruitment rate
+::: {#fig-pVR layout-ncol="2"}
+![Median survival probabilities](Figures/Avg_pSurv_Map-1.png){#fig-pVRA}
 
-Annual recruitment rates showed a spatial pattern (@fig-rRep) that was inverse relative to that of annual survival (@fig-pSurv), with lower recruitment rates co-occurring with higher survival rates and vice-versa.
-This is consistent with basic life history theory [@stearns1992] and with previous studies finding relatively slower life histories in alpine / high altitude areas [e.g.
-@sandercock2005; @bears2009; @alice2016].
+![Median recruitment rates](Figures/Avg_rRep_Map-1.png){#fig-pVRB}
 
-::: {#fig-rRep layout-ncol="2"}
-![Median recruitment estimates](Figures/rRep_Median.png){#fig-rRepA}
-
-![Uncertainty in recruitment estimates](Figures/rRep_SD.png){#fig-rRepB}
-
-Median (a) and standard deviation (b) of posterior estimates of average annual recruitment rate across 41 reporting areas in Norway. Darker colors indicate higher median values and higher uncertainty.
+Posterior medians of average annual survival probabilities (a) and recruitment rates (b) across 41 reporting areas (summarised at the municipality level) in Norway.
+Darker colors indicate higher median values.
+Measures for corresponding uncertainty in estimates are visualized in SFs "Avg_pSurv_Map.pdf" and "Avg_rRep_Map.pdf" for survival and recruitment, respectively.
 :::
@@
 -->
+
+Annual recruitment rates showed a spatial pattern that was inverse relative to that of annual survival, with lower recruitment rates co-occurring with higher survival rates and vice-versa.
+This is consistent with basic life history theory [@stearns1992] and with previous studies finding relatively slower life histories in alpine / high altitude areas [e.g.
+@sandercock2005; @bears2009; @alice2016].
 

### 2023-12-29 Manuscript_Nater_etal.qmd
Completed text for vital rate section

@@
 lower recruitment rates co-occurring with higher survival rates and vice-versa [@fig-pVRB; @fig-paramCorrs B).
-Across all areas and year, average recruitment rate was `r R.hyper.mean`.
-Unlike for survival, the model predicted similar magnitudes of spatial and temporal variation (random effect SDs of `r R.sd.spatial` and `r R.sd.temporal`, respectively) and about twice as much residual variation (`r R.sd.residual`).
+Across all areas and years, average recruitment rate was `r R.hyper.mean`.
+Unlike for survival, the model predicted similar magnitudes of spatial and temporal variation (random effect SDs of `r R.sd.spatial` and `r R.sd.temporal`, respectively), and about twice as much residual variation (`r R.sd.residual`).
 
@@
 :::
+
+The MCMC chains for many of the area-specific average survival probabilities and recruitment rates, as well as for the global averages (intercepts) for both vital rates, were mixing rather poorly.
+Despite that, mixing was good and resulting posteriors well defined for the area- and year-specific estimates of survival and recruitment (SF "PostDens_tS_tR.pdf").
+There was substantial variation in both vital rates across time (SFs "TimeSeries_pSurv.pdf" and "TimeSeries_rRep.pdf").
+In many areas, the years 2011, 2014, and 2018 not only sported high population densities (see above) but were also characterized by both high recruitment and low subsequent survival.
+The overall low density years 2012 and 2015, conversely, often featured lower recruitment and, in some cases, higher survival.
+Notably, there were also years with very little spatial synchrony, i.e.
+very different relative yearly survival probabilities and recruitment rates (e.g.
+2010 and 2020 for survival and 2013, 2016, and 2017 for recruitment).
 

### 2023-12-29 Manuscript_Nater_etal.qmd
Small adjustments for correct render

@@
 The latter means that information about spatio-temporal variation in demographic rates and population dynamics is particularly important in order to design sustainable harvest strategies.
-In addition, as a sentinel species the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is included in both the main national biodiversity [Nature Index for Norway; @jakobsson2020; <https://www.naturindeks.no/Indicators/lirype>] and ecosystem [Assessment of the Ecological Condition; @framstad2022] assessments.
+In addition, as a sentinel species the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is included in both the main national biodiversity [Nature Index for Norway, @jakobsson2020, <https://www.naturindeks.no/Indicators/lirype>] and ecosystem [Assessment of the Ecological Condition, @framstad2022] assessments.
 
@@
 
+  if(!("Area" %in% colnames(table))){
+    table$Area <- "None"
+  }
+
   print.out <- list(
@@
 # Rodent effects stats
-#betaR.hyper.mean
-#betaR.sd.spatial
-#betaR.sd.temporal
-#betaR.sd.residual
+betaR.hyper.mean <- printPostSum(table = post.sum$hParams.sum,
+                             row = which(post.sum$hParams.sum$Parameter == "h.Mu.betaR.R"))
+betaR.sd.spatial <- printPostSum(table = post.sum$hParams.sum,
+                             row = which(post.sum$hParams.sum$Parameter == "h.sigma.betaR.R"))
 betaR.min <- printPostSum(table = post.sum$betaR.sum, 
@@
 lower recruitment rates co-occurring with higher survival rates and vice-versa [@fig-pVRB; @fig-paramCorrs B).
-Across all areas and years, average recruitment rate was `r R.hyper.mean`.
-Unlike for survival, the model predicted similar magnitudes of spatial and temporal variation (random effect SDs of `r R.sd.spatial` and `r R.sd.temporal`, respectively), and about twice as much residual variation (`r R.sd.residual`).
+Across all areas and years, average recruitment rate was `r R.hyper.mean$post.sum`.
+Unlike for survival, the model predicted similar magnitudes of spatial and temporal variation (random effect SDs of `r R.sd.spatial$post.sum` and `r R.sd.temporal$post.sum`, respectively), and about twice as much residual variation (`r R.sd.residual$post.sum`).
 
@@
 
-The model predicted substantial variation in the effect of the rodent occupancy covariate on recruitment (@fig-betaR).
-Areas in the north and south west generally had recruitment increasing with rodent occupancy while the remaining areas were a mosaic of positive and negative effects, albeit with some spatial clustering.
-
-::: {#fig-betaR layout-ncol="2"}
-![Median rodent effect estimates](Figures/betaR_Median.png){#fig-betaRA}
-
-![Uncertainty in rodent effect estimates](Figures/betaR_SD.png){#fig-betaRB}
-
-Median (a) and standard deviation (b) of posterior estimates of effects of rodent occupancy on recruitment rate across 41 reporting areas in Norway.
-In a), blueish colors indicate negative effects while reddish colors indicate positive effects (yellow = no effect).
-In b) darker colors indicate higher median values and higher uncertainty.
-:::
+The model predicted a positive global effect of rodent occupancy on recruitment rate (average slope on the log scale = `r betaR.hyper.mean$post.sum`).
+Nonetheless, spatial variation in the rodent effect was substantial (random effect SD = `r betaR.sd.spatial$post.sum`).
+This resulted in negative median effects in `r n.neg.betaR` areas, positive median effects in `r n.pos.betaR` areas, and a range of effect sizes from `r betaR.min$post.sum` (area "`r betaR.min$area.name`") to `r betaR.max$post.sum` (area "`r betaR.max$area.name`", @fig-paramCorrs C).
+The largest positive rodent effects were estimated for areas in the very North of Norway, as well as in the mountaineous regions in the central and southwestern parts of the country (SF "betaR_Map.pdf").
+Effects with negative posterior medians were located mostly at intermediate latitudes, but we note that all of these had posterior distributions featuring substantial overlap with 0 (@fig-paramCorrs C).
 

### 2023-12-30 Manuscript_Nater_etal.qmd
Added text on variance decomposition results

@@
 
+```{r, echo = FALSE}
+# Posterior summaries
+varDec.sum <- readRDS("Results/PosteriorSummaries_VarDecomp.rds")
+
+# Printing function
+printPostSum2 <- function(table, Parameter, VarComponent){
+
+  idx <- which(table$Parameter == Parameter & table$VarComponent == VarComponent)
+
+  print.out <- paste0(
+      round(table$median[idx]*100, digits = 1), 
+      " [",
+      round(table$lCI[idx]*100, digits = 1),
+      ", ",
+      round(table$uCI[idx]*100, digits = 1),
+      "]"
+  )
+  return(print.out)
+}
+
+# All parameter summaries
+Rec_A <- printPostSum2(varDec.sum, "Recruitment", "Area")
+Rec_Y <- printPostSum2(varDec.sum, "Recruitment", "Year")
+Rec_R <- printPostSum2(varDec.sum, "Recruitment", "Residual")
+Rec_Cov <- printPostSum2(varDec.sum, "Recruitment", "RodentOcc")
+
+Surv_A <- printPostSum2(varDec.sum, "Survival", "Area")
+Surv_Y <- printPostSum2(varDec.sum, "Survival", "Year")
+Surv_R <- printPostSum2(varDec.sum, "Survival", "Residual")
+
+Det_A <- printPostSum2(varDec.sum, "Detection", "Area")
+Det_Y <- printPostSum2(varDec.sum, "Detection", "Year")
+Det_R <- printPostSum2(varDec.sum, "Detection", "Residual")
+```
+
+The relative importance of different components for explaining parameter variation differed among recruitment rate, survival probability, and detection decay (@fig-varDecomp).
+The largest portion of variation in recruitment was attributed to residual variation (`r Rec_R` %), followed by spatial (`r Rec_A` %) and temporal (`r Rec_Y` %) variation.
+Rodent occupancy, which contains both a spatial and a temporal dimension, explained `r Rec_Cov` % of the total variation.
+For survival, there was large uncertainty in the estimated proportions of variance explained by different components.
+The model predicted similar potential contributions from temporal (`r Surv_Y` %) and residual (`r Surv_R` %) variation and suggested that spatial variation was only responsible for `r Surv_A` % of the total variance.
+Total variance in detection decay was attributed evenly to spatial, temporal, and residual variation at `r Det_A` %, `r Det_Y` %, and `r Det_R` %, respectively.
+
+![Posterior distributions for the proportions of parameter variance explained by spatial (blue), temporal (orange), and residual (green) variation, as well as by effects of rodent occupancy (yellow).](Figures/VarDecomposition_overall.png){#fig-varDecomp}
+
 # Discussion

### 2023-12-31 Manuscript_Nater_etal.qmd
Added results for detection parameters

@@
 # Printing function
-printPostSum <- function(table, row = NULL, area = NULL){
+printPostSum <- function(table, row = NULL, area = NULL, exp = FALSE, pCalc = FALSE){
 
@@
     table$Area <- "None"
+  }
+
+  if(exp){
+    table$Median <- exp(table$Median)
+    table$lCI <- exp(table$lCI)
+    table$uCI <- exp(table$uCI)
+  }
+
+  if(pCalc){
+     pi <- 3.141593
+     W <- 200
+     table$Median <- sqrt((pi*table$Median^2)/2)/W
+     table$lCI <- sqrt((pi*table$lCI^2)/2)/W
+     table$uCI <- sqrt((pi*table$uCI^2)/2)/W
   }
@@
 # Detection stats
-#Det.hyper.mean
-#Det.sd.spatial
-#Det.sd.temporal
-#Det.sd.residual
-#Det.min
-#Det.max
+Det.hyper.mean <- printPostSum(table = post.sum$hParams.sum, 
+                              row = which(post.sum$hParams.sum$Parameter == "h.mu.dd"),
+                              exp = TRUE)
+pDet.hyper.mean <- printPostSum(table = post.sum$hParams.sum, 
+                              row = which(post.sum$hParams.sum$Parameter == "h.mu.dd"),
+                              exp = TRUE, pCalc = TRUE)
+
+Det.sd.spatial <- printPostSum(table = post.sum$hParams.sum,
+                             row = which(post.sum$hParams.sum$Parameter == "h.sigma.dd"))
+Det.sd.temporal <- printPostSum(table = post.sum$hParams.sum,
+                             row = which(post.sum$hParams.sum$Parameter == "sigmaT.dd"))
+Det.sd.residual <- printPostSum(table = post.sum$hParams.sum,
+                             row = which(post.sum$hParams.sum$Parameter == "sigmaR.dd"))
+Det.min <- printPostSum(table = post.sum$detect.sum, 
+                      row = which(post.sum$detect.sum$Median == min(post.sum$detect.sum$Median)))
+
+pDet.min <- printPostSum(table = post.sum$detect.sum, 
+                      row = which(post.sum$detect.sum$Median == min(post.sum$detect.sum$Median)),
+                      pCalc = TRUE)
+
+Det.max <- printPostSum(table = post.sum$detect.sum,
+                      row = which(post.sum$detect.sum$Median == max(post.sum$detect.sum$Median)))
+
+pDet.max <- printPostSum(table = post.sum$detect.sum,
+                      row = which(post.sum$detect.sum$Median == max(post.sum$detect.sum$Median)),
+                      pCalc = TRUE)
 
@@
 of annual survival, i.e.
-lower recruitment rates co-occurring with higher survival rates and vice-versa [@fig-pVRB; @fig-paramCorrs B).
+lower recruitment rates co-occurring with higher survival rates and vice-versa (@fig-pVRB; @fig-paramCorrs B).
 Across all areas and years, average recruitment rate was `r R.hyper.mean$post.sum`.
@@
 
-TBA
+Detection decay parameters, which determine detection probability in distance sampling surveys, varied across areas from between `r Det.min$post.sum` in "`r Det.min$area.name`" to `r Det.max$post.sum` in "`r Det.max$area.name`, resulting in detection probabilities over the transect sites (truncation distance = 200 m) from `r pDet.min$post.sum` to `r pDet.max$post.sum`, respectively.
+The global average detection decay was `r Det.hyper.mean$post.sum` (detection probability = `r pDet.hyper.mean$post.sum`), and in general, higher values were more common in the Southern half of the country than the Northern half (SF "Avg_detect_Map.pdf").
+Variation in detection over time was modest (SF "TimeSeries_pDetect.pdf"), and among-year variation in detection decay (random effect log SD = `r Det.sd.temporal$post.sum`) was almost identical in magnitude to spatial (`r Det.sd.spatial$post.sum`) and residual (`r Det.sd.residual$post.sum`) variation.
 
@@
 # Discussion
-
-TBA
 

### 2024-01-04 Manuscript_Nater_etal.qmd
minor edit

@@
 
-There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly and requires trained specialists, thus limiting the amount of data that can be collected, while participatory monitoring, i.e.
+There is often a trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly and requires trained specialists, thus limiting the amount of data that can be collected, while participatory monitoring, i.e.
 the collection of ecological data by members of the public [also called citizen science, @fraisl2022], allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias [@johnston2023].
@@
 
-Recruitment rates varied between `r R.min$post.sum` (area "`r R.min$area.name`") and `r R.max$post.sum` (area "`r R.max$area.name`") and displayed a spatial pattern opposite to that 
-of annual survival, i.e.
+Recruitment rates varied between `r R.min$post.sum` (area "`r R.min$area.name`") and `r R.max$post.sum` (area "`r R.max$area.name`") and displayed a spatial pattern opposite to that of annual survival, i.e.
 lower recruitment rates co-occurring with higher survival rates and vice-versa (@fig-pVRB; @fig-paramCorrs B).

### 2024-01-05 Manuscript_Nilsen_Nater.qmd
Added discussion section

@@
 Posterior distributions for parameters estimated in three model fits to each of 10 simulated data sets are shown in @fig-sim-detection and @fig-sim-demo.
-Overall, the IDSM was able to corretly estimate both detection parameters and demographic rate parameters from all 10 simulated datasets without any systematic bias.
+Overall, the IDSM was able to correctly estimate both detection parameters and demographic rate parameters from all 10 simulated datasets without any systematic bias.
 The replicate runs for each dataset resulted in very similar posterior distributions, demonstrating that the models converged towards the same posterior distributions irrespective of starting values.
@@
 
--   In general, the model does a very good job in recreating the underlying parameters when fitted to simulated data where the true values are known.
+We developed a model that use line transect distance sampling survey data combined with data from marked individuals to estimate population abundance, survival probabilities and recruitment rates in an open population setting.
+We first used simulated data to examine models ability to recover the underlying parameters when they were known.
+Then we fitted the model to data from an ongoing field study on willow ptarmigan in Norway.
+In general, the model did a very good job in recreating the underlying parameters when fitted to simulated data where the true values are known.
 For the simulated data, a relative wide range of parameter values were allowed, and we did not detect any particular problems or biases with respect to this variation in the underlying parameters.
--   Both simulated data tests and real data application have shown that mixing is suboptimal for the average recruitment rate parameter (Mu.R).
-Future work should perhaps look into optimizing sampling?
-(For biological interpretation that is not a big problem as the yearly recruitment rate estimates are fine...
-but have to keep in mind when interpreting) <!-- I leave it to you Erlend to decide where, whether, and how much you want to go into this one -->
--   The fact that this formulation allow us to separately estimate the underlying demographic rates (survival and recruitment) is a clear improvement compared to previous open population DS models, as this allow us both to obtain estimates of the rates, but also to assess the effect of environmental variation directly on demographic rates and not just on population growth rate ($\lambda$) .
--   Nevertheless, we did not detect a clear effect of rodent abundance (indexed by the occurrence of rodents on the transect lines) on recruitment rate.
+When the input data are unbiased (with respect to the underlying model formulation) we therefore consider the model to be able to extract unbiased and meaningful demographic parameters from distance sampling data.
+
+Open population formulations of the distance sampling model has previously been presented (refs), and applied to various ecological systems (refs).
+Our model extends those previous applications by formulating the underlying process model as a stage-structured matrix model where the matrix elements are represented by annual survival probabilities and recruitment rates.
+While such models have been available for a range of other modelling frameworks previously (in particular), this is to our knowledge the first explicit use of such a state model in a distance sampling framework.
+Such a formulation allow us to both to obtain estimates of the underlying demographic rates, but also to assess the effect of environmental variation directly on demographic rates and not just on population growth rate ($\lambda$).
+As an example of such environmental variation, we investigated the effects of small rodent abundance on the recruitment rates.
+The fact that we were able to successfully recover the covariate effect when the model was applied to simulated data, suggest that in fact the model is able to estimate the effect when there is a clear signal in the data.
+When applied to real data, we did not detect a clear effect of rodent abundance (indexed by the occurrence of rodents on the transect lines) on recruitment rate.
 This might be somewhat surprising, given that such a pattern has been reported repeatedly in the literature (see e.g.
 @Bowler2020).
-However, the data on rodent abundance was heavily zero-inflated, and the annual variation in the index was rather small.
-\[A few sentences about fading rodent cycles, and perhaps a weakened link to ptarmigan dynamics\].
-\[Also: a time-series of 15 years is still relatively short, so it is to be expected that we have low statistical power to detect temporal covariate effects --\> more years of data, or space-for-time substitution, will help)
--   In our process model, the population growth rate ($\lambda$) is determined by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, creating a dependence between the demographic parameters.
+We speculate that there are (at least) three potential explenations to this result.
+First, for our specific case study the data on rodent abundance was heavily zero-inflated, and the annual variation in the index was rather small.
+It is well known that the amplitude and regularity of the rodent cycles has been fading in recent decades (refs), and our study area might be no exception.
+Lack of peak rodent years in the time series to which we fitted the model.
+In addition, previous research has shown that ptarmigan recruitment is also sensitive to the weather in the late winter and spring (before and during the breeding season); we did not fit any weather covariates to the model so there is a possibility that strong effects of spring conditions in certain years have masked any remaining effects of small rodent abundance.
+Finally, the data set we used is relative short (15 years), leaving us with somewhat limited statistical power.
+We therefore do not consider our a particularily strong test of the underlying effect of small rodent fluctuations on ptarmigan recruitment rates.
+In future applications, a stronger test based on more years of data and data from additional areas is encouraged.
+In e.g.
+[@Bowler2020] we used data from the same sampling program but from more areas, and detected a very clear signal from small rodent abundance on ptarmigan population growth rate.
+Our new model allow us to investigate to which extent similar results emerge when we can link the covariate effect to the actual demographic rate and not only to the population growth rate.
+
+One potential source of bias in our model framework is related to the way that we use the distance sampling data to estimate survival and recruitment rates.
+In our process model, the population growth rate ($\lambda$) is determined by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, and thsi create a dependence between the demographic parameters.
 If the age ratio in the data are biased or contain frequent misclassifications, this is likely to affect the relative contribution of survival and recruitment to the growth rate.
@@
 Thus biases in the age ratio will affect the demographic rates, but not so much the estimated density.
-Compared to previous studies on ptarmigan, we could have expected a higher survival rate indicating that there is some misclassification in the data with respect to age specification.
--   The model is flexible, and can be extended with additional auxiliary data, such as separate data on recruitment rates that could improve the precision of the demographic rate estimates.
-Another one might be data that allows to distinguish between true and apparent survival<!-- or is that accounted for in the telemetry data, i.e.
-are the individuals that left hte study area censored?
--->.
-It is also possible to extent the model with additional covariates, to gain a better understanding of how species respond to environmental variation through the effects on vital rates.
-\[See my point above on limited statistical power with 15 years.
-Maybe could be good to link to the space-for-time substitution idea here and "tease" the extension to multi-area since the monitoring programme, at least for line transects, covers many more areas).
+Compared to previous studies on ptarmigan (see e.g.
+[@Israelsen2020][@Sandercock2011]), we could have expected a higher survival rate indicating that there is some misclassification in the data with respect to age specification.
+In future developments of the model, explicit accounting for such misclassification will be considered.
+
+-   Both simulated data tests and real data application have shown that mixing is suboptimal for the average recruitment rate parameter (Mu.R).
+Future work should perhaps look into optimizing sampling?
+(For biological interpretation that is not a big problem as the yearly recruitment rate estimates are fine...
+but have to keep in mind when interpreting) <!-- I leave it to you Erlend to decide where, whether, and how much you want to go into this one -->
+
+In summary, we conclude that the model presented here is useful in many ecological systems where the sampling survey data include information about the age-ratio of the observed individuals.
+Following the integrated modelling philosophy, it also allow for the integration of auxiliary data.
+In our application here, we integrated data from marked birds by formulating an explicit (known-fate) mark-recapture model that explicitly supported the estimation of survival probabilities in the state process.
+The model is flexible, and can be extended with additional auxiliary data, such as separate data on recruitment rates that could improve the precision of the demographic rate estimates.
+Moreover, the model can be extended with additional covariates, to gain a better understanding of how species respond to environmental variation through the effects on vital rates.
+Finally, as the underlying model structure is based on a hierarchical model the framework readily support multi-area extensions, where the covariate-effects are efficiently estimated by time-for-space substitution.
 

### 2024-01-05 Manuscript_Nilsen_Nater.qmd
Edits to abstract
I've tried to highlight general usefulness a bit more, and be a bit more precise about what we found (not just what we did).
I hope you like it.

@@
 
-*Estimation of abundance and demographic rates for populations of wild vertebrate species is a challenging but fundamental issue for both management and research into ecology and evolution.
-One approach that has been used extensively to estimate abundance of wild living populations is Distance Sampling (DS) methods based on line transect survey data.
-Historically, DS models were only available as open population models, and did not allow for the direct estimation of changes in abundance through time.
-The advent of open population formulations based on the DS framework greatly extended the scope of the models, but so far models that estimate both temporal dynamics in abundance as well as the underlying demographic rates has not been implemented.
-Here, we present an integrated distance sampling approach, that utilize age-structured survey data and auxiliary data from marked individuals to jointly estimate population dynamics and the temporal variation in the demographic rates (recruitment rate and survival probability) that determine the temporal transition.
-The underlying process model is based on a two-stage transition matrix (*$A_t$*), with matrix entries given by survival probabilities (*$S$*) and time-specific recruitment rates (*$R_t$*).
-This framework allow us to make full use of the available data, and to effectively integrate the two data sources in an integrated modelling framework.
-Moreover, demographic rates often respond to environmental variation, and our approach allow us to directly estimate the effect of such environmental covariates on demographic rate variation.
-We first fit the model to simulated data, to assess it's ability to recover the underlying population dynamics when the true values are known.
-Then, we use data from a study of willow ptarmigan (Lagopus lagopus) in Norway as a case study.*
+*Estimation of abundance and demographic rates for populations of wild species is a challenging but fundamental issue for both management and research into ecology and evolution.
+One set of approaches that has been used extensively to estimate abundance of wildlife populations is Distance Sampling (DS) methods for line or point transect survey data.
+Historically, DS models were only available as closed* <!-- I think you mean "closed" here?
+--> *population models, and did not allow for the estimation of changes in abundance through time.
+The advent of open population formulations based on the DS framework greatly extended the scope of the models, but much untapped potential remains in models that estimate temporal dynamics not only in abundance but also in the underlying demographic rates.
+Here, we present an integrated distance sampling approach that utilize age-structured survey data and auxiliary data from marked individuals to jointly estimate population dynamics and the demographic rates (recruitment rate and survival probability) that drive temporal changes .
+The core of the resulting model is, in essence, equivalent to both an integrated population model and a matrix population model with two age classes: juveniles and adults.
+The integrated framework allows making full use of the available data by effectively combining line transect and telemetry data, and can easily be adapted to include additional and/or different data types.
+Moreover, as demographic rates often respond to environmental variation, our approach also supports direct estimation of the effects of such environmental covariates on demographic rates.
+Through a comprehensive simulation study we show that the model is able to adequately recover true population and vital rate dynamics.
+Subsequent application to data from a study of willow ptarmigan (Lagopus lagopus) in Norway showcases the frameworks ability to recover both fluctuations and trends in population dynamics and highlights its potential applicability to a wide range of species surveyed using distance sampling approaches.*
 

### 2024-01-05 Manuscript_Nilsen_Nater.qmd
Edits to introduction and methods

@@
 
+<!-- Just a note: we seems to also still be missing quite a few references in the introduction -->
+
 Estimating abundance and demographic rates for wildlife populations is an integral part of basic and applied ecology [@Skalski2005, @williams2002].
@@
 
-Until recently, joint estimation of population dynamics and demography has relied mostly on data from marked individuals through application of open-population capture-mark-recapture models [@schaub2021].
+Until recently, joint estimation of population dynamics and demography has relied mostly on data from marked individuals and associated open-population capture-mark-recapture models [@schaub2021].
 While such methods can provide valuable information for both ecological research and management, collecting the necessary data is typically costly and logistically challenging to implement over large areas.
-Monitoring focusing on abundance trends over larger areas, on the other hand, are typically based on data from unmarked animals (ref).
+Monitoring programmes focusing on abundance trends over larger areas, on the other hand, are typically based on data from unmarked animals (ref).
 A common approach for such surveys is to structure data collection around the distance sampling (DS) method.
@@
 
-The most common implementations of DS models have long used close-population formulations, and thus do not allow formulating a process model that projects abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates @buckland2015.
+The most common implementations of DS models have long used closed-population formulations, and thus do not allow formulating a process model that projects abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates @buckland2015.
 In recent years, DS approaches have been extended in many ways, including applications that estimate changes in abundance over time in open populations via a hidden state model representing population dynamics [@moore2011; @sollmann2015].
@@
 #### 2.1.1 Age-structured population model
+
+<!-- Note: missing ref below. Maybe good to "search" the whole text for "ref" to make sure we have replaced the placeholder with appropriate references everywhere -->
 
@@
 
-Vital rates (survival probabilities $S$, recruitment rates $R$), detection parameters (half-normal detection parameters $\sigma$), and initial population densities can all be modelled as time-dependent in our framework.
+Vital rates (survival probabilities $S$, recruitment rates $R$) and detection parameters (half-normal detection parameters $\sigma$) can all be modelled as time-dependent in our framework.
 For both the tests with simulated data and the case study described below, we implemented log-normally distributed random year effects on all parameters except survival, which was set to be constant.
@@
 We generated a total of 10 simulated datasets in five steps.
-First, we simulated 15-year time-series of survival and recruitment rates from biologically plausible values for averages of and -- in the case of recruitment -- among-year variation in demographic rates (survival was held constant across years).
+First, we simulated 15-year time-series of survival and recruitment rates from biologically plausible values for averages and -- in the case of recruitment -- among-year variation in demographic rates (survival was held constant across years).
 Second, we used the yearly demographic rate and realistic initial population densities to simulate stochastic population dynamics in 50 distinct sites.
@@
 Fourth, we assigned a distance from transect line to each group and simulated the line transect survey in all 50 sites across 15 years.
-Finally, we simulated 5-year time-series of radio-telemetry data (= survival from one year to the next) for a subset of individuals (30 on average) using the simulated survival probabilities for each relevant year.
+Finally, we simulated 5-year time-series of radio-telemetry data (= survival from one year to the next) for a subset of individuals (30 per year on average) using the simulated survival probabilities for each relevant year.
 We then fit the IDSM to each of the 10 simulated datasets three times, using distinct seeds for both simulating initial values and initiating and running the MCMC.
@@
 For the model fits to simulated and real data we ran 3 and 4 MCMC chains with NIMBLE's standard samples for 500k and 100k iterations, respectively.
-300k and 40k and thereof were discarded as burn-in prior to thinning with factors 5 and 20 , leaving us with 40k and 3k posterior samples per chain (total of 120k and 12k samples per run), repectively.
+300k and 40k and thereof were discarded as burn-in prior to thinning with factors 5 and 20 , leaving us with 40k and 3k posterior samples per chain (total of 120k and 12k samples per run), respectively.
 

### 2024-01-05 Manuscript_Nilsen_Nater.qmd
Edited/rewrote/commented on discussion

@@
 Given the available data, the IDSM was not able to estimate a clear effect of small rodent abundance on ptarmigan recruitment (slope-paramater for the z-standardized rodent occurrence data = `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$Mean, 3)` ; 95% C.I.
-= `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$lower, 3)` - `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$upper, 3)`).
+= `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$lower, 3)` - `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$upper, 3)`, Supplementary Figure "Rep_betaR.R.png").
 
@@
 
-We developed a model that use line transect distance sampling survey data combined with data from marked individuals to estimate population abundance, survival probabilities and recruitment rates in an open population setting.
-We first used simulated data to examine models ability to recover the underlying parameters when they were known.
-Then we fitted the model to data from an ongoing field study on willow ptarmigan in Norway.
-In general, the model did a very good job in recreating the underlying parameters when fitted to simulated data where the true values are known.
-For the simulated data, a relative wide range of parameter values were allowed, and we did not detect any particular problems or biases with respect to this variation in the underlying parameters.
-When the input data are unbiased (with respect to the underlying model formulation) we therefore consider the model to be able to extract unbiased and meaningful demographic parameters from distance sampling data.
+We developed an integrated model that jointly analyses line transect distance sampling survey data and data from marked individuals to estimate population abundance, survival probabilities ,and recruitment rates over time.
+We first used simulated data to examine model's ability to recover the underlying parameters when they were known.
+We then fitted the model to data from an ongoing field study on willow ptarmigan in Norway to showcase its applicability to real wildlife monitoring data.
 
-Open population formulations of the distance sampling model has previously been presented (refs), and applied to various ecological systems (refs).
-Our model extends those previous applications by formulating the underlying process model as a stage-structured matrix model where the matrix elements are represented by annual survival probabilities and recruitment rates.
-While such models have been available for a range of other modelling frameworks previously (in particular), this is to our knowledge the first explicit use of such a state model in a distance sampling framework.
-Such a formulation allow us to both to obtain estimates of the underlying demographic rates, but also to assess the effect of environmental variation directly on demographic rates and not just on population growth rate ($\lambda$).
-As an example of such environmental variation, we investigated the effects of small rodent abundance on the recruitment rates.
-The fact that we were able to successfully recover the covariate effect when the model was applied to simulated data, suggest that in fact the model is able to estimate the effect when there is a clear signal in the data.
-When applied to real data, we did not detect a clear effect of rodent abundance (indexed by the occurrence of rodents on the transect lines) on recruitment rate.
-This might be somewhat surprising, given that such a pattern has been reported repeatedly in the literature (see e.g.
+Open population formulations of the distance sampling model have previously been presented (refs), and applied to various ecological systems (refs).
+The model presented here extends those previous applications by formulating the underlying population process model as a stage-structured matrix model [@caswell2000] in which the matrix elements are represented by annual survival probabilities and recruitment rates.
+While this has been the common approach for a range of other statistical modelling frameworks, including the growing suite of models falling into the category of integrated population models [@schaub2022], the integration of mechanistic population models into distance sampling frameworks is rather new.
+The resulting modelling framework allow us to make maximum use of distance sampling data in combination with auxiliary information both from the distance sampling survey itself (i.e.
+information on age, sex, etc.
+of observed animals) and from other types of monitoring, and enables estimation not only of changes in population density but also of underlying vital rates over time.
+
+In general, the model did a very good job at recreating the underlying parameters when fitted to simulated data where the true values are known.
+The simulated data sets were based on relatively wide ranges of parameter values, yet we did not find the model to have any particular problems or biases recovering the original parameters used in simulations.
+When the input data are unbiased (with respect to the underlying model formulation) we therefore consider the model to be able to extract unbiased and meaningful demographic parameters from age-structured distance sampling data.
+
+Real data, however, are likely to be subject to certain biases, and one type of bias that is likely to be common for age-structured distance sampling data arises from failure to (correctly) classify the age class of observed individuals.
+<!-- Maybe add one illustrative sentence here about how/why this happens using our ptarmigan monitoring setup as an example?
+--> Observations with incorrectly classified age have the potential to introduce bias in the IDSMs relative estimates of survival and recruitment.
+This is due to the way it usese the distance sampling data to estimate survival and recruitment rates.
+In our process model, the population growth rate ($\lambda$) is determined by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, and this creates a dependence between the demographic parameters.
+If the age ratio in the data are biased or contain frequent misclassifications, this is likely to affect the relative contribution of survival and recruitment to the growth rate.
+To get an idea of the potential effect of this on parameter estimation, we checked the sensitivity of the output of the model fit to real data with regard to the treatment of birds classified as "unknown sex and age" by the field personnel (see **Methods**).
+In the model version presented in the results section, we made the assumption that these birds were in fact juveniles.
+<!-- Restate here why this is likely / probably a good guess --> Comparing estimates to an alternative scenario in which we discarded all birds classified as "unknown sex and age" (Appendix) we found that -- as expected -- estimated population density was virtually unaffected by the treatment of "unknown sex and age" observations, while the annual demographic parameters shifted proportional to the amount of "unknown sex and age" observations in the given year (towards higher recruitment and lower survival).
+Thus, biases in the reported age ratios may affect estimated of demographic rates, but not so much population density.
+Since the proportion of "unknown sex and age" observations in our ptarmigan case study was rather low, potential biases in estimates resulting from age misclassification are expected to be small.
+Nonetheless, future developments of the IDSM modelling framework should focus on ways of accounting explicitly for misclassification of age class in the field.
+
+<!-- I suggest we start this next paragraph with a couple of general sentences about the density estimates, even if it is just stating that the density estimates correspond closely to what we know/have estimated with other models previously.
+Maybe also add one sentence or so about the possible biology behind the recent increase.
+Then moving on to survival and recruitment -->
+
+Compared to previous studies on ptarmigan (see e.g.
+[@Israelsen2020; @Sandercock2011]), we could have expected somewhat higher estimates of survival probability.
+One potential reason for the overall lower estimates obtained here is that our IDSM analysis assumed constant survival over the period 2007-2021 while survival, in reality, may have changed over time.
+If survival in more recent years, when telemetry data was collected [and the study of e.g.
+@@Israelsen2020 was carried out] was higher than in earlier years -- something that seems likely given population increase over recent years -- an average over the entire time period is expected to be lower.
+On a different note, we can also not exclude the possibility of a small degree of bias in survival estimates due to misclassification of age in the data (see above), especially seeing as the IDSM's recruitment rate estimates also appear somewhat high compared to other studies on ptarmigan (+refs).
+<!-- If we have any "biological" thoughts about why somewhat high recruitment rates may not be that ridiculous here, then that could be added here -->
+
+In addition to estimating demographic rates from line transect data, the IDSM also allows including relevant environmental effects on the demographic rates themselves, and not just on population growth rate as a whole ($\lambda$).
+In the ptarmigan case study we thus attempted to investigate the effect of small rodent abundance (approximated as the proportion of transect lines on which rodents were reported each year) on recruitment rate.
+<!-- Deleted this because we did NOT include the rodent effect in the simulated data tests --> Unfortunately, we were not able to detect a clear effect of rodent abundance due to large uncertainty associated with the estimate (Supplementary Figure "Rep_betaR.R.png").
+This may seem somewhat surprising at first glance, given that such a pattern has been reported repeatedly in the literature (see e.g.
 @Bowler2020).
-We speculate that there are (at least) three potential explenations to this result.
-First, for our specific case study the data on rodent abundance was heavily zero-inflated, and the annual variation in the index was rather small.
-It is well known that the amplitude and regularity of the rodent cycles has been fading in recent decades (refs), and our study area might be no exception.
-Lack of peak rodent years in the time series to which we fitted the model.
-In addition, previous research has shown that ptarmigan recruitment is also sensitive to the weather in the late winter and spring (before and during the breeding season); we did not fit any weather covariates to the model so there is a possibility that strong effects of spring conditions in certain years have masked any remaining effects of small rodent abundance.
-Finally, the data set we used is relative short (15 years), leaving us with somewhat limited statistical power.
-We therefore do not consider our a particularily strong test of the underlying effect of small rodent fluctuations on ptarmigan recruitment rates.
-In future applications, a stronger test based on more years of data and data from additional areas is encouraged.
-In e.g.
-[@Bowler2020] we used data from the same sampling program but from more areas, and detected a very clear signal from small rodent abundance on ptarmigan population growth rate.
-Our new model allow us to investigate to which extent similar results emerge when we can link the covariate effect to the actual demographic rate and not only to the population growth rate.
+We speculate that there are (at least) three potential explanations to this.
+The first is that our covariate data may not have been well suited for estimating effects on recruitment.
+The data on rodent abundance was heavily zero-inflated, and the annual variation in the index was rather small otherwise, making for a covariate with relatively little information content.
+While this may be partially a consequence of how these data are collected, it is also well known that the amplitude and regularity of the rodent cycles has been fading in recent decades (refs), and our study area might be no exception.
+Lack of peak rodent years in the time series to which we fitted the model may thus also have contributed to making effect estimation challenging.
+Second, it is possible that rodent effects were obscured by other, potentially stronger, covariate effects.
+Previous research has shown that ptarmigan recruitment is also sensitive to the weather in the late winter and spring (before and during the breeding season); as we did not fit any weather covariates to the model so there is a possibility that strong unaccounted for effects of spring conditions in certain years may have masked any remaining effects of small rodent abundance.
+Finally, the data set used in this analysis is relatively short (15 years), leaving us with somewhat limited statistical power to detect effects of temporal covariates .
+Taken together we therefore do not consider this study as a particularly strong test of the underlying effect of small rodent fluctuations on ptarmigan recruitment rates.
+It is worth noting that future applications could increase statistical power by including either more years of data or capitalizing on space-for-time substitution as the Norwegian ptarmigan monitoring programme spans many more locations beyond Lierne.
+@Bowler2020, for example, used data from the same sampling program but from more areas using a simpler open population DS model, and detected a very clear signal from small rodent abundance on ptarmigan population growth rate.
+An extension of our IDSM to include data from multiple areas therefore constitutes a promising approach for investigating to which extent similar results emerge by linking environmental covariates to the actual demographic rates and not only just to the resulting population growth rates.
 
-One potential source of bias in our model framework is related to the way that we use the distance sampling data to estimate survival and recruitment rates.
-In our process model, the population growth rate ($\lambda$) is determined by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, and thsi create a dependence between the demographic parameters.
-If the age ratio in the data are biased or contain frequent misclassifications, this is likely to affect the relative contribution of survival and recruitment to the growth rate.
-As a test of the phenomena, we also assessed how sensitive the model output was with regard to the treatment of birds classified as "unknown sex and age" by the field personnel (see **Methods**).
-In the version presented in the results section, we made the assumption that these birds were in fact juveniles.
-In an alternative scenario, we discarded all birds classified as unknowns, and rerun the model (Appendix).
-As expected the estimated population density was virtually unaffected by this decision, whereas the effect on the demographic parameters were somewhat more affected.
-Thus biases in the age ratio will affect the demographic rates, but not so much the estimated density.
-Compared to previous studies on ptarmigan (see e.g.
-[@Israelsen2020][@Sandercock2011]), we could have expected a higher survival rate indicating that there is some misclassification in the data with respect to age specification.
-In future developments of the model, explicit accounting for such misclassification will be considered.
-
--   Both simulated data tests and real data application have shown that mixing is suboptimal for the average recruitment rate parameter (Mu.R).
-Future work should perhaps look into optimizing sampling?
-(For biological interpretation that is not a big problem as the yearly recruitment rate estimates are fine...
-but have to keep in mind when interpreting) <!-- I leave it to you Erlend to decide where, whether, and how much you want to go into this one -->
-
-In summary, we conclude that the model presented here is useful in many ecological systems where the sampling survey data include information about the age-ratio of the observed individuals.
-Following the integrated modelling philosophy, it also allow for the integration of auxiliary data.
-In our application here, we integrated data from marked birds by formulating an explicit (known-fate) mark-recapture model that explicitly supported the estimation of survival probabilities in the state process.
-The model is flexible, and can be extended with additional auxiliary data, such as separate data on recruitment rates that could improve the precision of the demographic rate estimates.
-Moreover, the model can be extended with additional covariates, to gain a better understanding of how species respond to environmental variation through the effects on vital rates.
-Finally, as the underlying model structure is based on a hierarchical model the framework readily support multi-area extensions, where the covariate-effects are efficiently estimated by time-for-space substitution.
+The new IDSM framework presented here is relevant for many wildlife populations that are surveyed using line transect sampling that includes additional information on age, sex, and/or life stages of the observed individuals.
+Following the integrated modelling philosophy, the IDSM also allows for the integration of auxiliary data.
+In our application here, we integrated data from radio-telemetry of marked birds, which explicitly supported the estimation of survival probabilities.
+The IDSM framework is very flexible, however, and open to the inclusion of additional/other auxiliary data that contains information on demographic rates or population size/density.
+Moreover, hierarchical nature of the model makes is straight forward to adapt to different species and to include different suites of environmental covariates on the demographic rates.
+Finally, it constitutes a modelling framework that is well suited for extension to multiple areas and thus able to capitalize on space-for-time substitution to produce large-scale and spatially explicit estimates of population density, demographic rates, and environmental effects from large-scale (participatory) monitoring.
 

### 2024-01-05 Manuscript_Nilsen_Nater.qmd
Edited acknowledgements

@@
 Fjellstyrene i Lierne contributed both to the line transect program and the field work related to the marked willow ptarmigan.
-Norwegian Environment Agency funded the projects (grant nrs **TO BE ADDED**).
-\[+ additional people contributing to the various projects\]
-
-\[Should we thank James for discussions etc.?\]
+We also thank James A.
+Martin for constructive discussions during the later stages of this work.
+Both data collection and analytical work were financed by the Norwegian Environment Agency funded the projects (grant nrs **TO BE ADDED**).
 

### 2024-01-05 Manuscript_Nilsen_Nater.qmd
Updated acknowledgments, data/code availability, added code ref

@@
 If survival in more recent years, when telemetry data was collected [and the study of e.g.
-@@Israelsen2020 was carried out] was higher than in earlier years -- something that seems likely given population increase over recent years -- an average over the entire time period is expected to be lower.
+\@ @Israelsen2020 was carried out] was higher than in earlier years -- something that seems likely given population increase over recent years -- an average over the entire time period is expected to be lower.
 On a different note, we can also not exclude the possibility of a small degree of bias in survival estimates due to misclassification of age in the data (see above), especially seeing as the IDSM's recruitment rate estimates also appear somewhat high compared to other studies on ptarmigan (+refs).
@@
 ### Data and code availability
+
+The raw data from the line transect surveys is deposited on GBIF and can be accessed freely via the Living Norway Data Portal (https://data.livingnorway.no/).
+The work here is based on version 1.7 of the dataset "Tetraonid line transect surveys from Norway: Data from Fjellstyrene" (key b49a2978-0e30-4748-a99f-9301d17ae119).
+
+The auxiliary radio-telemetry data and rodent occupancy data, and all code used for wrangling, analysing, and visualizing data and results can be found in the project's repository on GitHub: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.
+The results presented in this paper were created using version 1.3 of the code [@IDSMCode_2024].
+
+Supplementary figures and posterior samples from the model run on the real data are archived on OSF: (**TBA**)<!-- Add link/reference to OSF storage -->

### 2024-01-07 Manuscript_Nater_etal.qmd
Minor additions to intro and results

--- 2023-12-30 Manuscript_Nater_etal.qmd
@@
 ---
-title: "Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird"
+title: "Large-scale spatiotemporal variation in vital rates of an alpine bird"
 format: html
@@
 There is growing demand for biodiversity indicators from international unions, national government, local management bodies, and corporate and industry actors.
-Such indicators ideally represent a wide range of biodiversity's states and functions [e.g.
-Essential Biodiversity Variables, @pereira2013; @jetz2019], yet the development of suitable indicators for certain qualities, such as species abundance and demography, has been more difficult than for others [@schmeller2018; @waldock2022].
+Such indicators ideally represent a wide range of biodiversity's states and functions (e.g.
++ ref on basic EBV idea), yet the development of suitable indicators for certain attributes, such as species abundance and demography, has been more difficult than for others (+ref).
 This is at least partially due to challenging requirements regarding spatial scales of useful biodiversity indicators.
-On one hand, indicators need to be representative at large geographic scales, for example, in the context of countries' reporting towards biodiversity targets [e.g.
-@feld2009].
-On the other hand, indicators also ideally have good spatial resolution as the scales relevant for local-level management and planning are often much smaller [@stevenson2021].
+On one hand, indicators need to be representative at large geographic scales, for example, in the context of countries' reporting towards biodiversity targets (+ ref).
+On the other hand, indicators also ideally have good spatial resolution as the scales relevant for local-level management and planning are often much smaller (+ ref).
 This latter requirement is particularly crucial for (infrastructure) development strategies and for species management and conservation, both of which tend to require knowledge on species abundance and population dynamics (i.e.
-demographic rates) that is relevant for their county- or municipality-level decision making (+ refs?).
-Another reason why abundance and population indicators ideally come with good spatial resolution is that there can be substantial amounts of variation in population dynamics and life history of species across space [e.g.
-@robinson2014; @horswill2019].
+demographic rates) that is relevant for their county- or municipality-level decision making (+ refs).
+Another reason why abundance and population indicators ideally come with good spatial resolution is that there can be substantial amounts of variation in population dynamics and life history of species across space (+ref).
 This is variation that needs to be accounted for in order to develop successful and sustainable strategies for area use, (harvest) management, and species and biodiversity conservation [@williams2002].
 
-While large-scale, spatially-explicit indicators for species abundance and populations are clearly needed, development and practical implementation are greatly limited due to the reliance of such indicators on the availability of data from large-scale, long-term monitoring programmes [@proencca2017].
+While large-scale, spatially-explicit indicators for species abundance and populations are clearly needed, development and practical implementation are greatly limited due to the reliance of such indicators on the availability of data from large-scale, long-term monitoring programmes (+ref).
 Consequently, many countries have been working on setting up, maintaining, and improving such monitoring programmes over the last decades.
-Many now well-established programmes focus on breeding birds and butterflies, and examples include the North American Breeding Bird Survey (<https://www.pwrc.usgs.gov/bbs/>), UK Butterfly Monitoring Scheme (<https://ukbms.org/>), Swiss Biodiversity Monitoring (<https://www.biodiversitymonitoring.ch/>), and many more.
+Many now well-established programmes focus on breeding birds and butterflies, and examples include the North American Breeding Bird Survey (<https://www.pwrc.usgs.gov/bbs/>), UK Butterfly Monitoring Scheme (<https://ukbms.org/>), the Game and Wildlife Conservation Trust Partridge Count Scheme [@aebischer2010], and Swiss Biodiversity Monitoring (<https://www.biodiversitymonitoring.ch/>).
 <!-- is there any other one we want to name here specifically?
@@
 There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly and requires trained specialists, thus limiting the amount of data that can be collected, while participatory monitoring, i.e.
-the collection of ecological data by members of the public [also called citizen science, @fraisl2022], allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias [@johnston2023].
-Consequently, many large-scale monitoring programmes are often limited to presence-absence or very simple count observations, making them suitable for the development of indicators of distribution and population trends, but usually not for indicators of abundance, population dynamics, and demographic rates (+ example refs).
+the collection of ecological data by members of the public (also called citizen or community science, +ref), allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias (+ref).
+Consequently, many large-scale monitoring programmes are limited to presence-absence or very simple count observations, making them suitable for the development of indicators of distribution and population trends, but usually not for indicators of abundance, population dynamics, and demographic rates (+ ref).
 The exception here are monitoring programmes that succeed in making use of a large number of citizen scientists that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol.
+For example, in the United States hunters participate in the collection of bands and wings from harvested American Woodcock (*Scopolax minor)* to estimate survival and age ratios [@zimmerman2010].
 <!-- can we think of 1-2 internationally well-known examples?
@@
 The latter means that information about spatio-temporal variation in demographic rates and population dynamics is particularly important in order to design sustainable harvest strategies.
-In addition, as a sentinel species the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is included in both the main national biodiversity [Nature Index for Norway, @jakobsson2020, <https://www.naturindeks.no/Indicators/lirype>] and ecosystem [Assessment of the Ecological Condition, @framstad2022] assessments.
+In addition, as a sentinel species the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is included in both the main national biodiversity [Nature Index for Norway; @jakobsson2020; <https://www.naturindeks.no/Indicators/lirype>] and ecosystem [Assessment of the Ecological Condition; @framstad2022] assessments.
 
@@
 In addition, the local study designs are reviewed by staff at NINA, and common recommendations for study design are provided.
-However, due to the fact that participation by stakeholders is voluntary, the spatial distribution of transects and sampling effort is not homogeneous.
+However, because participation by stakeholders is voluntary, the spatial distribution of transects and sampling effort is not homogeneous.
 In general, the effort is higher in South-Eastern and Central Norway, medium in Northern Norway, and low in Western and Southern Norway.
 
-In this study we used all publicly available data for the period 2007-2021, which included a total of 2225 transects in 41 different reporting districts spanning 9 counties and 50 municipalities.
+In this study we used all publicly available data for the period 2007-2021, which included a total of 2225 transects in 41 different survey areas spanning 9 counties and 50 municipalities.
 Transects on which no willow ptarmigan were ever observed (i.e.
@@
 
-@nilsen2023 recently developed an integrated distance sampling model (IDSM) which jointly analyses line transect and radio-telemetry data and applied it to willow ptarmigan in the Western part of Lierne municipality in Norway.
-The model consists of a deterministic population model with two age classes (juveniles and adults) and four data likelihoods: 1) likelihood for observation distances from transect lines for estimating detection probability; 2) likelihood for age-specific counts on transect surveys for estimating numbers of juveniles and adults present; 3) likelihood for juvenile to adult ratios observed at the locality level to provide estimate recruitment rate (as juveniles/adult); and 4) likelihood for known-fate telemetry data to estimate seasonal and annual survival.
-Below, we describe our new extension of this model to include data from several as opposed to just one.
-For more detailed information on the single-site model, including tests of model performance, see @nilsen2023.
+-   Introduction of integrated Distance sampling model (IDSM) from MS1
 
@@
 
-For applying the ptarmigan IDSM across all 41 reporting districts we included an area index in all model parameters (@fig-model) and enabled sharing of information among areas by explicitly modelling spatial variation alongside shared temporal and residual variation in vital rates and detection parameters.
+-   Addition of spatial dimension
+-   Hyper-parameter distributions for spatial variation
+-   Temporal variation (shared year RE and "residual" variation)
 
-The spatially-explicit formulation of the two age-class population model can be written as:
+The model developed as part of this project and presented in @nilsen2023 was designed with the potential of extension to run on data from multiple areas simultaneously in mind, and this extension involves two steps in practice.
 
-$$
-\begin{aligned}
-  D_{juv, x, j, t+1} & = D_{ad, x, j, t+1} * R_{x, t+1} \\
-  D_{ad, x, j, t+1} & = S_{x, t} * (D_{juv, x, j, t} + D_{ad, x, j, t})  
-\end{aligned}
-$$
-
-Here, $D_{juv, j, x, t}$ and $D_{ad, j, x, t}$ are the densities of juvenile and adult ptarmigan in survey site (= transect) $j$ of area $x$ in year $t$, respectively.
-Both juveniles and adult survive from year $t$ to $t+1$ with an area- ($x$) and year- ($t$) specific survival probability $S_{x, t}$, and survivors produce the next generation of juveniles according to an area- and year-specific recruitment rate ($R_{x,t}$).
+The first step is the inclusion of an additional dimension, area, to enable using data collected from multiple areas (@fig-model).
+The area-dimension is added across all model components with the exception of the known-fate telemetry data (and seasonal decomposition of survival estimated from it), as that is only available for the Lierne area.
 
@@
 R\[x,t\] = recruitment rate in year t.
-S\[x,t\] = survival probability from year t to t+1.
-Note that the additional site (=transect) dimension, "j", is omitted for the sake of illustration.](Figures/RypeIDSM_multiArea_structure.png){#fig-model}
+S\[x,t\] = survival probability from year t to t+1.](Figures/RypeIDSM_multiArea_structure.png){#fig-model}
 
-The initial densities, $D_{juv, x, j, 1}$ and $D_{ad, x, j, 1}$ are modelled for each site (= transect) as random realizations of log normal distributions with area-specific log means ($\mu^{D1}_x$) and log standard deviations ($\sigma^{D1}_x$).
-Survival ($S_{x,t}$) and recruitment ($R_{x,t}$), on the other hand, are assumed to be the same for all sites $j$ within a given area $x$ and were modelled as:
+The second step is making adjustments to capitalize on the integration of data from multiple areas to make maximal use of available information.
+There are several not mutually exclusive options for approaching this, and one of them is the use of "hyper-parameters" for sharing information across areas [e.g.
+@horswill2019; @horswill2021].
+As an illustrative example, take the area-specific average annual survival probability, $\mu^S_x$.
+Instead of assigning independent non-informative priors to each $\mu^S_x$, we now assume that the variation in average survival across space can be described by a common (logit) normal distribution such that:
 
 $$
-\begin{aligned}
-  logit(S_{x, t}) & = logit(\mu^S) + \varepsilon^{X.S}_{x} + \varepsilon^{T.S}_{t} + \varepsilon^{R.S}_{x,t} \\
-    log(R_{x, t}) & = log(\mu^R) + \beta_x*rodentOcc_{x,t} +  \varepsilon^{X.R}_{x} + \varepsilon^{T.R}_{t} + \varepsilon^{R.R}_{x,t}
-\end{aligned}
-$$ The global means ,$\mu$, and the normally distributed spatial random effects, $\varepsilon^X$, represent the equivalent of what is elsewhere referred to as "hyper-parameter distributions" for sharing information on demographic rates across areas [e.g.
-@horswill2019; @horswill2021].
-We also used this same approach for defining the area-specific effects ($\beta_x$) of local yearly rodent occupancy ($rodentOcc_{x,t}$) on recruitment.
-In addition to spatial variation in survival and recruitment, we also included large-scale temporal variation through random year effects that were shared by across all areas ($\varepsilon^T_t$) and otherwise unaccounted for variation through year- and area-specific residual random effects ($\varepsilon^R_{x,t}$).
-Spatial, temporal, and residual random effects were modelled as normally distributed with globally defined (= shared) standard deviations.
+\begin{equation}
+\mu^S_x \sim Normal(logit(h.\mu^S), h.\sigma^S)
+\end{equation}
+$$ where $h.\mu^S$ is the "hyper mean" and $h.\sigma^S$ is the "hyper standard deviation". Non-informative priors are then assigned to the hyper-parameters instead of the area-specific realizations:
 
-The three likelihoods for data resulting from the line transect sampling (observation distances, age-specific counts, and juvenile to adult ratios, see above) were also formulated as spatially explicit, with year- and area-specific distance sampling detection parameters modelled in the same way as survival and recruitment (bar the effect of rodent occupancy, @fig-model).
-For the known-fate telemetry data (and the seasonal decomposition of survival estimated from it), on the other hand, we did not add an additional area dimension as this data was only available for the Lierne area.
+$$
+\begin{align}
+h.\mu^S & \sim Uniform(0,1) \\
+h.\sigma^S & \sim Uniform(0, 5)
+\end{align}
+$$
+
+For the multi-area implementation of the IDSM, we used the equivalent of the above approach for parameters representing average survival probabilities, baseline recruitment rates, rodent effect on recruitment, and average distance sampling detection.
+Initial population density, on the other hand, was estimated independently for each area.
+Random effects standard deviations were assumed to be the same for all areas, effectively resulting in space-for-time subtitution for random effect estimation.
+
+Another approach to maximising information gain when jointly analysing data from different areas is explicit modelling of spatial and spatio-temporal variation.
 
@@
 
-We implemented our multi-area IDSM in a Bayesian framework using NIMBLE version 1.0.1 [@valpine2017] in R version 4.3.1 [@R2023].
-For the likelihood for line transect observation distances we used a custom half-normal distribution developed by Michael Scroggie in the "nimbleDistance" package (https://github.com/scrogster/nimbleDistance).
-We used non-informative uniform priors for all parameters and including biologically sensible boundaries where possible.
-We simulated complete sets of initial values for all model nodes prior to model running and using pre-defined seeds to ensure reproducibility.
-Using NIMBLE's standard samplers, we then ran 4 MCMC chains of 100k iterations each.
-We discarded the first 40k samples of each chain as burn-in, and thinned the remainder by a factor 20, resulting in a final joint posterior containing a total of 4 x 3k = 120k samples (note that high thinning rates were necessary to constrain memory load of the joint posterior, which included 314568 monitored parameters).
+-   Bayesian, NIMBLE
+-   Priors
+-   Default samplers
+-   Chains, iterations, burn-in, thinning, posterior samples
 
-## Post-hoc variance decomposition
+## Variance decomposition
 
-Following model fitting, we calculated posterior distributions for the proportions of variance in survival probabilities, recruitment rates, and detection decay explained by 1) spatial variation ($var_{area}$), 2) temporal variation ($var_{year}$), 3) residual variation ($var_{residual}$), and 4) variation in rodent occupancy ($var_{rodent}$).
-To obtain the proportion variance explained by each of the component, we divided it by the sum of all other components ($var_{area} + var_{year} + var_{residual} + var_{rodent}$).
-The spatial, temporal, and residual variance components were defined as the square of the estimated corresponding random effects standard deviation from the model while $var_{rodent}$ was calculated as the variance of all area- and year-specific $\beta_x*rodentOcc_{x,t}$ products.
+-   How variance decomposition was done
 
@@
 
-Reproducibility and ease of repeating analyses was a key focus when developing the multi-area IDSM.To that end, we set up the workflow as a "targets pipeline", implemented through the R package "targets" [@landau2021].
-The pipeline contains a variety of options for controlling modelling decisions in the workflow such as the year range of data to consider, the level of spatial aggregation (i.e.
-reporting district vs.
-survey locality), whether to model time variation in survival and/or effects of rodent abundance, whether to run MCMC chains sequentially or in parallel, etc.
-A visual representation of the pipeline is also displayed in @fig-workflow and for more details on pipeline implementation and options, we refer the reader to the GitHub repository: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.
+Reproducibility and ease of repeating analyses was a key focus when developing the multi-area IDSM.To that end, we set up the workflow as a "targets pipeline", implemented throught the R packages "targets" [@landau2021].
+A visual representation of the pipeline is displayed in @fig-workflow.
 
@@
 
-```{r, echo = FALSE}
-# Posterior summaries
-post.sum <- readRDS("Results/PosteriorSummaries_byArea.rds")
-
-# Printing function
-printPostSum <- function(table, row = NULL, area = NULL){
-
-  if(!missing(area)){
-    row <- which(table$Area == area)
-  }
-
-  if(!("Area" %in% colnames(table))){
-    table$Area <- "None"
-  }
-
-  print.out <- list(
-    area.name = table$Area[row],
-    post.sum = paste0(
-      round(table$Median[row], digits = 3), 
-      " [",
-      round(table$lCI[row], digits = 3),
-      ", ",
-      round(table$uCI[row], digits = 3),
-      "]"
-    )
-  )
-  return(print.out)
-}
-
-# Detection stats
-#Det.hyper.mean
-#Det.sd.spatial
-#Det.sd.temporal
-#Det.sd.residual
-#Det.min
-#Det.max
-
-# Survival stats
-S.hyper.mean <- printPostSum(table = post.sum$hParams.sum, 
-                              row = which(post.sum$hParams.sum$Parameter == "h.Mu.S"))
-S.sd.spatial <- printPostSum(table = post.sum$hParams.sum,
-                             row = which(post.sum$hParams.sum$Parameter == "h.sigma.S"))
-S.sd.temporal <- printPostSum(table = post.sum$hParams.sum,
-                             row = which(post.sum$hParams.sum$Parameter == "sigmaT.S"))
-S.sd.residual <- printPostSum(table = post.sum$hParams.sum,
-                             row = which(post.sum$hParams.sum$Parameter == "sigmaR.S"))
-S.min <- printPostSum(table = post.sum$pSurv.sum, 
-                      row = which(post.sum$pSurv.sum$Median == min(post.sum$pSurv.sum$Median)))
-S.max <- printPostSum(table = post.sum$pSurv.sum,
-                      row = which(post.sum$pSurv.sum$Median == max(post.sum$pSurv.sum$Median)))
-
-# Recruitment stats
-R.hyper.mean <- printPostSum(table = post.sum$hParams.sum,
-                             row = which(post.sum$hParams.sum$Parameter == "h.Mu.R"))
-R.sd.spatial <- printPostSum(table = post.sum$hParams.sum,
-                             row = which(post.sum$hParams.sum$Parameter == "h.sigma.R"))
-R.sd.temporal <- printPostSum(table = post.sum$hParams.sum,
-                             row = which(post.sum$hParams.sum$Parameter == "sigmaT.R"))
-R.sd.residual <- printPostSum(table = post.sum$hParams.sum,
-                             row = which(post.sum$hParams.sum$Parameter == "sigmaR.R"))
-R.min <- printPostSum(table = post.sum$rRep.sum, 
-                      row = which(post.sum$rRep.sum$Median == min(post.sum$rRep.sum$Median)))
-R.max <- printPostSum(table = post.sum$rRep.sum,
-                      row = which(post.sum$rRep.sum$Median == max(post.sum$rRep.sum$Median)))
-
-# Rodent effects stats
-betaR.hyper.mean <- printPostSum(table = post.sum$hParams.sum,
-                             row = which(post.sum$hParams.sum$Parameter == "h.Mu.betaR.R"))
-betaR.sd.spatial <- printPostSum(table = post.sum$hParams.sum,
-                             row = which(post.sum$hParams.sum$Parameter == "h.sigma.betaR.R"))
-betaR.min <- printPostSum(table = post.sum$betaR.sum, 
-                       row = which(post.sum$betaR.sum$Median == min(post.sum$betaR.sum$Median)))
-betaR.max <- printPostSum(table = post.sum$betaR.sum,
-                      row = which(post.sum$betaR.sum$Median == max(post.sum$betaR.sum$Median)))
-n.pos.betaR <- length(which(post.sum$betaR.sum$Median > 0))
-n.neg.betaR <- length(which(post.sum$betaR.sum$Median < 0))
-
-# Population density stats
-popDens.sum <- subset(post.sum$popDens.sum, SummaryPeriod == "2018-2021")
-dens.min <- printPostSum(table = popDens.sum, 
-                      row = which(popDens.sum$Median == min(popDens.sum$Median)))
-dens.max <- printPostSum(table = popDens.sum,
-                      row = which(popDens.sum$Median == max(popDens.sum$Median)))
-
-# Population growth rate stats
-lambda.sum <- subset(post.sum$lambda.sum, SummaryPeriod == "2018-2021")
-lambda.min <- printPostSum(table = lambda.sum, 
-                      row = which(lambda.sum$Median == min(lambda.sum$Median)))
-lambda.max <- printPostSum(table = lambda.sum,
-                      row = which(lambda.sum$Median == max(lambda.sum$Median)))
-n.pos.lambda <- length(which(lambda.sum$Median > 1))
-n.neg.lambda <- length(which(lambda.sum$Median < 1))
-
-```
-
-All numerical results in the following are presented as median \[95% credible interval\] unless otherwise indicated.
-Posterior summaries (median, 95% credible interval, mean, standard deviation, coefficient of variation) for all main parameters are also provided in the supplementary file "PosteriorSummaries_byAreas.csv".
-Supplementary figures (SF) are provided as .pdf files with captions in "SuppFigures_Captions.txt".
-
 ## Population density
 
-Only during the most recent four years (2018-2021) has data been collected regularly for all reporting areas included in the analyses.
-During this period, estimated population densities varied between `r dens.min$post.sum` birds/km\^2 in the area "`r dens.min$area.name`" close to the coast in central Norway to `r dens.max$post.sum` birds/km\^2 in "`r dens.max$area.name`" further south near the Swedish border.
-In general, population density appeared to be lowest in the Northern parts of Norway and highest in the eastern part of central Norway (@fig-densA).
-Uncertainty in density estimates was relatively consistent, with a few areas (inlcuding the one with the lowest estimated density, "`r dens.min$area.name`") sticking out by having substantially less precise estimates (@fig-densB).
-Populations fluctuated substantially over time in any given area (SF "TimeSeries_popDens1.pdf") and some years seemed to be indicative of relatively high (e.g.
-2011, 2014, 2018) or low (e.g.
-2012, 2015) densities across a large number of areas.
+@fig-densA shows that population density is generally lowest in the Northern parts of Norway and highest in central Norway, close to the Swedish borders.
+Higher median density estimates also seem to be associated with higher estimation uncertainty (@fig-densB) and therefore have to be interpreted with care.
 
 ::: {#fig-dens layout-ncol="2"}
-![Median density estimates](Figures/Avg_popDens_Map-4.png){#fig-densA}
+![Median density estimates](Figures/density_2018-2020_Median.png){#fig-densA}
 
-![Uncertainty in density estimates](Figures/Avg_popDens_Map-6.png){#fig-densB}
+![Uncertainty in density estimates](Figures/density_2018-2020_SD.png){#fig-densB}
 
-Median (a) and coefficient of variation (standard deviation / mean) (b) of posterior estimates of average ptarmigan density in the four most recent years (2018-2021) across 41 reporting areas (summarised at the municipality level) in Norway.
+Median (a) and standard deviation (b) of posterior estimates of average ptarmigan density in the three most recent years (2018-2020) across 41 reporting areas in Norway.
 Darker colors indicate higher median values and higher uncertainty.
@@
 
-Average population growth rates over the last four years (2018-2021) ranged from moderate declines (`r lambda.min$post.sum` in the "`r lambda.min$area.name`" area) to \> 50% increase (`r lambda.max$post.sum` in the "`r lambda.max$area.name`" area).
-In the majority of reporting areas (`r n.pos.lambda` out of 41), populations of willow ptarmigan have been increasing over the period 2018-2021 (@fig-lam).
-Some areas -- predominantly in central Norway -- also had declining populations, but many of those declines followed upon periods of increase between the start of data collection in 2007 and sometime between 2016 and 2018 (SF "TimeSeries_popDens1.pdf").
+In most areas, populations of willow ptarmigan have been increasing over the period 2018-2020, potentially indicating a positive effect of more restricted hunting in recent years (@fig-lam).
+Some areas -- predominantly in central Norway -- also had declining populations, but many of those declines followed upon a period of increase between the start of data colection in 2007 and sometime between 2016 and 2018.
 
 ::: {#fig-lam layout-ncol="2"}
-![Median population growth rate estimates](Figures/Avg_lambda_Map-4.png){#fig-lamA}
+![Median population growth rate estimates](Figures/lambda_2018-2020_Median.png){#fig-lamA}
 
-![Uncertainty in population growth rate estimates](Figures/Avg_lambda_Map-5.png){#fig-lamB}
+![Uncertainty in population growth rate estimates](Figures/lambda_2018-2020_SD.png){#fig-lamB}
 
-Median (a) and standard deviation (b) of posterior estimates of average annual population growth rate over the four most recent years (2018-2021) across 41 reporting areas (summarised at the municipality level) in Norway.
+Median (a) and standard deviation (b) of posterior estimates of average annual population growth rate over the three most recent years (2018-2020) across 41 reporting areas in Norway.
 In a), pinkish colors indicate declining populations while greenish colors indicate growing populations (white = stable populations).
-In b) darker colors indicate higher uncertainty.
+In b) darker colors indicate higher median values and higher uncertainty.
 :::
 
-The highest population growth rates were estimated for areas with relatively low population densities across latitudes [@fig-paramCorrs A] but we did not find evidence for a strong association between population growth rates and population densities overall.
+## Survival probability
 
-![Posterior summaries (points = posterior medians, lines = 95% credible intervals) of area-specific population growth rate vs.
-population density (A), recruitment rate vs.
-survival probability (B) and rodent effect on recruitment along a longitudinal gradient (C).
-Color indicates latitude of the midpoint of each area.](Figures/AllParams_Latitude.png){#fig-paramCorrs}
+Annual survival probabilities ranged from aroung 35% to 50% across reporting areas in Norway, with the highest values occurring in the far north and in the mountains in the south (@fig-pSurv).
 
-## Survival probabilities and recruitment rates
+::: {#fig-pSurv layout-ncol="2"}
+![Median survival estimates](Figures/pSurv_Median.png){#fig-pSurvA}
 
-Annual survival probabilities ranged from `r S.min$post.sum` (area "`r S.min$area.name`") to `r S.max$post.sum` (area "`r S.max$area.name`") across reporting areas in Norway, with the highest values occurring in the far north and in the mountains in the south (@fig-pVRA).
-The global average survival probability across all areas and years ($\mu^S$) was estimated at `r S.hyper.mean`.
-Spatial variation in survival (random effect SD = `r S.sd.spatial$post.sum`) was relatively low compared to temporal (`r S.sd.temporal$post.sum`) and residual (`r S.sd.residual$post.sum`) variation.
+![Uncertainty in survival estimates](Figures/pSurv_SD.png){#fig-pSurvB}
 
-Recruitment rates varied between `r R.min$post.sum` (area "`r R.min$area.name`") and `r R.max$post.sum` (area "`r R.max$area.name`") and displayed a spatial pattern opposite to that 
-of annual survival, i.e.
-lower recruitment rates co-occurring with higher survival rates and vice-versa [@fig-pVRB; @fig-paramCorrs B).
-Across all areas and years, average recruitment rate was `r R.hyper.mean$post.sum`.
-Unlike for survival, the model predicted similar magnitudes of spatial and temporal variation (random effect SDs of `r R.sd.spatial$post.sum` and `r R.sd.temporal$post.sum`, respectively), and about twice as much residual variation (`r R.sd.residual$post.sum`).
-
-::: {#fig-pVR layout-ncol="2"}
-![Median survival probabilities](Figures/Avg_pSurv_Map-1.png){#fig-pVRA}
-
-![Median recruitment rates](Figures/Avg_rRep_Map-1.png){#fig-pVRB}
-
-Posterior medians of average annual survival probabilities (a) and recruitment rates (b) across 41 reporting areas (summarised at the municipality level) in Norway.
-Darker colors indicate higher median values.
-Measures for corresponding uncertainty in estimates are visualized in SFs "Avg_pSurv_Map.pdf" and "Avg_rRep_Map.pdf" for survival and recruitment, respectively.
+Median (a) and standard deviation (b) of posterior estimates of average annual survival probability across 41 reporting areas in Norway.
+Darker colors indicate higher median values and higher uncertainty.
 :::
 
-The MCMC chains for many of the area-specific average survival probabilities and recruitment rates, as well as for the global averages (intercepts) for both vital rates, were mixing rather poorly.
-Despite that, mixing was good and resulting posteriors well defined for the area- and year-specific estimates of survival and recruitment (SF "PostDens_tS_tR.pdf").
-There was substantial variation in both vital rates across time (SFs "TimeSeries_pSurv.pdf" and "TimeSeries_rRep.pdf").
-In many areas, the years 2011, 2014, and 2018 not only sported high population densities (see above) but were also characterized by both high recruitment and low subsequent survival.
-The overall low density years 2012 and 2015, conversely, often featured lower recruitment and, in some cases, higher survival.
-Notably, there were also years with very little spatial synchrony, i.e.
-very different relative yearly survival probabilities and recruitment rates (e.g.
-2010 and 2020 for survival and 2013, 2016, and 2017 for recruitment).
+## Recruitment rate
+
+Annual recruitment rates showed a spatial pattern (@fig-rRep) that was inverse relative to that of annual survival (@fig-pSurv), with lower recruitment rates co-occurring with higher survival rates and vice-versa.
+This is consistent with basic life history theory [@stearns1992] and with previous studies finding relatively slower life histories in alpine / high altitude areas [e.g.
+@sandercock2005; @bears2009; @alice2016].
+
+::: {#fig-rRep layout-ncol="2"}
+![Median recruitment estimates](Figures/rRep_Median.png){#fig-rRepA}
+
+![Uncertainty in recruitment estimates](Figures/rRep_SD.png){#fig-rRepB}
+
+Median (a) and standard deviation (b) of posterior estimates of average annual recruitment rate across 41 reporting areas in Norway. Darker colors indicate higher median values and higher uncertainty.
+:::
 
@@
 
-The model predicted a positive global effect of rodent occupancy on recruitment rate (average slope on the log scale = `r betaR.hyper.mean$post.sum`).
-Nonetheless, spatial variation in the rodent effect was substantial (random effect SD = `r betaR.sd.spatial$post.sum`).
-This resulted in negative median effects in `r n.neg.betaR` areas, positive median effects in `r n.pos.betaR` areas, and a range of effect sizes from `r betaR.min$post.sum` (area "`r betaR.min$area.name`") to `r betaR.max$post.sum` (area "`r betaR.max$area.name`", @fig-paramCorrs C).
-The largest positive rodent effects were estimated for areas in the very North of Norway, as well as in the mountaineous regions in the central and southwestern parts of the country (SF "betaR_Map.pdf").
-Effects with negative posterior medians were located mostly at intermediate latitudes, but we note that all of these had posterior distributions featuring substantial overlap with 0 (@fig-paramCorrs C).
+The model predicted substantial variation in the effect of the rodent occupancy covariate on recruitment (@fig-betaR).
+Areas in the north and south west generally had recruitment increasing with rodent occupancy while the remaining areas were a mosaic of positive and negative effects, albeit with some spatial clustering.
+
+::: {#fig-betaR layout-ncol="2"}
+![Median rodent effect estimates](Figures/betaR_Median.png){#fig-betaRA}
+
+![Uncertainty in rodent effect estimates](Figures/betaR_SD.png){#fig-betaRB}
+
+Median (a) and standard deviation (b) of posterior estimates of effects of rodent occupancy on recruitment rate across 41 reporting areas in Norway.
+In a), blueish colors indicate negative effects while reddish colors indicate positive effects (yellow = no effect).
+In b) darker colors indicate higher median values and higher uncertainty.
+:::
 
@@
 
-```{r, echo = FALSE}
-# Posterior summaries
-varDec.sum <- readRDS("Results/PosteriorSummaries_VarDecomp.rds")
-
-# Printing function
-printPostSum2 <- function(table, Parameter, VarComponent){
-
-  idx <- which(table$Parameter == Parameter & table$VarComponent == VarComponent)
-
-  print.out <- paste0(
-      round(table$median[idx]*100, digits = 1), 
-      " [",
-      round(table$lCI[idx]*100, digits = 1),
-      ", ",
-      round(table$uCI[idx]*100, digits = 1),
-      "]"
-  )
-  return(print.out)
-}
-
-# All parameter summaries
-Rec_A <- printPostSum2(varDec.sum, "Recruitment", "Area")
-Rec_Y <- printPostSum2(varDec.sum, "Recruitment", "Year")
-Rec_R <- printPostSum2(varDec.sum, "Recruitment", "Residual")
-Rec_Cov <- printPostSum2(varDec.sum, "Recruitment", "RodentOcc")
-
-Surv_A <- printPostSum2(varDec.sum, "Survival", "Area")
-Surv_Y <- printPostSum2(varDec.sum, "Survival", "Year")
-Surv_R <- printPostSum2(varDec.sum, "Survival", "Residual")
-
-Det_A <- printPostSum2(varDec.sum, "Detection", "Area")
-Det_Y <- printPostSum2(varDec.sum, "Detection", "Year")
-Det_R <- printPostSum2(varDec.sum, "Detection", "Residual")
-```
-
-The relative importance of different components for explaining parameter variation differed among recruitment rate, survival probability, and detection decay (@fig-varDecomp).
-The largest portion of variation in recruitment was attributed to residual variation (`r Rec_R` %), followed by spatial (`r Rec_A` %) and temporal (`r Rec_Y` %) variation.
-Rodent occupancy, which contains both a spatial and a temporal dimension, explained `r Rec_Cov` % of the total variation.
-For survival, there was large uncertainty in the estimated proportions of variance explained by different components.
-The model predicted similar potential contributions from temporal (`r Surv_Y` %) and residual (`r Surv_R` %) variation and suggested that spatial variation was only responsible for `r Surv_A` % of the total variance.
-Total variance in detection decay was attributed evenly to spatial, temporal, and residual variation at `r Det_A` %, `r Det_Y` %, and `r Det_R` %, respectively.
-
-![Posterior distributions for the proportions of parameter variance explained by spatial (blue), temporal (orange), and residual (green) variation, as well as by effects of rodent occupancy (yellow).](Figures/VarDecomposition_overall.png){#fig-varDecomp}
+TBA
 
@@
 -   Biological results (interpretations and comparisons with other case studies / species / systems)
+
+[@Wilson_Martin_2011] Found x y z
+
 -   With some focus on what we are able to get insights on thanks to the integration across time and space (i.e.
@@
 -->
-
-Annual recruitment rates showed a spatial pattern that was inverse relative to that of annual survival, with lower recruitment rates co-occurring with higher survival rates and vice-versa.
-This is consistent with basic life history theory [@stearns1992] and with previous studies finding relatively slower life histories in alpine / high altitude areas [e.g.
-@sandercock2005; @bears2009; @alice2016].
 

### 2024-01-07 Manuscript_Nilsen_Nater.qmd
Submit version of manuscript + bib-file

--- 2024-01-05 Manuscript_Nilsen_Nater.qmd
@@
 title: "An integrated open population distance sampling approach for modelling age-structured populations"
-format: html
+format: docx
 author: Erlend B. Nilsen & Chloé R. Nater
@@
 
-## Abstract
+### Abstract
 
@@
 One set of approaches that has been used extensively to estimate abundance of wildlife populations is Distance Sampling (DS) methods for line or point transect survey data.
-Historically, DS models were only available as closed* <!-- I think you mean "closed" here?
---> *population models, and did not allow for the estimation of changes in abundance through time.
+Historically, DS models were only available as closed* *population models, and did not allow for the estimation of changes in abundance through time.
 The advent of open population formulations based on the DS framework greatly extended the scope of the models, but much untapped potential remains in models that estimate temporal dynamics not only in abundance but also in the underlying demographic rates.
-Here, we present an integrated distance sampling approach that utilize age-structured survey data and auxiliary data from marked individuals to jointly estimate population dynamics and the demographic rates (recruitment rate and survival probability) that drive temporal changes .
+Here, we present an integrated distance sampling approach that utilize age-structured survey data and auxiliary data from marked individuals to jointly estimate population density and the demographic rates (recruitment rate and survival probability) that drive temporal changes density.
 The core of the resulting model is, in essence, equivalent to both an integrated population model and a matrix population model with two age classes: juveniles and adults.
@@
 
-<!-- Just a note: we seems to also still be missing quite a few references in the introduction -->
-
-Estimating abundance and demographic rates for wildlife populations is an integral part of basic and applied ecology [@Skalski2005, @williams2002].
+Estimating abundance and demographic rates for wildlife populations is an integral part of basic and applied ecology [@Skalski2005; @williams2002].
 Over the last few decades, tremendous progress has been made towards this end.
-This progress is partly driven the development and application of new field data collection methods and approaches, such as citizen science data (ref), camera trap data (ref) and the collection of environmental DNA data (ref).
-In addition, developments of novel statistical methods alongside decreases in computational costs now allow researchers to estimate abundance and demographic rates in situations where it was not feasible before (ref).
+This progress is partly driven the development and application of new field data collection methods and approaches, such as citizen science data [@Danielsen2022], camera trap data [@Hamel2013] and the collection of environmental DNA data [@Beng2020].
+In addition, developments of novel statistical methods alongside decreases in computational costs now allow researchers to estimate abundance and demographic rates in situations where it was not feasible before [@Zipkin2021].
 Combined, these advances put us in a much better position for estimating quantities needed for population management [@williams2002] and indices relevant for large scale policy applications, e.g.
@@
 While such methods can provide valuable information for both ecological research and management, collecting the necessary data is typically costly and logistically challenging to implement over large areas.
-Monitoring programmes focusing on abundance trends over larger areas, on the other hand, are typically based on data from unmarked animals (ref).
-A common approach for such surveys is to structure data collection around the distance sampling (DS) method.
+Monitoring programmes focusing on abundance trends over larger areas, on the other hand, are typically based on data from unmarked animals.
+One often used approach for such surveys is to structure data collection around the distance sampling (DS) method.
 DS has been used for estimating animal abundance in a wide range of contexts and for a variety of taxa [@buckland2015].
@@
 The most common implementations of DS models have long used closed-population formulations, and thus do not allow formulating a process model that projects abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates @buckland2015.
-In recent years, DS approaches have been extended in many ways, including applications that estimate changes in abundance over time in open populations via a hidden state model representing population dynamics [@moore2011; @sollmann2015].
+In recent years, DS approaches have been extended in many ways, including applications that estimate changes in abundance over time in open populations via a hidden state model representing population dynamics [@moore2011; @sollmann2015; @schaub2021].
 This has greatly extended the potential of DS approaches for studying ecological dynamics across time and space.
@@
 #### 2.1.1 Age-structured population model
-
-<!-- Note: missing ref below. Maybe good to "search" the whole text for "ref" to make sure we have replaced the placeholder with appropriate references everywhere -->
 
@@
 The willow ptarmigan has a circumpolar distribution (Fuglei et al., 2020), and lives year-round in heterogeneous alpine and artic ecosystems.
-In Norway, there has been a long-term decline in the willow ptarmigan abundance across more than a century (ref), but in the last few decades abundance trends has fluctuated both in time and space.
-In Scandinavia, willow ptarmigan is a valued game species (ref), and there have been several long-term research projects devoted to understanding how they respond to environmental variation and harvest management [@Israelsen2020; @Sandercock2011].
+In Norway, there has been a long-term decline in the willow ptarmigan abundance across more than a century [@Hjeljord2022], but in the last few decades abundance trends has fluctuated both in time and space.
+In Scandinavia, willow ptarmigan is a valued game species (see e.g.
+@Andersen2014 ), and there have been several long-term research projects devoted to understanding how they respond to environmental variation and harvest management [@Israelsen2020; @Sandercock2011].
 A key insight from across several study areas is the the annual recruitment rate (i.e.
 $R_t$ in our model, as outlined above) is highly variable, and is affected both by spring conditions [@Eriksen2023] and the abundance of small rodents, which constitute alternative prey for common predators (i.e.
-the Alternative Prey Hypothesis ref).
+the Alternative Prey Hypothesis; see @Hagen1952; @Kausrud2008; @Bowler2020).
 Adult survival show less inter-annual fluctuations [@Israelsen2020], although spatial (and potentially temporal) variation due to e.g.
@@
 As the surveys are conducted in early August, juveniles can be distinguished from adults by their smaller body size.
-<!-- maybe also mention how males and females are distinguished?
-Also, is it just body size for the juveniles or also plumage?
---> Nevertheless, mis-identification can occur, and in addition a proportion of the birds are registered as "unknown" age and/or sex.
-Since 2019, data has been collected through a mobile app tailormade for this project <!-- Add name + maybe link to download in app store?
---> .
+Males and females are manly distinguished by the sound (males often make a characteristic sound when being flushed.
+Nevertheless, mis-identification can occur, and in addition a proportion of the birds are registered as "unknown" age and/or sex.
+Since 2019, data has been collected through a mobile app tailormade for this project, which is available trhough App Store (for ios phones) and Google Play (for Android phones).
 Before 2019, field workers reported their data through a dedicated web portal.
 After data are collected and reported, they undergo several steps of quality control: first by local contacts and subsequently by personnel at the Norwegian Institute for Nature Research (NINA).
-The data are then standardized based on the Darwin-Core standard (ref), and made publicly available as a sampling-event data set published through GBIF [@Nilsen2023a].
-For additional description of the data collection procedures, see [@Bowler2020]; [@Kvasnes2018][@Nilsen2023a].
+The data are then standardized based on the Darwin-Core standard [@Wieczorek2012], and made publicly available as a sampling-event data set published through GBIF [@Nilsen2023a].
+For additional description of the data collection procedures, see [@Bowler2020; @Kvasnes2018; @Nilsen2023a].
 
-The individual longitudinal study based on radio collared willow ptarmigan was conducted in 2015-2020.
-<!-- Above you stated that radio telemetry data collection is until 2021 and ongoing.
-Until what year was it?
---> Each winter (in February-March), willow ptarmigan were located at night using snowmobiles and large hand nets with prolonged handles, as described in [@Israelsen2020].
+The individual longitudinal study based on radio collared willow ptarmigan was conducted in 2015-2021.
+Each winter (in February-March), willow ptarmigan were located at night using snowmobiles and large hand nets with prolonged handles, as described in [@Israelsen2020].
 To prevent birds from flying off before the field personnel were close enough to capture them, a high-powered head lamp was used to dazzle the birds.
@@
 300k and 40k and thereof were discarded as burn-in prior to thinning with factors 5 and 20 , leaving us with 40k and 3k posterior samples per chain (total of 120k and 12k samples per run), respectively.
+Posterior samples from the model fitted to real data are available at @Nilsen2024 (in folder PosteriorSamples_LierneCaseStudy).
 
@@
 The replicate runs for each dataset resulted in very similar posterior distributions, demonstrating that the models converged towards the same posterior distributions irrespective of starting values.
-Estimates of population sizes / densities were also adequate, and are presented in the Supplementary Materials (see folder SimCheck_byDataSet).
+Estimates of population sizes / densities were also adequate, and are presented in the supplementary materials (available at @Nilsen2024 in folder SimCheck_byDataSet).
 
@@
 Given the available data, the IDSM was not able to estimate a clear effect of small rodent abundance on ptarmigan recruitment (slope-paramater for the z-standardized rodent occurrence data = `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$Mean, 3)` ; 95% C.I.
-= `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$lower, 3)` - `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$upper, 3)`, Supplementary Figure "Rep_betaR.R.png").
+= `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$lower, 3)` - `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$upper, 3)`, see supplementary figure "Rep_betaR.R.png" in @Nilsen2024 ).
 
@@
 
-We developed an integrated model that jointly analyses line transect distance sampling survey data and data from marked individuals to estimate population abundance, survival probabilities ,and recruitment rates over time.
-We first used simulated data to examine model's ability to recover the underlying parameters when they were known.
+We developed an integrated populaiton model that jointly analyses line transect distance sampling survey data and data from marked individuals to estimate population abundance, survival probabilities, and recruitment rates over time.
+We first used simulated data to examine the model's ability to recover the underlying parameters when they were known.
 We then fitted the model to data from an ongoing field study on willow ptarmigan in Norway to showcase its applicability to real wildlife monitoring data.
 
-Open population formulations of the distance sampling model have previously been presented (refs), and applied to various ecological systems (refs).
+Open population formulations of the distance sampling model have previously been presented [@sollmann2015; @Bowler2020; @moore2011], and in these studies also applied to various ecological systems.
 The model presented here extends those previous applications by formulating the underlying population process model as a stage-structured matrix model [@caswell2000] in which the matrix elements are represented by annual survival probabilities and recruitment rates.
-While this has been the common approach for a range of other statistical modelling frameworks, including the growing suite of models falling into the category of integrated population models [@schaub2022], the integration of mechanistic population models into distance sampling frameworks is rather new.
+While this has been the common approach for a range of other statistical modelling frameworks, including the growing suite of models falling into the category of integrated population models [@schaub2021], the integration of mechanistic population models into distance sampling frameworks is rather new.
 The resulting modelling framework allow us to make maximum use of distance sampling data in combination with auxiliary information both from the distance sampling survey itself (i.e.
@@
 
-Real data, however, are likely to be subject to certain biases, and one type of bias that is likely to be common for age-structured distance sampling data arises from failure to (correctly) classify the age class of observed individuals.
-<!-- Maybe add one illustrative sentence here about how/why this happens using our ptarmigan monitoring setup as an example?
---> Observations with incorrectly classified age have the potential to introduce bias in the IDSMs relative estimates of survival and recruitment.
-This is due to the way it usese the distance sampling data to estimate survival and recruitment rates.
+Real data, however, are likely to be subject to certain biases, and such biases might result in biased parameter estimates unless accounted for.
+One type of bias that is likely to be common for age-structured distance sampling data arises from failure to (correctly) classify the age class of observed individuals.
+In our case study on willow ptarmigan in Norway, such misclassification is likely to happen at an unknown rate, even if the size difference between adult and juvenile birds are quite substantial during the survey.
+Moreover, The probability for misclassificiation might be related to both the timing of the survey (e.g.
+mid August rather than early August), it might vary between observers, and even by survey conditions.
+Observations with incorrectly classified age have the potential to introduce bias in the IDSMs relative estimates of survival and recruitment.
+This is due to the way it useses the distance sampling data to estimate survival and recruitment rates.
 In our process model, the population growth rate ($\lambda$) is determined by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, and this creates a dependence between the demographic parameters.
 If the age ratio in the data are biased or contain frequent misclassifications, this is likely to affect the relative contribution of survival and recruitment to the growth rate.
-To get an idea of the potential effect of this on parameter estimation, we checked the sensitivity of the output of the model fit to real data with regard to the treatment of birds classified as "unknown sex and age" by the field personnel (see **Methods**).
+To get an idea of the potential effect of this on parameter estimation, we checked the sensitivity of the output of the model fit to real data with regard to the treatment of birds classified as "unknown sex and age" by the field personnel (see Methods).
 In the model version presented in the results section, we made the assumption that these birds were in fact juveniles.
-<!-- Restate here why this is likely / probably a good guess --> Comparing estimates to an alternative scenario in which we discarded all birds classified as "unknown sex and age" (Appendix) we found that -- as expected -- estimated population density was virtually unaffected by the treatment of "unknown sex and age" observations, while the annual demographic parameters shifted proportional to the amount of "unknown sex and age" observations in the given year (towards higher recruitment and lower survival).
+Comparing estimates to an alternative scenario in which we discarded all birds classified as "unknown sex and age" (see supplementary figures in @Nilsen2024 ) we found that -- as expected -- estimated population density was virtually unaffected by the treatment of "unknown sex and age" observations, while the annual demographic parameters shifted proportional to the amount of "unknown sex and age" observations in the given year (towards higher recruitment and lower survival).
 Thus, biases in the reported age ratios may affect estimated of demographic rates, but not so much population density.
@@
 
-<!-- I suggest we start this next paragraph with a couple of general sentences about the density estimates, even if it is just stating that the density estimates correspond closely to what we know/have estimated with other models previously.
-Maybe also add one sentence or so about the possible biology behind the recent increase.
-Then moving on to survival and recruitment -->
-
+The density estimates that we derived from the case study on Willow ptarmigan in Norway is comparable to previous estimates from across Norway (see e.g.
+@Sandercock2011; @Kvasnes2014b).
+Throughout the study period from 2007 - 2021, the density increased markedly, but the reason for this increase is not known.
 Compared to previous studies on ptarmigan (see e.g.
@@
 One potential reason for the overall lower estimates obtained here is that our IDSM analysis assumed constant survival over the period 2007-2021 while survival, in reality, may have changed over time.
-If survival in more recent years, when telemetry data was collected [and the study of e.g.
-@@Israelsen2020 was carried out] was higher than in earlier years -- something that seems likely given population increase over recent years -- an average over the entire time period is expected to be lower.
-On a different note, we can also not exclude the possibility of a small degree of bias in survival estimates due to misclassification of age in the data (see above), especially seeing as the IDSM's recruitment rate estimates also appear somewhat high compared to other studies on ptarmigan (+refs).
-<!-- If we have any "biological" thoughts about why somewhat high recruitment rates may not be that ridiculous here, then that could be added here -->
+If survival in more recent years, when telemetry data was collected (and the study of e.g.
+@Israelsen2020 was carried out) was higher than in earlier years -- something that seems likely given population increase over recent years -- an average over the entire time period is expected to be lower.
+On a different note, we can also not exclude the possibility of a small degree of bias in survival estimates due to misclassification of age in the data (see above), especially seeing as the IDSM's recruitment rate estimates also appear somewhat high compared to other studies on willow ptarmigan [@Eriksen2023; @Kvasnes2014a; @Steen1988].
+Note however that the last period of the study period in Lierne, there was substantial increase in the density (ref Figure 2), which might indicate that the recruitment and/or survival rates were high in this period.
 
@@
 In the ptarmigan case study we thus attempted to investigate the effect of small rodent abundance (approximated as the proportion of transect lines on which rodents were reported each year) on recruitment rate.
-<!-- Deleted this because we did NOT include the rodent effect in the simulated data tests --> Unfortunately, we were not able to detect a clear effect of rodent abundance due to large uncertainty associated with the estimate (Supplementary Figure "Rep_betaR.R.png").
-This may seem somewhat surprising at first glance, given that such a pattern has been reported repeatedly in the literature (see e.g.
+Unfortunately, we were not able to detect a clear effect of rodent abundance due to large uncertainty associated with the estimate (see supplementary figures in @Nilsen2024).
+This may seem somewhat surprising given that such a pattern has been reported repeatedly in the literature (see e.g.
 @Bowler2020).
-We speculate that there are (at least) three potential explanations to this.
+We speculate that there are at least three potential (non-mutual) explanations to this result.
 The first is that our covariate data may not have been well suited for estimating effects on recruitment.
 The data on rodent abundance was heavily zero-inflated, and the annual variation in the index was rather small otherwise, making for a covariate with relatively little information content.
-While this may be partially a consequence of how these data are collected, it is also well known that the amplitude and regularity of the rodent cycles has been fading in recent decades (refs), and our study area might be no exception.
+While this may be partially a consequence of how these data are collected, it is also well known that the amplitude and regularity of the rodent cycles has been fading in recent decades [@Kausrud2008; @Cornulier2013], and our study area might be no exception.
 Lack of peak rodent years in the time series to which we fitted the model may thus also have contributed to making effect estimation challenging.
@@
 We are grateful to field workers that collected the line transect data through the Hønsefuglportalen program.
-Fjellstyrene i Lierne contributed both to the line transect program and the field work related to the marked willow ptarmigan.
+Fjellstyrene i Lierne and many colleagues from NINA and Nord University contributed both to the line transect program and the field work related to the marked willow ptarmigan.
 We also thank James A.
 Martin for constructive discussions during the later stages of this work.
-Both data collection and analytical work were financed by the Norwegian Environment Agency funded the projects (grant nrs **TO BE ADDED**).
+
+### Funding
+
+Both data collection and analytical work were financed by the Norwegian Environment Agency funded the projects (grant nrs 17010522, 19047014 and 22047004).
+
+### Conflict of interest disclosure
+
+The authors declare that they comply with the PCI rule of having no financial conflicts of interest in relation to the content of the article
 
 ### Data and code availability
+
+The raw data from the line transect surveys is deposited on GBIF and can be accessed freely via the Living Norway Data Portal (https://data.livingnorway.no).
+The work here is based on version 1.7 of the dataset "Tetraonid line transect surveys from Norway: Data from Fjellstyrene" [@Nilsen2023a].
+
+The auxiliary radio-telemetry data and rodent occupancy data, and all code used for wrangling, analysing, and visualizing data and results can be found in the project's repository on GitHub: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.
+The results presented in this paper were created using version 1.3 of the code [@IDSMCode_2024].
+
+Supplementary figures and posterior samples from the model run on the real data are available as a time-stamped open archived on Open Science Framework [@Nilsen2024].
+
+## References

### 2024-01-08 R/expandSeed_MCMC.R
Updated iseed expansion

@@ -1,16 +1,15 @@
 expandSeed_MCMC <- function(seed, nchains){
   
-  if(nchains > 6){
-    stop("Seed expansion is currently only supported for up to 6 chains.")
+  if(nchains > 5){
+    stop("Seed expansion is currently only supported for up to 5 chains.")
   }
   
   seed.list <- c(
     seed,
     round((seed*70)/4),
-    round(seed/3)+1,
-    seed + 88,
+    (seed + 10)*7,
     round(seed*5/3),
-    seed + 37
+    seed + 37*2
   )
   
   return(seed.list[1:nchains])

### 2024-01-08 Analysis_RealData.R
Changed origin seed

@@ -8,7 +8,7 @@ library(nimble)
 #-------#
 
 ## Set seed
-mySeed <- 37
+mySeed <- 32
 set.seed(mySeed)
 
 ## Source all functions in "R" folder

### 2024-01-08 _targets.R

@@ -11,7 +11,7 @@ library(tmap)
 library(parallel)
 
 ## Set seed
-mySeed <- 37
+mySeed <- 32
 set.seed(mySeed)
 
 ## Source all functions in "R" folder

### 2024-01-08 R/setupModel.R
Fixed how seed is set for simulating initial values

@@ -58,7 +58,7 @@ setupModel <- function(modelCode, customDist,
   }
   
   ## Simulate initial values
-  set.seed(initVals.seed)
+  #set.seed(initVals.seed)
   initVals <- list()
   for(c in 1:nchains){
     
@@ -75,7 +75,8 @@ setupModel <- function(modelCode, customDist,
                                      nim.constants = nim.constants, 
                                      R_perF = R_perF, 
                                      survVarT = survVarT,
-                                     fitRodentCov = fitRodentCov)
+                                     fitRodentCov = fitRodentCov,
+                                     initVals.seed = initVals.seed[c])
     }
 
   }

### 2024-01-08 R/simulateInits.R

@@ -13,7 +13,9 @@
 #'
 #' @examples
 
-simulateInits <- function(nim.data, nim.constants, R_perF, survVarT, fitRodentCov){
+simulateInits <- function(nim.data, nim.constants, R_perF, survVarT, fitRodentCov, initVals.seed){
+  
+  set.seed(initVals.seed)
   
   # Limits and constants #
   #----------------------#

### 2024-01-08 R/simulateInits.R
Updated initial value simulation parameters for hyper S

@@ -57,8 +57,8 @@ simulateInits <- function(nim.data, nim.constants, R_perF, survVarT, fitRodentCo
   #-------------#
   
   ## Area-specific survival parameters
-  h.Mu.S <- runif(1, 0.3, 0.5) 
-  h.sigma.S <- runif(1, 0.05, 0.25)
+  h.Mu.S <- runif(1, 0.40, 0.45) 
+  h.sigma.S <- runif(1, 0.05, 0.2)
   Mu.S1 <- runif(1, 0.5, 0.7)
   
   mu.S <- EnvStats::rnormTrunc(N_areas, qlogis(h.Mu.S), sd = h.sigma.S, max = qlogis(Mu.S1))

### 2024-01-08 Analysis_RealData.R
Added global definition of nchains & separate seeds for each MCMC

@@ -11,6 +11,9 @@ library(nimble)
 mySeed <- 32
 set.seed(mySeed)
 
+## Set number of chains
+nchains <- 5
+
 ## Source all functions in "R" folder
 sourceDir <- function(path, trace = TRUE, ...) {
   for (nm in list.files(path, pattern = "[.][RrSsQq]$")) {
@@ -125,6 +128,9 @@ input_data <- prepareInputData(d_trans = LT_data$d_trans,
 ## Write model code
 modelCode <- writeModelCode(survVarT = survVarT)
 
+## Expand seed to get MCMC seeds
+MCMC.seeds <- expandSeed_MCMC(seed = mySeed, nchains = nchains)
+
 ## Setup for model using nimbleDistance::dHN
 model_setup <- setupModel(modelCode = modelCode,
                           R_perF = R_perF,
@@ -132,13 +138,9 @@ model_setup <- setupModel(modelCode = modelCode,
                           fitRodentCov = fitRodentCov,
                           nim.data = input_data$nim.data,
                           nim.constants = input_data$nim.constants,
-                          testRun = TRUE, 
-                          nchains = 4,
-                          initVals.seed = mySeed)
-
-
-## Expand seed to get MCMC seeds
-MCMC.seeds <- expandSeed_MCMC(seed = mySeed, nchains = model_setup$mcmcParams$nchains)
+                          testRun = FALSE, 
+                          nchains = nchains,
+                          initVals.seed = MCMC.seeds)
 
 
 # MODEL (TEST) RUN #

### 2024-01-08 _targets.R

@@ -14,6 +14,9 @@ library(parallel)
 mySeed <- 32
 set.seed(mySeed)
 
+## Set number of chains
+nchains <- 5
+
 ## Source all functions in "R" folder
 sourceDir <- function(path, trace = TRUE, ...) {
   for (nm in list.files(path, pattern = "[.][RrSsQq]$")) {
@@ -116,6 +119,13 @@ list(
     writeModelCode(survVarT = survVarT)
   ),
   
+  
+  tar_target(
+    MCMC.seeds,
+    expandSeed_MCMC(seed = mySeed, 
+                    nchains = nchains)
+  ),
+  
   tar_target(
     model_setup,
     setupModel(modelCode = modelCode,
@@ -124,15 +134,9 @@ list(
                nim.constants = input_data$nim.constants,
                survVarT = survVarT,
                fitRodentCov = fitRodentCov,
-               testRun = TRUE, 
-               nchains = 4,
-               initVals.seed = mySeed)
-  ),
-  
-  tar_target(
-    MCMC.seeds,
-    expandSeed_MCMC(seed = mySeed, 
-                    nchains = model_setup$mcmcParams$nchains)
+               testRun = FALSE, 
+               nchains = nchains,
+               initVals.seed = MCMC.seeds)
   ),
   
   tar_target(

### 2024-01-08 R/setupModel.R
Increased number of iterations

@@ -28,7 +28,7 @@
 setupModel <- function(modelCode, customDist,
                        nim.data, nim.constants,
                        R_perF, survVarT, fitRodentCov, addDummyDim = TRUE,
-                       niter = 100000, nthin = 20, nburn = 40000, nchains = 3,
+                       niter = 150000, nthin = 25, nburn = 75000, nchains = 3,
                        testRun = FALSE, initVals.seed){
   
   require('nimble')

### 2024-01-08 Manuscript_Nater_etal.qmd
Small edits

--- 2023-12-31 Manuscript_Nater_etal.qmd
@@
 the collection of ecological data by members of the public [also called citizen science, @fraisl2022], allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias [@johnston2023].
-Consequently, many large-scale monitoring programmes are often limited to presence-absence or very simple count observations, making them suitable for the development of indicators of distribution and population trends, but usually not for indicators of abundance, population dynamics, and demographic rates (+ example refs).
+Consequently, many large-scale monitoring programmes are often limited to presence-absence or very simple count observations, making them suitable for the development of indicators of species distributions and perhaps population trends, but usually not of abundance, population dynamics, and demographic rates (+ example refs).
 The exception here are monitoring programmes that succeed in making use of a large number of citizen scientists that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol.
@@
 Only during the most recent four years (2018-2021) has data been collected regularly for all reporting areas included in the analyses.
-During this period, estimated population densities varied between `r dens.min$post.sum` birds/km\^2 in the area "`r dens.min$area.name`" close to the coast in central Norway to `r dens.max$post.sum` birds/km\^2 in "`r dens.max$area.name`" further south near the Swedish border.
+During this period, estimated population densities varied between `r dens.min$post.sum` birds/km^2^ in the area "`r dens.min$area.name`" close to the coast in central Norway to `r dens.max$post.sum` birds/km^2^ in "`r dens.max$area.name`" further south near the Swedish border.
 In general, population density appeared to be lowest in the Northern parts of Norway and highest in the eastern part of central Norway (@fig-densA).
@@
 Annual survival probabilities ranged from `r S.min$post.sum` (area "`r S.min$area.name`") to `r S.max$post.sum` (area "`r S.max$area.name`") across reporting areas in Norway, with the highest values occurring in the far north and in the mountains in the south (@fig-pVRA).
-The global average survival probability across all areas and years ($\mu^S$) was estimated at `r S.hyper.mean`.
+The global average survival probability across all areas and years ($\mu^S$) was estimated at `r S.hyper.mean$post.sum`.
 Spatial variation in survival (random effect SD = `r S.sd.spatial$post.sum`) was relatively low compared to temporal (`r S.sd.temporal$post.sum`) and residual (`r S.sd.residual$post.sum`) variation.
 
-Recruitment rates varied between `r R.min$post.sum` (area "`r R.min$area.name`") and `r R.max$post.sum` (area "`r R.max$area.name`") and displayed a spatial pattern opposite to that 
-of annual survival, i.e.
+Recruitment rates varied between `r R.min$post.sum` (area "`r R.min$area.name`") and `r R.max$post.sum` (area "`r R.max$area.name`") and displayed a spatial pattern opposite to that of annual survival, i.e.
 lower recruitment rates co-occurring with higher survival rates and vice-versa (@fig-pVRB; @fig-paramCorrs B).
@@
 Nonetheless, spatial variation in the rodent effect was substantial (random effect SD = `r betaR.sd.spatial$post.sum`).
-This resulted in negative median effects in `r n.neg.betaR` areas, positive median effects in `r n.pos.betaR` areas, and a range of effect sizes from `r betaR.min$post.sum` (area "`r betaR.min$area.name`") to `r betaR.max$post.sum` (area "`r betaR.max$area.name`", @fig-paramCorrs C).
+This resulted in negative median effects in `r n.neg.betaR` areas, positive median effects in `r n.pos.betaR` areas, and a range of effect sizes from `r betaR.min$post.sum` (area "`r betaR.min$area.name`") to `r betaR.max$post.sum` (area "`r betaR.max$area.name`", @fig-paramCorrs C, SF "Rep_betaR.R.pdf").
 The largest positive rodent effects were estimated for areas in the very North of Norway, as well as in the mountaineous regions in the central and southwestern parts of the country (SF "betaR_Map.pdf").
@@
 
+## Reproducible workflows for a sustainable future
+
+-   Structure / setup of this workflow (from data collection to analytical results)
+
+-   Why this matters: transparency and credibility, feedback to volunteer data collectors, cost-effective to re-do analyses for annual/regular reporting, easy to re-use and build upon, publicly accessible and well documented to allow more effective resource use in follow-up projects (sustainable use of funding), etc.
+
+-   Also: useful to link up to different reporting routes and/or more general (multi-indicator) projects <!-- @EVERYONE: Do we know of some good references for this?
+PRseumably we cannot cite our chapter in the Technical Manual yet :-P -->
+
 -   Outlook: integration with / coupling to models for harvest to develop tools for (automatic) calculation of e.g.
@@
 -->
-
-## Reproducible workflows for a sustainable future
-
--   Structure / setup of this workflow (from data collection to analytical results)
--   Why this matters: transparency and credibility, feedback to volunteer data collectors, cost-effective to re-do analyses for annual/regular reporting, easy to re-use and build upon, publicly accessible and well documented to allow more effective resource use in follow-up projects (sustainable use of funding), etc.
--   Also: useful to link up to different reporting routes and/or more general (multi-indicator) projects <!-- @EVERYONE: Do we know of some good references for this?
-PRseumably we cannot cite our chapter in the Technical Manual yet :-P -->
 

### 2024-01-08 Manuscript_Nater_etal.qmd

@@
 Such indicators ideally represent a wide range of biodiversity's states and functions [e.g.
-Essential Biodiversity Variables, @pereira2013; @jetz2019], yet the development of suitable indicators for certain qualities, such as species abundance and demography, has been more difficult than for others [@schmeller2018; @waldock2022].
+Essential Biodiversity Variables, @pereira2013; @jetz2019], yet the development of suitable indicators for certain attributes, such as species abundance and demography, has been more difficult than for others [@schmeller2018; @waldock2022].
 This is at least partially due to challenging requirements regarding spatial scales of useful biodiversity indicators.
@@
 Consequently, many countries have been working on setting up, maintaining, and improving such monitoring programmes over the last decades.
-Many now well-established programmes focus on breeding birds and butterflies, and examples include the North American Breeding Bird Survey (<https://www.pwrc.usgs.gov/bbs/>), UK Butterfly Monitoring Scheme (<https://ukbms.org/>), Swiss Biodiversity Monitoring (<https://www.biodiversitymonitoring.ch/>), and many more.
+Many now well-established programmes focus on breeding birds and butterflies, and examples include the North American Breeding Bird Survey (<https://www.pwrc.usgs.gov/bbs/>), UK Butterfly Monitoring Scheme (<https://ukbms.org/>), the Game and Wildlife Conservation Trust Partridge Count Scheme [@aebischer2010], and the Swiss Biodiversity Monitoring (<https://www.biodiversitymonitoring.ch/>).
 <!-- is there any other one we want to name here specifically?
@@
 The exception here are monitoring programmes that succeed in making use of a large number of citizen scientists that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol.
+For example, in the United States hunters participate in the collection of bands and wings from harvested American Woodcock (*Scopolax minor)* to estimate survival and age ratios [@zimmerman2010].
 <!-- can we think of 1-2 internationally well-known examples?
@@
 In addition, the local study designs are reviewed by staff at NINA, and common recommendations for study design are provided.
-However, due to the fact that participation by stakeholders is voluntary, the spatial distribution of transects and sampling effort is not homogeneous.
+However, because participation by stakeholders is voluntary, the spatial distribution of transects and sampling effort is not homogeneous.
 In general, the effort is higher in South-Eastern and Central Norway, medium in Northern Norway, and low in Western and Southern Norway.
@@
 ## Implications for management
-
 <!-- preliminary section title) -->

### 2024-01-08 Manuscript_Nater_etal.qmd
Added more detailed bullet point structure for discussion

@@
 
--   Biological results (interpretations and comparisons with other case studies / species / systems)
-
-[@Wilson_Martin_2011] Found x y z
-
--   With some focus on what we are able to get insights on thanks to the integration across time and space (i.e.
-better estimates through data sharing, relative importance of different kinds of variation, etc.
-) <!-- @ERLEND & @JAMES, once the "new" results are in: thoughts on what we see that is interesting, cool, expected, unexpected etc.
+<!-- @ERLEND & @JAMES, once the "new" results are in: thoughts on what we see that is interesting, cool, expected, unexpected etc.
 Can we compare to other large scale studies, either on other grouse-like creatures or harvested species more generally?
 -->
+
+<!-- Spatial perspective -->
+
+-   2-55 birds/km^2^, lowest in North, highest in eastern central Norway (close to Swedish border)
+
+-   Most uncertainty in density estimates in areas with relatively low population sizes, low number of transects, and recent start of data collection
+
+-   The model also does "hindcasts", i.e.
+predictions for years prior to data collection, for areas where data collection started after 2007.
+But uncertainty increases quickly, so no point looking \> 2 years back.
+
+-   Over half of areas have population increases in recent years.
+
+-   Recent declines: mountains in central Norway, typically following increases up to around 2016
+
+-   Across populations, no clear relationship between density and growth rates (within populations?)
+
+-   There is a spatial association lower density / lower reproduction / higher survival ("slower life history"?) versus higher density / higher reproduction / lower survival ("faster life history?")
+
+-   Survival recruitment trade-off: high survival and low recruitment in far north and southern mountains
 
@@
 
+<!-- Temporal perspective -->
+
+-   There are some years with relatively high densities in most areas: 2011, 2013, 2018
+
+-   These same years are also characterized by high recruitment, and subsequent low survival, and the match up with the years with big rodent peaks in many areas
+
+-   Many rodents -\> predator release -\> higher recruitment -\> higher densities -\> negative density feedbacks on survival -\> lower survival -\> low density next year?
+
+-   Also: many rodents -\> good year for predators -\> high predation in following season/year -\> lower survival -\> low density next year?
+
+-   This seems to support that the rodent abundance is an important driver -\> supported by the model: rodent effects estimated positive or inconclusive across areas
+
+-   This is well established (for Norwegian ptarmigan and elsewhere?) and was expected
+
+-   BUT: the model was unable to recover the positive relationship between rodent abundance and recruitment when applied only to data from a single location.
+But with this multi-area model, estimation and inference is much improved -\> space-for-time substitution works!
+
+-   This also means we have potential for including other covariates
+
+-   Examples of what else could matter and why
+
+<!-- Variation beyond rodents -->
+
+-   Including more and/or other covariates may also help further explaining vital rate variation, not least because for both survival and recruitment, residual variation was the largest component.
+
+-   However, looking at spatial vs.
+temporal variation, we found very different patterns: for recruitment, we have more spatial than temporal variation, but they are not vastly different.
+For survival, spatial variation was quite low compared to the rather large temporal variation
+
+-   Some hypotheses about large- vs. local-scale drivers?
+
+[@Wilson_Martin_2011] Found x y z
+
 ## Implications for management
+
 <!-- preliminary section title) -->
@@
 
+-   Management also includes optimization of monitoring
+
+-   Detection was less variable across time and space than vital rates, but we do see it tends to be higher in the southern half of the country
+
+-   Discuss this related to landscape features and/or observer pool? Thoughts on what could be done to improve, potentially?
+
 ## Model limitations and outlook
@@
 -   Some limitations of "base model" that need mentioning (e.g. age assignment, ignoring potential effects of flock size on detection)
-
--   Limited usefulness of auxiliary radio-telemetry data as only from one area (big potential for including additional datasets from other small-scale studies, telemetry but also other, e.g.
-nest monitoring?) <!-- @ERLEND: Any thoughts on what the status here is?
-Might there be other data available from beyond Lierne?
-Are there exciting future prospects e.g.
-linked to automated monitoring with cameras at a larger scale?
--->
 
@@
 
+-   Limited usefulness of auxiliary radio-telemetry data as only from one area (big potential for including additional datasets from other small-scale studies, telemetry but also other, e.g.
+nest monitoring?) <!-- @ERLEND: Any thoughts on what the status here is?
+Might there be other data available from beyond Lierne?
+Are there exciting future prospects e.g.
+linked to automated monitoring with cameras/sound boxes at a larger scale?
+-->
+
+-   Outlook: integration with / coupling to models for harvest to develop tools for (automatic) calculation of e.g.
+hunting quotas <!-- @JAMES: This one is for you.
+Some ruminations on how and why it would be great to formally model harvest, integration with hunting data, maybe eventually develop a tool to calculate quotas (I know this has been developed for other game species), etc.
+-->
+
 ## Reproducible workflows for a sustainable future
@@
 PRseumably we cannot cite our chapter in the Technical Manual yet :-P -->
-
--   Outlook: integration with / coupling to models for harvest to develop tools for (automatic) calculation of e.g.
-hunting quotas <!-- @JAMES: This one is for you.
-Some ruminations on how and why it would be great to formally model harvest, integration with hunting data, maybe eventually develop a tool to calculate quotas (I know this has been developed for other game species), etc.
--->
 

### 2024-01-09 Manuscript_Nater_etal.qmd
Updated MCMC parameters in text

@@
 We simulated complete sets of initial values for all model nodes prior to model running and using pre-defined seeds to ensure reproducibility.
-Using NIMBLE's standard samplers, we then ran 4 MCMC chains of 100k iterations each.
-We discarded the first 40k samples of each chain as burn-in, and thinned the remainder by a factor 20, resulting in a final joint posterior containing a total of 4 x 3k = 120k samples (note that high thinning rates were necessary to constrain memory load of the joint posterior, which included 314568 monitored parameters).
+Using NIMBLE's standard samplers, we then ran 5 MCMC chains of 150k iterations each.
+We discarded the first 75k samples of each chain as burn-in, and thinned the remainder by a factor 25, resulting in a final joint posterior containing a total of 5 x 3k = 15k samples (note that high thinning rates were necessary to constrain memory load of the joint posterior, which included 314568 monitored parameters).
 

### 2024-01-09 Manuscript_Nater_etal.qmd
updated disc and bib

@@
 @robinson2014; @horswill2019].
-This is variation that needs to be accounted for in order to develop successful and sustainable strategies for area use, (harvest) management, and species and biodiversity conservation [@williams2002].
+This variation that needs to be accounted for to develop successful and sustainable strategies for area use, (harvest) management, and species and biodiversity conservation [@williams2002].
 
@@
 There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly and requires trained specialists, thus limiting the amount of data that can be collected, while participatory monitoring, i.e.
-the collection of ecological data by members of the public [also called citizen science, @fraisl2022], allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias [@johnston2023].
+the collection of ecological data by members of the public [also called citizen or community science, @fraisl2022], allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias [@johnston2023].
 Consequently, many large-scale monitoring programmes are often limited to presence-absence or very simple count observations, making them suitable for the development of indicators of species distributions and perhaps population trends, but usually not of abundance, population dynamics, and demographic rates (+ example refs).
@@
 @bowler2020; @nilsen2020] but large-scale estimation of demographic rates underlying these trends has thus far remained as untapped potential of the dataset.
-@nilsen2023 recently developed a novel integrated distance sampling model (IDSM) which successfully uses age information in line transect data alongside radio-telemetry data to estimate not just abundance, but also survival and recruitment across years.
+@nilsen2023 recently developed a novel integrated distance sampling model (IDSM) which successfully uses the age of idividuals detected along line transects data coupled with radio-telemetry data to estimate not just abundance, but also survival and recruitment across years.
 In this study, we adapt the model of @nilsen2023 to run not just on a single site but on all areas with publicly available line transect data from "Hønsefuglportalen" simultaneously.
@@
 We then apply the resulting multi-area IDSM to "Hønsefuglportalen" data on willow ptarmigan to estimate population size, age-structure, survival, recruitment, and impacts of small rodent abundance across 41 reporting districts and 15 years (2007-2021) for this popular small-game species.
-We further embed the modelling workflow in a reproducible, semi-automated pipeline that will greatly facilitate the repeated calculation of abundance and population indicators at different spatial scales as new data is added every year.
+We further embed the modelling workflow in a reproducible, semi-automated pipeline that will greatly facilitate the repeated calculation of abundance and population indicators at different spatial scales as new data are added every year.
 
@@
 
-The willow ptarmigan (*Lagopus lagopus*) is a tetraoid bird with a circumpolar distribution, mainly inhabiting sub-alpine and arctic ecosystems [see e.g.
+The willow ptarmigan is a tetraoid bird with a circumpolar distribution, mainly inhabiting sub-alpine and arctic ecosystems [see e.g.
 @fuglei2020].
@@
 Their population dynamics are characterized by large inter-annual fluctuations in abundance [@hjeljord2022], and previous research has tied these fluctuations to rodent cycle through shared predators [@hagen1952; @Bowler2020].
-This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis and has been a central part of Fennoscandian research on grouse population dynamics for many decades [@hagen1952; @Linden1988: @Steen1985].
-In addition, spring weather conditions and phenology is known to have considerable effects on breeding success and recruitment rates [@Eriksen2023, @Steen1985].
+This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis and has been a central part of Fennoscandian research on grouse population dynamics for many decades [@hagen1952; @Linden1988; @Steen1985].
+In addition, spring weather conditions and phenology is known to have considerable effects on breeding success and recruitment rates [@Eriksen2023; @Steen1985].
 Across their distributional range, they are iconic species with a high cultural value, partly linked to their popularity as game species.
@@
 
-In this study we used all publicly available data for the period 2007-2021, which included a total of 2225 transects in 41 different reporting districts spanning 9 counties and 50 municipalities.
-Transects on which no willow ptarmigan were ever observed (i.e.
-species absence likely due to low habitat suitability) were dropped, resulting in 2077 transects used in the analyses.
-
 ### Radio-telemetry study in Lierne
@@
 The radio-telemetry study is described in detail in @Israelsen2020 and in @Arnekleiv2022.
-
-In this study we used data from years 2015 - 2020, and the total sample size across these years was 139 birds for the Aug-Jan period and 258 birds for the Feb-Jul period.
 
@@
 
+<!-- move this here but can go either way on location. -->
+
+We used all publicly available data for the period 2007-2021, which included a total of 2225 transects in 41 different reporting districts spanning 9 counties and 50 municipalities.
+Transects on which no willow ptarmigan were ever observed (i.e.
+species absence likely due to low habitat suitability) were dropped, resulting in 2077 transects used in the analyses.
+For the radio-telemetry aspect of the model, we used data from years 2015 - 2020, and the total sample size across these years was 139 birds for the Aug-Jan period and 258 birds for the Feb-Jul period.
+
 ## Population density
@@
 
-The highest population growth rates were estimated for areas with relatively low population densities across latitudes [@fig-paramCorrs A] but we did not find evidence for a strong association between population growth rates and population densities overall.
+The highest population growth rates were estimated for areas with relatively low population densities across latitudes but we did not find evidence for a strong association between population growth rates and population densities overall ([@fig-paramCorrs] A).
 
@@
 
-Recruitment rates varied between `r R.min$post.sum` (area "`r R.min$area.name`") and `r R.max$post.sum` (area "`r R.max$area.name`") and displayed a spatial pattern opposite to that of annual survival, i.e.
-lower recruitment rates co-occurring with higher survival rates and vice-versa (@fig-pVRB; @fig-paramCorrs B).
+Recruitment rates varied between `r R.min$post.sum` (area "`r R.min$area.name`") and `r R.max$post.sum` (area "`r R.max$area.name`") and displayed a spatial pattern opposite to that of annual survival, i.e., lower recruitment rates co-occurring with higher survival rates and vice-versa (@fig-pVRB; @fig-paramCorrs B).
 Across all areas and years, average recruitment rate was `r R.hyper.mean$post.sum`.
@@
 Detection decay parameters, which determine detection probability in distance sampling surveys, varied across areas from between `r Det.min$post.sum` in "`r Det.min$area.name`" to `r Det.max$post.sum` in "`r Det.max$area.name`, resulting in detection probabilities over the transect sites (truncation distance = 200 m) from `r pDet.min$post.sum` to `r pDet.max$post.sum`, respectively.
-The global average detection decay was `r Det.hyper.mean$post.sum` (detection probability = `r pDet.hyper.mean$post.sum`), and in general, higher values were more common in the Southern half of the country than the Northern half (SF "Avg_detect_Map.pdf").
+The global average detection decay was `r Det.hyper.mean$post.sum` (detection probability = `r pDet.hyper.mean$post.sum`), and in general, higher values were more common in the Southern half of the country than the Northern half (SF"Avg_detect_Map.pdf").
 Variation in detection over time was modest (SF "TimeSeries_pDetect.pdf"), and among-year variation in detection decay (random effect log SD = `r Det.sd.temporal$post.sum`) was almost identical in magnitude to spatial (`r Det.sd.spatial$post.sum`) and residual (`r Det.sd.residual$post.sum`) variation.
@@
 
+Building on the work of @nilsen2023, we used a structured community science monitoring program and a novel integrated population model to estimate spatial and temporal variation in demography of a culturally important game bird species.
+Although our study was exploratory in nature, the patterns observed generally conformed to those expected from ecological and life history theory.
+Apparent trade-offs between survival and recruitment was observed.
+Support was also found for the benefits of alternative prey to ptarmigan (i.e., alternative prey hypothesis).
+A muted relationship between population density and growth rates was found, however, this was possibly due to omitted environmental gradients that would have better described the availability of resources across the large spatial extent of the study.
+
 ## Abundance and vital rates across space and time
@@
 <!-- Spatial perspective -->
+
+The rich spatial distribution of sampling afforded us the opportunity to explore variation in density and vital rates across a relatively large spatial extent.
+Density followed the expected pattern of lower density at northern latitudes and western longitudes.
+Within the spatial extent of our study, these areas experience the most extreme climatic conditions including coldest temperatures and shortest growing season.
+Ptarmigan are income breeders using food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005].
+Reduced primary production limits food availability which limits carrying capacity relative to more southern populations.
+
+We did not find a clear relationship between density and growth rates.
+However, this is unsurprising as it appears to be substantial variation in demography which suggests carrying capacity varies over space.
+This notion is further supported by the residual variation unaccounted for in the model which is likely environmental conditions that govern resource availability.
+
+[@Wilson_Martin_2011] found greater fecundity at greater latitudes compared to the southern terminus of their distribution in the USA; however, annual survival was less.
 
@@
 <!-- Temporal perspective -->
+
+The Alternative Prey Hypothesis (APH) has been well-supported throughout the literature across multiple taxa [e.g., @hagen1952; @Kjellander_Nordström_2003; @Reif_Tornberg_Jungell_Korpimäki_2001] and for willow ptarmigan [@bowler2020].
+We also found support for APH but the positive effect of rodents on recruitment were more pronounced at northern latitudes while equivocal elsewhere.
+Predator-prey dynamics are notoriously complex and interact with a multitude of environmental variables.
+@Nielsen_1999 found that gyrfalcon predation on ptarmigan was greatest during years of low ptarmigan abundance which suggests that the availability of alternative prey is most important at low densities as exhibited at northern latitudes in our study.
 
@@
 
--   Management also includes optimization of monitoring
+Management decisions rarely and shouldn't occur at the resolution of large geo-political boundaries (e.g., Norway) because of the variation in demographic processes that occur.
+Our study suggests decisions should be spatially refined to capture "local" population processes that will interact with management.
+
+Management also includes optimization of monitoring
 

### 2024-01-10 Manuscript_Nater_etal.qmd
Moved info on data used back into methods section

@@
 
+In this study we used all publicly available data for the period 2007-2021, which included a total of 2225 transects in 41 different reporting districts spanning 9 counties and 50 municipalities.
+Transects on which no willow ptarmigan were ever observed (i.e.
+species absence likely due to low habitat suitability) were dropped, resulting in 2077 transects used in the analyses.
+
 ### Radio-telemetry study in Lierne
@@
 The radio-telemetry study is described in detail in @Israelsen2020 and in @Arnekleiv2022.
+
+In this study we used data from years 2015 - 2020, and the total sample size across these years was 139 birds for the Aug-Jan period and 258 birds for the Feb-Jul period.
 
@@
 Supplementary figures (SF) are provided as .pdf files with captions in "SuppFigures_Captions.txt".
-
-<!-- move this here but can go either way on location. -->
-
-We used all publicly available data for the period 2007-2021, which included a total of 2225 transects in 41 different reporting districts spanning 9 counties and 50 municipalities.
-Transects on which no willow ptarmigan were ever observed (i.e.
-species absence likely due to low habitat suitability) were dropped, resulting in 2077 transects used in the analyses.
-For the radio-telemetry aspect of the model, we used data from years 2015 - 2020, and the total sample size across these years was 139 birds for the Aug-Jan period and 258 birds for the Feb-Jul period.
 

### 2024-01-10 Manuscript_Nater_etal.qmd
Fixed spacing in MS and .bib

@@
 Detection decay parameters, which determine detection probability in distance sampling surveys, varied across areas from between `r Det.min$post.sum` in "`r Det.min$area.name`" to `r Det.max$post.sum` in "`r Det.max$area.name`, resulting in detection probabilities over the transect sites (truncation distance = 200 m) from `r pDet.min$post.sum` to `r pDet.max$post.sum`, respectively.
-The global average detection decay was `r Det.hyper.mean$post.sum` (detection probability = `r pDet.hyper.mean$post.sum`), and in general, higher values were more common in the Southern half of the country than the Northern half (SF"Avg_detect_Map.pdf").
+The global average detection decay was `r Det.hyper.mean$post.sum` (detection probability = `r pDet.hyper.mean$post.sum`), and in general, higher values were more common in the Southern half of the country than the Northern half (SF "Avg_detect_Map.pdf").
 Variation in detection over time was modest (SF "TimeSeries_pDetect.pdf"), and among-year variation in detection decay (random effect log SD = `r Det.sd.temporal$post.sum`) was almost identical in magnitude to spatial (`r Det.sd.spatial$post.sum`) and residual (`r Det.sd.residual$post.sum`) variation.

### 2024-01-10 Manuscript_Nater_etal.qmd
Edited new "overview" paragraph
I note that this paragraph has "abstract vibes" or perhaps "conclusion" vibes to me... I may move it later.

@@
 
-Building on the work of @nilsen2023, we used a structured community science monitoring program and a novel integrated population model to estimate spatial and temporal variation in demography of a culturally important game bird species.
-Although our study was exploratory in nature, the patterns observed generally conformed to those expected from ecological and life history theory.
-Apparent trade-offs between survival and recruitment was observed.
-Support was also found for the benefits of alternative prey to ptarmigan (i.e., alternative prey hypothesis).
-A muted relationship between population density and growth rates was found, however, this was possibly due to omitted environmental gradients that would have better described the availability of resources across the large spatial extent of the study.
+Building on the work of @nilsen2023, we applied a novel integrated population model to data collected through a national-scale participatory monitoring programme to estimate spatial and temporal variation in demography of a culturally important game bird species, the willow ptarmigan.
+While our study was exploratory in nature, it recovered patterns consistent with ecological and life-history theory including trade-offs between survival and recruitment, and a tendency towards slower life histories at higher latitudes and altitudes.
+Space-fot-time substitution also provided the statistical power necessary for the analysis to provide evidence for the alternative prey hypothesis, i.e.
+ptarmigans benefitting from high abundance of alternative prey for their predators.
 

### 2024-01-10 Manuscript_Nater_etal.qmd
Completed "skeleton text" for first chapter of Discussion
Much (biology, especially) remains to be added. See HTML comments.

@@
 While our study was exploratory in nature, it recovered patterns consistent with ecological and life-history theory including trade-offs between survival and recruitment, and a tendency towards slower life histories at higher latitudes and altitudes.
-Space-fot-time substitution also provided the statistical power necessary for the analysis to provide evidence for the alternative prey hypothesis, i.e.
-ptarmigans benefitting from high abundance of alternative prey for their predators.
+Space-for-time substitution also provided the statistical power necessary for the analysis to provide evidence for the alternative prey hypothesis, i.e.
+ptarmigans benefiting from high abundance of alternative prey for their predators.
+<!-- To me, this little paragraph has "abstract vibes", or maybe "conclusion vibes".
+I'll leave it here for now but note that it may get moved eventually.
 
@@
 
-<!-- @ERLEND & @JAMES, once the "new" results are in: thoughts on what we see that is interesting, cool, expected, unexpected etc.
+<!-- @ERLEND & @JAMES thoughts on what we see that is interesting, cool, expected, unexpected etc.
 Can we compare to other large scale studies, either on other grouse-like creatures or harvested species more generally?
@@
 
-The rich spatial distribution of sampling afforded us the opportunity to explore variation in density and vital rates across a relatively large spatial extent.
-Density followed the expected pattern of lower density at northern latitudes and western longitudes.
-Within the spatial extent of our study, these areas experience the most extreme climatic conditions including coldest temperatures and shortest growing season.
-Ptarmigan are income breeders using food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005].
-Reduced primary production limits food availability which limits carrying capacity relative to more southern populations.
+The wide spatial distribution of the line transect monitoring afforded us the opportunity to explore variation in population density and vital rates across a relatively large spatial extent.
+Ptarmigan densities across the 41 reporting districts included in our analyses varied from around 2 birds/km^2^ to 55 birds/km^2^, with the lowest densities occurring far north in the country, as well as on the west coast and in the mountains in central Norway [@fig-densA].
+These areas are characterized by more extreme climatic conditions, boasting cold temperatures and – in the North – a short growing season.
+Resulting reduced primary production limits food availability and as ptarmigan are income breeders using food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005], lower carrying capacity in such areas is to be expected.
+<!-- @ERLEND do we have some more thoughts to add here?
+-->
 
-We did not find a clear relationship between density and growth rates.
-However, this is unsurprising as it appears to be substantial variation in demography which suggests carrying capacity varies over space.
-This notion is further supported by the residual variation unaccounted for in the model which is likely environmental conditions that govern resource availability.
+We found increasing population trends over recent years in over half of the reporting districts [@fig-lamA].
+<!-- relate this to changes in hunting management in recent years?
+Any other factors we think could have contributed?
+--> Nonetheless, recent population declines were also evident in some areas, and particularly in the mountains in central Norway [@fig-lamA].
+<!-- What could be going on here?
+It's also interesting that while some of these areas have declines over the most recent years, that usually follows on a period of increase, or at least stability, up to around 2016 -->
 
-[@Wilson_Martin_2011] found greater fecundity at greater latitudes compared to the southern terminus of their distribution in the USA; however, annual survival was less.
-
--   2-55 birds/km^2^, lowest in North, highest in eastern central Norway (close to Swedish border)
-
--   Most uncertainty in density estimates in areas with relatively low population sizes, low number of transects, and recent start of data collection
-
--   The model also does "hindcasts", i.e.
-predictions for years prior to data collection, for areas where data collection started after 2007.
-But uncertainty increases quickly, so no point looking \> 2 years back.
-
--   Over half of areas have population increases in recent years.
-
--   Recent declines: mountains in central Norway, typically following increases up to around 2016
-
--   Across populations, no clear relationship between density and growth rates (within populations?)
-
--   There is a spatial association lower density / lower reproduction / higher survival ("slower life history"?) versus higher density / higher reproduction / lower survival ("faster life history?")
-
--   Survival recruitment trade-off: high survival and low recruitment in far north and southern mountains
-
-Annual recruitment rates showed a spatial pattern that was inverse relative to that of annual survival, with lower recruitment rates co-occurring with higher survival rates and vice-versa.
-This is consistent with basic life history theory [@stearns1992] and with previous studies finding relatively slower life histories in alpine / high altitude areas [e.g.
-@sandercock2005; @bears2009; @alice2016].
+While we found the highest population growth rates in areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A], our analysis did not provide evidence for a strong relationship between density and growth rates across space.
+However, this is unsurprising as we did find substantial spatial differences in demography, which suggests that carrying capacity too varies over space.
+Overall, annual recruitment rates showed a spatial pattern that was inversely related to that of annual survival, with lower recruitment rates co-occurring with higher survival rates (and lower densities) and vice-versa [@fig-paramCorrs B], as we would expecte based on life history theory [@stearns1992].
+Furthermore, the areas with relatively lower recruitment and higher survival were primarily found in the northern parts of Norway and in the mountainous region in the central part of the country, which aligns with previous studies finding relatively slower bird life histories in alpine / high altitude areas [e.g.
+@sandercock2005; @bears2009; @Wilson_Martin_2011; @alice2016].
+<!-- What about the latitude?
+Do we know of any studies looking at latitudinal variation in life histories?
+When I last searched, I only found studies focusing on altitude -->
 
 <!-- Temporal perspective -->
+Population densities in all areas were subject to substantial variation across years, featuring periods of stability, increase, and decrease (SF "TimeSeries_PopDens1.pdf").
+In most areas, there were also strong year-by-year fluctuations in population density on top of longer-term trends.
+Some of the resulting "high density years" were highly synchronized across large spatial scales, such as the years 2011, 2013, and 2018.
+Taking a closer look, we find that these are years that are characterized by high recruitment (SF "TimeSeries_rRep.pdf") and low subsequent survival (SF "TimeSeries_pSurv.pdf"), which often resulted in steep population declines towards the following year.
+The fact that these same years also match up with observed peaks in rodent abundance in many areas provides evidence for the Alternative Prey Hypothesis (APH).
+The AHP stipulates that high abundance of alternative prey (rodents, in this case) for common predators leads to population growth, and is well-supported throughout the literature for a range of taxa [e.g., @hagen1952; @Kjellander_Nordström_2003; @Reif_Tornberg_Jungell_Korpimäki_2001], including willow ptarmigan [@bowler2020].
+<!-- Maybe another sentence or two about the low survival following the peak, which could be either due to negative density-dependence leading to resource competition among ptarmigan, or due to the rodent peak allowing high recruitment also for predators, which then in turn will have increased impacts on all prey populations in the subsequent year?
+-->
 
-The Alternative Prey Hypothesis (APH) has been well-supported throughout the literature across multiple taxa [e.g., @hagen1952; @Kjellander_Nordström_2003; @Reif_Tornberg_Jungell_Korpimäki_2001] and for willow ptarmigan [@bowler2020].
-We also found support for APH but the positive effect of rodents on recruitment were more pronounced at northern latitudes while equivocal elsewhere.
-Predator-prey dynamics are notoriously complex and interact with a multitude of environmental variables.
-@Nielsen_1999 found that gyrfalcon predation on ptarmigan was greatest during years of low ptarmigan abundance which suggests that the availability of alternative prey is most important at low densities as exhibited at northern latitudes in our study.
-
--   There are some years with relatively high densities in most areas: 2011, 2013, 2018
-
--   These same years are also characterized by high recruitment, and subsequent low survival, and the match up with the years with big rodent peaks in many areas
-
--   Many rodents -\> predator release -\> higher recruitment -\> higher densities -\> negative density feedbacks on survival -\> lower survival -\> low density next year?
-
--   Also: many rodents -\> good year for predators -\> high predation in following season/year -\> lower survival -\> low density next year?
-
--   This seems to support that the rodent abundance is an important driver -\> supported by the model: rodent effects estimated positive or inconclusive across areas
-
--   This is well established (for Norwegian ptarmigan and elsewhere?) and was expected
-
--   BUT: the model was unable to recover the positive relationship between rodent abundance and recruitment when applied only to data from a single location.
-But with this multi-area model, estimation and inference is much improved -\> space-for-time substitution works!
-
--   This also means we have potential for including other covariates
-
--   Examples of what else could matter and why
+Taken together, this alludes to the large role small rodents play in shaping ptarmigan demography and population dynamics and is further supported by the largely positive effects of rodent abundance on recruitment estimated by our model (@fig-paramCorrs C).
+Once more, the highest latitude and highest altitude areas stood out, sporting the strongest rodent effects (SF "betaR_Map.pdf").
+This may be linked to the overall low ptarmigan densities in these areas, as e.g.
+@Nielsen_1999 found evidence that gyrfalcon predation on ptarmigan was most impactful when ptarmigan densities were low.
+<!-- This previous sentence is adapted from a suggestion by James, but I feel like we need some more explanation/arguments here to clarify this.
+Perhaps also some other and/or more specific hypotheses= -->
+Notably, the conclusive estimation of overall positive effects of rodent abundance on recruitment in our model was only possible thanks to the integration and sharing of data across multiple areas.
+When @nilsen2023 fit the IDSM to data from only a single area, they were unable to obtain a reliable estimate for the rodent effect due to limited statistical power.
+Consequently, the space-for-time substitution that comes with extending the model across mulitple area allows estimation of covariate effects that otherwise cannot be estimated, and opens up for future possibilities for studying effects of not just rodents, but also other environmental drivers on ptarmigan population dynamics.
+For example, [...] <!-- @ERLEND here we could add some thoughts on other covariates that may matter and are realistic/relevant/realistic to include in future studies.
+Best if it kind of naturally leads over to (and does not overlap too much with) the next paragraph  -->
 
 <!-- Variation beyond rodents -->
-
--   Including more and/or other covariates may also help further explaining vital rate variation, not least because for both survival and recruitment, residual variation was the largest component.
-
--   However, looking at spatial vs.
-temporal variation, we found very different patterns: for recruitment, we have more spatial than temporal variation, but they are not vastly different.
-For survival, spatial variation was quite low compared to the rather large temporal variation
-
--   Some hypotheses about large- vs. local-scale drivers?
-
-[@Wilson_Martin_2011] Found x y z
+Extending the present analysis by including additional covariates may also help with better understanding the mechanisms underlying the large portion of demographic rate variation that could only be attributed to random variation so far.
+This is the case especially for the relatively large residual variation [@fig-varDecomp] but also relevant for constant spatial and shared temporal variation.
+Interestingly, we found that a larger portion of variation in recruitment was allocated to spatial than to temporal factors, while the opposite was the case for survival.
+[...]
+<!-- Any hypotheses/thoughts related to large- vs. local-scale drivers to explain this? -->
 
@@
 -   Discuss this related to landscape features and/or observer pool? Thoughts on what could be done to improve, potentially?
+
+-   Most uncertainty in density estimates in areas with relatively low population sizes, low number of transects, and recent start of data collection
+
+-   The model also does "hindcasts", i.e.
+predictions for years prior to data collection, for areas where data collection started after 2007.
+But uncertainty increases quickly, so no point looking \> 2 years back.
 

### 2024-01-10 Manuscript_Nater_etal.qmd
Added "skeleton text" for second chapter of Discussion
Here, too, plenty to add

@@
 
-<!-- preliminary section title) -->
+Management decisions made at the resolution of large geo-political boundaries (e.g., Norway) run a high risk of being inadequate when there is substantial spatial variation in demographic processes and population dynamics, as is the case for willow ptarmigan.
+In Norway, willow ptarmigan -- and small game in general -- is managed [...] <!-- Brief summary of how and at what spatial scales patrmigan/small game management works in Norway -->
+The results from ur study suggests that management decisions should be spatially refined and ideally informed by knowledge about the "local" population processes.
+[...] <!-- Some specific implications of the results for management of hunting and of habitat (e.g.
+better managed hunting is good because we do see increases in majority of local populations, but the mountain areas have had recent declines that may call for additional local measures, these measures may need to be adjusted since the mountain populations appear to have slower life histories and likely different sensitivities, etc.) -->
 
--   What results mean for management (of habitat and of harvest)
-
-Management decisions rarely and shouldn't occur at the resolution of large geo-political boundaries (e.g., Norway) because of the variation in demographic processes that occur.
-Our study suggests decisions should be spatially refined to capture "local" population processes that will interact with management.
-
-Management also includes optimization of monitoring
-
--   Detection was less variable across time and space than vital rates, but we do see it tends to be higher in the southern half of the country
-
--   Discuss this related to landscape features and/or observer pool? Thoughts on what could be done to improve, potentially?
-
--   Most uncertainty in density estimates in areas with relatively low population sizes, low number of transects, and recent start of data collection
-
--   The model also does "hindcasts", i.e.
-predictions for years prior to data collection, for areas where data collection started after 2007.
-But uncertainty increases quickly, so no point looking \> 2 years back.
+Besides pointers on how to adjust management of ptarmigan and their habitat, our results also provided some insights into the value, and possibly opportunities for improving the monitoring programme.
+First and foremost, our study demonstrates the tremendous potential lying in coordinating structured monitoring using a common protocol and training programme, as well as data processing pipeline.
+These were indeed the prerequisites that allowed us to easily and efficiently integrate data collected across the entire country in a joint analysis, and draw inference on spatio-temporal variation in demography and population dynamics at a large scale but also with good spatial resolution.
+While overall less variable across space and time than vital rates, differences in detection probabilities were nonetheless evident (SFs "Avg_detect_Map.pdf" and "TimeSeries_pDetect.pdf") and may help with mapping out potential for improvement in the monitoring programme.
+Particularly, we found generally lower detection probabilities in the northern half of Norway.
+[...] <!-- Some thoughts on why that might be.
+Could be related to landscape features and/or available observer pool.
+Any suggestions on what could be done to improve detection in the North, or at least "make up" for it?
+-->
+Ultimately, increased detection probability would contribute to obtaining more precise estimates of both population density and demographic rates, which -- in turn -- would be of great value in particular in areas with relatively low population densities, low number of transects, and less years of data.
 
@@
 -   Also: useful to link up to different reporting routes and/or more general (multi-indicator) projects <!-- @EVERYONE: Do we know of some good references for this?
-PRseumably we cannot cite our chapter in the Technical Manual yet :-P -->
+Prseumably we cannot cite our chapter in the Technical Manual yet :-P -->
 

### 2024-01-11 Manuscript_Nater_etal.qmd
Added abstract, text to discussion, and some bibs.

@@
 
-TBA
+Quantifying the temporal and spatial variation in animal demography is a central theme of ecology and important for directing policy.
+However, field sampling at large spatial extents for long periods of time is cost prohibitive and often politically untennable.
+Citizen science programs can alleviate these issues by harnessing stakeholders to increase the spatial and temporal resolution of datasets.
+Yet, many citizen science programs lack data on individual animals making it difficult to model demography.
+Integrated population models (IPM) offer promise to use multiple streams of data on populations or multiple aspects of a single dataset to estimate key vital rates for populations or communities.
+We leveraged a large-scale citizen science monitoring program---the "Hønsefuglportalen"--- to estimate spatial and temporal variation in demography for willow ptarmigan (*Lagopus lagopus*).
+The core field sampling approach was line-transect distance sampling with age-specific detections---the latter critical to modelling the population dynamics in the IPM.
+We explored latitudinal and longitudinal variation in recruitment, survival, and density; patterns of density dependence; and the relationship between recruitment and rodent occupancy using an IPM in a Bayesian framework following open science principles.
+Overall results were consistent with expectations of ecological theory.
+Slower life history patterns were observed at higher latitudes and altitudes portending differential effects of climate change on ptarmigan across their range.
+Observed patterns between rodents and recruitment were consistent with the alternative prey hypothesis (APH); however, spatial variation in the directionality and magnitude of effect were substantial.
+The robustness of the hierarchical structure of our model was instrumental in elucidating these patterns.
+Our study highlights the power of citizen science, IPMs, and open science principles to estimate spatial and temporal variation in the demography for a culturally important species.
 
@@
 I'll leave it here for now but note that it may get moved eventually.
-
+<!-- Matter of style but this is typically how I do it but your call. :-) >
 ## Abundance and vital rates across space and time
@@
 Ptarmigan densities across the 41 reporting districts included in our analyses varied from around 2 birds/km^2^ to 55 birds/km^2^, with the lowest densities occurring far north in the country, as well as on the west coast and in the mountains in central Norway [@fig-densA].
-These areas are characterized by more extreme climatic conditions, boasting cold temperatures and – in the North – a short growing season.
+These areas are characterized by more extreme climatic conditions, boasting cold temperatures and -- in the North -- a short growing season.
 Resulting reduced primary production limits food availability and as ptarmigan are income breeders using food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005], lower carrying capacity in such areas is to be expected.
 <!-- @ERLEND do we have some more thoughts to add here?
--->
-
-We found increasing population trends over recent years in over half of the reporting districts [@fig-lamA].
+-->We found increasing population trends over recent years in over half of the reporting districts [@fig-lamA].
 <!-- relate this to changes in hunting management in recent years?
@@
 
-While we found the highest population growth rates in areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A], our analysis did not provide evidence for a strong relationship between density and growth rates across space.
+<!-- Erlend will probably have better ideas than I.-->
+
+Density dependence is an important process to understand in ecology but more so for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; @Aanes_Engen_SÆther_Willebrand_Marcström_2002; @willebrand2001].
+Negative density dependence has been found in numerous gallinaceous birds such as northern bobwhites [*Colinus virginianus*; [@mcconnell2018]\], \[*Perdix perdix*; [@bro2003]\], and wild turkeys \[*Meleagris gallopavo*; [@mcghee2007]]*.* <!--# Got the tasty ones in here for you -->While we found the highest population growth rates in areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A], our analysis did not provide evidence for a strong relationship between density and growth rates across space.
 However, this is unsurprising as we did find substantial spatial differences in demography, which suggests that carrying capacity too varies over space.
-Overall, annual recruitment rates showed a spatial pattern that was inversely related to that of annual survival, with lower recruitment rates co-occurring with higher survival rates (and lower densities) and vice-versa [@fig-paramCorrs B], as we would expecte based on life history theory [@stearns1992].
+Also, support for density dependence in willow ptarmigan has been mixed.
+@myrberget1988 observed no change in productivity despite a 50% decrease in abundance.
+Conversely, @pedersen2004 found strong overwinter negative density dependence and they posited that immigration was the vital rate that responded to changes in density.
+Low survival rates appeared to have a one-year lag following abundance peaks.
+This pattern could be a result of negative density dependence that went undetected when looking at the ensemble of samples across the 40 areas.
+Negative density dependence has been shown in willow ptarmigan populations which may explain explain reduced survival following years with elevated densities [@pedersen2004].
+
+<!--# Not happy with the DD para yet but I can work on Jan 11 -->
+
+<!-- Maybe another sentence or two about the low survival following the peak, which could be either due to negative density-dependence leading to resource competition among ptarmigan, or due to the rodent peak allowing high recruitment also for predators, which then in turn will have increased impacts on all prey populations in the subsequent year?
+-->
+
+Overall, annual recruitment rates showed a spatial pattern that was inversely related to that of annual survival, with lower recruitment rates co-occurring with higher survival rates (and lower densities) and vice-versa [@fig-paramCorrs B], as we would expect based on life history theory [@stearns1992].
 Furthermore, the areas with relatively lower recruitment and higher survival were primarily found in the northern parts of Norway and in the mountainous region in the central part of the country, which aligns with previous studies finding relatively slower bird life histories in alpine / high altitude areas [e.g.
@@
 <!-- Temporal perspective -->
+
 Population densities in all areas were subject to substantial variation across years, featuring periods of stability, increase, and decrease (SF "TimeSeries_PopDens1.pdf").
@@
 The AHP stipulates that high abundance of alternative prey (rodents, in this case) for common predators leads to population growth, and is well-supported throughout the literature for a range of taxa [e.g., @hagen1952; @Kjellander_Nordström_2003; @Reif_Tornberg_Jungell_Korpimäki_2001], including willow ptarmigan [@bowler2020].
-<!-- Maybe another sentence or two about the low survival following the peak, which could be either due to negative density-dependence leading to resource competition among ptarmigan, or due to the rodent peak allowing high recruitment also for predators, which then in turn will have increased impacts on all prey populations in the subsequent year?
--->
+Conversely, @Nyström_etal_2006 suggested that gyrfalcons (a specialized ptarmigan predator) did not respond to rodent populations and did not switch to alternative prey when ptarmigan populations are low.
+Nonetheless, other predators such as fox could prey switch in response to changing availability of rodents and ptarmigan as shown by @Breisjøberget_etal_2018 and @bowler2020.
+However, statistical associations in our study and others is not causal and alternative explanations for this pattern are plausible.
+The positive association between rodents and ptarmigan could be caused by bottom-up processes such as food availability unrelated to predation.
+Future work should develop causal models and manipulative experiments to establish causal relationships among top-down and bottom-up processes.
 
@@
 <!-- This previous sentence is adapted from a suggestion by James, but I feel like we need some more explanation/arguments here to clarify this.
-Perhaps also some other and/or more specific hypotheses= -->
-Notably, the conclusive estimation of overall positive effects of rodent abundance on recruitment in our model was only possible thanks to the integration and sharing of data across multiple areas.
+Perhaps also some other and/or more specific hypotheses= --> Notably, the conclusive estimation of overall positive effects of rodent abundance on recruitment in our model was only possible thanks to the integration and sharing of data across multiple areas.
 When @nilsen2023 fit the IDSM to data from only a single area, they were unable to obtain a reliable estimate for the rodent effect due to limited statistical power.
-Consequently, the space-for-time substitution that comes with extending the model across mulitple area allows estimation of covariate effects that otherwise cannot be estimated, and opens up for future possibilities for studying effects of not just rodents, but also other environmental drivers on ptarmigan population dynamics.
-For example, [...] <!-- @ERLEND here we could add some thoughts on other covariates that may matter and are realistic/relevant/realistic to include in future studies.
+Consequently, the space-for-time substitution that comes with extending the model across multiple area allows estimation of covariate effects that otherwise cannot be estimated, and opens up for future possibilities for studying effects of not just rodents, but also other environmental drivers on ptarmigan population dynamics.
+For example, \[...\] <!-- @ERLEND here we could add some thoughts on other covariates that may matter and are realistic/relevant/realistic to include in future studies.
 Best if it kind of naturally leads over to (and does not overlap too much with) the next paragraph  -->
@@
 <!-- Variation beyond rodents -->
+
 Extending the present analysis by including additional covariates may also help with better understanding the mechanisms underlying the large portion of demographic rate variation that could only be attributed to random variation so far.
@@
 Interestingly, we found that a larger portion of variation in recruitment was allocated to spatial than to temporal factors, while the opposite was the case for survival.
-[...]
-<!-- Any hypotheses/thoughts related to large- vs. local-scale drivers to explain this? -->
+\[...\] <!-- Any hypotheses/thoughts related to large- vs.
+local-scale drivers to explain this?
+-->
+
+<!-- This is interesting I will do some reading on this and add some text on Jan 11-->
 
@@
 Management decisions made at the resolution of large geo-political boundaries (e.g., Norway) run a high risk of being inadequate when there is substantial spatial variation in demographic processes and population dynamics, as is the case for willow ptarmigan.
-In Norway, willow ptarmigan -- and small game in general -- is managed [...] <!-- Brief summary of how and at what spatial scales patrmigan/small game management works in Norway -->
-The results from ur study suggests that management decisions should be spatially refined and ideally informed by knowledge about the "local" population processes.
-[...] <!-- Some specific implications of the results for management of hunting and of habitat (e.g.
+In Norway, willow ptarmigan -- and small game in general -- is managed \[...\] <!-- Brief summary of how and at what spatial scales patrmigan/small game management works in Norway --> The results from ur study suggests that management decisions should be spatially refined and ideally informed by knowledge about the "local" population processes.
+\[...\] <!-- Some specific implications of the results for management of hunting and of habitat (e.g.
 better managed hunting is good because we do see increases in majority of local populations, but the mountain areas have had recent declines that may call for additional local measures, these measures may need to be adjusted since the mountain populations appear to have slower life histories and likely different sensitivities, etc.) -->
@@
 Particularly, we found generally lower detection probabilities in the northern half of Norway.
-[...] <!-- Some thoughts on why that might be.
+\[...\] <!-- Some thoughts on why that might be.
 Could be related to landscape features and/or available observer pool.
 Any suggestions on what could be done to improve detection in the North, or at least "make up" for it?
--->
-Ultimately, increased detection probability would contribute to obtaining more precise estimates of both population density and demographic rates, which -- in turn -- would be of great value in particular in areas with relatively low population densities, low number of transects, and less years of data.
+--> Ultimately, increased detection probability would contribute to obtaining more precise estimates of both population density and demographic rates, which -- in turn -- would be of great value in particular in areas with relatively low population densities, low number of transects, and less years of data.
 
@@
 
--   Outlook: integration with / coupling to models for harvest to develop tools for (automatic) calculation of e.g.
-hunting quotas <!-- @JAMES: This one is for you.
+-   Outlook: integration with / coupling to models for harvest to develop tools for (automatic) calculation of e.g. hunting quotas
+
+    The framework used in our study can be adapted to include the effects of management actions and harvest.
+Harvest effects on willow ptarmigan has been studied yet uncertainties still exist [@sandercock_etal2011; @Aanes_Engen_SÆther_Willebrand_Marcström_2002; @pedersen2004; @willebrand2001].
+Also, there is a dearth of empirical information about the effects of harvest on many exploited species despite decades of continued exploitation.
+
+-   <!-- @JAMES: This one is for you.
 Some ruminations on how and why it would be great to formally model harvest, integration with hunting data, maybe eventually develop a tool to calculate quotas (I know this has been developed for other game species), etc.
 -->
+
+-   <!-- Will add more Jan 11 -->
 
@@
 TBA ...
+We are grateful to the participants in the "Hønsefuglportalen" and their furry friends that collected ptarmigan observations.
 Additionally, we would like to thank Bernardo Brandão Niebuhr dos Santos for his help with spatial data visualization.
-<!-- @ERLEND: you maybe know best all the people/institutions we need to thank here --> <!-- @JAMES: should we be acknowledging your fullbright scholarship which made the collab possible?
-If so, please add details.
--->
+J.
+A.
+Martin was supported by a Fulbright Fellowship and the Norwegian Institute for Nature Research that facilitated this collaboration.
+<!-- @ERLEND: you maybe know best all the people/institutions we need to thank here -->
 

### 2024-01-11 Manuscript_Nater_etal.qmd
Added draft text for "Limitations and outlook" section

@@
 
-Density dependence is an important process to understand in ecology but more so for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; @Aanes_Engen_SÆther_Willebrand_Marcström_2002; @willebrand2001].
-Negative density dependence has been found in numerous gallinaceous birds such as northern bobwhites [*Colinus virginianus*; [@mcconnell2018]\], \[*Perdix perdix*; [@bro2003]\], and wild turkeys \[*Meleagris gallopavo*; [@mcghee2007]]*.* <!--# Got the tasty ones in here for you -->While we found the highest population growth rates in areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A], our analysis did not provide evidence for a strong relationship between density and growth rates across space.
+Density dependence is an important process to understand in ecology but more so for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; aanes2002; @willebrand2001].
+Negative density dependence has been found in numerous gallinaceous birds such as northern bobwhites *Colinus virginianus* [@mcconnell2018], *Perdix perdix* [@bro2003], and wild turkeys *Meleagris gallopavo* [@mcghee2007]).
+
+While we found the highest population growth rates in areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A], our analysis did not provide evidence for a strong relationship between density and growth rates across space.
 However, this is unsurprising as we did find substantial spatial differences in demography, which suggests that carrying capacity too varies over space.
@@
 <!--# Not happy with the DD para yet but I can work on Jan 11 -->
+<!--# @JAMES: Feel free to add more, but I feel like if we want to go for a DD deep-dive, I need to run a test for evidence of WITHIN area DD first. Because there may well be a pattern there... -->
 
@@
 
--   Some limitations of "base model" that need mentioning (e.g. age assignment, ignoring potential effects of flock size on detection)
+The primary focus of this work was placed on developing an effective pipeline for integrating data and modelling population dynamics across a large number of areas.
+Consequently, many additional opportunities for improving and refining the modelling framework itself remain.
+First, the precision and accuracy of model estimates might be increased through better accounting for heterogeneity and potential biases in detection of birds during the line transect surveys.
+In an earlier study analyzing data from the same monitoring programme, @bowler2020, for example, found that detection probability was not independent of the size of group birds were part of, resulting in birds in larger groups being more likely to be detected, especially at larger distances.
+When birds are observed in larger groups, it is also not unlikely that human observers may miscount, i.e.
+that there is some observation error in the number reported.
+This could be accounted for by adding an additional layer of hierarchy to observation process (see e.g.
+REF), which could further be used to also account for error in judging the observation distance [e.g.
+@marques2004].
+Another potential source of bias in our IDSM is related to failure to correctly assign the age class of observed birds.
+@nilsen2023 showed that incorrect age assessment can bias (relative) estimates of survival and recruitment, and while they only found a weak bias in their case study on a single area, the problem may be larger in a multi-area setting that may contain areas with different proportions of misclassified observations.
+If misclassification happened at random, mixture models could be used to determine the likely age class of individuals to whom no age class was assigned during observations (+ref).
+In our case, however, juveniles are likely more prone to misclassification due to [...] <!-- @ERLEND can you add reasons?
+-->.
+Future studies should investigate to what degree available information on e.g.
+group composition could be used for this, and what kind of auxiliary data would need to be collected to reliably model misclassification error.
 
--   Outlook: improved modelling of spatio-temporal patterns through correlated REs
+The second (and perhaps most attractive) aspect of our modelling framework in the context of future work is its spatio-temporal hierarchical structure.
+While we included spatial, temporal, and residual variation in our framework here, we treated them as independent.
+Alternatively, spatial (and temporal) correlations among parameters can be modelled explicitly, something that is commonly done e.g.
+for modern species distribution models [e.g.
+@pacifici2017; @guelat2018].
+For demographic models, this has rarely been implemented so far, not least due to the fact that few demographic models have sufficient spatial resolution [@schaub2021].
+The ptarmigan IDSM presented in this study, however, does have sufficient resolution and our results do indeed support that there is spatial clustering in both overall and time-dependent demographic parameters (e.g.
+@fig-pVR, SFs "Avg_pSurv_Map.pdf" & "Avg_rRep_Map.pdf").
+Furthermore, we did find that mixing of several of the global and area-specific intercept parameters in the current model was suboptimal, suggesting that there may be much to gain from additional structuring, as well as from development of more efficient MCMC sampling strategies for the resulting extended model.
+One promising framework for approaching this are conditionally autoregressive models (CARs).
+Such models have been used repeatedly for modelling spatial autocorrelation in species occupancy and demographic rates [e.g.
+@saracco2010; @saracco2012; @guelat2018]
+and are straightforward to implement using NIMBLE [@lawson2020].
+One possible challenge with using CAR models to explicitly model spatial correlations within our ptarmigan IDSM is that CAR models rely on "neighborhood" relationships between discrete areas and many "neighbors" are missing in our ptarmigan data (e.g.
+@fig-dens).
+While estimation of latent parameters in missing areas may possible (Perry de Valpine, personal communication; @schaub2021 chapter 19), this also provides an opportunity for inclusion of additional data.
+The line transect survey data included in this study constitutes just the publicly available part of the data collected through "Hønsefuglportalen" but the programme also includes additional surveys on private land.
+[...] <!-- @ERLEND: Can we add a very brief ocverview over what the coverage (in terms of municipalities) would be if we had access to all the Hønsefuglportalen data?
+--> Exploring to what degree additional data from Hønsefuglportalen could be included in future studies employing an extended IDSM with additional spatial structuring is therefore a worthwhile endeavor.
 
-Another approach to maximising information gain when jointly analysing data from different areas is explicit modelling of spatial and spatio-temporal variation.
-While this is commonly done for modern species distribution models [e.g.
-@pacifici2017; @guelat2018], it has rarely been implemented in demographic models.
-This is not least due to the fact that few demographic models span large spatial extents.
-Since our IDSM and the willow ptarmigan data do actually do that though, extending the framework to account for spatial and spatiotemporal correlations may be feasible.
-Per now, we have not yet attempted this, but are planning to do so in the near future.
-(BUT: we do already have some computational limitations now, so likely to need development of more efficient sampling strategies.
-Which is fortunately possible with NIMBLE but there is a chance that another platform, e.g.
-INLA, may be required) <!-- @ERLEND: related to this, I want to discuss the potential of including the not publicly available data.
-Is there an overview somewhere over what the coverage (in terms of municipalities) would be if we had access to all the Hønsefuglportalen data?
-This might matter not just for increasing sample size, but perhaps also making CAR modelling (spatial structure through neighborhood relations) possible -->
-
-<!-- @MYSELF: Have a closer look at chapter 19 in IPM book to see what they did there and also what other work they reference -->
-
--   Limited usefulness of auxiliary radio-telemetry data as only from one area (big potential for including additional datasets from other small-scale studies, telemetry but also other, e.g.
-nest monitoring?) <!-- @ERLEND: Any thoughts on what the status here is?
-Might there be other data available from beyond Lierne?
+Finally, including further data beyond the line transect surveys may be relevant in the future, and in particular in the context of informing and improving management of ptarmigan hunting. 
+In the present study, we have used auxiliary radio-telemetry data to supplement information on survival, but since this data was available for only one out of 41 areas, its influence was likely small.
+Nonetheless, this illustrates a way for how smaller datasets from single or subsets of areas can be integrated into a large-scale modelling framework.
+Other relevant data could be included using the same approach [...]  <!-- @ERLEND: Any thoughts on what the status here is?
+Might there be other data available from Lierne and/or beyond (other telemetry data, nest monitoring, etc.?
 Are there exciting future prospects e.g.
@@
 -->
-
--   Outlook: integration with / coupling to models for harvest to develop tools for (automatic) calculation of e.g. hunting quotas
-
-    The framework used in our study can be adapted to include the effects of management actions and harvest.
-Harvest effects on willow ptarmigan has been studied yet uncertainties still exist [@sandercock_etal2011; @Aanes_Engen_SÆther_Willebrand_Marcström_2002; @pedersen2004; @willebrand2001].
-Also, there is a dearth of empirical information about the effects of harvest on many exploited species despite decades of continued exploitation.
-
--   <!-- @JAMES: This one is for you.
-Some ruminations on how and why it would be great to formally model harvest, integration with hunting data, maybe eventually develop a tool to calculate quotas (I know this has been developed for other game species), etc.
+The most relevant source of data to be included into the IDSM framework in the near future, however, is harvest data.
+<!-- @ERLEND 1-2 sentence summary of what kind of harvest data is available, and at what spatial covarage and resolution --> As the IDSM framework is, in essence, an IPM, harvest can be modelled through partitioning of survival into cause-specific mortality in the process model and inclusion of relevant harvest data likelihoods [e.g.
+@gamelon2021; @nater2021].
+Harvest effects on willow ptarmigan have been studied previously yet uncertainties persist [@sandercock_etal2011; aanes2002; @pedersen2004; @willebrand2001].
+[...] <!-- @JAMES & Erlend: Maybe we can be a bit more specific about what the "uncertainties" are regarding ptarmigan harvest?
+Might be good to keep the "spatially explicit" perspective a bit, i.e.
+why would it be great to look at this with this kind of model, instead of just aggregate numbers across large regions?
+Also, maybe we should highlight that up-to-date (easilty repeatable) analyses are important because things can change, and harvest effects may interact with other (emergent) factors such as climate change and habitat degradation.
+Finally, it could also be nice to link this to some studies on other species...
 -->
-
--   <!-- Will add more Jan 11 -->
 
@@
 
+-   Outlook: integration with / coupling to models for harvest to develop tools for (automatic) calculation of e.g. hunting quotas
+
 -   Also: useful to link up to different reporting routes and/or more general (multi-indicator) projects <!-- @EVERYONE: Do we know of some good references for this?
-Prseumably we cannot cite our chapter in the Technical Manual yet :-P -->
+Presumably we cannot cite our chapter in the Technical Manual yet :-P -->
 
@@
 TBA ...
-We are grateful to the participants in the "Hønsefuglportalen" and their furry friends that collected ptarmigan observations.
+We are grateful to the participants in the "Hønsefuglportalen" and their dog helpers that collected ptarmigan observations.
 Additionally, we would like to thank Bernardo Brandão Niebuhr dos Santos for his help with spatial data visualization.

### 2024-01-12 Manuscript_Nater_etal.qmd
edits to qdm and bib

@@
 Quantifying the temporal and spatial variation in animal demography is a central theme of ecology and important for directing policy.
-However, field sampling at large spatial extents for long periods of time is cost prohibitive and often politically untennable.
+However, field sampling at large spatial extents for long periods of time is cost prohibitive and often politically untenable.
 Citizen science programs can alleviate these issues by harnessing stakeholders to increase the spatial and temporal resolution of datasets.
@@
 
-Building on the work of @nilsen2023, we applied a novel integrated population model to data collected through a national-scale participatory monitoring programme to estimate spatial and temporal variation in demography of a culturally important game bird species, the willow ptarmigan.
-While our study was exploratory in nature, it recovered patterns consistent with ecological and life-history theory including trade-offs between survival and recruitment, and a tendency towards slower life histories at higher latitudes and altitudes.
-Space-for-time substitution also provided the statistical power necessary for the analysis to provide evidence for the alternative prey hypothesis, i.e.
-ptarmigans benefiting from high abundance of alternative prey for their predators.
-<!-- To me, this little paragraph has "abstract vibes", or maybe "conclusion vibes".
-I'll leave it here for now but note that it may get moved eventually.
-<!-- Matter of style but this is typically how I do it but your call. :-) >
 ## Abundance and vital rates across space and time
@@
 
-Density dependence is an important process to understand in ecology but more so for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; aanes2002; @willebrand2001].
+Density dependence is an important process to understand in ecology but more so for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; @aanes2002; @willebrand2001].
 Negative density dependence has been found in numerous gallinaceous birds such as northern bobwhites *Colinus virginianus* [@mcconnell2018], *Perdix perdix* [@bro2003], and wild turkeys *Meleagris gallopavo* [@mcghee2007]).
-
 While we found the highest population growth rates in areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A], our analysis did not provide evidence for a strong relationship between density and growth rates across space.
@@
 Low survival rates appeared to have a one-year lag following abundance peaks.
-This pattern could be a result of negative density dependence that went undetected when looking at the ensemble of samples across the 40 areas.
-Negative density dependence has been shown in willow ptarmigan populations which may explain explain reduced survival following years with elevated densities [@pedersen2004].
+This pattern could be a result of negative density dependence that went undetected when looking at the ensemble of samples across the 41 areas.
+Future work should explore different time-lags for density and population growth and the possibility for non-linear effects.
 
-<!--# Not happy with the DD para yet but I can work on Jan 11 -->
 <!--# @JAMES: Feel free to add more, but I feel like if we want to go for a DD deep-dive, I need to run a test for evidence of WITHIN area DD first. Because there may well be a pattern there... -->
 
-<!-- Maybe another sentence or two about the low survival following the peak, which could be either due to negative density-dependence leading to resource competition among ptarmigan, or due to the rodent peak allowing high recruitment also for predators, which then in turn will have increased impacts on all prey populations in the subsequent year?
--->
+<!--# Yeah I agree. We already have a lot going on so I am ambivalent about adding another aspect. -->
 
@@
 
+<!--# Yeah I wasn't able to find much. -->
+
 <!-- Temporal perspective -->
@@
 Taking a closer look, we find that these are years that are characterized by high recruitment (SF "TimeSeries_rRep.pdf") and low subsequent survival (SF "TimeSeries_pSurv.pdf"), which often resulted in steep population declines towards the following year.
-The fact that these same years also match up with observed peaks in rodent abundance in many areas provides evidence for the Alternative Prey Hypothesis (APH).
+The fact that these same years also match up with observed peaks in rodent abundance in many areas provides evidence for the Alternative Prey Hypothesis (APH; [@Israelsen2020]).
 The AHP stipulates that high abundance of alternative prey (rodents, in this case) for common predators leads to population growth, and is well-supported throughout the literature for a range of taxa [e.g., @hagen1952; @Kjellander_Nordström_2003; @Reif_Tornberg_Jungell_Korpimäki_2001], including willow ptarmigan [@bowler2020].
@@
 Consequently, the space-for-time substitution that comes with extending the model across multiple area allows estimation of covariate effects that otherwise cannot be estimated, and opens up for future possibilities for studying effects of not just rodents, but also other environmental drivers on ptarmigan population dynamics.
-For example, \[...\] <!-- @ERLEND here we could add some thoughts on other covariates that may matter and are realistic/relevant/realistic to include in future studies.
+For example, the cover of ericaceous shrubs as a proxy for food availability or spatially-explicit spring green up dates both measured using remote sensing are ecologicaly plausible and feasibly acquired factors.
+<!-- @ERLEND here we could add some thoughts on other covariates that may matter and are realistic/relevant/realistic to include in future studies.
 Best if it kind of naturally leads over to (and does not overlap too much with) the next paragraph  -->
+
+<!--# I added two that came to my mind  -->
 
@@
 Interestingly, we found that a larger portion of variation in recruitment was allocated to spatial than to temporal factors, while the opposite was the case for survival.
-\[...\] <!-- Any hypotheses/thoughts related to large- vs.
-local-scale drivers to explain this?
--->
+Contrastingly, ptarmigan nest failure is almost entirely caused by predation which has been shown to vary greatly across years and less across space [@Munkebye_Pedersen_Steen_Brøseth_2003; @kvasnes2015; @kvasnes2014] whereas adult survival has been shown to not very greatly across years [@Israelsen2020].
+Also, immigration, included in our recruitment term <!--# this true?
+-->, has been hypothesized to be a spatial process [@kvasnes2015].
+This conundrum could be a result of the limited spatial replication of many previous studies that we were able to address with our approach.
+Future research could help rectify these discrepancies by including more mechanistic relationships between environmental factors and vital rates.
+
+<!-- Any hypotheses/thoughts related to large- vs. local-scale drivers to explain this? -->
 
 <!-- This is interesting I will do some reading on this and add some text on Jan 11-->
+
+<!-- I don't know if I helped or not. It is an interesting result. Erlend, you got thoughts? -->
 
@@
 Management decisions made at the resolution of large geo-political boundaries (e.g., Norway) run a high risk of being inadequate when there is substantial spatial variation in demographic processes and population dynamics, as is the case for willow ptarmigan.
-In Norway, willow ptarmigan -- and small game in general -- is managed \[...\] <!-- Brief summary of how and at what spatial scales patrmigan/small game management works in Norway --> The results from ur study suggests that management decisions should be spatially refined and ideally informed by knowledge about the "local" population processes.
+In Norway, willow ptarmigan -- and small game in general -- is managed \[...\] <!-- Brief summary of how and at what spatial scales patrmigan/small game management works in Norway --> The results from our study suggests that management decisions should be spatially refined and ideally informed by knowledge about the "local" population processes.
 \[...\] <!-- Some specific implications of the results for management of hunting and of habitat (e.g.
@@
 Any suggestions on what could be done to improve detection in the North, or at least "make up" for it?
---> Ultimately, increased detection probability would contribute to obtaining more precise estimates of both population density and demographic rates, which -- in turn -- would be of great value in particular in areas with relatively low population densities, low number of transects, and less years of data.
+-->
+
+<!--# They obviously use a lesser breed of bird dog up that way.Probably use more german shorthairs or vislas. :-)  -->
+
+Ultimately, increased detection probability would contribute to obtaining more precise estimates of both population density and demographic rates, which -- in turn -- would be of great value in particular in areas with relatively low population densities, low number of transects, and less years of data.
 
@@
 If misclassification happened at random, mixture models could be used to determine the likely age class of individuals to whom no age class was assigned during observations (+ref).
-In our case, however, juveniles are likely more prone to misclassification due to [...] <!-- @ERLEND can you add reasons?
+In our case, however, juveniles are likely more prone to misclassification due to \[...\] <!-- @ERLEND can you add reasons?
 -->.
@@
 Such models have been used repeatedly for modelling spatial autocorrelation in species occupancy and demographic rates [e.g.
-@saracco2010; @saracco2012; @guelat2018]
-and are straightforward to implement using NIMBLE [@lawson2020].
+@saracco2010; @saracco2012; @guelat2018] and are straightforward to implement using NIMBLE [@lawson2020].
 One possible challenge with using CAR models to explicitly model spatial correlations within our ptarmigan IDSM is that CAR models rely on "neighborhood" relationships between discrete areas and many "neighbors" are missing in our ptarmigan data (e.g.
@@
 The line transect survey data included in this study constitutes just the publicly available part of the data collected through "Hønsefuglportalen" but the programme also includes additional surveys on private land.
-[...] <!-- @ERLEND: Can we add a very brief ocverview over what the coverage (in terms of municipalities) would be if we had access to all the Hønsefuglportalen data?
+\[...\] <!-- @ERLEND: Can we add a very brief ocverview over what the coverage (in terms of municipalities) would be if we had access to all the Hønsefuglportalen data?
 --> Exploring to what degree additional data from Hønsefuglportalen could be included in future studies employing an extended IDSM with additional spatial structuring is therefore a worthwhile endeavor.
@@
 linked to automated monitoring with cameras/sound boxes at a larger scale?
--->
-The most relevant source of data to be included into the IDSM framework in the near future, however, is harvest data.
+--> The most relevant source of data to be included into the IDSM framework in the near future, however, is harvest data.
 <!-- @ERLEND 1-2 sentence summary of what kind of harvest data is available, and at what spatial covarage and resolution --> As the IDSM framework is, in essence, an IPM, harvest can be modelled through partitioning of survival into cause-specific mortality in the process model and inclusion of relevant harvest data likelihoods [e.g.
 @gamelon2021; @nater2021].
-Harvest effects on willow ptarmigan have been studied previously yet uncertainties persist [@sandercock_etal2011; aanes2002; @pedersen2004; @willebrand2001].
-[...] <!-- @JAMES & Erlend: Maybe we can be a bit more specific about what the "uncertainties" are regarding ptarmigan harvest?
+Harvest effects on willow ptarmigan have been studied previously yet uncertainties persist [@sandercock_etal2011; @aanes2002; @pedersen2004; @willebrand2001].
+For example, the strength and shape of density dependence and how harvest may induce population sinks because of Allee effects [@kvasnes2015]).
+Modelling allee effects, or better yet predicting where they are likely to occur if over-harvest occurs, requires a spatially-explicit understanding of harvest and its impact on local populations.
+Because harvest effort, thus rates, can vary dramatically over time and space [@breisjøberget2018] easily repeatable analyses that can be done continually would prevent prolonged over-exploitation.
+Harvest effects may interact with other (emergent) factors such as climate change and habitat degradation making the need to be identify triggers for policy changes paramount [@gamelon2019].
+
+<!-- @JAMES & Erlend: Maybe we can be a bit more specific about what the "uncertainties" are regarding ptarmigan harvest?
 Might be good to keep the "spatially explicit" perspective a bit, i.e.
@@
 -->
+
+<!--# That do it? -->
 
@@
 ## Conclusion
+
+Building on the work of @nilsen2023, we applied a novel integrated population model to data collected through a national-scale participatory monitoring programme to estimate spatial and temporal variation in demography of a culturally important game bird species, the willow ptarmigan.
+While our study was exploratory in nature, it recovered patterns consistent with ecological and life-history theory including trade-offs between survival and recruitment, and a tendency towards slower life histories at higher latitudes and altitudes.
+Space-for-time substitution also provided the statistical power necessary for the analysis to provide evidence for the alternative prey hypothesis, i.e.
+ptarmigans benefiting from high abundance of alternative prey for their predators.
+
+<!-- To me, this little paragraph has "abstract vibes", or maybe "conclusion vibes". I'll leave it here for now but note that it may get moved eventually. -->
 

### 2024-01-12 Manuscript_Nater_etal.qmd
Post-merge clean-up

@@
 This pattern could be a result of negative density dependence that went undetected when looking at the ensemble of samples across the 41 areas.
-Future work should explore different time-lags for density and population growth and the possibility for non-linear effects.
+Negative density dependence has been shown in willow ptarmigan populations which may explain explain reduced survival following years with elevated densities [@pedersen2004].Future work should explore different time-lags for density and population growth and the possibility for non-linear effects.
+
+<!--# Not happy with the DD para yet but I can work on Jan 11 -->
 
@@
 <!--# They obviously use a lesser breed of bird dog up that way.Probably use more german shorthairs or vislas. :-)  -->
+<!-- Funnily enough, I have been suggesting to Erlend that we should have a master student to analyze the effect of dog breed on detection probability :-P -->
 

### 2024-01-12 Manuscript_Nater_etal.qmd
Changed comment
Added inspirational text on reproducible workflows

@@
 
--   Structure / setup of this workflow (from data collection to analytical results)
+Producing a transparent and reproducible workflow for the analysis presented here was a central objective in this study.
+We have done this by setting up "targets" pipeline [@landau2021], which allows (re-)running the complete workflow from downloading the publicly available data to visualizing the results produced by the IDSM [@fig-workflow].
+Modern applied ecology needs research to be published not just as manuscripts, but as reproducible and well documented workflows [@lewis2018].
+This is particularly crucial for research that is (to be) closely tied to management and/or used to create biodiversity indicators that are to be reported nationally or internationally, or to be used by industrial partners.
+This is both because of the enhanced transparency and credibility provided by openly available reproducible workflows and because of their cost-effectiveness, which allows for more sustainable use of funding in the mid- to long-term.
+Finally, open and reproducible workflows facilitate collaboration and inclusion of stakeholders in the research process, paving the path for the translational science that is required for society to tackle the the biodiversity crisis [@rubert2021].
+It is our hope that this study can serve as an example of where to start.
 
--   Why this matters: transparency and credibility, feedback to volunteer data collectors, cost-effective to re-do analyses for annual/regular reporting, easy to re-use and build upon, publicly accessible and well documented to allow more effective resource use in follow-up projects (sustainable use of funding), etc.
-
--   Outlook: integration with / coupling to models for harvest to develop tools for (automatic) calculation of e.g. hunting quotas
-
--   Also: useful to link up to different reporting routes and/or more general (multi-indicator) projects <!-- @EVERYONE: Do we know of some good references for this?
-Presumably we cannot cite our chapter in the Technical Manual yet :-P -->
+ <!-- @EVERYONE: Do we know of some good references for this section? Presumably we cannot cite our chapter in the Technical Manual yet :-P -->
 
@@
 
-<!-- To me, this little paragraph has "abstract vibes", or maybe "conclusion vibes". I'll leave it here for now but note that it may get moved eventually. -->
+<!--James originally intended this paragraph for the start of the discussion. I'll figure out where it fits best once the rest is mostly in place! -->
 

### 2024-01-12 Manuscript_Nater_etal.qmd

@@
 
+Building on the work of @nilsen2023, we applied a novel integrated population model to data collected through a national-scale participatory monitoring programme to estimate spatial and temporal variation in demography of a culturally important game bird species, the willow ptarmigan.
+While our study was exploratory in nature, it recovered patterns consistent with ecological and life-history theory including trade-offs between survival and recruitment, and a tendency towards slower life histories at higher latitudes and altitudes.
+Space-for-time substitution also provided the statistical power necessary for the analysis to provide evidence for the alternative prey hypothesis, i.e.
+ptarmigans benefiting from high abundance of alternative prey for their predators.
+Taken together, the results highlight the potential of integrating demographic data across large spatial scales in the contexts of both informing management and creating biodiversity indicators for higher-level reporting.
+
 ## Abundance and vital rates across space and time
@@
 
-## Conclusion
-
-Building on the work of @nilsen2023, we applied a novel integrated population model to data collected through a national-scale participatory monitoring programme to estimate spatial and temporal variation in demography of a culturally important game bird species, the willow ptarmigan.
-While our study was exploratory in nature, it recovered patterns consistent with ecological and life-history theory including trade-offs between survival and recruitment, and a tendency towards slower life histories at higher latitudes and altitudes.
-Space-for-time substitution also provided the statistical power necessary for the analysis to provide evidence for the alternative prey hypothesis, i.e.
-ptarmigans benefiting from high abundance of alternative prey for their predators.
-
-<!--James originally intended this paragraph for the start of the discussion. I'll figure out where it fits best once the rest is mostly in place! -->
-
--   Summarise neatly
--   Things worth highlighting: inference across large spatial scale but with "useful" resolution for management, advantages of integration across space (we can estimate better and more), integration of small locality-specific datasets with large national-scale monitoring, importance of reproducible workflows
-
 # Acknowledgements

### 2024-01-15 Manuscript_Nater_etal.qmd
minor edits

@@
 title: "Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird"
-format: html
+format: docx
 author: Chloé R. Nater, James A. Martin, Erlend B. Nilsen
@@
 <!--# They obviously use a lesser breed of bird dog up that way.Probably use more german shorthairs or vislas. :-)  -->
+
 <!-- Funnily enough, I have been suggesting to Erlend that we should have a master student to analyze the effect of dog breed on detection probability :-P -->

### 2024-01-15 Analysis_RealData.R
Added post-hoc check for density-dependence

@@ -11,9 +11,6 @@ library(nimble)
 mySeed <- 32
 set.seed(mySeed)
 
-## Set number of chains
-nchains <- 5
-
 ## Source all functions in "R" folder
 sourceDir <- function(path, trace = TRUE, ...) {
   for (nm in list.files(path, pattern = "[.][RrSsQq]$")) {
@@ -128,9 +125,6 @@ input_data <- prepareInputData(d_trans = LT_data$d_trans,
 ## Write model code
 modelCode <- writeModelCode(survVarT = survVarT)
 
-## Expand seed to get MCMC seeds
-MCMC.seeds <- expandSeed_MCMC(seed = mySeed, nchains = nchains)
-
 ## Setup for model using nimbleDistance::dHN
 model_setup <- setupModel(modelCode = modelCode,
                           R_perF = R_perF,
@@ -139,8 +133,12 @@ model_setup <- setupModel(modelCode = modelCode,
                           nim.data = input_data$nim.data,
                           nim.constants = input_data$nim.constants,
                           testRun = FALSE, 
-                          nchains = nchains,
-                          initVals.seed = MCMC.seeds)
+                          nchains = 6,
+                          initVals.seed = mySeed)
+
+
+## Expand seed to get MCMC seeds
+MCMC.seeds <- expandSeed_MCMC(seed = mySeed, nchains = model_setup$mcmcParams$nchains)
 
 
 # MODEL (TEST) RUN #
@@ -270,6 +268,17 @@ if(fitRodentCov){
                     area_names = input_data$nim.constant$area_names,
                     fitRodentCov = fitRodentCov)
 }
+
+
+# OPTIONAL: CHECK WITHIN-AREA DENSITY DEPENDENCE #
+#------------------------------------------------#
+
+checkDD(mcmc.out = IDSM.out.tidy, 
+        N_areas = input_data$nim.constant$N_areas, 
+        area_names = input_data$nim.constant$area_names, 
+        N_sites = input_data$nim.constant$N_sites, 
+        min_years = input_data$nim.constant$min_years, 
+        max_years = input_data$nim.constant$max_years)
 
 
 # OPTIONAL: CALCULATE AND PLOT VARIANCE DECOMPOSITION #

### 2024-01-15 R/checkDD.R

@@ -1,0 +1,75 @@
+#' Calculate posterior summaries of parameter correlations with density
+#'
+#' @param mcmc.out an mcmc list containing posterior samples from a model run.
+#' @param N_areas integer. Number of areas included in analyses. 
+#' @param area_names character vector containing area/location names.
+#' @param N_sites matrix containing the number of sites per area/location.
+#' @param min_years integer vector. Indices of first year with available data for each area/location. 
+#' @param max_years integer vector. Indices of last year with available data for each area/location. 
+#'
+#' @return a vector of pdf plot names. The plots can be found in Plots/TimeSeries.
+#' @export
+#'
+#' @examples
+
+checkDD <- function(mcmc.out, 
+                    N_areas, area_names, N_sites, 
+                    min_years, max_years){
+  
+  ## Convert posterior samples to matrix format
+  out.mat <- as.matrix(mcmc.out)
+
+  ## Set up data frame for posterior summaries
+  DD_corrs <- data.frame()
+  
+  ## Calculate posterior correlation coefficients 
+  for(x in 1:N_areas){
+    
+    # Determine area-specific year range
+    area_yearIdxs <- (min_years[x]:max_years[x])
+    
+    # Set up vector for storing correlation coefficients
+    corr_lambda <- corr_rep <- corr_surv <- rep(NA, nrow(out.mat))
+    
+    for(i in 1:nrow(out.mat)){
+      
+      # Extract parameter time series
+      densT <- rep(NA, length(area_yearIdxs))
+
+      for(t in 1:length(area_yearIdxs)){
+        densT[t] <- mean(out.mat[i, paste0("Density[",  x, ", 1, ", 1:N_sites[x], ", ", area_yearIdxs[t], "]")] + 
+          out.mat[i, paste0("Density[",  x, ", 2, ", 1:N_sites[x], ", ", area_yearIdxs[t], "]")])
+      }
+      
+      lambdaT <- c(densT[2:length(densT)]/densT[1:(length(densT)-1)], NA)
+      repT <- unname(out.mat[i, paste0("R_year[",  x, ", ", area_yearIdxs, "]")])
+      survT <- unname(out.mat[i, paste0("S[",  x, ", ", area_yearIdxs, "]")])
+      
+      # Calculate correlation coefficients
[…]

### 2024-01-15 _targets.R

@@ -14,8 +14,7 @@ library(parallel)
 mySeed <- 32
 set.seed(mySeed)
 
-## Set number of chains
-nchains <- 5
+nchains <- 6
 
 ## Source all functions in "R" folder
 sourceDir <- function(path, trace = TRUE, ...) {
@@ -212,6 +211,17 @@ list(
                       N_areas = input_data$nim.constant$N_areas, 
                       area_names = input_data$nim.constant$area_names,
                       fitRodentCov = fitRodentCov)
+  ),
+  
+  tar_target(
+    DD.checks,
+    checkDD(mcmc.out = IDSM.out.tidy, 
+            N_areas = input_data$nim.constant$N_areas, 
+            area_names = input_data$nim.constant$area_names, 
+            N_sites = input_data$nim.constant$N_sites, 
+            min_years = input_data$nim.constant$min_years, 
+            max_years = input_data$nim.constant$max_years),
+    format = "file"
   ),
   
   tar_target(

### 2024-01-16 Manuscript_Nater_etal.qmd
Text edits and additions to intro, methods and results, and .bib file

@@
 title: "Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird"
-format: docx
+format: html
 author: Chloé R. Nater, James A. Martin, Erlend B. Nilsen
@@
 Citizen science programs can alleviate these issues by harnessing stakeholders to increase the spatial and temporal resolution of datasets.
-Yet, many citizen science programs lack data on individual animals making it difficult to model demography.
+Yet, most citizen science programs lack data from marked animals making it difficult to model demography.
 Integrated population models (IPM) offer promise to use multiple streams of data on populations or multiple aspects of a single dataset to estimate key vital rates for populations or communities.
@@
 There is growing demand for biodiversity indicators from international unions, national government, local management bodies, and corporate and industry actors.
-Such indicators ideally represent a wide range of biodiversity's states and functions [e.g.
+Such indicators should ideally represent a wide range of biodiversity's states and functions [e.g.
 Essential Biodiversity Variables, @pereira2013; @jetz2019], yet the development of suitable indicators for certain attributes, such as species abundance and demography, has been more difficult than for others [@schmeller2018; @waldock2022].
@@
 @feld2009].
-On the other hand, indicators also ideally have good spatial resolution as the scales relevant for local-level management and planning are often much smaller [@stevenson2021].
+On the other hand, indicators also ideally have high spatial resolution, as the scales relevant for local-level management and planning are often much smaller [@stevenson2021].
 This latter requirement is particularly crucial for (infrastructure) development strategies and for species management and conservation, both of which tend to require knowledge on species abundance and population dynamics (i.e.
@@
 @robinson2014; @horswill2019].
-This variation that needs to be accounted for to develop successful and sustainable strategies for area use, (harvest) management, and species and biodiversity conservation [@williams2002].
+This variation needs to be accounted for to develop successful and sustainable strategies for area use, (harvest) management, and species and biodiversity conservation [@williams2002].
 
@@
 Consequently, many countries have been working on setting up, maintaining, and improving such monitoring programmes over the last decades.
-Many now well-established programmes focus on breeding birds and butterflies, and examples include the North American Breeding Bird Survey (<https://www.pwrc.usgs.gov/bbs/>), UK Butterfly Monitoring Scheme (<https://ukbms.org/>), the Game and Wildlife Conservation Trust Partridge Count Scheme [@aebischer2010], and the Swiss Biodiversity Monitoring (<https://www.biodiversitymonitoring.ch/>).
+Many now well-established programmes focus on breeding birds and butterflies, and examples include the North American Breeding Bird Survey (<https://www.pwrc.usgs.gov/bbs/>), the PanEuropean Common Bird Monitoring Scheme (<https://pecbms.info/>[- PECBMS](https://pecbms.info/)), the UK Butterfly Monitoring Scheme ([https://ukbms.org/](#0){.uri style="font-size: 11pt;"}), the Game and Wildlife Conservation Trust Partridge Count Scheme [@aebischer2010], and the Swiss Biodiversity Monitoring ([https://www.biodiversitymonitoring.ch/](#0){.uri style="font-size: 11pt;"}).
 <!-- is there any other one we want to name here specifically?
@@
 
-There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly and requires trained specialists, thus limiting the amount of data that can be collected, while participatory monitoring, i.e.
+There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly, requires trained specialists, and depends on at least some top-down control of the program.
+This often limit the amount of data that can be collected, while participatory monitoring, i.e.
 the collection of ecological data by members of the public [also called citizen or community science, @fraisl2022], allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias [@johnston2023].
@@
 For example, in the United States hunters participate in the collection of bands and wings from harvested American Woodcock (*Scopolax minor)* to estimate survival and age ratios [@zimmerman2010].
+At the European level, the recently established initiative "European Observatory of Wildlife" are offering common field- and analyses protocols and aim to establish a network of "observation points" to monitor wildlife populations at the European level (<https://wildlifeobservatory.org/>).
 <!-- can we think of 1-2 internationally well-known examples?
-Check out MAPS (Monitoring of Avian Productivity and Survivorship) and CES (Constant Effort Sites) --> In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", <https://honsefugl.nina.no/Innsyn/en>).
+Check out MAPS (Monitoring of Avian Productivity and Survivorship) and CES (Constant Effort Sites) --> In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", [https://honsefugl.nina.no/Innsyn/en](#0){.uri style="font-size: 11pt;"}).
 It is a line transect survey programme carried out annually in \>120 localities across the country (\>2000 transects) by trained volunteers using pointing dogs.
-The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata (+ref to protocol?) and established routines for quality control and annual releases of publicly available data via GBIF.
+The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata (+ref to protocol?) and established routines for quality control and annual releases of publicly available data via the Global Biodiversity Information Facility (GBIF).
 As such, it is particularly well suited to become part of a workflow for producing and updating abundance and population indicators on an annual basis.
@@
 The line transect data from "Hønsefuglportalen" has been used previously for estimating abundance trends of willow ptarmigan (*Lagopus lagopus*) across Norway [e.g.
-@bowler2020; @nilsen2020] but large-scale estimation of demographic rates underlying these trends has thus far remained as untapped potential of the dataset.
+@bowler2020; @nilsen2020], and to test relevant ecological hypotheses [@bowler2020; @Breisjøberget_etal_2018].
+However, large-scale estimation of demographic rates underlying these trends has thus far remained as untapped potential of the dataset.
 @nilsen2023 recently developed a novel integrated distance sampling model (IDSM) which successfully uses the age of idividuals detected along line transects data coupled with radio-telemetry data to estimate not just abundance, but also survival and recruitment across years.
-In this study, we adapt the model of @nilsen2023 to run not just on a single site but on all areas with publicly available line transect data from "Hønsefuglportalen" simultaneously.
+In this study, we adapt and extend the model of @nilsen2023 to run not just on a single site but on all areas with publicly available line transect data from "Hønsefuglportalen" simultaneously.
 Unlike several previous studies applying integrated models for population dynamics to multiple (sub-) populations separately and comparing results [e.g.
@@
 While the species is currently listed as Least Concern (LC) both in the global and Norwegian Red List of Species, it has undergone rather dramatic declines in abundance in Norway since the turn of the 20th century [@hjeljord2022].
-While the main reason for the long-term decline in abundance remain unresolved, willow ptarmigan are considered sentinel species that are sensitive to both climate change and land use changes [@Henden2017; @Storch2007].
+The main reason for the long-term decline in abundance remain unresolved, but the willow ptarmigan are considered sentinel species that are sensitive to both climate change and land use changes [@Henden2017; @Storch2007].
 Moreover, being one of only a handful of bird species that spend the winter in mountain ecosystems in Scandinavia, they are important components of the ecosystem as prey species for the gyrfalcon [@Alastair2020] <!-- do we have other "important" species to add here.
@@
 Some other endangered beasts perhaps?
--->.
+ERLEND: Well; there are more mammals, as they typically do not migrate long distances to other continents....
+I suggest we leave it like this -->.
 The willow ptarmigan has a relatively fast pace of life [@sandercock2005; @Steen1996], and can display substantial spatio-temporal variation in demographic rates [@Bowler2020].
 Their population dynamics are characterized by large inter-annual fluctuations in abundance [@hjeljord2022], and previous research has tied these fluctuations to rodent cycle through shared predators [@hagen1952; @Bowler2020].
-This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis and has been a central part of Fennoscandian research on grouse population dynamics for many decades [@hagen1952; @Linden1988; @Steen1985].
+This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis (APH) and has been a central part of Fennoscandian research on grouse population dynamics for many decades [@Elton1942; @hagen1952; @Linden1988; @Steen1985].
 In addition, spring weather conditions and phenology is known to have considerable effects on breeding success and recruitment rates [@Eriksen2023; @Steen1985].
-Across their distributional range, they are iconic species with a high cultural value, partly linked to their popularity as game species.
-The latter means that information about spatio-temporal variation in demographic rates and population dynamics is particularly important in order to design sustainable harvest strategies.
+Across their distributional range, willow ptarmigan are iconic species with a high cultural value, partly linked to their popularity as game species.
+The latter means that information about spatio-temporal variation in demographic rates and population dynamics is particularly important in order to design sustainable harvest strategies [@Eriksen2023].
 In addition, as a sentinel species the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is included in both the main national biodiversity [Nature Index for Norway, @jakobsson2020, <https://www.naturindeks.no/Indicators/lirype>] and ecosystem [Assessment of the Ecological Condition, @framstad2022] assessments.
@@
 The line transect survey data were collected through a structured participatory monitoring program called "Hønsefuglportalen" (<https://honsefugl.nina.no/Innsyn/en>).
-In August each year, trained volunteer fieldworkers collect observations of willow ptarmigan and other grouse species (rock ptarmigan *Lagopus muta*, black grouse *Lyrurus tetrix*, capercaillie *Tetrao urogallus*) along predefined line transects.
+In the first three weeks of August each year, trained volunteer fieldworkers collect observations of willow ptarmigan and other grouse species (rock ptarmigan *Lagopus muta*, black grouse *Lyrurus tetrix*, and capercaillie *Tetrao urogallus*) along predefined line transects.
 To increase the detection probability, each team of fieldworkers use a pointing dog to locate the birds.
@@
 The transect lines vary in length, but are typically between 1 and 8 km (range: 0.3-16.2 km, median: 3 km).
-When birds are observed, the exact location of observation is noted, along with its perpendicular distance from the transect line, as well as the age and sex of the birds.
+When birds are observed, the exact location of observation is reported, along with its perpendicular distance from the transect line, as well as the age and sex of the birds.
 An observation typically includes 1 - 12 birds (mean = 5.6), with groups \> 1 typically representing one brood (female and or male with young-of-the-year chicks).
@@
 Surveys are carried out on both public and private land.
-After an initial embargo period, all data from public land are published and made freely available as a sampling-event data set on GBIF (refs).
+After an initial embargo period, all data from public land are published and made freely available as a sampling-event data set on GBIF (<https://www.gbif.org/sampling-event-data>).
 The published datasets contain both metadata about the transect surveys (survey date, line transect length and location, study area ID, etc.) as well as bird observation data (species, number of birds of different categories (adult males, adult females, juveniles, and birds of unknown category), perpendicular distance to transect line, exact location, and time of observation).
@@
 In this study we used all publicly available data for the period 2007-2021, which included a total of 2225 transects in 41 different reporting districts spanning 9 counties and 50 municipalities.
-Transects on which no willow ptarmigan were ever observed (i.e.
-species absence likely due to low habitat suitability) were dropped, resulting in 2077 transects used in the analyses.
+Transects on which no willow ptarmigan were observed during the study period (i.e.
+species absence likely due to low habitat suitability) were not included.
+After this initial filtering, a total of 2077 transects were included in the analyses.
 
@@
 Can we have a tiny bit more detail here?
---> at least once a month.
+--> at least once a month by radio triangulation.
+Most of the fieldwork was conducted from the ground, but to avoid loss of data, the birds were also triangulated from helicopters in May, September and November.
 During the breeding and chick-rearing season (May to July) birds were monitored more often, and during December and January we obtained few observations due to challenging field work conditions.
@@
 @nilsen2023 recently developed an integrated distance sampling model (IDSM) which jointly analyses line transect and radio-telemetry data and applied it to willow ptarmigan in the Western part of Lierne municipality in Norway.
-The model consists of a deterministic population model with two age classes (juveniles and adults) and four data likelihoods: 1) likelihood for observation distances from transect lines for estimating detection probability; 2) likelihood for age-specific counts on transect surveys for estimating numbers of juveniles and adults present; 3) likelihood for juvenile to adult ratios observed at the locality level to provide estimate recruitment rate (as juveniles/adult); and 4) likelihood for known-fate telemetry data to estimate seasonal and annual survival.
+The model consists of a deterministic population model <!-- I think this depends on defenitions.
+It did contain temporal variation in recruitment, which is in fact environmental stochasticity.
+It did not contain demographic stochasticity --> with two age classes (juveniles and adults) and four data likelihoods: 1) likelihood for observation distances from transect lines for estimating detection probability; 2) likelihood for age-specific counts on transect surveys for estimating numbers of juveniles and adults present; 3) likelihood for juvenile to adult ratios observed at the locality level to provide estimate recruitment rate (as juveniles/adult); and 4) likelihood for known-fate telemetry data to estimate seasonal and annual survival.
 Below, we describe our new extension of this model to include data from several as opposed to just one.
@@
 The pink "x"'s indicate the added dimension for area.
-Juv\[x,t\] = juveniles in year t.
-Ad\[x,t\] = adults in year t.
-R\[x,t\] = recruitment rate in year t.
-S\[x,t\] = survival probability from year t to t+1.
+Juv\[x,t\] = juveniles in area x year t.
+Ad\[x,t\] = adults in area x in year t.
+R\[x,t\] = recruitment rate in area x in year t.
+S\[x,t\] = survival probability from year t to t+1 in area x.
 Note that the additional site (=transect) dimension, "j", is omitted for the sake of illustration.](Figures/RypeIDSM_multiArea_structure.png){#fig-model}
@@
 
-The three likelihoods for data resulting from the line transect sampling (observation distances, age-specific counts, and juvenile to adult ratios, see above) were also formulated as spatially explicit, with year- and area-specific distance sampling detection parameters modelled in the same way as survival and recruitment (bar the effect of rodent occupancy, @fig-model).
-For the known-fate telemetry data (and the seasonal decomposition of survival estimated from it), on the other hand, we did not add an additional area dimension as this data was only available for the Lierne area.
+The three likelihoods for data resulting from the line transect sampling (observation distances, age-specific counts, and juvenile to adult ratios, see above) were also formulated as spatially explicit, with year- and area-specific distance sampling detection parameters modelled in the same way as survival and recruitment (except the effect of rodent occupancy, @fig-model).
+For the known-fate telemetry data (and the seasonal decomposition of survival estimated from it), on the other hand, we did not add an additional area dimension as this data was only available for on study area (the Lierne area).
 
@@
 Following model fitting, we calculated posterior distributions for the proportions of variance in survival probabilities, recruitment rates, and detection decay explained by 1) spatial variation ($var_{area}$), 2) temporal variation ($var_{year}$), 3) residual variation ($var_{residual}$), and 4) variation in rodent occupancy ($var_{rodent}$).
-To obtain the proportion variance explained by each of the component, we divided it by the sum of all other components ($var_{area} + var_{year} + var_{residual} + var_{rodent}$).
+To obtain the proportion variance explained by each of the component, we divided it by the sum of all the components ($var_{area} + var_{year} + var_{residual} + var_{rodent}$).
 The spatial, temporal, and residual variance components were defined as the square of the estimated corresponding random effects standard deviation from the model while $var_{rodent}$ was calculated as the variance of all area- and year-specific $\beta_x*rodentOcc_{x,t}$ products.
+<!-- We might need to justify the latter a bit, as the rodent variance is calculated different than the other variances -->
 
@@
 
-Reproducibility and ease of repeating analyses was a key focus when developing the multi-area IDSM.To that end, we set up the workflow as a "targets pipeline", implemented through the R package "targets" [@landau2021].
+Reproducibility and ease of repeating analyses was a key focus when developing the multi-area IDSM.
+To that end, we set up the workflow as a "targets pipeline", implemented through the R package "targets" [@landau2021].
 The pipeline contains a variety of options for controlling modelling decisions in the workflow such as the year range of data to consider, the level of spatial aggregation (i.e.
@@
 During this period, estimated population densities varied between `r dens.min$post.sum` birds/km^2^ in the area "`r dens.min$area.name`" close to the coast in central Norway to `r dens.max$post.sum` birds/km^2^ in "`r dens.max$area.name`" further south near the Swedish border.
-In general, population density appeared to be lowest in the Northern parts of Norway and highest in the eastern part of central Norway (@fig-densA).
-Uncertainty in density estimates was relatively consistent, with a few areas (inlcuding the one with the lowest estimated density, "`r dens.min$area.name`") sticking out by having substantially less precise estimates (@fig-densB).
+In general, during our study period the population density appeared to be lowest in the Northern parts of Norway and highest in the eastern part of central Norway (@fig-densA).
+Uncertainty in density estimates was relatively consistent, with a few areas (including the one with the lowest estimated density, "`r dens.min$area.name`") sticking out by having substantially less precise estimates (@fig-densB).
 Populations fluctuated substantially over time in any given area (SF "TimeSeries_popDens1.pdf") and some years seemed to be indicative of relatively high (e.g.
@@
 The highest population growth rates were estimated for areas with relatively low population densities across latitudes but we did not find evidence for a strong association between population growth rates and population densities overall ([@fig-paramCorrs] A).
+<!-- Indicate that this is just for the four most recent years.
+Same for figure below -->
 
@@
 
-The MCMC chains for many of the area-specific average survival probabilities and recruitment rates, as well as for the global averages (intercepts) for both vital rates, were mixing rather poorly.
+The MCMC chains for many of the area-specific average survival probabilities and recruitment rates, as well as for the global averages for both vital rates, were mixing rather poorly.
 Despite that, mixing was good and resulting posteriors well defined for the area- and year-specific estimates of survival and recruitment (SF "PostDens_tS_tR.pdf").
 There was substantial variation in both vital rates across time (SFs "TimeSeries_pSurv.pdf" and "TimeSeries_rRep.pdf").
-In many areas, the years 2011, 2014, and 2018 not only sported high population densities (see above) but were also characterized by both high recruitment and low subsequent survival.
-The overall low density years 2012 and 2015, conversely, often featured lower recruitment and, in some cases, higher survival.
+In many areas, the years 2011, 2014, and 2018 not only supported high population densities (see above) but were also characterized by both high recruitment and low subsequent survival.
+<!-- Here and above: I would perhaps consider deleting the reference to 2011 as a "high density year", as there is really a lot of spatial variation in this pattern.
+I'm not so convinced that is the case in general --> The overall low density years 2012 and 2015, conversely, often featured lower recruitment and, in some cases, higher survival.
 Notably, there were also years with very little spatial synchrony, i.e.
@@
 Nonetheless, spatial variation in the rodent effect was substantial (random effect SD = `r betaR.sd.spatial$post.sum`).
-This resulted in negative median effects in `r n.neg.betaR` areas, positive median effects in `r n.pos.betaR` areas, and a range of effect sizes from `r betaR.min$post.sum` (area "`r betaR.min$area.name`") to `r betaR.max$post.sum` (area "`r betaR.max$area.name`", @fig-paramCorrs C, SF "Rep_betaR.R.pdf").
+This resulted in negative (median) effects in `r n.neg.betaR` areas, positive (median) effects in `r n.pos.betaR` areas, and a range of effect sizes from `r betaR.min$post.sum` (area "`r betaR.min$area.name`") to `r betaR.max$post.sum` (area "`r betaR.max$area.name`", @fig-paramCorrs C, SF "Rep_betaR.R.pdf").
 The largest positive rodent effects were estimated for areas in the very North of Norway, as well as in the mountaineous regions in the central and southwestern parts of the country (SF "betaR_Map.pdf").
 Effects with negative posterior medians were located mostly at intermediate latitudes, but we note that all of these had posterior distributions featuring substantial overlap with 0 (@fig-paramCorrs C).
+<!-- we sometimes refer to rodent abundance, sometimes occupancy.
+I think what we have is most correctly refered to as an index of abundance -->
 
@@
 Space-for-time substitution also provided the statistical power necessary for the analysis to provide evidence for the alternative prey hypothesis, i.e.
-ptarmigans benefiting from high abundance of alternative prey for their predators.
+ptarmigans benefiting from high abundance of alternative rodent prey for their predators.
 Taken together, the results highlight the potential of integrating demographic data across large spatial scales in the contexts of both informing management and creating biodiversity indicators for higher-level reporting.

### 2024-01-19 Manuscript_Nater_etal.qmd
Minor edit to text

@@
 The fact that these same years also match up with observed peaks in rodent abundance in many areas provides evidence for the Alternative Prey Hypothesis (APH; [@Israelsen2020]).
-The AHP stipulates that high abundance of alternative prey (rodents, in this case) for common predators leads to population growth, and is well-supported throughout the literature for a range of taxa [e.g., @hagen1952; @Kjellander_Nordström_2003; @Reif_Tornberg_Jungell_Korpimäki_2001], including willow ptarmigan [@bowler2020].
+The APH stipulates that high abundance of alternative prey (rodents, in this case) for common predators leads to population growth, and is well-supported throughout the literature for a range of taxa [e.g., @hagen1952; @Kjellander_Nordström_2003; @Reif_Tornberg_Jungell_Korpimäki_2001], including willow ptarmigan [@bowler2020].
 Conversely, @Nyström_etal_2006 suggested that gyrfalcons (a specialized ptarmigan predator) did not respond to rodent populations and did not switch to alternative prey when ptarmigan populations are low.
-Nonetheless, other predators such as fox could prey switch in response to changing availability of rodents and ptarmigan as shown by @Breisjøberget_etal_2018 and @bowler2020.
+Nonetheless, other predators such as red fox could prey switch in response to changing availability of rodents and ptarmigan as shown by @Breisjøberget_etal_2018 and @bowler2020.
 However, statistical associations in our study and others is not causal and alternative explanations for this pattern are plausible.
@@
 The work presented above is based on versions 1.7, 1.8, and 1.12 for the datasets from Fjellstyrene (key b49a2978-0e30-4748-a99f-9301d17ae119), Statskog (key 6a948a1c-7e23-4d99-b1c1-ec578d0d3159), and FeFo (key c47f13c1-7427-45a0-9f12-237aad351040), respectively.
+<!-- I thinbk it is better to cite the datasets (with doi and all that) rather than the key.
+That way, the citation and data use will be tracked by GBIF (and Living Norway) -->
 

### 2024-01-22 Manuscript_Nater_etal.qmd
Edits and text addition to the discussion. Additional references added to .bib file

@@
 
+<!-- ERLEND: I'm not sure if we want to go this way.
+The density the last few years is hard to interpret, given that there is so much temporal variation in densities.
+Also, growth during last four years might very well be related to cycle phase rather than spatial variation in demography.
+If we had included data until 2023, the density in Northern Norway would have come out as much higher, whereas many areas further south had very low density.
+I suggest we keep this part to a minimum, as there are many other intersting aspects to focus on -->
+
 <!-- Erlend will probably have better ideas than I.-->
@@
 Density dependence is an important process to understand in ecology but more so for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; @aanes2002; @willebrand2001].
-Negative density dependence has been found in numerous gallinaceous birds such as northern bobwhites *Colinus virginianus* [@mcconnell2018], *Perdix perdix* [@bro2003], and wild turkeys *Meleagris gallopavo* [@mcghee2007]).
+Negative density dependence has been found in several gallinaceous birds such as northern bobwhites *Colinus virginianus* [@mcconnell2018], *Perdix perdix* [@bro2003], and wild turkeys *Meleagris gallopavo* [@mcghee2007]).
 While we found the highest population growth rates in areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A], our analysis did not provide evidence for a strong relationship between density and growth rates across space.
@@
 Conversely, @pedersen2004 found strong overwinter negative density dependence and they posited that immigration was the vital rate that responded to changes in density.
+In @bowler2020, density dependence was included as a Gompertz form of density dependence (i.e.
+linear relationship betwee log popualtion size and log($\lambda$)), but without specifying which vital rate was affected by density dependence.
 Low survival rates appeared to have a one-year lag following abundance peaks.
 This pattern could be a result of negative density dependence that went undetected when looking at the ensemble of samples across the 41 areas.
-Negative density dependence has been shown in willow ptarmigan populations which may explain explain reduced survival following years with elevated densities [@pedersen2004].Future work should explore different time-lags for density and population growth and the possibility for non-linear effects.
+<!-- A suggested mechanism: If abundance peaks are related to years with high recruitment and high rodent abundance (in general), then predation on all life stages might be higher the year after due to chrash in the roden population.
+This is an indirect form for dd, but might still manifest as dd in the data.
+--> Negative density dependence has been shown in willow ptarmigan populations which may explain explain reduced survival following years with elevated densities [@pedersen2004].
+Future work should explore different forms of density dependence, diferent time-lags, and the possibility for non-linear effects.
 
@@
 
+<!-- ERLEND: In comparative studies, my impression is that there is a "faster life" at higher latitudes, but I did not find much about this when it comes to intraspecific comparisons -->
+
 <!--# Yeah I wasn't able to find much. -->
@@
 Some of the resulting "high density years" were highly synchronized across large spatial scales, such as the years 2011, 2013, and 2018.
-Taking a closer look, we find that these are years that are characterized by high recruitment (SF "TimeSeries_rRep.pdf") and low subsequent survival (SF "TimeSeries_pSurv.pdf"), which often resulted in steep population declines towards the following year.
+Taking a closer look, we find that these are years that are characterized by high recruitment (SF "TimeSeries_rRep.pdf"), follwef by a low survival the year after (SF "TimeSeries_pSurv.pdf").
+This often resulted in steep population declines towards the following year.
 The fact that these same years also match up with observed peaks in rodent abundance in many areas provides evidence for the Alternative Prey Hypothesis (APH; [@Israelsen2020]).
@@
 Nonetheless, other predators such as red fox could prey switch in response to changing availability of rodents and ptarmigan as shown by @Breisjøberget_etal_2018 and @bowler2020.
-However, statistical associations in our study and others is not causal and alternative explanations for this pattern are plausible.
+However, statistical associations in our study and others might not necessarily show causal relationships, and alternative explanations for this pattern are plausible.
 The positive association between rodents and ptarmigan could be caused by bottom-up processes such as food availability unrelated to predation.
 Future work should develop causal models and manipulative experiments to establish causal relationships among top-down and bottom-up processes.
+<!-- I sugest we delete the last sententence.
+It's not that I disagree, but this is an issue that has recieved so much attention, and there are indeed some experiments (in particular for the rodent cycling, but also grouse in UK).
+So if we go deeper into this discussion, we need to write a paragraph about htat as well.
+-->
 
@@
 <!-- This previous sentence is adapted from a suggestion by James, but I feel like we need some more explanation/arguments here to clarify this.
-Perhaps also some other and/or more specific hypotheses= --> Notably, the conclusive estimation of overall positive effects of rodent abundance on recruitment in our model was only possible thanks to the integration and sharing of data across multiple areas.
+Perhaps also some other and/or more specific hypotheses= -->
+
+<!-- ERLEND: I'm not sure about the gyrfalcon, but might be a contributing factor.
+In Bowler at al.
+we discuss the fact that the predator community tend to be larger in warmer areas (more species), including higher abundance of generatlists that migth persist at high densities also in years with few rodents.
+I think that might be a (more) plausible explanation.
+-->
+
+Notably, the conclusive estimation of overall positive effects of rodent abundance on recruitment in our model was only possible thanks to the integration and sharing of data across multiple areas.
 When @nilsen2023 fit the IDSM to data from only a single area, they were unable to obtain a reliable estimate for the rodent effect due to limited statistical power.
 Consequently, the space-for-time substitution that comes with extending the model across multiple area allows estimation of covariate effects that otherwise cannot be estimated, and opens up for future possibilities for studying effects of not just rodents, but also other environmental drivers on ptarmigan population dynamics.
+In previous work, both based on marked birds [@Eriksen2023] and unmarked birds [@bowler2020; @Henden2020; @Novoa2016], the spring conditions has come out as an important predictor of ptarmigan recruitment rates.
+In general, it is found warmer and earlier springs result in earlier breeding, larger clutch sizes [@Eriksen2023] and with resulting higher recruitment rates measured in the late summer and early fall (as in our study).
+In @bowler2020, it was reported that the strength of this relationship was not consistent in time and space, but was generally stronger in colder areas, similar to the effect of rodent abundance reported here.
+
 For example, the cover of ericaceous shrubs as a proxy for food availability or spatially-explicit spring green up dates both measured using remote sensing are ecologicaly plausible and feasibly acquired factors.
@@
 Also, immigration, included in our recruitment term <!--# this true?
+--> <!-- No,this is not the case.
+recruitment is measured in the late summer (august), before the birds start dispersing.
+It might be affected by very local immigration (100's of meeters), but not on a meaningful scale.
+Our survival rates on the other hand are indeed "apparent survival" and does include migration in and out of the sampling area.
 -->, has been hypothesized to be a spatial process [@kvasnes2015].
@@
 Management decisions made at the resolution of large geo-political boundaries (e.g., Norway) run a high risk of being inadequate when there is substantial spatial variation in demographic processes and population dynamics, as is the case for willow ptarmigan.
-In Norway, willow ptarmigan -- and small game in general -- is managed \[...\] <!-- Brief summary of how and at what spatial scales patrmigan/small game management works in Norway --> The results from our study suggests that management decisions should be spatially refined and ideally informed by knowledge about the "local" population processes.
+In Norway, willow ptarmigan -- and small game in general -- is managed at the local and/or regional scale, with rather limited national regulation beyond updating the season length every fourth year.
+In effect, both the management system, regulation type (quota type, season length, number of licences, bag limit etc) and quota size is decided by the local or regional stakeholders.
+Thus, while national estimates (abundance and/or temporal trend in abundance) might be important for red listing decisions and for setting the maximum season length, remaining decisions about harvest management are taken locally.
+<!-- I will add some relevant references in next iteration.
+We have no "go to" reference for the management system (we should write one), but I will see if I can piece together something.
+-->  [...] <!-- Brief summary of how and at what spatial scales patrmigan/small game management works in Norway --> The results from our study suggests that management decisions should be spatially refined and ideally informed by knowledge about the "local" population processes.
 \[...\] <!-- Some specific implications of the results for management of hunting and of habitat (e.g.
-better managed hunting is good because we do see increases in majority of local populations, but the mountain areas have had recent declines that may call for additional local measures, these measures may need to be adjusted since the mountain populations appear to have slower life histories and likely different sensitivities, etc.) -->
+better managed hunting is good because we do see increases in majority of local populations, but the mountain areas have had recent declines that may call for additional local measures, these measures may need to be adjusted since the mountain populations appear to have slower life histories and likely different sensitivities, etc.) ERLEND: As stated above, I think we should downplay the short-term local trends and difference in abundace, as these are hypervariable and might have given a very different picture had we included the last 2 years.
+We can still keep the main points; managers need to respond to local dymamics (and thus need local data or predictions), without pointing to specific areas that need specific concern (as this has allready changed somewhat since 2021).
+-->
 
@@
 <!-- Funnily enough, I have been suggesting to Erlend that we should have a master student to analyze the effect of dog breed on detection probability :-P -->
+
+<!-- Only speculation, and no references to show to.
+But much of the willow ptarmigan habiat in Northern Norway is located in birch forest, so perhaps the habitat is more closed compared to souther norway ptarmigan habitat?
+Second, if (not sure if this is the case) the habitat is more rugged (as might be on the coast in Northern Norway) detections are probably lower.
+Third - clusters are smaller in North (ref lower recruitement), which might affect detection probabilities.
+-->
 
@@
 If misclassification happened at random, mixture models could be used to determine the likely age class of individuals to whom no age class was assigned during observations (+ref).
-In our case, however, juveniles are likely more prone to misclassification due to \[...\] <!-- @ERLEND can you add reasons?
--->.
+In our case, we might suspect that it is more likely to classify an adult bird as juvenile rather than the other way around, or to classify it as "unknown" age.
+One reason for this might be that the observers look for specific signs (size, male sound) when classifying a bird as adult, and might classify it as juvenile or unknown if the signs are not clearly detected.
 Future studies should investigate to what degree available information on e.g.
@@
 The line transect survey data included in this study constitutes just the publicly available part of the data collected through "Hønsefuglportalen" but the programme also includes additional surveys on private land.
-\[...\] <!-- @ERLEND: Can we add a very brief ocverview over what the coverage (in terms of municipalities) would be if we had access to all the Hønsefuglportalen data?
+Including also private land would provide better coverage in souh-eastern and southern parts of Norway, partly from areas where there is very limited data from public land.
+<!-- @ERLEND: Can we add a very brief ocverview over what the coverage (in terms of municipalities) would be if we had access to all the Hønsefuglportalen data?
 --> Exploring to what degree additional data from Hønsefuglportalen could be included in future studies employing an extended IDSM with additional spatial structuring is therefore a worthwhile endeavor.
@@
 Nonetheless, this illustrates a way for how smaller datasets from single or subsets of areas can be integrated into a large-scale modelling framework.
-Other relevant data could be included using the same approach [...] <!-- @ERLEND: Any thoughts on what the status here is?
+Other relevant data could be included using the same approach would be to integrate data from ongoing nesting success monitoring (Nilsen, E.B.
+unpublished data), data from past studies of marked birds \[@sandercock_etal2011\], and data from other monitoring programs for breeding birds based on point counts (see e.g.
+the Norwegian Breeding Bird Monitoring: <https://hekkefuglovervakingen.nina.no/>) [...] <!-- @ERLEND: Any thoughts on what the status here is?
 Might there be other data available from Lierne and/or beyond (other telemetry data, nest monitoring, etc.?
@@
 --> The most relevant source of data to be included into the IDSM framework in the near future, however, is harvest data.
+Such data might be available with two different spatial and temporal resolution.
+First, at the municipality level there are data with national coverage collected annually by Statistics Norway (<https://www.ssb.no/>).
+Second, many public land owners have data with much higher temporal (daily) and spatial (harvest area) resolution, including both harvest bags and harvest effort (number of hunters pr area pr day).
 <!-- @ERLEND 1-2 sentence summary of what kind of harvest data is available, and at what spatial covarage and resolution --> As the IDSM framework is, in essence, an IPM, harvest can be modelled through partitioning of survival into cause-specific mortality in the process model and inclusion of relevant harvest data likelihoods [e.g.
@@
 For example, the strength and shape of density dependence and how harvest may induce population sinks because of Allee effects [@kvasnes2015]).
-Modelling allee effects, or better yet predicting where they are likely to occur if over-harvest occurs, requires a spatially-explicit understanding of harvest and its impact on local populations.
+Modelling allee effects, or better yet predicting where they are likely to occur if over-harvest occurs, requires a spatially-explicit understanding of harvest and its impact on local populations [@kvasnes2015].
 Because harvest effort, thus rates, can vary dramatically over time and space [@breisjøberget2018] easily repeatable analyses that can be done continually would prevent prolonged over-exploitation.

### 2024-01-25 Manuscript_Nater_etal.qmd
Edited the abstract

@@
 
-Quantifying the temporal and spatial variation in animal demography is a central theme of ecology and important for directing policy.
-However, field sampling at large spatial extents for long periods of time is cost prohibitive and often politically untenable.
-Citizen science programs can alleviate these issues by harnessing stakeholders to increase the spatial and temporal resolution of datasets.
-Yet, most citizen science programs lack data from marked animals making it difficult to model demography.
-Integrated population models (IPM) offer promise to use multiple streams of data on populations or multiple aspects of a single dataset to estimate key vital rates for populations or communities.
-We leveraged a large-scale citizen science monitoring program---the "Hønsefuglportalen"--- to estimate spatial and temporal variation in demography for willow ptarmigan (*Lagopus lagopus*).
-The core field sampling approach was line-transect distance sampling with age-specific detections---the latter critical to modelling the population dynamics in the IPM.
-We explored latitudinal and longitudinal variation in recruitment, survival, and density; patterns of density dependence; and the relationship between recruitment and rodent occupancy using an IPM in a Bayesian framework following open science principles.
-Overall results were consistent with expectations of ecological theory.
-Slower life history patterns were observed at higher latitudes and altitudes portending differential effects of climate change on ptarmigan across their range.
-Observed patterns between rodents and recruitment were consistent with the alternative prey hypothesis (APH); however, spatial variation in the directionality and magnitude of effect were substantial.
-The robustness of the hierarchical structure of our model was instrumental in elucidating these patterns.
-Our study highlights the power of citizen science, IPMs, and open science principles to estimate spatial and temporal variation in the demography for a culturally important species.
+Quantifying temporal and spatial variation in animal population size and demography is a central theme in ecological research and important for directing management and policy.
+However, this requires field sampling at large spatial extents and over long periods of time, which is not only prohibitively costly but also often politically untenable.
+Citizen science programs can alleviate these constraints by recruiting stakeholders and the public to increase the spatial and temporal resolution of monitoring effort and hence resulting data.
+While the majority of citizen science programs are limited by opportunistic sampling designs, we are starting to see the emergence of structures citizen science programs that employ trained volunteers to collect data according to standardized protocols.
+Simultaneously, there is much ongoing development of statistical models that are increasingly more powerful and able to make more efficient use of field data.
+Integrated population models (IPMs), for example, are able to use multiple streams of data from different field monitoring programmes and/or multiple aspects of single datasets to estimate population sizes and key vital rates.
+Here, we developed a multi-area version of a recently developed integrated distance sampling model (IDSM) and applied it to data from a large-scale citizen science monitoring program---the "Hønsefuglportalen"--- to estimate spatial and temporal variation in population dynamics of willow ptarmigan (*Lagopus lagopus*) in Norway.
+We set up an open and reproducible workflow for exploring temporal, spatial (latitudinal, longitudinal, altitudinal), and residual variation in recruitment, survival, and population density, as well as relationships between vital rates and relevant covariates and signals of density dependence.
+Recruitment rates varied more across space than over time, while the opposite was the case for survival.
+Slower life history patterns (higher survival, lower recruitment) appeared to be more common at higher latitudes and altitudes, portending differential effects of climate change on ptarmigan across their range.
+While there was variation in the magnitude of the effect small rodent abundance had on recruitment, the relationships were predominantly positive and thus consistent with the alternative prey hypothesis, which predicts higher reproductive output when there is plenty of alternative prey for shared predators.
+Notably, the accurate estimation of covariate effect was only made possible by integrating data from several monitoring areas for analysis.
+Our study thus highlights the potential of citizen science monitoring and integrated modelling approaches for estimating and understanding spatio-temporal patterns species abundance and demographic rates, and showcases how corresponding workflows can be set up in a reproducible and semi-automated way that increases their usefulness for informing management and regular reporting towards national and international biodiversity frameworks.
 

### 2024-01-25 Manuscript_Nater_etal.qmd
Revised manuscript

--- 2024-01-19 Manuscript_Nater_etal.qmd
@@
 
-Quantifying the temporal and spatial variation in animal demography is a central theme of ecology and important for directing policy.
-However, field sampling at large spatial extents for long periods of time is cost prohibitive and often politically untenable.
-Citizen science programs can alleviate these issues by harnessing stakeholders to increase the spatial and temporal resolution of datasets.
-Yet, most citizen science programs lack data from marked animals making it difficult to model demography.
-Integrated population models (IPM) offer promise to use multiple streams of data on populations or multiple aspects of a single dataset to estimate key vital rates for populations or communities.
-We leveraged a large-scale citizen science monitoring program---the "Hønsefuglportalen"--- to estimate spatial and temporal variation in demography for willow ptarmigan (*Lagopus lagopus*).
-The core field sampling approach was line-transect distance sampling with age-specific detections---the latter critical to modelling the population dynamics in the IPM.
-We explored latitudinal and longitudinal variation in recruitment, survival, and density; patterns of density dependence; and the relationship between recruitment and rodent occupancy using an IPM in a Bayesian framework following open science principles.
-Overall results were consistent with expectations of ecological theory.
-Slower life history patterns were observed at higher latitudes and altitudes portending differential effects of climate change on ptarmigan across their range.
-Observed patterns between rodents and recruitment were consistent with the alternative prey hypothesis (APH); however, spatial variation in the directionality and magnitude of effect were substantial.
-The robustness of the hierarchical structure of our model was instrumental in elucidating these patterns.
-Our study highlights the power of citizen science, IPMs, and open science principles to estimate spatial and temporal variation in the demography for a culturally important species.
+Quantifying temporal and spatial variation in animal population size and demography is a central theme in ecological research and important for directing management and policy.
+However, this requires field sampling at large spatial extents and over long periods of time, which is not only prohibitively costly but also often politically untenable.
+Participatory monitoring programs (also called citizen science programmes) can alleviate these constraints by recruiting stakeholders and the public to increase the spatial and temporal resolution of sampling effort and hence resulting data.
+While the majority of participatory monitoring programs are limited by opportunistic sampling designs, we are starting to see the emergence of structures citizen science programs that employ trained volunteers to collect data according to standardized protocols.
+Simultaneously, there is much ongoing development of statistical models that are increasingly more powerful and able to make more efficient use of field data.
+Integrated population models (IPMs), for example, are able to use multiple streams of data from different field monitoring programmes and/or multiple aspects of single datasets to estimate population sizes and key vital rates.
+Here, we developed a multi-area version of a recently developed integrated distance sampling model (IDSM) and applied it to data from a large-scale participatory monitoring program -- the "Hønsefuglportalen" -- to estimate spatial and temporal variation in population dynamics of willow ptarmigan (*Lagopus lagopus*) in Norway.
+We set up an open and reproducible workflow for exploring temporal, spatial (latitudinal, longitudinal, altitudinal), and residual variation in recruitment, survival, and population density, as well as relationships between vital rates and relevant covariates and signals of density dependence.
+Recruitment rates varied more across space than over time, while the opposite was the case for survival.
+Slower life history patterns (higher survival, lower recruitment) appeared to be more common at higher latitudes and altitudes, portending differential effects of climate change on ptarmigan across their range.
+While there was variation in the magnitude of the effect small rodent occupancy had on recruitment, the relationships were predominantly positive and thus consistent with the alternative prey hypothesis, which predicts higher reproductive output when there is plenty of alternative prey for shared predators.
+Notably, the accurate estimation of covariate effect was only made possible by integrating data from several monitoring areas for analysis.
+Our study thus highlights the potential of participatory monitoring and integrated modelling approaches for estimating and understanding spatio-temporal patterns species abundance and demographic rates, and showcases how corresponding workflows can be set up in a reproducible and semi-automated way that increases their usefulness for informing management and regular reporting towards national and international biodiversity frameworks.
 
 # Introduction
-
-<!-- @JAMES & ERLEND: First draft is relatively "compact". Input on extending it and/or addition of relevant references appreciated -->
 
@@
 This latter requirement is particularly crucial for (infrastructure) development strategies and for species management and conservation, both of which tend to require knowledge on species abundance and population dynamics (i.e.
-demographic rates) that is relevant for their county- or municipality-level decision making (+ refs?).
+demographic rates) that is relevant for their county- or municipality-level decision making [@christie2020].
 Another reason why abundance and population indicators ideally come with good spatial resolution is that there can be substantial amounts of variation in population dynamics and life history of species across space [e.g.
@@
 Consequently, many countries have been working on setting up, maintaining, and improving such monitoring programmes over the last decades.
-Many now well-established programmes focus on breeding birds and butterflies, and examples include the North American Breeding Bird Survey (<https://www.pwrc.usgs.gov/bbs/>), the PanEuropean Common Bird Monitoring Scheme (<https://pecbms.info/>[- PECBMS](https://pecbms.info/)), the UK Butterfly Monitoring Scheme ([https://ukbms.org/](#0){.uri style="font-size: 11pt;"}), the Game and Wildlife Conservation Trust Partridge Count Scheme [@aebischer2010], and the Swiss Biodiversity Monitoring ([https://www.biodiversitymonitoring.ch/](#0){.uri style="font-size: 11pt;"}).
-<!-- is there any other one we want to name here specifically?
--->
+Many now well-established programmes focus on breeding birds and butterflies, and examples include the North American Breeding Bird Survey (<https://www.pwrc.usgs.gov/bbs/>), the PanEuropean Common Bird Monitoring Scheme (<https://pecbms.info/>), the UK Butterfly Monitoring Scheme (<https://ukbms.org/>), the Game and Wildlife Conservation Trust Partridge Count Scheme [@aebischer2010], and the Swiss Biodiversity Monitoring (<https://www.biodiversitymonitoring.ch/>).
 
 There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly, requires trained specialists, and depends on at least some top-down control of the program.
-This often limit the amount of data that can be collected, while participatory monitoring, i.e.
+This often limits the amount of data that can be collected, while participatory monitoring, i.e.
 the collection of ecological data by members of the public [also called citizen or community science, @fraisl2022], allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias [@johnston2023].
 Consequently, many large-scale monitoring programmes are often limited to presence-absence or very simple count observations, making them suitable for the development of indicators of species distributions and perhaps population trends, but usually not of abundance, population dynamics, and demographic rates (+ example refs).
-The exception here are monitoring programmes that succeed in making use of a large number of citizen scientists that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol.
+The exception here are monitoring programmes that succeed in making use of a large number of volunteers that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol.
 For example, in the United States hunters participate in the collection of bands and wings from harvested American Woodcock (*Scopolax minor)* to estimate survival and age ratios [@zimmerman2010].
 At the European level, the recently established initiative "European Observatory of Wildlife" are offering common field- and analyses protocols and aim to establish a network of "observation points" to monitor wildlife populations at the European level (<https://wildlifeobservatory.org/>).
-<!-- can we think of 1-2 internationally well-known examples?
-Check out MAPS (Monitoring of Avian Productivity and Survivorship) and CES (Constant Effort Sites) --> In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", [https://honsefugl.nina.no/Innsyn/en](#0){.uri style="font-size: 11pt;"}).
+In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", [https://honsefugl.nina.no/Innsyn/en](#0){.uri style="font-size: 11pt;"}).
 It is a line transect survey programme carried out annually in \>120 localities across the country (\>2000 transects) by trained volunteers using pointing dogs.
@@
 The line transect data from "Hønsefuglportalen" has been used previously for estimating abundance trends of willow ptarmigan (*Lagopus lagopus*) across Norway [e.g.
-@bowler2020; @nilsen2020], and to test relevant ecological hypotheses [@bowler2020; @Breisjøberget_etal_2018].
-However, large-scale estimation of demographic rates underlying these trends has thus far remained as untapped potential of the dataset.
-@nilsen2023 recently developed a novel integrated distance sampling model (IDSM) which successfully uses the age of idividuals detected along line transects data coupled with radio-telemetry data to estimate not just abundance, but also survival and recruitment across years.
+@bowler2020; @nilsen2020], and to test a range of relevant ecological hypotheses [@bowler2020; @Breisjøberget_etal_2018].
+However, large-scale estimation of demographic rates underlying abundance trends has thus far remained as untapped potential of the dataset.
+@nilsen2023 recently developed a novel integrated distance sampling model (IDSM) which successfully uses the age of individuals detected along line transects data coupled with radio-telemetry data to jointly estimate abundance, survival, and recruitment across years.
 In this study, we adapt and extend the model of @nilsen2023 to run not just on a single site but on all areas with publicly available line transect data from "Hønsefuglportalen" simultaneously.
@@
 @horswill2019; @morrison2022].
-We then apply the resulting multi-area IDSM to "Hønsefuglportalen" data on willow ptarmigan to estimate population size, age-structure, survival, recruitment, and impacts of small rodent abundance across 41 reporting districts and 15 years (2007-2021) for this popular small-game species.
+We then apply the resulting multi-area IDSM to "Hønsefuglportalen" data on willow ptarmigan to estimate population size, age-structure, survival, recruitment, and impacts of small rodent occupancy across 41 reporting districts and 15 years (2007-2021) for this culturally important small-game species.
 We further embed the modelling workflow in a reproducible, semi-automated pipeline that will greatly facilitate the repeated calculation of abundance and population indicators at different spatial scales as new data are added every year.
@@
 The main reason for the long-term decline in abundance remain unresolved, but the willow ptarmigan are considered sentinel species that are sensitive to both climate change and land use changes [@Henden2017; @Storch2007].
-Moreover, being one of only a handful of bird species that spend the winter in mountain ecosystems in Scandinavia, they are important components of the ecosystem as prey species for the gyrfalcon [@Alastair2020] <!-- do we have other "important" species to add here.
-Arctic fox?
-Some other endangered beasts perhaps?
-ERLEND: Well; there are more mammals, as they typically do not migrate long distances to other continents....
-I suggest we leave it like this -->.
-The willow ptarmigan has a relatively fast pace of life [@sandercock2005; @Steen1996], and can display substantial spatio-temporal variation in demographic rates [@Bowler2020].
-Their population dynamics are characterized by large inter-annual fluctuations in abundance [@hjeljord2022], and previous research has tied these fluctuations to rodent cycle through shared predators [@hagen1952; @Bowler2020].
-This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis (APH) and has been a central part of Fennoscandian research on grouse population dynamics for many decades [@Elton1942; @hagen1952; @Linden1988; @Steen1985].
+Moreover, being one of only a handful of bird species that spend the winter in mountain ecosystems in Scandinavia, they are important components of the ecosystem as prey species for resident predators, such as the gyrfalcon [@Alastair2020].
+The willow ptarmigan has a relatively fast pace of life [@sandercock2005; @Steen1996], and can display substantial spatio-temporal variation in demographic rates [@bowler2020].
+Their population dynamics are characterized by large inter-annual fluctuations in abundance [@hjeljord2022], and previous research has tied these fluctuations to rodent cycle through shared predators [@hagen1952; @bowler2020].
+This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis (APH) and has been a central part of research on Fennoscandian grouse population dynamics for many decades [@Elton1942; @hagen1952; @Linden1988; @Steen1985].
 In addition, spring weather conditions and phenology is known to have considerable effects on breeding success and recruitment rates [@Eriksen2023; @Steen1985].
@@
 The latter means that information about spatio-temporal variation in demographic rates and population dynamics is particularly important in order to design sustainable harvest strategies [@Eriksen2023].
-In addition, as a sentinel species the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is included in both the main national biodiversity [Nature Index for Norway, @jakobsson2020, <https://www.naturindeks.no/Indicators/lirype>] and ecosystem [Assessment of the Ecological Condition, @framstad2022] assessments.
+In addition, being a sentinel species, the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is included in both the main national biodiversity [Nature Index for Norway, @jakobsson2020, <https://www.naturindeks.no/Indicators/lirype>] and ecosystem condition [Assessment of the Ecological Condition, @framstad2022] assessments.
 
@@
 To increase the detection probability, each team of fieldworkers use a pointing dog to locate the birds.
-Each survey team typically consist of at least two persons (one dog handler and one person responsible for following the transect line) and one dog.
+Each survey team typically consist of at least two people (one dog handler and one person responsible for following the transect line) and one dog.
 Often, more than one dog is used for a survey, but only one dog should be used at a time.
@@
 When the surveys are conducted in August, the chicks of the year are able to fly but can be distinguished from older birds as they are still of smaller body size.
-Since 2019, most of the data has been collected using a mobile app tailored made for the monitoring program, where the field workers can register and get access to the transect lines allocated to them by the local organizers.
+Since 2019, most of the data has been collected using a mobile app tailored to the monitoring program, where the field workers can register and get access to the transect lines allocated to them by the local organizers.
 Prior to 2019, data were collected on a dedicated fieldwork form, and entered manually in a web portal afterwards.
@@
 After an initial embargo period, all data from public land are published and made freely available as a sampling-event data set on GBIF (<https://www.gbif.org/sampling-event-data>).
-The published datasets contain both metadata about the transect surveys (survey date, line transect length and location, study area ID, etc.) as well as bird observation data (species, number of birds of different categories (adult males, adult females, juveniles, and birds of unknown category), perpendicular distance to transect line, exact location, and time of observation).
+The published datasets contain both metadata about the transect surveys (survey date, line transect length and location, study area ID, etc.) and bird observation data (species, number of birds of different categories (adult males, adult females, juveniles, and birds of unknown category), perpendicular distance to transect line, exact location, and time of observation).
 Formally, the data from public land is published as three distinct data sets, one for each of the main public land administrators (Statskog, FeFo and Fjellstyrene, respectively).
-<!-- I changed "owners" to "administrators", just because to me "public" implies it is not owned by anyone.
-But if this is wrong, please change it back -->
 
@@
 The marked birds were then monitored on a regular basis until they either i) died, ii) their transmitter's battery stopped working, or iii) we lost contact with the bird for other reasons.
-For most of the year, the birds were monitored <!-- how were they actually monitored?
-Can we have a tiny bit more detail here?
---> at least once a month by radio triangulation.
-Most of the fieldwork was conducted from the ground, but to avoid loss of data, the birds were also triangulated from helicopters in May, September and November.
-During the breeding and chick-rearing season (May to July) birds were monitored more often, and during December and January we obtained few observations due to challenging field work conditions.
+For most of the year, the birds were monitored at least once a month by radio triangulation.
+Most of the fieldwork was conducted from the ground, but to avoid data gaps, the birds were also triangulated from helicopters in May, September, and November.
+During the breeding and chick-rearing season (May to July) birds were monitored more often, and during December and January we obtained fewer observations due to challenging field work conditions.
 A proportion of the birds were harvest annually in the regular recreational harvest, and birds that were harvested were reported as shot to the field personnel.
@@
 
+### Rodent occupancy data
+
+As part of the line transect sampling (see above), observers are also requested to report whether they have seen any small rodents while surveying a transect.
+For each transect survey, this information is recorded as 1's (small rodents spotted at least once) and 0's (no small rodents spotted).
+We aggregated this data into area- and year-specific rodent occupancy covariates by averaging the 0 and 1 reports for all transect surveys within a given area and year and subsequently z-standardizing values.
+We note that while we refer to the covariate as "rodent occupancy" throughout the manuscript, it can be interpreted as an index for rodent abundance.
+<!-- @ERLEND: I have added this section as it was missing.
+Anything you would like to add here?
+-->
+
 ## National-scale integrated model
@@
 @nilsen2023 recently developed an integrated distance sampling model (IDSM) which jointly analyses line transect and radio-telemetry data and applied it to willow ptarmigan in the Western part of Lierne municipality in Norway.
-The model consists of a deterministic population model <!-- I think this depends on defenitions.
-It did contain temporal variation in recruitment, which is in fact environmental stochasticity.
-It did not contain demographic stochasticity --> with two age classes (juveniles and adults) and four data likelihoods: 1) likelihood for observation distances from transect lines for estimating detection probability; 2) likelihood for age-specific counts on transect surveys for estimating numbers of juveniles and adults present; 3) likelihood for juvenile to adult ratios observed at the locality level to provide estimate recruitment rate (as juveniles/adult); and 4) likelihood for known-fate telemetry data to estimate seasonal and annual survival.
-Below, we describe our new extension of this model to include data from several as opposed to just one.
+The model consists of a population model with two age classes (juveniles and adults) and four data likelihoods: 1) likelihood for observation distances from transect lines for estimating detection probability; 2) likelihood for age-specific counts on transect surveys for estimating numbers of juveniles and adults present; 3) likelihood for juvenile to adult ratios observed at the locality level to provide estimate recruitment rate (as juveniles/adult); and 4) likelihood for known-fate telemetry data to estimate seasonal and annual survival.
+Below, we describe our new extension of this model to include data from several areas as opposed to just one.
 For more detailed information on the single-site model, including tests of model performance, see @nilsen2023.
@@
 For the likelihood for line transect observation distances we used a custom half-normal distribution developed by Michael Scroggie in the "nimbleDistance" package (https://github.com/scrogster/nimbleDistance).
-We used non-informative uniform priors for all parameters and including biologically sensible boundaries where possible.
+We used non-informative uniform priors for all parameters, but used biologically sensible boundaries where possible.
 We simulated complete sets of initial values for all model nodes prior to model running and using pre-defined seeds to ensure reproducibility.
@@
 reporting district vs.
-survey locality), whether to model time variation in survival and/or effects of rodent abundance, whether to run MCMC chains sequentially or in parallel, etc.
+survey locality), whether to model time variation in survival and/or effects of rodent occupancy, whether to run MCMC chains sequentially or in parallel, etc.
 A visual representation of the pipeline is also displayed in @fig-workflow and for more details on pipeline implementation and options, we refer the reader to the GitHub repository: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.
@@
 During this period, estimated population densities varied between `r dens.min$post.sum` birds/km^2^ in the area "`r dens.min$area.name`" close to the coast in central Norway to `r dens.max$post.sum` birds/km^2^ in "`r dens.max$area.name`" further south near the Swedish border.
-In general, during our study period the population density appeared to be lowest in the Northern parts of Norway and highest in the eastern part of central Norway (@fig-densA).
+In general, recent population density appeared to be lowest in the Northern parts of Norway and highest in the eastern part of central Norway (@fig-densA).
 Uncertainty in density estimates was relatively consistent, with a few areas (including the one with the lowest estimated density, "`r dens.min$area.name`") sticking out by having substantially less precise estimates (@fig-densB).
@@
 2011, 2014, 2018) or low (e.g.
-2012, 2015) densities across a large number of areas.
+2012, 2015) densities across a substantial number of areas.
 
@@
 
-The highest population growth rates were estimated for areas with relatively low population densities across latitudes but we did not find evidence for a strong association between population growth rates and population densities overall ([@fig-paramCorrs] A).
-<!-- Indicate that this is just for the four most recent years.
-Same for figure below -->
+The highest recent population growth rates were estimated for areas with relatively low recent population densities across latitudes but we did not find evidence for a strong association between population growth rates and population densities overall ([@fig-paramCorrs] A).
 
 ![Posterior summaries (points = posterior medians, lines = 95% credible intervals) of area-specific population growth rate vs.
-population density (A), recruitment rate vs.
+population density over the four most recent years (2018-2021, A), recruitment rate vs.
 survival probability (B) and rodent effect on recruitment along a longitudinal gradient (C).
@@
 There was substantial variation in both vital rates across time (SFs "TimeSeries_pSurv.pdf" and "TimeSeries_rRep.pdf").
-In many areas, the years 2011, 2014, and 2018 not only supported high population densities (see above) but were also characterized by both high recruitment and low subsequent survival.
-<!-- Here and above: I would perhaps consider deleting the reference to 2011 as a "high density year", as there is really a lot of spatial variation in this pattern.
-I'm not so convinced that is the case in general --> The overall low density years 2012 and 2015, conversely, often featured lower recruitment and, in some cases, higher survival.
+In a substantial number of areas, the years 2011, 2014, and 2018 not only supported high population densities (see above) but were also characterized by both high recruitment and low subsequent survival.
+<!-- ERLEND: Here and above: I would perhaps consider deleting the reference to 2011 as a "high density year", as there is really a lot of spatial variation in this pattern.
+I'm not so convinced that is the case in general.
+CHLOE: I checked again in my "scoring sheet", and it's a clear density peak in 16 areas, which is not far from 2014 (19 areas) and 2018 (18 areas).
+Obviously there is variation, but I do think this is substantial.
+I would like to leave it in, but have changed the phrasing to "a substantial number of areas" instead of "many"/"a large number".
+Okay?
+--> The overall low density years 2012 and 2015, conversely, often featured lower recruitment and, in some cases, higher survival.
 Notably, there were also years with very little spatial synchrony, i.e.
@@
 Effects with negative posterior medians were located mostly at intermediate latitudes, but we note that all of these had posterior distributions featuring substantial overlap with 0 (@fig-paramCorrs C).
-<!-- we sometimes refer to rodent abundance, sometimes occupancy.
-I think what we have is most correctly refered to as an index of abundance -->
+<!-- ERLEND: we sometimes refer to rodent abundance, sometimes occupancy.
+I think what we have is most correctly refered to as an index of abundance.
+CHLOE: Good point about being consistent.
+I see that I have mostly used "occupancy" in text, figures, and code, and to me, that is also the most "precise" description.
+So I have standardized to primarily using "occupancy".
+Nonetheless, we should state that the way we treat occupancy becomes an abundance proxy.
+In fact, I noticed we were missing the description of rodent data.
+So added that, and made a mention there.
+Okay?
+-->
 
@@
 
-<!-- @ERLEND & @JAMES thoughts on what we see that is interesting, cool, expected, unexpected etc.
-Can we compare to other large scale studies, either on other grouse-like creatures or harvested species more generally?
--->
+The wide spatial distribution of the line transect monitoring afforded us the opportunity to explore variation in population density and vital rates across a relatively large spatial extent.
 
-<!-- Spatial perspective -->
+Ptarmigan densities across the 41 reporting districts included in our analyses varied from around 2 birds/km^2^ to 55 birds/km^2^, with the lowest densities occurring far north in the country, as well as on the west coast and in the mountains in central Norway [@fig-densA].
+The same spatial pattern was also evident at the level of the demographic rates: consistent with basic life history theory [@stearns1992], average recruitment rates were inversely related to average survival probabilities [@fig-paramCorrs B], and the slower life histories (higher survival and lower recruitment) were more common in the northern and mountaineous parts of the country.
+This aligns with previous studies reporting relatively slower bird life histories in alpine / high altitude areas [e.g.
+@sandercock2005; @bears2009; @Wilson_Martin_2011; @alice2016].
+In Norway, the northern and mountainous areas are characterized by more extreme climatic conditions, boasting cold temperatures and short growing seasons.
+Resulting reduced primary production limits food availability and as ptarmigan are income breeders that use food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005], lower carrying capacity in such areas is to be expected.
 
-The wide spatial distribution of the line transect monitoring afforded us the opportunity to explore variation in population density and vital rates across a relatively large spatial extent.
-Ptarmigan densities across the 41 reporting districts included in our analyses varied from around 2 birds/km^2^ to 55 birds/km^2^, with the lowest densities occurring far north in the country, as well as on the west coast and in the mountains in central Norway [@fig-densA].
-These areas are characterized by more extreme climatic conditions, boasting cold temperatures and -- in the North -- a short growing season.
-Resulting reduced primary production limits food availability and as ptarmigan are income breeders using food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005], lower carrying capacity in such areas is to be expected.
-<!-- @ERLEND do we have some more thoughts to add here?
--->We found increasing population trends over recent years in over half of the reporting districts [@fig-lamA].
-<!-- relate this to changes in hunting management in recent years?
-Any other factors we think could have contributed?
---> Nonetheless, recent population declines were also evident in some areas, and particularly in the mountains in central Norway [@fig-lamA].
-<!-- What could be going on here?
-It's also interesting that while some of these areas have declines over the most recent years, that usually follows on a period of increase, or at least stability, up to around 2016 -->
-
-<!-- Erlend will probably have better ideas than I.-->
-
-Density dependence is an important process to understand in ecology but more so for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; @aanes2002; @willebrand2001].
-Negative density dependence has been found in numerous gallinaceous birds such as northern bobwhites *Colinus virginianus* [@mcconnell2018], *Perdix perdix* [@bro2003], and wild turkeys *Meleagris gallopavo* [@mcghee2007]).
-While we found the highest population growth rates in areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A], our analysis did not provide evidence for a strong relationship between density and growth rates across space.
-However, this is unsurprising as we did find substantial spatial differences in demography, which suggests that carrying capacity too varies over space.
-Also, support for density dependence in willow ptarmigan has been mixed.
-@myrberget1988 observed no change in productivity despite a 50% decrease in abundance.
-Conversely, @pedersen2004 found strong overwinter negative density dependence and they posited that immigration was the vital rate that responded to changes in density.
-Low survival rates appeared to have a one-year lag following abundance peaks.
-This pattern could be a result of negative density dependence that went undetected when looking at the ensemble of samples across the 41 areas.
-Negative density dependence has been shown in willow ptarmigan populations which may explain explain reduced survival following years with elevated densities [@pedersen2004].Future work should explore different time-lags for density and population growth and the possibility for non-linear effects.
-
-<!--# Not happy with the DD para yet but I can work on Jan 11 -->
-
-<!--# @JAMES: Feel free to add more, but I feel like if we want to go for a DD deep-dive, I need to run a test for evidence of WITHIN area DD first. Because there may well be a pattern there... -->
-
-<!--# Yeah I agree. We already have a lot going on so I am ambivalent about adding another aspect. -->
-
-Overall, annual recruitment rates showed a spatial pattern that was inversely related to that of annual survival, with lower recruitment rates co-occurring with higher survival rates (and lower densities) and vice-versa [@fig-paramCorrs B], as we would expect based on life history theory [@stearns1992].
-Furthermore, the areas with relatively lower recruitment and higher survival were primarily found in the northern parts of Norway and in the mountainous region in the central part of the country, which aligns with previous studies finding relatively slower bird life histories in alpine / high altitude areas [e.g.
-@sandercock2005; @bears2009; @Wilson_Martin_2011; @alice2016].
-<!-- What about the latitude?
-Do we know of any studies looking at latitudinal variation in life histories?
-When I last searched, I only found studies focusing on altitude -->
-
-<!--# Yeah I wasn't able to find much. -->
-
-<!-- Temporal perspective -->
-
-Population densities in all areas were subject to substantial variation across years, featuring periods of stability, increase, and decrease (SF "TimeSeries_PopDens1.pdf").
+We found increasing population trends over recent years in over half of the reporting districts, but population declines were also evident in some areas, particularly in the mountains in central Norway [@fig-lamA].
+While we may speculate that recent population trends may be linked to changes in harvest regulations and/or climatic conditions, considering the whole time-series (2007-2021) illustrated that population densities in all areas were subject to substantial variation across years, featuring periods of stability, increase, and decrease (SF "TimeSeries_PopDens1.pdf").
 In most areas, there were also strong year-by-year fluctuations in population density on top of longer-term trends.
 Some of the resulting "high density years" were highly synchronized across large spatial scales, such as the years 2011, 2013, and 2018.
-Taking a closer look, we find that these are years that are characterized by high recruitment (SF "TimeSeries_rRep.pdf") and low subsequent survival (SF "TimeSeries_pSurv.pdf"), which often resulted in steep population declines towards the following year.
-The fact that these same years also match up with observed peaks in rodent abundance in many areas provides evidence for the Alternative Prey Hypothesis (APH; [@Israelsen2020]).
+Taking a closer look, we find that these are years that are characterized by high recruitment (SF "TimeSeries_rRep.pdf"), follwed by a low survival the year after (SF "TimeSeries_pSurv.pdf").
+This often resulted in steep population declines towards the following year.
+The fact that these same years also match up with observed peaks in rodent abundance in many areas, together with the largely positive effects of rodent occupancy on recruitment estimated by our model (@fig-paramCorrs C), provides evidence for the Alternative Prey Hypothesis [APH; @Israelsen2020].
 The APH stipulates that high abundance of alternative prey (rodents, in this case) for common predators leads to population growth, and is well-supported throughout the literature for a range of taxa [e.g., @hagen1952; @Kjellander_Nordström_2003; @Reif_Tornberg_Jungell_Korpimäki_2001], including willow ptarmigan [@bowler2020].
-Conversely, @Nyström_etal_2006 suggested that gyrfalcons (a specialized ptarmigan predator) did not respond to rodent populations and did not switch to alternative prey when ptarmigan populations are low.
-Nonetheless, other predators such as red fox could prey switch in response to changing availability of rodents and ptarmigan as shown by @Breisjøberget_etal_2018 and @bowler2020.
-However, statistical associations in our study and others is not causal and alternative explanations for this pattern are plausible.
-The positive association between rodents and ptarmigan could be caused by bottom-up processes such as food availability unrelated to predation.
-Future work should develop causal models and manipulative experiments to establish causal relationships among top-down and bottom-up processes.
+While @Nyström_etal_2006 suggested that gyrfalcons, which are specialized ptarmigan predators, do not respond to rodent populations or switch to alternative prey when ptarmigan populations are low, generalis predators, such as red foxes, are likely to shift from preying on ptarmigans to rodents when the latter become abundant [e.g.
+@Breisjøberget_etal_2018; @bowler2020].
+Taking a spatial perspective, the highest latitude and highest altitude areas stood out once more, sporting the strongest rodent effects (SF "betaR_Map.pdf").
+This could be related to warmer areas generally having larger predator guilds, and consequently more generalists that are able to maintain relatively stable populations irrespective of small rodent abundance [@bowler2020].
 
-Taken together, this alludes to the large role small rodents play in shaping ptarmigan demography and population dynamics and is further supported by the largely positive effects of rodent abundance on recruitment estimated by our model (@fig-paramCorrs C).
-Once more, the highest latitude and highest altitude areas stood out, sporting the strongest rodent effects (SF "betaR_Map.pdf").
-This may be linked to the overall low ptarmigan densities in these areas, as e.g.
-@Nielsen_1999 found evidence that gyrfalcon predation on ptarmigan was most impactful when ptarmigan densities were low.
-<!-- This previous sentence is adapted from a suggestion by James, but I feel like we need some more explanation/arguments here to clarify this.
-Perhaps also some other and/or more specific hypotheses= --> Notably, the conclusive estimation of overall positive effects of rodent abundance on recruitment in our model was only possible thanks to the integration and sharing of data across multiple areas.
+Notably, the conclusive estimation of overall positive effects of rodent occupancy on recruitment in our model was only possible thanks to the integration and sharing of data across multiple areas.
 When @nilsen2023 fit the IDSM to data from only a single area, they were unable to obtain a reliable estimate for the rodent effect due to limited statistical power.
 Consequently, the space-for-time substitution that comes with extending the model across multiple area allows estimation of covariate effects that otherwise cannot be estimated, and opens up for future possibilities for studying effects of not just rodents, but also other environmental drivers on ptarmigan population dynamics.
-For example, the cover of ericaceous shrubs as a proxy for food availability or spatially-explicit spring green up dates both measured using remote sensing are ecologicaly plausible and feasibly acquired factors.
-<!-- @ERLEND here we could add some thoughts on other covariates that may matter and are realistic/relevant/realistic to include in future studies.
-Best if it kind of naturally leads over to (and does not overlap too much with) the next paragraph  -->
+Doing so may also help with better understanding the mechanisms underlying the large portion of demographic rate variation that could only be attributed to random variation so far.
+This is the case especially for the relatively large residual variation [@fig-varDecomp] but also relevant for constant spatial and shared temporal variation.
+In previous work, both based on marked birds [@Eriksen2023] and unmarked birds [@bowler2020; @Henden2020; @Novoa2016], spring conditions has come out as an important predictor of ptarmigan recruitment rates.
+In general, it is found warmer and earlier springs result in earlier breeding, larger clutch sizes [@Eriksen2023], and resulting higher recruitment rates measured in the late summer and early fall (as in our study).
+In @bowler2020, it was reported that the strength of this relationship was not consistent in time and space, but was generally stronger in colder areas, similar to the effect of rodent occupancy reported here.
+In practice, measures of the cover of ericaceous shrubs (a proxy for food availability) or spatially-explicit spring green up dates derived from remote-sensing data constitute relevant candidate covariates for future work alongside temperature.
 
-<!--# I added two that came to my mind  -->
+Another important determinant of vital rate variation is density dependence, in particular for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; @aanes2002; @willebrand2001].
+Negative density dependence has been found in several gallinaceous birds such as northern bobwhites *Colinus virginianus* [@mcconnell2018], *Perdix perdix* [@bro2003], and wild turkeys *Meleagris gallopavo* [@mcghee2007]).
+For willow ptarmigan, evidence for density-dependent population regulation has been mixed.
+@myrberget1988, for example, observed no change in productivity despite a 50% decrease in abundance, while @pedersen2004 reported strong negative density-dependence over winter and posited that dispersal may be the vital rate that responded to changes in density most strongly.
+While we did not explicitly model density dependence in this study, our results can provide some preliminary insights into potential density feedbacks from both a spatial (cross-population) and a temporal (within-population) angle.
+Comparing average population densities and growth rate across areas did not provide evidence for strong density dependence, but there was a tendency towards the highest population growth rates appearing areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A].
+When considering density dependence across years within select areas, however, we found that higher density years were associated with higher recruitment the same year, but followed by lower apparent survival probabilities and, consequently, lower population growth rates (as determined by post-hoc Pearson correlation coefficients, supplementary file "DD_corrCoef.csv").
+While this seems to support the notion of negative density-dependence, testing for this post-hoc gives results that are confounded with sampling correlation.
+Hence, formally modelling density-dependence, possibly using different forms and time-lags, could prove to be a promising extension of our modelling framework in the future.
 
-<!-- Variation beyond rodents -->
-
-Extending the present analysis by including additional covariates may also help with better understanding the mechanisms underlying the large portion of demographic rate variation that could only be attributed to random variation so far.
-This is the case especially for the relatively large residual variation [@fig-varDecomp] but also relevant for constant spatial and shared temporal variation.
-Interestingly, we found that a larger portion of variation in recruitment was allocated to spatial than to temporal factors, while the opposite was the case for survival.
-Contrastingly, ptarmigan nest failure is almost entirely caused by predation which has been shown to vary greatly across years and less across space [@Munkebye_Pedersen_Steen_Brøseth_2003; @kvasnes2015; @kvasnes2014] whereas adult survival has been shown to not very greatly across years [@Israelsen2020].
-Also, immigration, included in our recruitment term <!--# this true?
--->, has been hypothesized to be a spatial process [@kvasnes2015].
-This conundrum could be a result of the limited spatial replication of many previous studies that we were able to address with our approach.
-Future research could help rectify these discrepancies by including more mechanistic relationships between environmental factors and vital rates.
-
-<!-- Any hypotheses/thoughts related to large- vs. local-scale drivers to explain this? -->
-
-<!-- This is interesting I will do some reading on this and add some text on Jan 11-->
-
-<!-- I don't know if I helped or not. It is an interesting result. Erlend, you got thoughts? -->
-
-## Implications for management
+## Implications for management and monitoring
 
 Management decisions made at the resolution of large geo-political boundaries (e.g., Norway) run a high risk of being inadequate when there is substantial spatial variation in demographic processes and population dynamics, as is the case for willow ptarmigan.
-In Norway, willow ptarmigan -- and small game in general -- is managed \[...\] <!-- Brief summary of how and at what spatial scales patrmigan/small game management works in Norway --> The results from our study suggests that management decisions should be spatially refined and ideally informed by knowledge about the "local" population processes.
-\[...\] <!-- Some specific implications of the results for management of hunting and of habitat (e.g.
-better managed hunting is good because we do see increases in majority of local populations, but the mountain areas have had recent declines that may call for additional local measures, these measures may need to be adjusted since the mountain populations appear to have slower life histories and likely different sensitivities, etc.) -->
+In Norway, willow ptarmigan -- and small game in general -- is managed at the local and/or regional scale, with rather limited national regulation beyond updating the season length every fourth year.
+In effect, both the management system, regulation type (quota type, season length, number of licences, bag limit etc) and quota size is decided by the local or regional stakeholders.
+Thus, while national estimates (abundance and/or temporal trend in abundance) might be important for red listing decisions and for setting the maximum season length, remaining decisions about harvest management are taken locally.
+<!-- ERLEND: I will add some relevant references in next iteration.
+We have no "go to" reference for the management system (we should write one), but I will see if I can piece together something.
+--> The results from our study highlight a large degree of spatio-temporal variation in both ptarmigan densities and demographic rates, suggesting that it is indeed suitable for management decisions to be spatially refined and ideally informed by up-to-date knowledge about recent "local" population processes.
+Accessible and easily repeatable modelling workflows, such as the one we have developed in this study, can thus become a valuable source of information for local decision-makers.
 
@@
 Particularly, we found generally lower detection probabilities in the northern half of Norway.
-\[...\] <!-- Some thoughts on why that might be.
-Could be related to landscape features and/or available observer pool.
-Any suggestions on what could be done to improve detection in the North, or at least "make up" for it?
--->
-
-<!--# They obviously use a lesser breed of bird dog up that way.Probably use more german shorthairs or vislas. :-)  -->
-
-<!-- Funnily enough, I have been suggesting to Erlend that we should have a master student to analyze the effect of dog breed on detection probability :-P -->
-
+This may be related to habitat features, as the transects in the North are primarily located in birch forests and more rugged terrain, which may hamper detectability.
+Additionally, the slower life histories in the northern areas are reflected as generally smaller bird clusters as well, and smaller clusters have previously been shown to have a lower detectability than larger ones [e.g.
+@bowler2020, see also next section].
+Our modelling framework can be easily adapted for studying the impact of different variables on detectability.
+Additionally, our results could motivate taking a closer look at monitoring challenges and potential improvements in the northern part of the country in particular.
 Ultimately, increased detection probability would contribute to obtaining more precise estimates of both population density and demographic rates, which -- in turn -- would be of great value in particular in areas with relatively low population densities, low number of transects, and less years of data.
@@
 that there is some observation error in the number reported.
-This could be accounted for by adding an additional layer of hierarchy to observation process (see e.g.
-REF), which could further be used to also account for error in judging the observation distance [e.g.
-@marques2004].
+This could be incorporated by including an additional layer of hierarchy to the observation process (see e.g.
+REF), and possibly further extended to also account for error in judging the observation distance [e.g., @marques2004].
 Another potential source of bias in our IDSM is related to failure to correctly assign the age class of observed birds.
@@
 If misclassification happened at random, mixture models could be used to determine the likely age class of individuals to whom no age class was assigned during observations (+ref).
-In our case, however, juveniles are likely more prone to misclassification due to \[...\] <!-- @ERLEND can you add reasons?
--->.
+In our case, we might suspect that an observed is more likely to classify an adult bird as juvenile rather than the other way around, and more likely to assingn "unknown" age class to juveniles than adults.
+One reason for this is that observers look for specific signs to classifying a bird as adult (e.g.
+size, male sound), and might default to juvenile or unknown if the signs are not clearly detected.
 Future studies should investigate to what degree available information on e.g.
@@
 Alternatively, spatial (and temporal) correlations among parameters can be modelled explicitly, something that is commonly done e.g.
-for modern species distribution models [e.g.
+in modern species distribution models [e.g.
 @pacifici2017; @guelat2018].
@@
 The line transect survey data included in this study constitutes just the publicly available part of the data collected through "Hønsefuglportalen" but the programme also includes additional surveys on private land.
-\[...\] <!-- @ERLEND: Can we add a very brief ocverview over what the coverage (in terms of municipalities) would be if we had access to all the Hønsefuglportalen data?
---> Exploring to what degree additional data from Hønsefuglportalen could be included in future studies employing an extended IDSM with additional spatial structuring is therefore a worthwhile endeavor.
+Extending to data from private land would provide better coverage especially in south-eastern and southern parts of Norway, which includes areas where only very limited amounts of data are collected on public land.
+Exploring to what degree additional data from Hønsefuglportalen could be included in future studies employing an extended IDSM with additional spatial structuring is therefore a worthwhile endeavor.
 
@@
 Nonetheless, this illustrates a way for how smaller datasets from single or subsets of areas can be integrated into a large-scale modelling framework.
-Other relevant data could be included using the same approach [...] <!-- @ERLEND: Any thoughts on what the status here is?
-Might there be other data available from Lierne and/or beyond (other telemetry data, nest monitoring, etc.?
-Are there exciting future prospects e.g.
-linked to automated monitoring with cameras/sound boxes at a larger scale?
---> The most relevant source of data to be included into the IDSM framework in the near future, however, is harvest data.
-<!-- @ERLEND 1-2 sentence summary of what kind of harvest data is available, and at what spatial covarage and resolution --> As the IDSM framework is, in essence, an IPM, harvest can be modelled through partitioning of survival into cause-specific mortality in the process model and inclusion of relevant harvest data likelihoods [e.g.
-@gamelon2021; @nater2021].
-Harvest effects on willow ptarmigan have been studied previously yet uncertainties persist [@sandercock_etal2011; @aanes2002; @pedersen2004; @willebrand2001].
-For example, the strength and shape of density dependence and how harvest may induce population sinks because of Allee effects [@kvasnes2015]).
-Modelling allee effects, or better yet predicting where they are likely to occur if over-harvest occurs, requires a spatially-explicit understanding of harvest and its impact on local populations.
-Because harvest effort, thus rates, can vary dramatically over time and space [@breisjøberget2018] easily repeatable analyses that can be done continually would prevent prolonged over-exploitation.
-Harvest effects may interact with other (emergent) factors such as climate change and habitat degradation making the need to be identify triggers for policy changes paramount [@gamelon2019].
-
-<!-- @JAMES & Erlend: Maybe we can be a bit more specific about what the "uncertainties" are regarding ptarmigan harvest?
-Might be good to keep the "spatially explicit" perspective a bit, i.e.
-why would it be great to look at this with this kind of model, instead of just aggregate numbers across large regions?
-Also, maybe we should highlight that up-to-date (easilty repeatable) analyses are important because things can change, and harvest effects may interact with other (emergent) factors such as climate change and habitat degradation.
-Finally, it could also be nice to link this to some studies on other species...
--->
-
-<!--# That do it? -->
+Other relevant data could be included using the same approach, for example data from ongoing nesting success monitoring (Nilsen, E.B.
+unpublished data), data from past studies of marked birds [@sandercock_etal2011], and data from other monitoring programs for breeding birds based on point counts (see e.g.
+the Norwegian Breeding Bird Monitoring: <https://hekkefuglovervakingen.nina.no/>).
+The most relevant source of data to be included into the IDSM framework in the near future, however, is harvest data.
+Such data might be available with two different spatial and temporal resolution.
+First, at the municipality level there are data with national coverage collected annually by Statistics Norway (<https://www.ssb.no/>).
+Second, many public land owners have data with much higher temporal (daily) and spatial (harvest area) resolution, including both harvest bags and harvest effort (number of hunters pr area pr day).
+As the IDSM framework is, in essence, an IPM, harvest can be modelled through partitioning of survival into cause-specific mortality in the process model and inclusion of relevant harvest data likelihoods [e.g., @gamelon2021; @nater2021].
+While harvest effects on willow ptarmigan have been studied previously, much uncertainty remains [@sandercock_etal2011; @aanes2002; @pedersen2004; @willebrand2001].
+For example, little is known about how harvest pressure and density feedbacks interact on differente temporal and spatial scales [@kvasnes2015], despite this knowledge being crucial for preventing over-exploitation and ensuring sustainable harvest [@williams2002; @breisjøberget2018].
+Additionally, harvest effects often interact with other (emergent) factors such as climate change and habitat degradation, making predictive models that account for harvest alongside other mechanisms invaluable for informing policy changes [@gamelon2019].
 
@@
 Modern applied ecology needs research to be published not just as manuscripts, but as reproducible and well documented workflows [@lewis2018].
-This is particularly crucial for research that is (to be) closely tied to management and/or used to create biodiversity indicators that are to be reported nationally or internationally, or to be used by industrial partners.
+This is particularly crucial for research that is (to be) closely tied to management and/or used to create biodiversity indicators that are to be reported nationally or internationally, or to be used by industrial partners [@powers2019].
 This is both because of the enhanced transparency and credibility provided by openly available reproducible workflows and because of their cost-effectiveness, which allows for more sustainable use of funding in the mid- to long-term.
@@
 
-<!-- @EVERYONE: Do we know of some good references for this section? Presumably we cannot cite our chapter in the Technical Manual yet :-P -->
+# Author contributions
+
+**Chloé R. Nater:** Conceptualization, Methodology, Software, Formal analysis, Writing - Original Draft, Writing - Review and editing, Visualization.
+
+**James A. Martin:** Conceptualization, Writing - Original Draft, Writing - review and editing.
+
+**Erlend B. Nilsen:** Conceptualization, Methodology, Data curation, Writing - Original Draft, Writing - Review and editing, Project administration, Funding acquisition
 
@@
 
-TBA ...
-We are grateful to the participants in the "Hønsefuglportalen" and their dog helpers that collected ptarmigan observations.
+We are grateful to the numerous volunteer observers (and their canine assistant) who collected ptarmigan observations as part of the "Hønsefuglportalen" monitoring effort.
 Additionally, we would like to thank Bernardo Brandão Niebuhr dos Santos for his help with spatial data visualization.
@@
 A.
-Martin was supported by a Fulbright Fellowship and the Norwegian Institute for Nature Research that facilitated this collaboration.
-<!-- @ERLEND: you maybe know best all the people/institutions we need to thank here -->
+Martin was supported by a Fulbright Fellowship and the Norwegian Institute for Nature Research, which facilitated this collaboration.
+
+# Funding
+
+Funding for both data collection and analytical work was provided by the Norwegian Environmental Agency (grant numbers 17010522, 19047014, and 22047004).
+
+# Conflict of interest disclosure
+
+The authors declare that they comply with the PCI rule of having no financial conflicts of interest in relation to the content of the article.
 
@@
 The raw data from the line transect surveys is deposited on GBIF and can be accessed freely via the Living Norway Data Portal (https://data.livingnorway.no/).
-The work presented above is based on versions 1.7, 1.8, and 1.12 for the datasets from Fjellstyrene (key b49a2978-0e30-4748-a99f-9301d17ae119), Statskog (key 6a948a1c-7e23-4d99-b1c1-ec578d0d3159), and FeFo (key c47f13c1-7427-45a0-9f12-237aad351040), respectively.
-<!-- I thinbk it is better to cite the datasets (with doi and all that) rather than the key.
-That way, the citation and data use will be tracked by GBIF (and Living Norway) -->
+The work presented above is based on versions 1.7, 1.8, and 1.12 for the datasets from Fjellstyrene [@fjellstyre2022], Statskog [@statskog2022], and FeFo [@fefo2022], respectively.
 

### 2024-01-26 Manuscript_Nater_etal.qmd
Added missing references

@@
 the collection of ecological data by members of the public [also called citizen or community science, @fraisl2022], allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias [@johnston2023].
-Consequently, many large-scale monitoring programmes are often limited to presence-absence or very simple count observations, making them suitable for the development of indicators of species distributions and perhaps population trends, but usually not of abundance, population dynamics, and demographic rates (+ example refs).
+Consequently, many large-scale monitoring programmes are often limited to presence(-absence) or very simple count observations, making them suitable for the development of indicators of species distributions and perhaps population trends, but usually not of abundance, population dynamics, and demographic rates [@dickinson2010; @johnston2023].
 The exception here are monitoring programmes that succeed in making use of a large number of volunteers that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol.
@@
 The spatial, temporal, and residual variance components were defined as the square of the estimated corresponding random effects standard deviation from the model while $var_{rodent}$ was calculated as the variance of all area- and year-specific $\beta_x*rodentOcc_{x,t}$ products.
-<!-- We might need to justify the latter a bit, as the rodent variance is calculated different than the other variances -->
+This approach for variance decomposition is equivalent to that used by @nater2018 and inspired by @nakagawa2013.
 
@@
 that there is some observation error in the number reported.
-This could be incorporated by including an additional layer of hierarchy to the observation process (see e.g.
-REF), and possibly further extended to also account for error in judging the observation distance [e.g., @marques2004].
+This could be incorporated by including an additional layer of hierarchy to the observation process [see e.g., @hamilton2018], and possibly further extended to also account for error in judging the observation distance [e.g., @marques2004].
 Another potential source of bias in our IDSM is related to failure to correctly assign the age class of observed birds.
 @nilsen2023 showed that incorrect age assessment can bias (relative) estimates of survival and recruitment, and while they only found a weak bias in their case study on a single area, the problem may be larger in a multi-area setting that may contain areas with different proportions of misclassified observations.
-If misclassification happened at random, mixture models could be used to determine the likely age class of individuals to whom no age class was assigned during observations (+ref).
+If misclassification happened at random, mixture models could be used to determine the likely age class of individuals to whom no age class was assigned during observations [@mccrea2013].
 In our case, we might suspect that an observed is more likely to classify an adult bird as juvenile rather than the other way around, and more likely to assingn "unknown" age class to juveniles than adults.
@@
 Furthermore, we did find that mixing of several of the global and area-specific intercept parameters in the current model was suboptimal, suggesting that there may be much to gain from additional structuring, as well as from development of more efficient MCMC sampling strategies for the resulting extended model.
-One promising framework for approaching this are conditionally autoregressive models (CARs).
+One promising framework for approaching this are conditionally autoregressive models [CARs, @verhoef2018].
 Such models have been used repeatedly for modelling spatial autocorrelation in species occupancy and demographic rates [e.g.

### 2024-01-26 Analysis_RealData.R
Updated to new shape file that does not include ocean

@@ -296,7 +296,7 @@ plotVarDecomposition(mcmc.out = IDSM.out.tidy,
 #---------------------#
 
 ## Make map of Norwegian municipalities ("fylke")
-NorwayMunic.map <- setupMap_NorwayMunic(shp.path = "data/Kommuner_2018_WGS84/Kommuner_2018_WGS84.shp",
+NorwayMunic.map <- setupMap_NorwayMunic(shp.path = "data/norway_municipalities/norway_municipalities.shp",
                                         d_trans = LT_data$d_trans,
                                         areas = areas, areaAggregation = TRUE)
 

### 2024-01-26 R/setupMap_NorwayMunic.R

@@ -25,6 +25,7 @@ setupMap_NorwayMunic <- function(shp.path, d_trans,
   
   ## Reading in map of Norwegian municipalities
   mapNM <- sf::st_read(shp.path)
+  names(mapNM)[which(names(mapNM) == "navn")] <- "KOMMUNENAV"
   
   ## Extract list of localities and areas with corresponding municipality names
   areaMunic <- d_trans %>%

### 2024-01-26 _targets.R

@@ -235,7 +235,7 @@ list(
   
   tar_target(
     NorwayMunic.map,
-    setupMap_NorwayMunic(shp.path = "data/Kommuner_2018_WGS84/Kommuner_2018_WGS84.shp",
+    setupMap_NorwayMunic(shp.path = "data/norway_municipalities/norway_municipalities.shp",
                          d_trans = LT_data$d_trans,
                          areas = areas, areaAggregation = areaAggregation)
   ),

### 2024-01-26 Manuscript_Nater_etal.qmd
Added OSF citation

@@
 
-The auxiliary radio-telemetry data, rodent occupancy data, and all code including the "targets pipeline" can be found in the project's repository on GitHub: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.
+The auxiliary radio-telemetry data, rodent occupancy data, posterior summaries, and supplementary figures are archived on OSF [@Nater_Nilsen_Martin_2024].
+
+All code including the "targets pipeline" can be found in the project's repository on GitHub: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.
 The results presented in this paper were created using version 2.0 of the code (Zenodo citation TBA).

### 2024-01-29 Manuscript_Nater_etal.qmd
Added reference to new code release
Edits and additions to manuscript and .bib file

@@
 
-There is growing demand for biodiversity indicators from international unions, national government, local management bodies, and corporate and industry actors.
+There is growing demand for biodiversity indicators from international unions, national governments, local management bodies, and corporate and industry actors.
 Such indicators should ideally represent a wide range of biodiversity's states and functions [e.g.
@@
 On the other hand, indicators also ideally have high spatial resolution, as the scales relevant for local-level management and planning are often much smaller [@stevenson2021].
-This latter requirement is particularly crucial for (infrastructure) development strategies and for species management and conservation, both of which tend to require knowledge on species abundance and population dynamics (i.e.
+This latter requirement is particularly crucial for infrastructure development strategies and for species management and conservation, both of which tend to require knowledge on species abundance and population dynamics (i.e.
 demographic rates) that is relevant for their county- or municipality-level decision making [@christie2020].
@@
 @robinson2014; @horswill2019].
-This variation needs to be accounted for to develop successful and sustainable strategies for area use, (harvest) management, and species and biodiversity conservation [@williams2002].
+This variation needs to be accounted for to develop successful and sustainable strategies for area use, harvest management, and species and biodiversity conservation [@williams2002].
 
@@
 It is a line transect survey programme carried out annually in \>120 localities across the country (\>2000 transects) by trained volunteers using pointing dogs.
-The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata (+ref to protocol?) and established routines for quality control and annual releases of publicly available data via the Global Biodiversity Information Facility (GBIF).
+The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata (+ref to protocol?) <!-- well, we do not really have a good citation (i.e.
+nothing published with a doi), but this is something we will release soon.
+The best option is to cite one of the papers where we have used the data (e.g.
+Bowler et al), or one of the GBIF data sets (where we describe the protocol in detail).
+--> and established routines for quality control and annual releases of publicly available data via the Global Biodiversity Information Facility (GBIF).
 As such, it is particularly well suited to become part of a workflow for producing and updating abundance and population indicators on an annual basis.
@@
 Their population dynamics are characterized by large inter-annual fluctuations in abundance [@hjeljord2022], and previous research has tied these fluctuations to rodent cycle through shared predators [@hagen1952; @bowler2020].
-This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis (APH) and has been a central part of research on Fennoscandian grouse population dynamics for many decades [@Elton1942; @hagen1952; @Linden1988; @Steen1985].
-In addition, spring weather conditions and phenology is known to have considerable effects on breeding success and recruitment rates [@Eriksen2023; @Steen1985].
+This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis (APH) and has been a central part of research on Fennoscandian grouse population dynamics for many decades [@Elton1942; @hagen1952; @Linden1988; @Steen1988].
+In addition, spring weather conditions and phenology is known to have considerable effects on breeding success and recruitment rates [@Eriksen2023; @Steen1988].
 Across their distributional range, willow ptarmigan are iconic species with a high cultural value, partly linked to their popularity as game species.
@@
 The published datasets contain both metadata about the transect surveys (survey date, line transect length and location, study area ID, etc.) and bird observation data (species, number of birds of different categories (adult males, adult females, juveniles, and birds of unknown category), perpendicular distance to transect line, exact location, and time of observation).
+As part of the sampling protocol, the field workers are also requested to report if they observed a selected set of mammalian species (small rodents, red foxes *Vulpes vulpes* and mountain hare *Lepus timidus)* while conducting the sampling*.* The small rodent observations from the line transect surveys will form the basis for our rodent co-variate.
 Formally, the data from public land is published as three distinct data sets, one for each of the main public land administrators (Statskog, FeFo and Fjellstyrene, respectively).
@@
 In addition, the local study designs are reviewed by staff at NINA, and common recommendations for study design are provided.
-However, because participation by stakeholders is voluntary, the spatial distribution of transects and sampling effort is not homogeneous.
+However, because participation by stakeholders is voluntary, the spatial distribution of transects and sampling effort is not homogeneous in space.
 In general, the effort is higher in South-Eastern and Central Norway, medium in Northern Norway, and low in Western and Southern Norway.
@@
 -->
+
+<!-- I have added some text above - and think this is sufficient -->
 
@@
 Okay?
---> The overall low density years 2012 and 2015, conversely, often featured lower recruitment and, in some cases, higher survival.
+ERLEND: Ok--> The overall low density years 2012 and 2015, conversely, often featured lower recruitment and, in some cases, higher survival.
 Notably, there were also years with very little spatial synchrony, i.e.
@@
 Okay?
--->
+ERLEND: Ok-->
 
@@
 Resulting reduced primary production limits food availability and as ptarmigan are income breeders that use food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005], lower carrying capacity in such areas is to be expected.
+Note, however, that we did not conduct any formal test of spatial variation in carrying capacity.
 
@@
 While we may speculate that recent population trends may be linked to changes in harvest regulations and/or climatic conditions, considering the whole time-series (2007-2021) illustrated that population densities in all areas were subject to substantial variation across years, featuring periods of stability, increase, and decrease (SF "TimeSeries_PopDens1.pdf").
+At a national scale, recent analysis by @nilsen2020 indicates that the ptarmigan population in Norway increased between 2009 and 2020.
 In most areas, there were also strong year-by-year fluctuations in population density on top of longer-term trends.
@@
 This often resulted in steep population declines towards the following year.
-The fact that these same years also match up with observed peaks in rodent abundance in many areas, together with the largely positive effects of rodent occupancy on recruitment estimated by our model (@fig-paramCorrs C), provides evidence for the Alternative Prey Hypothesis [APH; @Israelsen2020].
+The fact that these same years also match up with observed peaks in rodent abundance in many areas, together with the largely positive effects of rodent occupancy on recruitment estimated by our model (@fig-paramCorrs C), provides evidence for the Alternative Prey Hypothesis [APH; @hagen1952].
 The APH stipulates that high abundance of alternative prey (rodents, in this case) for common predators leads to population growth, and is well-supported throughout the literature for a range of taxa [e.g., @hagen1952; @Kjellander_Nordström_2003; @Reif_Tornberg_Jungell_Korpimäki_2001], including willow ptarmigan [@bowler2020].
-While @Nyström_etal_2006 suggested that gyrfalcons, which are specialized ptarmigan predators, do not respond to rodent populations or switch to alternative prey when ptarmigan populations are low, generalis predators, such as red foxes, are likely to shift from preying on ptarmigans to rodents when the latter become abundant [e.g.
+While @Nyström_etal_2006 suggested that gyrfalcons, which are specialized ptarmigan predators, do not respond to rodent populations or switch to alternative prey when ptarmigan populations are low, generalist predators, such as red foxes, are likely to shift from preying on ptarmigans to rodents when the latter become abundant [e.g.
 @Breisjøberget_etal_2018; @bowler2020].
-Taking a spatial perspective, the highest latitude and highest altitude areas stood out once more, sporting the strongest rodent effects (SF "betaR_Map.pdf").
+Taking a spatial perspective, the highest latitude and highest altitude areas stood out once more, sporting the strongest effects of rodent occupancy (SF "betaR_Map.pdf").
 This could be related to warmer areas generally having larger predator guilds, and consequently more generalists that are able to maintain relatively stable populations irrespective of small rodent abundance [@bowler2020].
@@
 @myrberget1988, for example, observed no change in productivity despite a 50% decrease in abundance, while @pedersen2004 reported strong negative density-dependence over winter and posited that dispersal may be the vital rate that responded to changes in density most strongly.
+At a similar note, @Henden2020 reported negative density dependence when using a Gompertz-model to examine how density and a range of environmental co-variates affeced willow ptarmigan population dynamics in the northernmost parts of Norway.
 While we did not explicitly model density dependence in this study, our results can provide some preliminary insights into potential density feedbacks from both a spatial (cross-population) and a temporal (within-population) angle.
@@
 When considering density dependence across years within select areas, however, we found that higher density years were associated with higher recruitment the same year, but followed by lower apparent survival probabilities and, consequently, lower population growth rates (as determined by post-hoc Pearson correlation coefficients, supplementary file "DD_corrCoef.csv").
-While this seems to support the notion of negative density-dependence, testing for this post-hoc gives results that are confounded with sampling correlation.
+While this seems to support the notion of negative density-dependence, testing for this post-hoc gives results that are confounded with sampling correlation [@Freckleton2006].
 Hence, formally modelling density-dependence, possibly using different forms and time-lags, could prove to be a promising extension of our modelling framework in the future.
@@
 In Norway, willow ptarmigan -- and small game in general -- is managed at the local and/or regional scale, with rather limited national regulation beyond updating the season length every fourth year.
-In effect, both the management system, regulation type (quota type, season length, number of licences, bag limit etc) and quota size is decided by the local or regional stakeholders.
+In effect, both the management system, regulation type (quota type, season length, number of licences, bag limit etc) and quota size is decided by the local or regional stakeholders [@Eriksen2018; @breisjøberget2018].
 Thus, while national estimates (abundance and/or temporal trend in abundance) might be important for red listing decisions and for setting the maximum season length, remaining decisions about harvest management are taken locally.
-<!-- ERLEND: I will add some relevant references in next iteration.
-We have no "go to" reference for the management system (we should write one), but I will see if I can piece together something.
---> The results from our study highlight a large degree of spatio-temporal variation in both ptarmigan densities and demographic rates, suggesting that it is indeed suitable for management decisions to be spatially refined and ideally informed by up-to-date knowledge about recent "local" population processes.
+The results from our study highlight a large degree of spatio-temporal variation in both ptarmigan densities and demographic rates, suggesting that it is indeed suitable for management decisions to be spatially refined and ideally informed by up-to-date knowledge about recent "local" population processes.
 Accessible and easily repeatable modelling workflows, such as the one we have developed in this study, can thus become a valuable source of information for local decision-makers.
 
-Besides pointers on how to adjust management of ptarmigan and their habitat, our results also provided some insights into the value, and possibly opportunities for improving the monitoring programme.
+Our results also provided some insights into the value, and possibly opportunities for improving the monitoring programme.
 First and foremost, our study demonstrates the tremendous potential lying in coordinating structured monitoring using a common protocol and training programme, as well as data processing pipeline.
@@
 Particularly, we found generally lower detection probabilities in the northern half of Norway.
-This may be related to habitat features, as the transects in the North are primarily located in birch forests and more rugged terrain, which may hamper detectability.
+This may be related to habitat features, as the transects in the North might be to a larger extent located in birch forests and more rugged terrain, which may hamper detectability.
 Additionally, the slower life histories in the northern areas are reflected as generally smaller bird clusters as well, and smaller clusters have previously been shown to have a lower detectability than larger ones [e.g.
@@
 While harvest effects on willow ptarmigan have been studied previously, much uncertainty remains [@sandercock_etal2011; @aanes2002; @pedersen2004; @willebrand2001].
-For example, little is known about how harvest pressure and density feedbacks interact on differente temporal and spatial scales [@kvasnes2015], despite this knowledge being crucial for preventing over-exploitation and ensuring sustainable harvest [@williams2002; @breisjøberget2018].
+For example, little is known about how harvest pressure and density feedbacks interact on different temporal and spatial scales [@kvasnes2015], despite this knowledge being crucial for preventing over-exploitation and ensuring sustainable harvest [@williams2002; @breisjøberget2018].
 Additionally, harvest effects often interact with other (emergent) factors such as climate change and habitat degradation, making predictive models that account for harvest alongside other mechanisms invaluable for informing policy changes [@gamelon2019].
@@
 
-Funding for both data collection and analytical work was provided by the Norwegian Environmental Agency (grant numbers 17010522, 19047014, and 22047004).
+Funding for both data collection and analytic work was provided by the Norwegian Environmental Agency (grant numbers 17010522, 19047014, and 22047004).
 
@@
 All code including the "targets pipeline" can be found in the project's repository on GitHub: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.
-The results presented in this paper were created using version 2.0 of the code (Zenodo citation TBA).
+The results presented in this paper were created using version 2.0 of the code [@chloernater_2024_10572340].
 

### 2024-01-29 Manuscript_Nater_etal.qmd

--- 2024-01-26 Manuscript_Nater_etal.qmd
@@
 Quantifying temporal and spatial variation in animal population size and demography is a central theme in ecological research and important for directing management and policy.
-However, this requires field sampling at large spatial extents and over long periods of time, which is not only prohibitively costly but also often politically untenable.
+However, this requires field sampling at large spatial extents and over long periods of time, which is not only prohibitively costly but often politically untenable.
 Participatory monitoring programs (also called citizen science programmes) can alleviate these constraints by recruiting stakeholders and the public to increase the spatial and temporal resolution of sampling effort and hence resulting data.
@@
 Here, we developed a multi-area version of a recently developed integrated distance sampling model (IDSM) and applied it to data from a large-scale participatory monitoring program -- the "Hønsefuglportalen" -- to estimate spatial and temporal variation in population dynamics of willow ptarmigan (*Lagopus lagopus*) in Norway.
-We set up an open and reproducible workflow for exploring temporal, spatial (latitudinal, longitudinal, altitudinal), and residual variation in recruitment, survival, and population density, as well as relationships between vital rates and relevant covariates and signals of density dependence.
+We constructed an open and reproducible workflow for exploring temporal, spatial (latitudinal, longitudinal, altitudinal), and residual variation in recruitment, survival, and population density, as well as relationships between vital rates and relevant covariates and signals of density dependence.
 Recruitment rates varied more across space than over time, while the opposite was the case for survival.
 Slower life history patterns (higher survival, lower recruitment) appeared to be more common at higher latitudes and altitudes, portending differential effects of climate change on ptarmigan across their range.
-While there was variation in the magnitude of the effect small rodent occupancy had on recruitment, the relationships were predominantly positive and thus consistent with the alternative prey hypothesis, which predicts higher reproductive output when there is plenty of alternative prey for shared predators.
+While there was variation in the magnitude of the effect small rodent occupancy had on recruitment, the relationships were predominantly positive and thus consistent with the alternative prey hypothesis.
 Notably, the accurate estimation of covariate effect was only made possible by integrating data from several monitoring areas for analysis.
-Our study thus highlights the potential of participatory monitoring and integrated modelling approaches for estimating and understanding spatio-temporal patterns species abundance and demographic rates, and showcases how corresponding workflows can be set up in a reproducible and semi-automated way that increases their usefulness for informing management and regular reporting towards national and international biodiversity frameworks.
+Our study highlights the potential of participatory monitoring and integrated modelling approaches for estimating and understanding spatio-temporal patterns species abundance and demographic rates, and showcases how corresponding workflows can be set up in a reproducible and semi-automated way that increases their usefulness for informing management and regular reporting towards national and international biodiversity frameworks.
+
+<!--# removed a few words/phrases that I thought were superfulous-->
 
@@
 During this period, estimated population densities varied between `r dens.min$post.sum` birds/km^2^ in the area "`r dens.min$area.name`" close to the coast in central Norway to `r dens.max$post.sum` birds/km^2^ in "`r dens.max$area.name`" further south near the Swedish border.
-In general, recent population density appeared to be lowest in the Northern parts of Norway and highest in the eastern part of central Norway (@fig-densA).
+In general, recent population density appeared to be lowest in the northern Norway and highest in the eastern part of central Norway (@fig-densA).
 Uncertainty in density estimates was relatively consistent, with a few areas (including the one with the lowest estimated density, "`r dens.min$area.name`") sticking out by having substantially less precise estimates (@fig-densB).
@@
 The wide spatial distribution of the line transect monitoring afforded us the opportunity to explore variation in population density and vital rates across a relatively large spatial extent.
-
 Ptarmigan densities across the 41 reporting districts included in our analyses varied from around 2 birds/km^2^ to 55 birds/km^2^, with the lowest densities occurring far north in the country, as well as on the west coast and in the mountains in central Norway [@fig-densA].
@@
 Another important determinant of vital rate variation is density dependence, in particular for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; @aanes2002; @willebrand2001].
-Negative density dependence has been found in several gallinaceous birds such as northern bobwhites *Colinus virginianus* [@mcconnell2018], *Perdix perdix* [@bro2003], and wild turkeys *Meleagris gallopavo* [@mcghee2007]).
+Negative density dependence has been found in several gallinaceous birds such as northern bobwhites *Colinus virginianus* [@mcconnell2018], *Perdix perdix* [@bro2003], and wild turkeys *Meleagris gallopavo* [@mcghee2007].
 For willow ptarmigan, evidence for density-dependent population regulation has been mixed.
@@
 First, at the municipality level there are data with national coverage collected annually by Statistics Norway (<https://www.ssb.no/>).
-Second, many public land owners have data with much higher temporal (daily) and spatial (harvest area) resolution, including both harvest bags and harvest effort (number of hunters pr area pr day).
+Second, many public land owners have data with much higher temporal (daily) and spatial (harvest area) resolution, including both harvest bags and harvest effort (number of hunters per area per day).
 As the IDSM framework is, in essence, an IPM, harvest can be modelled through partitioning of survival into cause-specific mortality in the process model and inclusion of relevant harvest data likelihoods [e.g., @gamelon2021; @nater2021].
 While harvest effects on willow ptarmigan have been studied previously, much uncertainty remains [@sandercock_etal2011; @aanes2002; @pedersen2004; @willebrand2001].
-For example, little is known about how harvest pressure and density feedbacks interact on differente temporal and spatial scales [@kvasnes2015], despite this knowledge being crucial for preventing over-exploitation and ensuring sustainable harvest [@williams2002; @breisjøberget2018].
+For example, little is known about how harvest pressure and density feedbacks interact on different temporal and spatial scales [@kvasnes2015], despite this knowledge being crucial for preventing over-exploitation and ensuring sustainable harvest [@williams2002; @breisjøberget2018].
 Additionally, harvest effects often interact with other (emergent) factors such as climate change and habitat degradation, making predictive models that account for harvest alongside other mechanisms invaluable for informing policy changes [@gamelon2019].

### 2024-01-29 Manuscript_Nater_etal.qmd

--- 2024-01-29 Manuscript_Nater_etal.qmd
@@
 Resulting reduced primary production limits food availability and as ptarmigan are income breeders that use food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005], lower carrying capacity in such areas is to be expected.
-Note, however, that we did not conduct any formal test of spatial variation in carrying capacity.
 
@@
 @myrberget1988, for example, observed no change in productivity despite a 50% decrease in abundance, while @pedersen2004 reported strong negative density-dependence over winter and posited that dispersal may be the vital rate that responded to changes in density most strongly.
-At a similar note, @Henden2020 reported negative density dependence when using a Gompertz-model to examine how density and a range of environmental co-variates affeced willow ptarmigan population dynamics in the northernmost parts of Norway.
+Similarly, @Henden2020 reported negative density dependence when using a Gompertz-model to examine how density and a range of environmental co-variates affeced willow ptarmigan population dynamics in the northernmost parts of Norway.
 While we did not explicitly model density dependence in this study, our results can provide some preliminary insights into potential density feedbacks from both a spatial (cross-population) and a temporal (within-population) angle.

### 2024-01-30 Manuscript_Nater_etal.qmd
Minor edits (processing comments/feedback)

@@
 Integrated population models (IPMs), for example, are able to use multiple streams of data from different field monitoring programmes and/or multiple aspects of single datasets to estimate population sizes and key vital rates.
-Here, we developed a multi-area version of a recently developed integrated distance sampling model (IDSM) and applied it to data from a large-scale participatory monitoring program -- the "Hønsefuglportalen" -- to estimate spatial and temporal variation in population dynamics of willow ptarmigan (*Lagopus lagopus*) in Norway.
+Here, we developed a multi-area version of a recently developed integrated distance sampling model (IDSM) and applied it to data from a large-scale participatory monitoring program -- the "Hønsefuglportalen" -- to study spatio-temporal variation in population dynamics of willow ptarmigan (*Lagopus lagopus*) in Norway.
 We constructed an open and reproducible workflow for exploring temporal, spatial (latitudinal, longitudinal, altitudinal), and residual variation in recruitment, survival, and population density, as well as relationships between vital rates and relevant covariates and signals of density dependence.
@@
 Notably, the accurate estimation of covariate effect was only made possible by integrating data from several monitoring areas for analysis.
-Our study highlights the potential of participatory monitoring and integrated modelling approaches for estimating and understanding spatio-temporal patterns species abundance and demographic rates, and showcases how corresponding workflows can be set up in a reproducible and semi-automated way that increases their usefulness for informing management and regular reporting towards national and international biodiversity frameworks.
-
-<!--# removed a few words/phrases that I thought were superfulous-->
+Our study highlights the potential of participatory monitoring and integrated modelling approaches for estimating and understanding spatio-temporal patterns in species abundance and demographic rates, and showcases how corresponding workflows can be set up in a reproducible and semi-automated way that increases their usefulness for informing management and regular reporting towards national and international biodiversity frameworks.
 
@@
 There is growing demand for biodiversity indicators from international unions, national governments, local management bodies, and corporate and industry actors.
-Such indicators should ideally represent a wide range of biodiversity's states and functions [e.g.
+Indicators should ideally represent a wide range of biodiversity's states and functions [e.g.
 Essential Biodiversity Variables, @pereira2013; @jetz2019], yet the development of suitable indicators for certain attributes, such as species abundance and demography, has been more difficult than for others [@schmeller2018; @waldock2022].
@@
 @feld2009].
-On the other hand, indicators also ideally have high spatial resolution, as the scales relevant for local-level management and planning are often much smaller [@stevenson2021].
+On the other hand, indicators also ideally have good spatial resolution, as the scales relevant for local-level management and planning are often much smaller [@stevenson2021].
 This latter requirement is particularly crucial for infrastructure development strategies and for species management and conservation, both of which tend to require knowledge on species abundance and population dynamics (i.e.
-demographic rates) that is relevant for their county- or municipality-level decision making [@christie2020].
+demographic rates) that is relevant for county- or municipality-level decision making [@christie2020].
 Another reason why abundance and population indicators ideally come with good spatial resolution is that there can be substantial amounts of variation in population dynamics and life history of species across space [e.g.
@@
 
-There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly, requires trained specialists, and depends on at least some top-down control of the program.
+There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly, requires trained specialists, and hinges on a sufficient degree of top-down control of the programme.
 This often limits the amount of data that can be collected, while participatory monitoring, i.e.
@@
 For example, in the United States hunters participate in the collection of bands and wings from harvested American Woodcock (*Scopolax minor)* to estimate survival and age ratios [@zimmerman2010].
-At the European level, the recently established initiative "European Observatory of Wildlife" are offering common field- and analyses protocols and aim to establish a network of "observation points" to monitor wildlife populations at the European level (<https://wildlifeobservatory.org/>).
-In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", [https://honsefugl.nina.no/Innsyn/en](#0){.uri style="font-size: 11pt;"}).
+At the European level, the recently established initiative "European Observatory of Wildlife" is offering common field- and analyses protocols and aims to establish a network of "observation points" for monitoring wildlife populations at the European level (<https://wildlifeobservatory.org/>).
+In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", <https://honsefugl.nina.no/Innsyn/en>).
 It is a line transect survey programme carried out annually in \>120 localities across the country (\>2000 transects) by trained volunteers using pointing dogs.
-The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata (+ref to protocol?) <!-- well, we do not really have a good citation (i.e.
-nothing published with a doi), but this is something we will release soon.
-The best option is to cite one of the papers where we have used the data (e.g.
-Bowler et al), or one of the GBIF data sets (where we describe the protocol in detail).
---> and established routines for quality control and annual releases of publicly available data via the Global Biodiversity Information Facility (GBIF).
+The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata and established routines for quality control and annual releases of publicly available data via the Global Biodiversity Information Facility (GBIF).
 As such, it is particularly well suited to become part of a workflow for producing and updating abundance and population indicators on an annual basis.
@@
 In addition, spring weather conditions and phenology is known to have considerable effects on breeding success and recruitment rates [@Eriksen2023; @Steen1988].
-Across their distributional range, willow ptarmigan are iconic species with a high cultural value, partly linked to their popularity as game species.
+Across their distributional range, willow ptarmigan are an iconic species with a high cultural value, partly linked to their popularity as game species.
 The latter means that information about spatio-temporal variation in demographic rates and population dynamics is particularly important in order to design sustainable harvest strategies [@Eriksen2023].
@@
 In the first three weeks of August each year, trained volunteer fieldworkers collect observations of willow ptarmigan and other grouse species (rock ptarmigan *Lagopus muta*, black grouse *Lyrurus tetrix*, and capercaillie *Tetrao urogallus*) along predefined line transects.
-To increase the detection probability, each team of fieldworkers use a pointing dog to locate the birds.
-Each survey team typically consist of at least two people (one dog handler and one person responsible for following the transect line) and one dog.
+To increase the detection probability, fieldworkers use pointing dogs to locate the birds.
+A survey team typically consist of at least two people (one dog handler and one person responsible for following the transect line) and one dog.
 Often, more than one dog is used for a survey, but only one dog should be used at a time.
@@
 The published datasets contain both metadata about the transect surveys (survey date, line transect length and location, study area ID, etc.) and bird observation data (species, number of birds of different categories (adult males, adult females, juveniles, and birds of unknown category), perpendicular distance to transect line, exact location, and time of observation).
-As part of the sampling protocol, the field workers are also requested to report if they observed a selected set of mammalian species (small rodents, red foxes *Vulpes vulpes* and mountain hare *Lepus timidus)* while conducting the sampling*.* The small rodent observations from the line transect surveys will form the basis for our rodent co-variate.
 Formally, the data from public land is published as three distinct data sets, one for each of the main public land administrators (Statskog, FeFo and Fjellstyrene, respectively).
@@
 In addition, the local study designs are reviewed by staff at NINA, and common recommendations for study design are provided.
-However, because participation by stakeholders is voluntary, the spatial distribution of transects and sampling effort is not homogeneous in space.
-In general, the effort is higher in South-Eastern and Central Norway, medium in Northern Norway, and low in Western and Southern Norway.
+However, because participation by stakeholders is voluntary, the spatial distribution of transects and sampling effort is not homogeneous across space.
+In general, sampling effort is higher in South-Eastern and Central Norway, intermediate in Northern Norway, and low in Western and Southern Norway.
 
@@
 We note that while we refer to the covariate as "rodent occupancy" throughout the manuscript, it can be interpreted as an index for rodent abundance.
-<!-- @ERLEND: I have added this section as it was missing.
-Anything you would like to add here?
--->
-
-<!-- I have added some text above - and think this is sufficient -->
 
@@
 Posterior summaries (median, 95% credible interval, mean, standard deviation, coefficient of variation) for all main parameters are also provided in the supplementary file "PosteriorSummaries_byAreas.csv".
-Supplementary figures (SF) are provided as .pdf files with captions in "SuppFigures_Captions.txt".
+Supplementary figures (SFs) are provided as .pdf files with captions in "SuppFigures_Captions.txt"; all files are deposited on OSF (<https://osf.io/7326r/>).
 
@@
 Effects with negative posterior medians were located mostly at intermediate latitudes, but we note that all of these had posterior distributions featuring substantial overlap with 0 (@fig-paramCorrs C).
-<!-- ERLEND: we sometimes refer to rodent abundance, sometimes occupancy.
-I think what we have is most correctly refered to as an index of abundance.
-CHLOE: Good point about being consistent.
-I see that I have mostly used "occupancy" in text, figures, and code, and to me, that is also the most "precise" description.
-So I have standardized to primarily using "occupancy".
-Nonetheless, we should state that the way we treat occupancy becomes an abundance proxy.
-In fact, I noticed we were missing the description of rodent data.
-So added that, and made a mention there.
-Okay?
-ERLEND: Ok-->
 
@@
 
-Detection decay parameters, which determine detection probability in distance sampling surveys, varied across areas from between `r Det.min$post.sum` in "`r Det.min$area.name`" to `r Det.max$post.sum` in "`r Det.max$area.name`, resulting in detection probabilities over the transect sites (truncation distance = 200 m) from `r pDet.min$post.sum` to `r pDet.max$post.sum`, respectively.
+Detection decay parameters, which determine detection probability in distance sampling surveys, varied across areas from between `r Det.min$post.sum` in "`r Det.min$area.name`" to `r Det.max$post.sum` in "`r Det.max$area.name`, resulting in detection probabilities over the transect sites ranging from `r pDet.min$post.sum` to `r pDet.max$post.sum`, respectively (truncation distance = 200 m) .
 The global average detection decay was `r Det.hyper.mean$post.sum` (detection probability = `r pDet.hyper.mean$post.sum`), and in general, higher values were more common in the Southern half of the country than the Northern half (SF"Avg_detect_Map.pdf").
@@
 The wide spatial distribution of the line transect monitoring afforded us the opportunity to explore variation in population density and vital rates across a relatively large spatial extent.
+
 Ptarmigan densities across the 41 reporting districts included in our analyses varied from around 2 birds/km^2^ to 55 birds/km^2^, with the lowest densities occurring far north in the country, as well as on the west coast and in the mountains in central Norway [@fig-densA].
@@
 We found increasing population trends over recent years in over half of the reporting districts, but population declines were also evident in some areas, particularly in the mountains in central Norway [@fig-lamA].
-While we may speculate that recent population trends may be linked to changes in harvest regulations and/or climatic conditions, considering the whole time-series (2007-2021) illustrated that population densities in all areas were subject to substantial variation across years, featuring periods of stability, increase, and decrease (SF "TimeSeries_PopDens1.pdf").
-At a national scale, recent analysis by @nilsen2020 indicates that the ptarmigan population in Norway increased between 2009 and 2020.
+Predominantly increasing population trends are consistent with a recent national-scale analysis by @nilsen2020 which found an overall increase in the Norwegian ptarmigan population between 2009 and 2020.
+While we may speculate that recent population trends could be linked to changes in harvest regulations and/or climatic conditions, considering the whole time-series (2007-2021) illustrated that population densities in all areas were subject to substantial variation across years, featuring periods of stability, increase, and decrease (SF "TimeSeries_PopDens1.pdf").
 In most areas, there were also strong year-by-year fluctuations in population density on top of longer-term trends.
 Some of the resulting "high density years" were highly synchronized across large spatial scales, such as the years 2011, 2013, and 2018.
-Taking a closer look, we find that these are years that are characterized by high recruitment (SF "TimeSeries_rRep.pdf"), follwed by a low survival the year after (SF "TimeSeries_pSurv.pdf").
+Taking a closer look, we find that these are years that are characterized by high recruitment (SF "TimeSeries_rRep.pdf"), followed by a low survival the year after (SF "TimeSeries_pSurv.pdf").
 This often resulted in steep population declines towards the following year.
@@
 This is the case especially for the relatively large residual variation [@fig-varDecomp] but also relevant for constant spatial and shared temporal variation.
-In previous work, both based on marked birds [@Eriksen2023] and unmarked birds [@bowler2020; @Henden2020; @Novoa2016], spring conditions has come out as an important predictor of ptarmigan recruitment rates.
-In general, it is found warmer and earlier springs result in earlier breeding, larger clutch sizes [@Eriksen2023], and resulting higher recruitment rates measured in the late summer and early fall (as in our study).
-In @bowler2020, it was reported that the strength of this relationship was not consistent in time and space, but was generally stronger in colder areas, similar to the effect of rodent occupancy reported here.
-In practice, measures of the cover of ericaceous shrubs (a proxy for food availability) or spatially-explicit spring green up dates derived from remote-sensing data constitute relevant candidate covariates for future work alongside temperature.
+In previous work based on both marked [@Eriksen2023] and unmarked birds [@bowler2020; @Henden2020; @Novoa2016], spring conditions has come out as an important predictor of ptarmigan recruitment rates.
+In general, warmer and earlier springs seem to favour earlier breeding, larger clutch sizes [@Eriksen2023], and resulting higher recruitment rates measured in the late summer and early fall.
+@bowler2020 further reported that the strength of this relationship was not consistent in time and space, but was generally stronger in colder areas, similar to what we found for the effect of rodent occupancy here.
+In practice, measures representing spring conditions, such as the cover of ericaceous shrubs (a proxy for food availability) or spatially-explicit spring green up dates derived from remote-sensing data, thus constitute relevant candidate covariates for future work alongside temperature.
 
@@
 @myrberget1988, for example, observed no change in productivity despite a 50% decrease in abundance, while @pedersen2004 reported strong negative density-dependence over winter and posited that dispersal may be the vital rate that responded to changes in density most strongly.
-Similarly, @Henden2020 reported negative density dependence when using a Gompertz-model to examine how density and a range of environmental co-variates affeced willow ptarmigan population dynamics in the northernmost parts of Norway.
+Similarly, @Henden2020 reported negative density dependence when using a Gompertz-model to examine how density and a range of environmental covariates affected willow ptarmigan population dynamics in the northernmost parts of Norway.
 While we did not explicitly model density dependence in this study, our results can provide some preliminary insights into potential density feedbacks from both a spatial (cross-population) and a temporal (within-population) angle.
@@
 
-Management decisions made at the resolution of large geo-political boundaries (e.g., Norway) run a high risk of being inadequate when there is substantial spatial variation in demographic processes and population dynamics, as is the case for willow ptarmigan.
-In Norway, willow ptarmigan -- and small game in general -- is managed at the local and/or regional scale, with rather limited national regulation beyond updating the season length every fourth year.
-In effect, both the management system, regulation type (quota type, season length, number of licences, bag limit etc) and quota size is decided by the local or regional stakeholders [@Eriksen2018; @breisjøberget2018].
-Thus, while national estimates (abundance and/or temporal trend in abundance) might be important for red listing decisions and for setting the maximum season length, remaining decisions about harvest management are taken locally.
+Management decisions made at the resolution of large geopolitical boundaries (e.g., Norway) run a high risk of being inadequate when there is substantial spatial variation in demographic processes and population dynamics, as is the case for willow ptarmigan.
+In Norway, willow ptarmigan -- and small game in general -- is managed at the local and/or regional scale, with rather limited national regulation beyond updating the length of the hunting season every fourth year.
+In effect, management system, regulation type (quota type, season length, number of licences, bag limit etc.), and quota size are governed by the local or regional stakeholders [@Eriksen2018; @breisjøberget2018].
+Thus, while national estimates (abundance and/or temporal trend in abundance) might be important for red listing decisions and for setting the maximum hunting season length, remaining decisions about harvest management are taken locally.
 The results from our study highlight a large degree of spatio-temporal variation in both ptarmigan densities and demographic rates, suggesting that it is indeed suitable for management decisions to be spatially refined and ideally informed by up-to-date knowledge about recent "local" population processes.
@@
 Our results also provided some insights into the value, and possibly opportunities for improving the monitoring programme.
-First and foremost, our study demonstrates the tremendous potential lying in coordinating structured monitoring using a common protocol and training programme, as well as data processing pipeline.
-These were indeed the prerequisites that allowed us to easily and efficiently integrate data collected across the entire country in a joint analysis, and draw inference on spatio-temporal variation in demography and population dynamics at a large scale but also with good spatial resolution.
+First and foremost, our study demonstrates the tremendous potential lying in coordinating structured monitoring that employs common sampling protocol, training programmes, and data processing pipelines.
+These were indeed the prerequisites that allowed us to easily and efficiently integrate data collected across the entire country in a joint analysis, and draw inference on fine-scale spatio-temporal variation in demography and population dynamics at across a large area.
 While overall less variable across space and time than vital rates, differences in detection probabilities were nonetheless evident (SFs "Avg_detect_Map.pdf" and "TimeSeries_pDetect.pdf") and may help with mapping out potential for improvement in the monitoring programme.
 Particularly, we found generally lower detection probabilities in the northern half of Norway.
-This may be related to habitat features, as the transects in the North might be to a larger extent located in birch forests and more rugged terrain, which may hamper detectability.
+This may be related to habitat features, as the transects in the North might be to a larger extent located in birch forests and rugged terrain, which may hamper detectability.
 Additionally, the slower life histories in the northern areas are reflected as generally smaller bird clusters as well, and smaller clusters have previously been shown to have a lower detectability than larger ones [e.g.
 @bowler2020, see also next section].
-Our modelling framework can be easily adapted for studying the impact of different variables on detectability.
+Our modelling framework can be easily adapted for studying the impact of these and other variables on detectability (see below).
 Additionally, our results could motivate taking a closer look at monitoring challenges and potential improvements in the northern part of the country in particular.
@@
 First, the precision and accuracy of model estimates might be increased through better accounting for heterogeneity and potential biases in detection of birds during the line transect surveys.
-In an earlier study analyzing data from the same monitoring programme, @bowler2020, for example, found that detection probability was not independent of the size of group birds were part of, resulting in birds in larger groups being more likely to be detected, especially at larger distances.
+In an earlier study analyzing data from the same monitoring programme, @bowler2020, found that detection probability was not independent of the size of group birds were part of, resulting in birds in larger groups being more likely to be detected, especially at larger distances.
 When birds are observed in larger groups, it is also not unlikely that human observers may miscount, i.e.
@@
 If misclassification happened at random, mixture models could be used to determine the likely age class of individuals to whom no age class was assigned during observations [@mccrea2013].
-In our case, we might suspect that an observed is more likely to classify an adult bird as juvenile rather than the other way around, and more likely to assingn "unknown" age class to juveniles than adults.
+In our case, we might suspect that an observed is more likely to classify an adult bird as juvenile rather than the other way around, and more likely to assign "unknown" age class to juveniles than adults.
 One reason for this is that observers look for specific signs to classifying a bird as adult (e.g.
@@
 Nonetheless, this illustrates a way for how smaller datasets from single or subsets of areas can be integrated into a large-scale modelling framework.
-Other relevant data could be included using the same approach, for example data from ongoing nesting success monitoring (Nilsen, E.B.
-unpublished data), data from past studies of marked birds [@sandercock_etal2011], and data from other monitoring programs for breeding birds based on point counts (see e.g.
+Other relevant data could be included using the same approach, for example data from ongoing nesting success monitoring, data from past studies of marked birds [@sandercock_etal2011], and data from other monitoring programs for breeding birds based on point counts (see e.g.
 the Norwegian Breeding Bird Monitoring: <https://hekkefuglovervakingen.nina.no/>).
 The most relevant source of data to be included into the IDSM framework in the near future, however, is harvest data.
-Such data might be available with two different spatial and temporal resolution.
+Such data might be available with different spatial and temporal resolutions.
 First, at the municipality level there are data with national coverage collected annually by Statistics Norway (<https://www.ssb.no/>).
-Second, many public land owners have data with much higher temporal (daily) and spatial (harvest area) resolution, including both harvest bags and harvest effort (number of hunters pr area pr day).
+Second, many public land owners have data with much higher temporal (daily) and spatial (harvest area) resolution, including both harvest bags and harvest effort (number of hunters per area per day).
 As the IDSM framework is, in essence, an IPM, harvest can be modelled through partitioning of survival into cause-specific mortality in the process model and inclusion of relevant harvest data likelihoods [e.g., @gamelon2021; @nater2021].
@@
 We have done this by setting up "targets" pipeline [@landau2021], which allows (re-)running the complete workflow from downloading the publicly available data to visualizing the results produced by the IDSM [@fig-workflow].
-Modern applied ecology needs research to be published not just as manuscripts, but as reproducible and well documented workflows [@lewis2018].
+Modern applied ecology needs research to be published not just as scientific papers, but as reproducible and well documented workflows [@lewis2018].
 This is particularly crucial for research that is (to be) closely tied to management and/or used to create biodiversity indicators that are to be reported nationally or internationally, or to be used by industrial partners [@powers2019].

### 2024-02-19 Analysis_SimData.R
Changed toggle

@@ -25,7 +25,7 @@ R_perF <- FALSE
 R_parent_drop0 <- TRUE
 
 # Time variation in survival
-survVarT <- TRUE
+survVarT <- FALSE
 
 # Rodent covariate on reproduction
 fitRodentCov <- FALSE

### 2024-02-26 Analysis_SimData_Replicates.R
Added package for spread_draws()

@@ -227,27 +227,27 @@ for(i in 1:length(simSeed.list)){
     ## Summarise posteriors to minimum necessary for plotting sim checks
     
     # Recruitment parameters
-    R_year <- IDSM.out %>% spread_draws(R_year[year])
-    Mu_R <- IDSM.out %>% spread_draws(Mu.R) %>% mutate(lab_code = "Mu.R")
-    sigmaT_R <- IDSM.out %>% spread_draws(sigmaT.R) %>% mutate(lab_code = "sigmaT.R")
+    R_year <- IDSM.out %>% tidybayes::spread_draws(R_year[year])
+    Mu_R <- IDSM.out %>% tidybayes::spread_draws(Mu.R) %>% mutate(lab_code = "Mu.R")
+    sigmaT_R <- IDSM.out %>% tidybayes::spread_draws(sigmaT.R) %>% mutate(lab_code = "sigmaT.R")
     
     # Survival parameters
-    Mu_S1 <- IDSM.out %>% spread_draws(Mu.S1) %>% mutate(Surv = "S1") %>% rename(S = Mu.S1) %>% select(S, Surv)
-    Mu_S <- IDSM.out %>% spread_draws(Mu.S) %>% mutate(Surv = "S") %>% rename(S = Mu.S) %>% select(S, Surv)
+    Mu_S1 <- IDSM.out %>% tidybayes::spread_draws(Mu.S1) %>% mutate(Surv = "S1") %>% rename(S = Mu.S1) %>% select(S, Surv)
+    Mu_S <- IDSM.out %>% tidybayes::spread_draws(Mu.S) %>% mutate(Surv = "S") %>% rename(S = Mu.S) %>% select(S, Surv)
     Mu_S_data <-  tibble(S = Mu_S$S/Mu_S1$S, Surv = "S2") %>% bind_rows(., Mu_S1, Mu_S)
     
     # Detection parameters
-    Mu_dd <- IDSM.out %>% spread_draws(mu.dd) %>% mutate(lab_code = "mu.dd")
-    sigmaT_dd <- IDSM.out %>% spread_draws(sigmaT.dd) %>% mutate(lab_code = "sigmaT.dd")
-    esw_year <- IDSM.out %>% spread_draws(esw[year])
-    p_year <- IDSM.out %>% spread_draws(p[year])
+    Mu_dd <- IDSM.out %>% tidybayes::spread_draws(mu.dd) %>% mutate(lab_code = "mu.dd")
[…]

### 2024-03-05 R/plotSimCheck_replicates.R
Fixed paths / directories

@@ -28,7 +28,7 @@ plotSimCheck_replicates <- function(plotColors = "customRainbow", thin = 1) {
   
   ## Load information on simulation and run seeds
   simSeeds <- readRDS("simData/seedList.rds")
-  runSeeds <- readRDS("simModelFits_sum/seedInfo.rds")
+  runSeeds <- readRDS("simModelFits/seedInfo.rds")
   
   ## Set up tibbles to collate simulated data and model posteriors
   N_simData <- D_simData <-  R_simData <- det_simData <- simParams <- tibble()
@@ -135,7 +135,14 @@ plotSimCheck_replicates <- function(plotColors = "customRainbow", thin = 1) {
   #---------------------#
   
   ## Plotting directory
-  ifelse(!dir.exists("Plots/SimCheck_replicates"), dir.create("Plots/SimCheck_replicates"), FALSE) ## Check if folder exists, if not create folder
+  if(!dir.exists("Plots")){
+    dir.create("Plots")
+    dir.create("Plots/SimCheck_replicates")
+  }
+  
+  if(!dir.exists("Plots/SimCheck_replicates")){
+    dir.create("Plots/SimCheck_replicates")
+  }
   
   ## List of plot names (required for targets integration)
   plot.paths <- c()

### 2024-06-26 Manuscript_Nilsen_Nater.qmd
Manuscript edits (revisions)

--- 2024-01-05 Manuscript_Nilsen_Nater.qmd
@@
 ---
-title: "An integrated open population distance sampling approach for modelling age-structured populations"
+title: "Open integrated distance sampling for modelling age-structured population dynamics"
 format: html
@@
 The advent of open population formulations based on the DS framework greatly extended the scope of the models, but much untapped potential remains in models that estimate temporal dynamics not only in abundance but also in the underlying demographic rates.
-Here, we present an integrated distance sampling approach that utilize age-structured survey data and auxiliary data from marked individuals to jointly estimate population dynamics and the demographic rates (recruitment rate and survival probability) that drive temporal changes .
-The core of the resulting model is, in essence, equivalent to both an integrated population model and a matrix population model with two age classes: juveniles and adults.
+Here, we present an integrated distance sampling approach that utilize age-structured survey data and auxiliary data from marked individuals to jointly estimate population dynamics and the demographic rates (recruitment rate and survival probability) that drive temporal changes.
+The resulting model is equivalent to an integrated population model with a two age classes: juveniles and adults.
 The integrated framework allows making full use of the available data by effectively combining line transect and telemetry data, and can easily be adapted to include additional and/or different data types.
@@
 Concurrent with the development of more sophisticated DS models, another group of models has emerged and rapidly gained popularity, not least for their ability to disentangle demographic processes underlying population dynamic: integrated population models [IPMs, @schaub2021].
-Through joint analysis of multiple datasets, IPMs allow simultaneous estimation of population size and composition, as well as all vital rates that form part of an underlying age- or stage-structured population model.
+Through joint analysis of multiple datasets (or multiple components of the same dataset), IPMs allow simultaneous estimation of population size and composition, as well as all vital rates that form part of an underlying age- or stage-structured population model.
 Since both DS models and IPMs estimate population size/density, a combination of the two frameworks has the potential to provide good estimates of both population- and demographic parameters by maximizing knowledge gained from transect surveys by augmenting them with other available data [e.g.
@@
 Both juveniles and adults survive from year $t$ census to year $t+1$ census with survival probability $S_t$.
-As ptarmigan can reproduce already as 1-year old, all survivors then produce offspring in late June which recruit into the population as juveniles just prior to the census in year $t+1$ according to a recruitment rate $R_{t+1}$.
+As ptarmigan can reproduce already as 1-year old, all survivors may then produce offspring in late June which recruit into the population as juveniles just prior to the census in year $t+1$ according to a recruitment rate $R_{t+1}$.
 The changes in densities (numbers) of juveniles and adults in the population, $D_{juv}$ and $D_{ad}$, can thus be expressed as
@@
 Vital rates (survival probabilities $S$, recruitment rates $R$) and detection parameters (half-normal detection parameters $\sigma$) can all be modelled as time-dependent in our framework.
-For both the tests with simulated data and the case study described below, we implemented log-normally distributed random year effects on all parameters except survival, which was set to be constant.
-In the case study, we additionally included an effect of rodent occupancy (see details below) on log recruitment rates, resulting in the following model:
+For both the tests with simulated data and the case study described below, we implemented log-normally distributed random year effects on recruitment rate and detection probability.
+Survival, on the other hand, was set to be constant.
+In the case study, we additionally included a covariate effect (see details below) on log recruitment rates, resulting in the following model:
 
@@
 \begin{equation}
-  log(R_t) = log(\mu_R) + \beta*RodentOcc_t + \epsilon_t
+  log(R_t) = log(\mu_R) + \beta*cov_t + \epsilon_t
 \end{equation}
-$$ where $\mu_R$ is the baseline recruitment rate, $\beta$ the slope of the effect of rodent occupancy, and $\epsilon_t$ the normally distributed random effects.
+$$ $\mu_R$ represents the mean recruitment rate if the covariate $cov_t$ is centered around 0 (e.g.
+z-standardized) or a baseline recruitment rate corresponding to $cov_t = 0$ if the covariate is not centered.
+$\beta$ the slope of the covariate effect, and $\epsilon_t$ the normally distributed random effects.
 
@@
 --> Nevertheless, mis-identification can occur, and in addition a proportion of the birds are registered as "unknown" age and/or sex.
+Besides bird observations, field workers also record whether (1) or not (0) they encounter small rodents on any transect line, allowing the proportion of transect lines with small rodent detection to be used as measure of rodent occupancy (covariate ranging from 0 to 1).
 Since 2019, data has been collected through a mobile app tailormade for this project <!-- Add name + maybe link to download in app store?
@@
 300k and 40k and thereof were discarded as burn-in prior to thinning with factors 5 and 20 , leaving us with 40k and 3k posterior samples per chain (total of 120k and 12k samples per run), respectively.
+MCMC parameters were chosen to yield a representative number of samples from converged chains, and convergence was determined based on visual inspection of trace plots.
 
@@
 For the latter, the yellow distribution is for the intercept, representing a baseline recruitment rate when rodent occupancy is low.
-The turquoise distributions are for year-specific estimates or recruitment rate, with darker colors indicating later years.](Figures/PostDens_Mu_S_tR.png){#fig-lierne-demo}
+The turquoise distributions are for year-specific estimates or recruitment rate, with darker colors indicating later years.
+For a visualization of the time-series of recruitment rates, see supplementary figure "TimeSeries_rRep" on OSF.](Figures/PostDens_Mu_S_tR.png){#fig-lierne-demo}
 
@@
 Thus, biases in the reported age ratios may affect estimated of demographic rates, but not so much population density.
-Since the proportion of "unknown sex and age" observations in our ptarmigan case study was rather low, potential biases in estimates resulting from age misclassification are expected to be small.
+Since the proportion of "unknown sex and age" observations in our ptarmigan case study was low (< 3% of observations), potential biases in estimates resulting from age misclassification are expected to be small.
 Nonetheless, future developments of the IDSM modelling framework should focus on ways of accounting explicitly for misclassification of age class in the field.
@@
 [@Israelsen2020; @Sandercock2011]), we could have expected somewhat higher estimates of survival probability.
-One potential reason for the overall lower estimates obtained here is that our IDSM analysis assumed constant survival over the period 2007-2021 while survival, in reality, may have changed over time.
+One potential reason for the overall lower estimates obtained here is that our IDSM analysis assumed constant survival over the period 2007-2021 while survival, in reality, must have varied over time.
 If survival in more recent years, when telemetry data was collected [and the study of e.g.
@@
 In the ptarmigan case study we thus attempted to investigate the effect of small rodent abundance (approximated as the proportion of transect lines on which rodents were reported each year) on recruitment rate.
-<!-- Deleted this because we did NOT include the rodent effect in the simulated data tests --> Unfortunately, we were not able to detect a clear effect of rodent abundance due to large uncertainty associated with the estimate (Supplementary Figure "Rep_betaR.R.png").
+<!-- Deleted this because we did NOT include the rodent effect in the simulated data tests --> We were not able to detect a clear effect of rodent abundance due to large uncertainty associated with the estimate (Supplementary Figure "Rep_betaR.R.png").
 This may seem somewhat surprising at first glance, given that such a pattern has been reported repeatedly in the literature (see e.g.
@@
 Second, it is possible that rodent effects were obscured by other, potentially stronger, covariate effects.
-Previous research has shown that ptarmigan recruitment is also sensitive to the weather in the late winter and spring (before and during the breeding season); as we did not fit any weather covariates to the model so there is a possibility that strong unaccounted for effects of spring conditions in certain years may have masked any remaining effects of small rodent abundance.
+Previous research has shown that ptarmigan recruitment is also sensitive to the weather in the late winter and spring (before and during the breeding season); as we did not fit any weather covariates to the model there is a possibility that effects of spring conditions in certain years may have masked any remaining effects of small rodent abundance.
 Finally, the data set used in this analysis is relatively short (15 years), leaving us with somewhat limited statistical power to detect effects of temporal covariates .
@@
 Moreover, hierarchical nature of the model makes is straight forward to adapt to different species and to include different suites of environmental covariates on the demographic rates.
-Finally, it constitutes a modelling framework that is well suited for extension to multiple areas and thus able to capitalize on space-for-time substitution to produce large-scale and spatially explicit estimates of population density, demographic rates, and environmental effects from large-scale (participatory) monitoring.
+Finally, it constitutes a modelling framework that is well suited for extension to multiple areas and thus able to capitalize on space-for-time substitution [@lovell2023] to produce large-scale and spatially explicit estimates of population density, demographic rates, and environmental effects from large-scale (participatory) monitoring.
 

### 2024-06-26 Manuscript_Nilsen_Nater.qmd

--- 2024-01-07 Manuscript_Nilsen_Nater.qmd
@@
 ---
-title: "An integrated open population distance sampling approach for modelling age-structured populations"
+title: "Open integrated distance sampling for modelling age-structured population dynamics"
 format: docx
@@
 One set of approaches that has been used extensively to estimate abundance of wildlife populations is Distance Sampling (DS) methods for line or point transect survey data.
-Historically, DS models were only available as closed* *population models, and did not allow for the estimation of changes in abundance through time.
+Historically, DS models were only available as closed population models, and did not allow for the estimation of changes in abundance through time.
 The advent of open population formulations based on the DS framework greatly extended the scope of the models, but much untapped potential remains in models that estimate temporal dynamics not only in abundance but also in the underlying demographic rates.
-Here, we present an integrated distance sampling approach that utilize age-structured survey data and auxiliary data from marked individuals to jointly estimate population density and the demographic rates (recruitment rate and survival probability) that drive temporal changes density.
-The core of the resulting model is, in essence, equivalent to both an integrated population model and a matrix population model with two age classes: juveniles and adults.
+Here, we present an integrated distance sampling approach that utilize age-structured survey data and auxiliary data from marked individuals to jointly estimate population density and the demographic rates (recruitment rate and survival probability) that drive temporal changes.
+The resulting model is equivalent to an integrated population model with a two age classes: juveniles and adults.
 The integrated framework allows making full use of the available data by effectively combining line transect and telemetry data, and can easily be adapted to include additional and/or different data types.
@@
 As the surveys are conducted in early August, juveniles can be distinguished from adults by their smaller body size.
-Males and females are manly distinguished by the sound (males often make a characteristic sound when being flushed.
+Males and females are manly distinguished by the sound (males often make a characteristic sound when being flushed).
 Nevertheless, mis-identification can occur, and in addition a proportion of the birds are registered as "unknown" age and/or sex.
-Since 2019, data has been collected through a mobile app tailormade for this project, which is available trhough App Store (for ios phones) and Google Play (for Android phones).
+Besides bird observations, field workers also record whether (1) or not (0) they encounter small rodents on any transect line, allowing the proportion of transect lines with small rodent detection to be used as measure of rodent occupancy (covariate ranging from 0 to 1).Since 2019, data has been collected through a mobile app tailormade for this project, which is available through App Store (for iOS phones) and Google Play (for Android phones).
 Before 2019, field workers reported their data through a dedicated web portal.
@@
 300k and 40k and thereof were discarded as burn-in prior to thinning with factors 5 and 20 , leaving us with 40k and 3k posterior samples per chain (total of 120k and 12k samples per run), respectively.
+MCMC parameters were chosen to yield a representative number of samples from converged chains, and convergence was determined based on visual inspection of trace plots.
 Posterior samples from the model fitted to real data are available at @Nilsen2024 (in folder PosteriorSamples_LierneCaseStudy).
@@
 
-Real data, however, are likely to be subject to certain biases, and such biases might result in biased parameter estimates unless accounted for.
+Real data, however, are likely to be subject to certain biases, which may result in biased parameter estimates unless accounted for.
 One type of bias that is likely to be common for age-structured distance sampling data arises from failure to (correctly) classify the age class of observed individuals.
@@
 
+Compared to previous studies on ptarmigan (see e.g.
+[@Israelsen2020; @Sandercock2011]), we could have expected somewhat higher estimates of survival probability.
+One potential reason for the overall lower estimates obtained here is that our IDSM analysis assumed constant survival over the period 2007-2021 while survival, in reality, must have varied over time.
+If survival in more recent years, when telemetry data was collected [and the study of e.g.
+\@ @Israelsen2020 was carried out] was higher than in earlier years -- something that seems likely given population increase over recent years -- an average over the entire time period is expected to be lower.
+On a different note, we can also not exclude the possibility of a small degree of bias in survival estimates due to misclassification of age in the data (see above), especially seeing as the IDSM's recruitment rate estimates also appear somewhat high compared to other studies on ptarmigan (+refs).
+<!-- If we have any "biological" thoughts about why somewhat high recruitment rates may not be that ridiculous here, then that could be added here -->
+
 In addition to estimating demographic rates from line transect data, the IDSM also allows including relevant environmental effects on the demographic rates themselves, and not just on population growth rate as a whole ($\lambda$).
 In the ptarmigan case study we thus attempted to investigate the effect of small rodent abundance (approximated as the proportion of transect lines on which rodents were reported each year) on recruitment rate.
-Unfortunately, we were not able to detect a clear effect of rodent abundance due to large uncertainty associated with the estimate (see supplementary figures in @Nilsen2024).
+We were not able to detect a clear effect of rodent abundance due to large uncertainty associated with the estimate (see Supplementary Figures "Rep_betaR.R.png" in @Nilsen2024).
 This may seem somewhat surprising given that such a pattern has been reported repeatedly in the literature (see e.g.
 @Bowler2020).
-We speculate that there are at least three potential (non-mutual) explanations to this result.
+We speculate that there are at least three potential and not mutually exclusive explanations to this result. 
 The first is that our covariate data may not have been well suited for estimating effects on recruitment.
@@
 Second, it is possible that rodent effects were obscured by other, potentially stronger, covariate effects.
-Previous research has shown that ptarmigan recruitment is also sensitive to the weather in the late winter and spring (before and during the breeding season); as we did not fit any weather covariates to the model so there is a possibility that strong unaccounted for effects of spring conditions in certain years may have masked any remaining effects of small rodent abundance.
+Previous research has shown that ptarmigan recruitment is also sensitive to the weather in the late winter and spring (before and during the breeding season); as we did not fit any weather covariates to the model so there is a possibility that effects of spring conditions in certain years may have masked any remaining effects of small rodent abundance.
 Finally, the data set used in this analysis is relatively short (15 years), leaving us with somewhat limited statistical power to detect effects of temporal covariates .

### 2024-06-26 Manuscript_Nilsen_Nater.qmd
Manuscript edits

@@
 The advent of open population formulations based on the DS framework greatly extended the scope of the models, but much untapped potential remains in models that estimate temporal dynamics not only in abundance but also in the underlying demographic rates.
-Here, we present an integrated distance sampling approach that utilize age-structured survey data and auxiliary data from marked individuals to jointly estimate population density and the demographic rates (recruitment rate and survival probability) that drive temporal changes.
+Here, we present an integrated distance sampling approach that utilizes age-structured survey data and auxiliary data from marked individuals to jointly estimate population density and the demographic rates (recruitment rate and survival probability) that drive temporal changes in density.
 The resulting model is equivalent to an integrated population model with a two age classes: juveniles and adults.
@@
 Through a comprehensive simulation study we show that the model is able to adequately recover true population and vital rate dynamics.
-Subsequent application to data from a study of willow ptarmigan (Lagopus lagopus) in Norway showcases the frameworks ability to recover both fluctuations and trends in population dynamics and highlights its potential applicability to a wide range of species surveyed using distance sampling approaches.*
+Subsequent application to data from a study of willow ptarmigan (_Lagopus lagopus_) in Norway showcases the frameworks ability to recover both fluctuations and trends in population dynamics and highlights its potential applicability to a wide range of species surveyed using distance sampling approaches.*
 
@@
 Over the last few decades, tremendous progress has been made towards this end.
-This progress is partly driven the development and application of new field data collection methods and approaches, such as citizen science data [@Danielsen2022], camera trap data [@Hamel2013] and the collection of environmental DNA data [@Beng2020].
+This progress is partly driven by development and application of new field data collection methods and approaches, such as citizen science data [@Danielsen2022], camera trap data [@Hamel2013] and the collection of environmental DNA data [@Beng2020].
 In addition, developments of novel statistical methods alongside decreases in computational costs now allow researchers to estimate abundance and demographic rates in situations where it was not feasible before [@Zipkin2021].
@@
 Monitoring programmes focusing on abundance trends over larger areas, on the other hand, are typically based on data from unmarked animals.
-One often used approach for such surveys is to structure data collection around the distance sampling (DS) method.
+One often used approach for such surveys is distance sampling (DS).
 DS has been used for estimating animal abundance in a wide range of contexts and for a variety of taxa [@buckland2015].
@@
 \end{equation}
-$$ $N_{juv,j,t}$ and $N_{ad,j,t}$ are then linked back to the population model by converting them to densities through multiplication with $2L_{j,t}W$ (where $L_{j,t}$ is length of transect $j$ in year $t$, and $W$ is the truncation distance).
+$$ $N_{juv,j,t}$ and $N_{ad,j,t}$ are then linked back to the population model by converting them to densities through multiplication by $2L_{j,t}W$ (where $L_{j,t}$ is length of transect $j$ in year $t$, and $W$ is the truncation distance).
 
@@
 
-We developed an integrated populaiton model that jointly analyses line transect distance sampling survey data and data from marked individuals to estimate population abundance, survival probabilities, and recruitment rates over time.
+We developed an integrated population model that jointly analyses line transect distance sampling survey data and data from marked individuals to estimate population abundance, survival probabilities, and recruitment rates over time.
 We first used simulated data to examine the model's ability to recover the underlying parameters when they were known.
@@
 
-Open population formulations of the distance sampling model have previously been presented [@sollmann2015; @Bowler2020; @moore2011], and in these studies also applied to various ecological systems.
+Open population formulations of the distance sampling model have previously been presented and applied to various ecological systems[@sollmann2015; @Bowler2020; @moore2011].
 The model presented here extends those previous applications by formulating the underlying population process model as a stage-structured matrix model [@caswell2000] in which the matrix elements are represented by annual survival probabilities and recruitment rates.
@@
 Observations with incorrectly classified age have the potential to introduce bias in the IDSMs relative estimates of survival and recruitment.
-This is due to the way it useses the distance sampling data to estimate survival and recruitment rates.
+This is due to the way it uses the distance sampling data to estimate survival and recruitment rates.
 In our process model, the population growth rate ($\lambda$) is determined by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, and this creates a dependence between the demographic parameters.
@@
 Thus, biases in the reported age ratios may affect estimated of demographic rates, but not so much population density.
-Since the proportion of "unknown sex and age" observations in our ptarmigan case study was low (< 3% of observations), potential biases in estimates resulting from age misclassification are expected to be small.
+Since the proportion of "unknown sex and age" observations in our ptarmigan case study was low (\< 3% of observations), potential biases in estimates resulting from age misclassification are expected to be small.
 Nonetheless, future developments of the IDSM modelling framework should focus on ways of accounting explicitly for misclassification of age class in the field.
@@
 Second, it is possible that rodent effects were obscured by other, potentially stronger, covariate effects.
-Previous research has shown that ptarmigan recruitment is also sensitive to the weather in the late winter and spring (before and during the breeding season); as we did not fit any weather covariates to the model so there is a possibility that effects of spring conditions in certain years may have masked any remaining effects of small rodent abundance.
+Previous research has shown that ptarmigan recruitment is also sensitive to the weather in the late winter and spring (before and during the breeding season); as we did not fit any weather covariates to the model, there is a possibility that effects of spring conditions in certain years may have masked any remaining effects of small rodent abundance.
 Finally, the data set used in this analysis is relatively short (15 years), leaving us with somewhat limited statistical power to detect effects of temporal covariates .
@@
 The IDSM framework is very flexible, however, and open to the inclusion of additional/other auxiliary data that contains information on demographic rates or population size/density.
-Moreover, hierarchical nature of the model makes is straight forward to adapt to different species and to include different suites of environmental covariates on the demographic rates.
+Moreover, the hierarchical nature of the model makes is straightforward to adapt to different species and to include different suites of environmental covariates on the demographic rates.
 Finally, it constitutes a modelling framework that is well suited for extension to multiple areas and thus able to capitalize on space-for-time substitution [@lovell2023] to produce large-scale and spatially explicit estimates of population density, demographic rates, and environmental effects from large-scale (participatory) monitoring.

### 2024-06-26 Manuscript_Nilsen_Nater.qmd
Reformatted authorship statement to CRediT

@@
 
-EBN Lead - Drafted the model code; Lead - Wrote first version of ms; Lead - Data collection; Contributed - Modelling and analyses
+**Erlend B.
+Nilsen**: Conceptualization, Methodology, Software, Formal analysis, Investigation, Data Curation, Writing - Original Draft, Writing - Review & Editing, Project administration, Funding acquisition.
 
-CRN Lead - Updated, further developed, and finalized model code; Lead - Modelling and analyses; Contributed - Wrote first version of ms
+**Chloé R. Nater**: Conceptualization, Methodology, Software, Validation, Formal analysis, Investigation, Data Curation, Writing - Original Draft, Writing - Review & Editing, Visualization.
 

### 2024-07-10 Manuscript_Nater_etal.qmd
Updated figure 1

@@
 
-![Simplified graphical representation of the ptarmigan life cycle with two age classes and the data sources included in the integrated distance sampling model.
-The pink "x"'s indicate the added dimension for area.
-Juv\[x,t\] = juveniles in area x year t.
-Ad\[x,t\] = adults in area x in year t.
-R\[x,t\] = recruitment rate in area x in year t.
-S\[x,t\] = survival probability from year t to t+1 in area x.
-Note that the additional site (=transect) dimension, "j", is omitted for the sake of illustration.](Figures/RypeIDSM_multiArea_structure.png){#fig-model}
+![Graphical representation of the ptarmigan life cycle with two age classes and the data sources, as well as joint distributions, included in the integrated distance sampling model.
+Indices $t$ denote year while $x$ represent the added dimension for area (corresponding to 41 reporting districts).
+Juvenile~x,t~ = juveniles in area $x$ year $t$.
+Adult~x,t~ = adults in area $x$ in year $t$.
+$R_{x,t}$ = recruitment rate in area $x$ in year $t$.
+$S_{x,t}$ = survival probability from year $t$ to $t+1$ in area $x$.
+Note that the additional site (=transect) dimension, $j$, is omitted for the sake of illustration.](Figures/RypeIDSM_multiArea_structure.png){#fig-model}
 

### 2024-07-10 Manuscript_Nater_etal.qmd
Smaller text edits

@@
 Participatory monitoring programs (also called citizen science programmes) can alleviate these constraints by recruiting stakeholders and the public to increase the spatial and temporal resolution of sampling effort and hence resulting data.
-While the majority of participatory monitoring programs are limited by opportunistic sampling designs, we are starting to see the emergence of structures citizen science programs that employ trained volunteers to collect data according to standardized protocols.
+While the majority of participatory monitoring programs are limited by opportunistic sampling designs, we are starting to see the emergence of structured citizen science programs that employ trained volunteers to collect data according to standardized protocols.
 Simultaneously, there is much ongoing development of statistical models that are increasingly more powerful and able to make more efficient use of field data.
@@
 While large-scale, spatially-explicit indicators for species abundance and populations are clearly needed, development and practical implementation are greatly limited due to the reliance of such indicators on the availability of data from large-scale, long-term monitoring programmes [@proencca2017].
-Consequently, many countries have been working on setting up, maintaining, and improving such monitoring programmes over the last decades.
-Many now well-established programmes focus on breeding birds and butterflies, and examples include the North American Breeding Bird Survey (<https://www.pwrc.usgs.gov/bbs/>), the PanEuropean Common Bird Monitoring Scheme (<https://pecbms.info/>), the UK Butterfly Monitoring Scheme (<https://ukbms.org/>), the Game and Wildlife Conservation Trust Partridge Count Scheme [@aebischer2010], and the Swiss Biodiversity Monitoring (<https://www.biodiversitymonitoring.ch/>).
+Consequently, many countries have been working on setting up, maintaining, and improving such monitoring programmes over the last few decades.
+Many now well-established programmes focus on breeding birds and butterflies, and examples include the North American Breeding Bird Survey (<https://www.pwrc.usgs.gov/bbs/>), the PanEuropean Common Bird Monitoring Scheme (<https://pecbms.info/>), the UK Butterfly Monitoring Scheme (<https://ukbms.org/>), the Game and Wildlife Conservation Trust Partridge Count Scheme [@aebischer2010], and the Swiss Biodiversity Monitoring program (<https://www.biodiversitymonitoring.ch/>).
 
@@
 @bowler2020; @nilsen2020], and to test a range of relevant ecological hypotheses [@bowler2020; @Breisjøberget_etal_2018].
-However, large-scale estimation of demographic rates underlying abundance trends has thus far remained as untapped potential of the dataset.
+However, large-scale estimation of demographic rates underlying abundance trends has thus far remained an untapped potential of the dataset.
 @nilsen2023 recently developed a novel integrated distance sampling model (IDSM) which successfully uses the age of individuals detected along line transects data coupled with radio-telemetry data to jointly estimate abundance, survival, and recruitment across years.
@@
 While the species is currently listed as Least Concern (LC) both in the global and Norwegian Red List of Species, it has undergone rather dramatic declines in abundance in Norway since the turn of the 20th century [@hjeljord2022].
-The main reason for the long-term decline in abundance remain unresolved, but the willow ptarmigan are considered sentinel species that are sensitive to both climate change and land use changes [@Henden2017; @Storch2007].
+The main reason for the long-term decline in abundance remain unresolved, but the willow ptarmigan is considered a sentinel species that is sensitive to both climate change and land use changes [@Henden2017; @Storch2007].
 Moreover, being one of only a handful of bird species that spend the winter in mountain ecosystems in Scandinavia, they are important components of the ecosystem as prey species for resident predators, such as the gyrfalcon [@Alastair2020].
 The willow ptarmigan has a relatively fast pace of life [@sandercock2005; @Steen1996], and can display substantial spatio-temporal variation in demographic rates [@bowler2020].
-Their population dynamics are characterized by large inter-annual fluctuations in abundance [@hjeljord2022], and previous research has tied these fluctuations to rodent cycle through shared predators [@hagen1952; @bowler2020].
+Their population dynamics are characterized by large inter-annual fluctuations in abundance [@hjeljord2022], and previous research has tied these fluctuations to rodent cycles through shared predators [@hagen1952; @bowler2020].
 This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis (APH) and has been a central part of research on Fennoscandian grouse population dynamics for many decades [@Elton1942; @hagen1952; @Linden1988; @Steen1988].
@@
 To increase the detection probability, fieldworkers use pointing dogs to locate the birds.
-A survey team typically consist of at least two people (one dog handler and one person responsible for following the transect line) and one dog.
+A survey team typically consists of at least two people (one dog handler and one person responsible for following the transect line) and one dog.
 Often, more than one dog is used for a survey, but only one dog should be used at a time.
@@
 
-The model of @nilsen2023 integrated line transect data with radio-telemetry data from from an ongoing field study of marked willow ptarmigans in Lierne municipality in Central Norway.
+The model of @nilsen2023 integrated line transect data with radio-telemetry data from an ongoing field study of marked willow ptarmigans in Lierne municipality in Central Norway.
 From 2015 to 2019, around 50 birds were captured in winter (late February or early March) each year and fitted with VHF collars.
@@
 A proportion of the birds were harvest annually in the regular recreational harvest, and birds that were harvested were reported as shot to the field personnel.
-In addition, as the collars had mortality switch, we were also able to locate and retrieve a high proportion of birds that died for natural causes, resulting in a known-fate mark-recapture dataset.
+In addition, as the collars had mortality switches, we were also able to locate and retrieve a high proportion of birds that died from natural causes, resulting in a known-fate mark-recapture dataset.
 The radio-telemetry study is described in detail in @Israelsen2020 and in @Arnekleiv2022.
@@
 The three likelihoods for data resulting from the line transect sampling (observation distances, age-specific counts, and juvenile to adult ratios, see above) were also formulated as spatially explicit, with year- and area-specific distance sampling detection parameters modelled in the same way as survival and recruitment (except the effect of rodent occupancy, @fig-model).
-For the known-fate telemetry data (and the seasonal decomposition of survival estimated from it), on the other hand, we did not add an additional area dimension as this data was only available for on study area (the Lierne area).
+For the known-fate telemetry data (and the seasonal decomposition of survival estimated from it), on the other hand, we did not add an additional area dimension as this data was only available for one study area (the Lierne area).
 
@@
 We implemented our multi-area IDSM in a Bayesian framework using NIMBLE version 1.0.1 [@valpine2017] in R version 4.3.1 [@R2023].
-For the likelihood for line transect observation distances we used a custom half-normal distribution developed by Michael Scroggie in the "nimbleDistance" package (https://github.com/scrogster/nimbleDistance).
+For the likelihood for line transect observation distances we used a custom half-normal distribution developed by Michael Scroggie in the "nimbleDistance" package (https://github.com/scrogster/nimbleDistance, see package vignette for specifics).
 We used non-informative uniform priors for all parameters, but used biologically sensible boundaries where possible.
@@
 We discarded the first 75k samples of each chain as burn-in, and thinned the remainder by a factor 25, resulting in a final joint posterior containing a total of 5 x 3k = 15k samples (note that high thinning rates were necessary to constrain memory load of the joint posterior, which included 314568 monitored parameters).
+In addition to the main model run, we implemented a second version of the model that did not use telemetry data to assess the potential impacts of auxiliary data from a single location on parameter estimates.
 
@@
 Following model fitting, we calculated posterior distributions for the proportions of variance in survival probabilities, recruitment rates, and detection decay explained by 1) spatial variation ($var_{area}$), 2) temporal variation ($var_{year}$), 3) residual variation ($var_{residual}$), and 4) variation in rodent occupancy ($var_{rodent}$).
-To obtain the proportion variance explained by each of the component, we divided it by the sum of all the components ($var_{area} + var_{year} + var_{residual} + var_{rodent}$).
+To obtain the proportion of variance explained by each of the component, we divided it by the sum of all the components ($var_{area} + var_{year} + var_{residual} + var_{rodent}$).
 The spatial, temporal, and residual variance components were defined as the square of the estimated corresponding random effects standard deviation from the model while $var_{rodent}$ was calculated as the variance of all area- and year-specific $\beta_x*rodentOcc_{x,t}$ products.
@@
 
+MCMC for all central model parameters converged within the given number of iterations, but chain mixing was remained suboptimal even at convergence for some parameters.
+The inclusion of telemetry data from a single location into the integrated model had did not substantially affect parameter estimates beyond the seasonal decomposition of survival in Lierne, where the telemetry data was collected (see online supplementary "Comparison_noTelemetry").
 All numerical results in the following are presented as median \[95% credible interval\] unless otherwise indicated.
@@
 During this period, estimated population densities varied between `r dens.min$post.sum` birds/km^2^ in the area "`r dens.min$area.name`" close to the coast in central Norway to `r dens.max$post.sum` birds/km^2^ in "`r dens.max$area.name`" further south near the Swedish border.
-In general, recent population density appeared to be lowest in the northern Norway and highest in the eastern part of central Norway (@fig-densA).
+In general, recent population density appeared to be lowest in northern Norway and highest in the eastern part of central Norway (@fig-densA).
 Uncertainty in density estimates was relatively consistent, with a few areas (including the one with the lowest estimated density, "`r dens.min$area.name`") sticking out by having substantially less precise estimates (@fig-densB).
@@
 The global average survival probability across all areas and years ($\mu^S$) was estimated at `r S.hyper.mean$post.sum`.
-Spatial variation in survival (random effect SD = `r S.sd.spatial$post.sum`) was relatively low compared to temporal (`r S.sd.temporal$post.sum`) and residual (`r S.sd.residual$post.sum`) variation.
+Spatial variation in survival (random effect SD on logit scale = `r S.sd.spatial$post.sum`) was relatively low compared to temporal (`r S.sd.temporal$post.sum`) and residual (`r S.sd.residual$post.sum`) variation.
 
@@
 Across all areas and years, average recruitment rate was `r R.hyper.mean$post.sum`.
-Unlike for survival, the model predicted similar magnitudes of spatial and temporal variation (random effect SDs of `r R.sd.spatial$post.sum` and `r R.sd.temporal$post.sum`, respectively), and about twice as much residual variation (`r R.sd.residual$post.sum`).
+Unlike for survival, the model predicted similar magnitudes of spatial and temporal variation (random effect SDs on log scale of `r R.sd.spatial$post.sum` and `r R.sd.temporal$post.sum`, respectively), and about twice as much residual variation (`r R.sd.residual$post.sum`).
 
@@
 The global average detection decay was `r Det.hyper.mean$post.sum` (detection probability = `r pDet.hyper.mean$post.sum`), and in general, higher values were more common in the Southern half of the country than the Northern half (SF"Avg_detect_Map.pdf").
-Variation in detection over time was modest (SF "TimeSeries_pDetect.pdf"), and among-year variation in detection decay (random effect log SD = `r Det.sd.temporal$post.sum`) was almost identical in magnitude to spatial (`r Det.sd.spatial$post.sum`) and residual (`r Det.sd.residual$post.sum`) variation.
+Variation in detection over time was modest on average but the degree of temporal changes varied by area, with some areas having nearly constant detection while others showed variation by factors larger than 1.5 (SF "TimeSeries_pDetect.pdf").
+The estimated average among-year variation in detection decay (random effect log SD = `r Det.sd.temporal$post.sum`) was almost identical in magnitude to spatial (`r Det.sd.spatial$post.sum`) and residual (`r Det.sd.residual$post.sum`) variation.
 
@@
 Doing so may also help with better understanding the mechanisms underlying the large portion of demographic rate variation that could only be attributed to random variation so far.
-This is the case especially for the relatively large residual variation [@fig-varDecomp] but also relevant for constant spatial and shared temporal variation.
+This is the case especially for the relatively large residual variation ([@fig-varDecomp]) but also relevant for constant spatial and shared temporal variation.
 In previous work based on both marked [@Eriksen2023] and unmarked birds [@bowler2020; @Henden2020; @Novoa2016], spring conditions has come out as an important predictor of ptarmigan recruitment rates.
@@
 While we did not explicitly model density dependence in this study, our results can provide some preliminary insights into potential density feedbacks from both a spatial (cross-population) and a temporal (within-population) angle.
-Comparing average population densities and growth rate across areas did not provide evidence for strong density dependence, but there was a tendency towards the highest population growth rates appearing areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A].
+Comparing average population densities and growth rate across areas did not provide evidence for strong density dependence, but there was a tendency towards the highest population growth rates appearing in areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A].
 When considering density dependence across years within select areas, however, we found that higher density years were associated with higher recruitment the same year, but followed by lower apparent survival probabilities and, consequently, lower population growth rates (as determined by post-hoc Pearson correlation coefficients, supplementary file "DD_corrCoef.csv").
@@
 Management decisions made at the resolution of large geopolitical boundaries (e.g., Norway) run a high risk of being inadequate when there is substantial spatial variation in demographic processes and population dynamics, as is the case for willow ptarmigan.
-In Norway, willow ptarmigan -- and small game in general -- is managed at the local and/or regional scale, with rather limited national regulation beyond updating the length of the hunting season every fourth year.
+In Norway, willow ptarmigan -- and small game in general -- are managed at the local and/or regional scale, with rather limited national regulation beyond updating the length of the hunting season every fourth year.
 In effect, management system, regulation type (quota type, season length, number of licences, bag limit etc.), and quota size are governed by the local or regional stakeholders [@Eriksen2018; @breisjøberget2018].
@@
 If misclassification happened at random, mixture models could be used to determine the likely age class of individuals to whom no age class was assigned during observations [@mccrea2013].
-In our case, we might suspect that an observed is more likely to classify an adult bird as juvenile rather than the other way around, and more likely to assign "unknown" age class to juveniles than adults.
+In our case, we might suspect that an observer is more likely to classify an adult bird as juvenile rather than the other way around, and more likely to assign "unknown" age class to juveniles than adults.
 One reason for this is that observers look for specific signs to classifying a bird as adult (e.g.
@@
 @fig-dens).
-While estimation of latent parameters in missing areas may possible (Perry de Valpine, personal communication; @schaub2021 chapter 19), this also provides an opportunity for inclusion of additional data.
+While estimation of latent parameters in missing areas may be possible (Perry de Valpine, personal communication; @schaub2021 chapter 19), this also provides an opportunity for inclusion of additional data.
 The line transect survey data included in this study constitutes just the publicly available part of the data collected through "Hønsefuglportalen" but the programme also includes additional surveys on private land.
@@
 Finally, including further data beyond the line transect surveys may be relevant in the future, and in particular in the context of informing and improving management of ptarmigan hunting.
-In the present study, we have used auxiliary radio-telemetry data to supplement information on survival, but since this data was available for only one out of 41 areas, its influence was likely small.
+In the present study, we have used auxiliary radio-telemetry data to supplement information on survival, but since this data was available for only one out of 41 areas, its influence was small.
 Nonetheless, this illustrates a way for how smaller datasets from single or subsets of areas can be integrated into a large-scale modelling framework.
@@
 
+We thank Mark Hewinson, Todd Arnold, and one anonymous reviewer for constructive comments during the review process.
 We are grateful to the numerous volunteer observers (and their canine assistant) who collected ptarmigan observations as part of the "Hønsefuglportalen" monitoring effort.

### 2024-07-10 Manuscript_Nilsen_Nater.qmd
Updated model diagram

@@
 Through a comprehensive simulation study we show that the model is able to adequately recover true population and vital rate dynamics.
-Subsequent application to data from a study of willow ptarmigan (_Lagopus lagopus_) in Norway showcases the frameworks ability to recover both fluctuations and trends in population dynamics and highlights its potential applicability to a wide range of species surveyed using distance sampling approaches.*
+Subsequent application to data from a study of willow ptarmigan (Lagopus lagopus*) in Norway showcases the frameworks ability to recover both fluctuations and trends in population dynamics and highlights its potential applicability to a wide range of species surveyed using distance sampling approaches.
 
@@
 
-![Simplified graphical representation of the ptarmigan life cycle with two age classes and the data sources included in the integrated distance sampling model.
-Juv\[t\] = juveniles in year t.
-Ad\[t\] = adults in year t.
-R\[t\] = recruitment rate in year t.
-S\[t\] = survival probability from year t to t+1.](Figures/RypeIDSM_structure.png){#fig-model}
+![Graphical representation of the annual ptarmigan life cycle with two age classes and the data sources included in the integrated distance sampling model.
+Juvenile~t~ = juveniles in year $t$.
+Adult~t~ = adults in year $t$.
+$R_t$ = recruitment rate in year $t$.
+$S_t$ = survival probability from year $t$ to $t+1$.](Figures/RypeIDSM_structure.png){#fig-model}
 

### 2024-07-23 Manuscript_Nilsen_Nater.qmd
Edit to figure 1 caption

@@
 
-![Graphical representation of the annual ptarmigan life cycle with two age classes and the data sources included in the integrated distance sampling model.
+![Graphical representation of the annual ptarmigan life cycle with two age classes under a post-breeding census and the data sources included in the integrated distance sampling model.
+Solid arrows represent relationships within the ptarmigan life cycle; dotted arrows visualize information flow from data sources to parameters.
+Blue and pink data nodes originate from distance sampling and line transect surveys, respectively.
 Juvenile~t~ = juveniles in year $t$.

### 2024-07-23 Manuscript_Nilsen_Nater.qmd
Clarified treatment of stochasticity in pop. model

@@
 Note that recruitment rate $R$ is defined as juveniles/adult (not juveniles/female).
+We assume no stochasticity beyond time-variation in vital rates in the model for population density itself, but instead treat local population sizes (numbers of birds in age class $a$ in year $t$ within the area of each transect $j$, $N_{a,j,t}$), as outcomes of a Poisson process with an expected average equaling $D_{a,t}$ times the transect area (see below).
 We also make the simplifying assumption that there is no age- or sex-dependence of vital rates, but this assumption could be relaxed by including additional auxiliary data [@Israelsen2020; @Sandercock2011].

### 2024-07-23 Manuscript_Nilsen_Nater.qmd
Fixed indexing for detection prob.

@@
 Our model includes three likelihoods for different components of the age-structured distance sampling data.
-First is the likelihood for the perpendicular detection distances from line transect, $y$, which are linked to distance-dependent detection probability $p$ through a half-normal detection function:
+First is the likelihood for the perpendicular detection distances from line transect, $y$, which are linked to distance-dependent detection probability $p_y$ through a half-normal detection function:
 
@@
 \begin{equation}
-p = exp(-\frac{y^2}{2\sigma^2})
+p_y = exp(-\frac{y^2}{2\sigma^2})
 \end{equation}

### 2024-07-23 Manuscript_Nater_etal.qmd
Addition to introduction

@@
 @robinson2014; @horswill2019].
-This variation needs to be accounted for to develop successful and sustainable strategies for area use, harvest management, and species and biodiversity conservation [@williams2002].
+Such variation can arise from differences in ecological contexts (including local habitat and weather conditions, hunting pressures, and interspecific interactions, e.g.
+@nilsen2009; @bond2021) and needs to be accounted for when developing sustainable strategies for area use, harvest management, and species and biodiversity conservation [@williams2002].
 

### 2024-07-23 Manuscript_Nater_etal.qmd
Addition to discussion

@@
 @fig-dens).
-While estimation of latent parameters in missing areas may be possible (Perry de Valpine, personal communication; @schaub2021 chapter 19), this also provides an opportunity for inclusion of additional data.
-The line transect survey data included in this study constitutes just the publicly available part of the data collected through "Hønsefuglportalen" but the programme also includes additional surveys on private land.
+Estimation of latent parameters in missing areas may be possible though (Perry de Valpine, personal communication; @schaub2021 chapter 19), and this may result in a unique opportunity for making predictions of ptarmigan population trends in unmonitored areas, provided that data for a sufficient number and range of areas are available.
+Here, we may benefit from the fact that the line transect survey data included in this study constitutes just the publicly available part of the data collected through "Hønsefuglportalen" but the programme also includes additional surveys on private land.
 Extending to data from private land would provide better coverage especially in south-eastern and southern parts of Norway, which includes areas where only very limited amounts of data are collected on public land.

### 2024-07-24 Analysis_RealData.R
Wrote function to plot average detection function per area

@@ -282,6 +282,15 @@ if(fitRodentCov){
 }
 
 
+# OPTIONAL: PLOT DETECTION FUNCTIONS #
+#------------------------------------#
+
+plotDetectFunction(mcmc.out = IDSM.out.tidy,
+                   maxDist = input_data$nim.constants$W,
+                   N_areas = input_data$nim.constant$N_areas, 
+                   area_names = input_data$nim.constant$area_names)
+
+  
 # OPTIONAL: CHECK WITHIN-AREA DENSITY DEPENDENCE #
 #------------------------------------------------#
 

### 2024-07-24 _targets.R

@@ -216,6 +216,14 @@ list(
   ),
   
   tar_target(
+    detect.funPlots,
+    plotDetectFunction(mcmc.out = IDSM.out.tidy,
+                       maxDist = input_data$nim.constants$W,
+                       N_areas = input_data$nim.constant$N_areas, 
+                       area_names = input_data$nim.constant$area_names)
+  ),
+  
+  tar_target(
     DD.checks,
     checkDD(mcmc.out = IDSM.out.tidy, 
             N_areas = input_data$nim.constant$N_areas, 

### 2024-07-24 Analysis_RealData_LierneVest.R
Added plotting function for detection probability

@@ -214,4 +214,11 @@ if(fitRodentCov){
 }
 
 
+# OPTIONAL: PLOT DETECTION FUNCTIONS #
+#------------------------------------#
+
+plotDetectFunction(mcmc.out = IDSM.out.tidy,
+                   maxDist = input_data$nim.constants$W,
+                   N_areas = input_data$nim.constant$N_areas, 
+                   area_names = input_data$nim.constant$area_names)
 

### 2024-07-24 R/plotDetectFunction.R

@@ -1,0 +1,66 @@
+#' Plot detection probability as a function of distance
+#'
+#'
+#' @param mcmc.out an mcmc list containing posterior samples from a model run.
+#' @param maxDist numeric. Maximum distance (in m) to plot detection for.
+#' @param N_areas integer. Number of areas included in analyses. 
+#' @param area_names character vector containing area/location names.
+#'
+#' @return a vector containing the pdf plot name. The plot can be found in Plots/DetectFunction.
+#' @export
+#'
+#' @examples
+
+plotDetectFunction <- function(mcmc.out,
+                               maxDist,
+                               N_areas, area_names){
+  
+  
+  ## Convert posterior samples to matrix
+  mcmc.mat <- as.matrix(mcmc.out)
+  
+  ## Make sequence of distances to predict for
+  dist <- seq(0, maxDist, length.out = 100)
+
+  ## Assemble dataframe for storing posterior summaries of predictions
+  pred.data <- data.frame()
+  
+  for(i in 1:N_areas){
+    
+    ## Extract posterior samples of relevant parameters
+    sigma_avg <- exp(mcmc.mat[, paste0("mu.dd[", i, "]")])
+
+    ## Make, summarise, and store predictions for each covariate value
+    for(x in 1:100){
+      p.pred <- exp(-((dist[x]^2)/(2*sigma_avg^2)))
+      pred.temp <- data.frame(Area = area_names[i], 
+                              distance = dist[x], 
+                              pred_Median = median(p.pred),
+                              pred_lCI = unname(quantile(p.pred, probs = 0.025)),
+                              pred_uCI = unname(quantile(p.pred, probs = 0.975)))
+      pred.data <- rbind(pred.data, pred.temp)
+    }
+  }
+  
+  ## Plot predictions
+  ifelse(!dir.exists("Plots/DetectFunction"), dir.create("Plots/DetectFunction"), FALSE) ## Check if folder exists, if not create folder
+  
+  pdf("Plots/DetectFunction/DetectionProb_distance.pdf", width = 6, height = 4) 
+  for(i in 1:N_areas){
+    print(
+      ggplot(subset(pred.data, Area == area_names[i]), aes(x = distance, y = pred_Median)) +
+        geom_line(color = "#856CEB") + 
+        geom_ribbon(aes(ymin = pred_lCI, ymax = pred_uCI), alpha = 0.5, fill = "#856CEB") + 
[…]

### 2024-07-24 Manuscript_Nilsen_Nater.qmd
Added reporting of results on detection to manuscript

@@
 Through a comprehensive simulation study we show that the model is able to adequately recover true population and vital rate dynamics.
-Subsequent application to data from a study of willow ptarmigan (Lagopus lagopus*) in Norway showcases the frameworks ability to recover both fluctuations and trends in population dynamics and highlights its potential applicability to a wide range of species surveyed using distance sampling approaches.
+Subsequent application to data from a study of willow ptarmigan (Lagopus lagopus) in Norway showcases the frameworks ability to recover both fluctuations and trends in population dynamics and highlights its potential applicability to a wide range of species surveyed using distance sampling approaches.*
 
@@
 
+The average probability of detecting individuals and groups of ptarmigan within transect line areas was `r round(filter(tidy_summary_stats, model_parameter =="p_mean")$Mean, 2)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="p_mean")$lower, 2)` - `r round(filter(tidy_summary_stats, model_parameter =="p_mean")$upper, 2)`), and estimated with a detection decay parameter $\sigma$ of `r round(filter(tidy_summary_stats, model_parameter =="sigma_mean")$Mean, 2)` (95% C.I = `r round(filter(tidy_summary_stats, model_parameter =="sigma_mean")$lower, 2)` - `r round(filter(tidy_summary_stats, model_parameter =="sigma_mean")$upper, 2)`).
+Detection probabilities were highest in the start of the study and in the period 2016-2019 and lowest from 2010-2012.
+The relationship between detection probability and distance and changes in detection over time are visualized in supplementary figures "DetectionProb_distance.png" and "TimeSeries_pDetect.png" in @Nilsen2024.
+
 ![Estimated density of willow ptarmigans in Lierne from 2007 to 2021.
@@
 Given the available data, the IDSM was not able to estimate a clear effect of small rodent abundance on ptarmigan recruitment (slope-paramater for the z-standardized rodent occurrence data = `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$Mean, 3)` ; 95% C.I.
-= `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$lower, 3)` - `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$upper, 3)`, see supplementary figure "Rep_betaR.R.png" in @Nilsen2024 ).
+= `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$lower, 3)` - `r round(filter(tidy_summary_stats, model_parameter =="betaR.R")$upper, 3)`, see supplementary figure "Rep_betaR.R.png" in @Nilsen2024).
 

### 2024-07-24 tidySummary.R

@@ -115,12 +115,37 @@ betaR.R_tidy <- out.mcmc %>% tidybayes::spread_draws(betaR.R[area],
             upper = quantile(betaR.R, probs = 0.975)) %>%
   mutate(model_parameter = "betaR.R")
 
+################################################################################
+### Summarize detection
+
+# By year
+detect_year_tidy <- out.mcmc %>% tidybayes::spread_draws(p[area, year], sep = "[,]", regex = FALSE) %>%
+  group_by(year) %>%
+  summarise(Mean = mean(p), 
+            std = sd(p), 
+            Median = median(p), 
+            lower = quantile(p, probs = 0.025), 
+            upper = quantile(p, probs = 0.975)) %>%
+  ungroup() %>%
+  mutate(model_parameter = "p_year")
+
+# Average parameters
+out.mat <- as.matrix(out.mcmc)
+p_avg <- rowMeans(out.mat[, paste0("p[1, ", 1:15, "]")])
+sigma_avg <- exp(out.mat[, "mu.dd[1]"])
+detect_avg_tidy <- data.frame(Mean = c(mean(p_avg), mean(sigma_avg)),
+                              std = c(sd(p_avg), sd(sigma_avg)),
+                              Median = c(median(p_avg), median(sigma_avg)), 
+                              lower = c(quantile(p_avg, probs = 0.025), quantile(sigma_avg, probs = 0.025)), 
+                              upper = c(quantile(p_avg, probs = 0.975), quantile(sigma_avg, probs = 0.975)),
+                              model_parameter = c("p_mean", "sigma_mean"))
 
 ################################################################################
 #### Putting together output
 
 pooled_results <- bind_rows(dens_tidy, R_year_tidy, muR_tidy, 
-                            muS_tidy, muS1_tidy, muS2_tidy, betaR.R_tidy)
+                            muS_tidy, muS1_tidy, muS2_tidy, betaR.R_tidy,
+                            detect_year_tidy, detect_avg_tidy)
 
 
 pooled_results

### 2024-08-26 Manuscript_Nilsen_Nater.qmd
Text edits following referee comments - Erlend

@@
 
-The most common implementations of DS models have long used closed-population formulations, and thus do not allow formulating a process model that projects abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates @buckland2015.
+The most common implementations of DS models have long used closed-population formulations by only estimating population density or abundance for one point in time, and not including an underlying process model that projects abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates @buckland2015.
 In recent years, DS approaches have been extended in many ways, including applications that estimate changes in abundance over time in open populations via a hidden state model representing population dynamics [@moore2011; @sollmann2015; @schaub2021].
@@
 In this study, we present a new IPM (IDSM, integrated distance sampling model) which integrates data from line transect distance sampling survey data and survival data from marked birds.
+The main motivation behind the study is to showcase how distance sampling models could be used to not only estimate population density but also demographic rates in an IPM setting.
+We believe the model would serve as a good building block for later applications that would focus more on ecological dynamics.
 The model's core is a stage-structured matrix population model that projects population size from one time step to the next based on underlying survival and recruitment rates.
@@
 
-Our open population integrated distance sampling model (IDSM) consists of two major components: a latent structured population model and a set of likelihoods for data originating from distance sampling surveys and auxiliary survival monitoring.
+Our open population integrated distance sampling model (IDSM) consists of two major components: a latent structured population model and a set of likelihoods for data originating from distance sampling surveys and auxiliary survival monitoring of marked birds.
 In the example case, these auxiliary data come from a radio-telemetry study, but in principle other types of capture-recapture data can also be used.
@@
 The population model follows a post-breeding census and includes two age classes: juveniles (young of the year) and adults (\> 1 year of age, @fig-model).
-This structure is inspired by earlier models for our focal species, the willow ptarmigan (+ref), and is commonly used for populations of passerine and game birds [@williams2002; @schaub2021].
-In the context of our willow ptarmigan case study (see below), the census falls into late summer and coincides with the annual distance-sampling survey in August.
+This structure commonly used for populations of passerine and game birds [@williams2002; @schaub2021].
+In the context of our willow ptarmigan case study (see below), the population is updated in late summer - which is when the annual distance-sampling survey takes place.
+At this time, the juvenile class is about 1 - 2 months old.
 
@@
 
-The implementation of the modelling framework we present makes three assumptions about the distance sampling survey: 1) the survey consists of line transects, 2) animals may be detected alone or in groups, and 3) juveniles and adults can be distinguished during surveys.
-These assumptions are inspired by our willow ptarmigan case study (details below).
+The implementation of the modelling framework we present assumes that the distance sampling survey data have the following characteristics: 1) the survey consists of line transects, 2) animals may be detected alone or in groups, and 3) juveniles and adults can be distinguished during surveys.
+These characteristics are inspired by our willow ptarmigan case study (details below).
 Our model includes three likelihoods for different components of the age-structured distance sampling data.
@@
 $$ Here, $Sk_t$ is the seasonal survival probability, and annual survival probability, $S_t$ is calculated as $S1_t * S2_t$.
+The motivation behind separating out the different seasons come from our ptarmigan case study (see below), but in cases where it is not relevant to distinguish between different seasons, $S_t$ could be estimated directly using the formulation above.
 
@@
 the Alternative Prey Hypothesis; see @Hagen1952; @Kausrud2008; @Bowler2020).
-Adult survival show less inter-annual fluctuations [@Israelsen2020], although spatial (and potentially temporal) variation due to e.g.
-harvest management is evident when comparing across studies [@Israelsen2020].
+Adult survival show less inter-annual fluctuations [@Israelsen2020], although variation due to e.g.
+harvest management is evident when comparing across studies [@Israelsen2020; @Sandercock2011].
 
@@
 Nevertheless, mis-identification can occur, and in addition a proportion of the birds are registered as "unknown" age and/or sex.
-Besides bird observations, field workers also record whether (1) or not (0) they encounter small rodents on any transect line, allowing the proportion of transect lines with small rodent detection to be used as measure of rodent occupancy (covariate ranging from 0 to 1).Since 2019, data has been collected through a mobile app tailormade for this project, which is available through App Store (for iOS phones) and Google Play (for Android phones).
+In the model version presented here, we have made the assumption that these birds were in fact juveniles.
+Besides bird observations, field workers also record whether (1) or not (0) they encounter small rodents on any transect line, allowing the proportion of transect lines with small rodent detection to be used as measure of rodent occupancy (covariate ranging from 0 to 1).
+Since 2019, data has been collected through a mobile app tailormade for this project, which is available through App Store (for iOS phones) and Google Play (for Android phones).
 Before 2019, field workers reported their data through a dedicated web portal.
@@
 
-Having evaluated the overall performance of our model on simulated data, we used data from our case study in Lierne as a case study to estimate abundance, vital rates and detection probabilities from a real-world data set.
+Having evaluated the overall performance of our model on simulated data, we used data from our case study in Lierne to estimate abundance, vital rates and detection probabilities from a real-world data set.
 Ptarmigan population density increased markedly across the study period, from \< 10 ptarmigan / $km^{2}$ in 2007 to \> 35 ptarmigan / $km^{2}$ in in 2021 (@fig-lierne-density).

### 2024-08-31 Manuscript_Nilsen_Nater.qmd
Edits to text + bib-file

@@
 One set of approaches that has been used extensively to estimate abundance of wildlife populations is Distance Sampling (DS) methods for line or point transect survey data.
-Historically, DS models were only available as closed population models, and did not allow for the estimation of changes in abundance through time.
+The first versions of DS models were only available as closed population models, and did not allow for the estimation of changes in abundance through time.
 The advent of open population formulations based on the DS framework greatly extended the scope of the models, but much untapped potential remains in models that estimate temporal dynamics not only in abundance but also in the underlying demographic rates.
@@
 
-The most common implementations of DS models have long used closed-population formulations by only estimating population density or abundance for one point in time, and not including an underlying process model that projects abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates @buckland2015.
+The most common implementations of DS models have long used closed-population formulations by only estimating population density or abundance for one point in time, and not including an explicit formulation of the process model that projects abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates @buckland2015.
 In recent years, DS approaches have been extended in many ways, including applications that estimate changes in abundance over time in open populations via a hidden state model representing population dynamics [@moore2011; @sollmann2015; @schaub2021].
@@
 \end{aligned}
-$$ The average detection probability $\hat{p_t}$ is an integral part of the second data likelihood which relates the observed number of animals in each age class $a$, $obsN_{a,j,t}$ ($j$ = transect) to the corresponding true number per transect, $N_{a,j,t}$:
+$$ The estimated average detection probability $\hat{p_t}$ is an integral part of the second data likelihood which relates the observed number of animals in each age class $a$, $obsN_{a,j,t}$ ($j$ = transect) to the corresponding true number per transect, $N_{a,j,t}$:
 
@@
 
-The third data likelihood focuses on the counts of adults ($obsAd_{j,t}$) and juveniles ($obsJuv_{j,t}$) observed during the distance sampling surveys and links them to year-specific recruitment rate:
+The third data likelihood focuses on the counts of adults ($obsAd_{j,t}$) and juveniles ($obsJuv_{j,t}$) observed during the distance sampling surveys and links them to the estimated year-specific recruitment rate:
 
@@
 \begin{equation}
-  obsJuv_{j,t} \sim Poisson(R_t*obsAd_{j,t})
+  obsJuv_{j,t} \sim Poisson(\hat{R_t}*obsAd_{j,t})
 \end{equation}
@@
 \begin{equation}
-  survivors_{k,t} \sim Binomial(released_{k,t}, Sk_t)
+  survivors_{k,t} \sim Binomial(released_{k,t}, \hat{Sk_t})
 \end{equation}
@@
 
-Vital rates (survival probabilities $S$, recruitment rates $R$) and detection parameters (half-normal detection parameters $\sigma$) can all be modelled as time-dependent in our framework.
+Vital rates (survival probabilities $S_t$, recruitment rates $R_t$) and detection parameters (half-normal detection parameters $\sigma_t$) can all be modelled as time-dependent in our framework.
 For both the tests with simulated data and the case study described below, we implemented log-normally distributed random year effects on recruitment rate and detection probability.
 Survival, on the other hand, was set to be constant.
+This was done because previous research has shown relatively modest interannual variation in survival [@Israelsen2020], but also to ensure that all parameters were estimable.
 In the case study, we additionally included a covariate effect (see details below) on log recruitment rates, resulting in the following model:

### 2024-09-01 Analysis_RealData_LierneVest.R
Re-run Lierne model with new parameterization

@@ -8,7 +8,7 @@ library(viridis)
 #-------#
 
 ## Define seed for initial value simulation and MCMC
-mySeed <- 0
+mySeed <- 32
 
 ## Source all functions in "R" folder
 sourceDir <- function(path, trace = TRUE, ...) {
@@ -157,17 +157,16 @@ IDSM.out <- nimbleMCMC(code = model_setup$modelCode,
                        setSeed = mySeed)
 Sys.time() - t.start
 
-saveRDS(IDSM.out, file = 'rypeIDSM_dHN_multiArea_realData_Lierne_alt.rds')
+saveRDS(IDSM.out, file = 'rypeIDSM_dHN_multiArea_realData_Lierne.rds')
 
 
 # COMPARE MODELS #
 #----------------#
 
-
-plotModelComparison(modelPaths = c("rypeIDSM_dHN_multiArea_realData_Lierne_alt.rds",
-                                   "P:/41201480_lirypedemografi_analyser_av_honsefuglportal_data/Lierne_ModelFit/rypeIDSM_dHN_multiArea_realData_Lierne.rds"), 
-                    modelChars = c("R_year[1]",
-                                   "ratio.JA1"), 
+plotModelComparison(modelPaths = c("rypeIDSM_dHN_multiArea_realData_Lierne.rds",
+                                   "rypeIDSM_dHN_multiArea_realData_Lierne_alt.rds"), 
+                    modelChars = c("Main model",
+                                   "Alternative model"), 
                     N_areas = input_data$nim.constants$N_areas, 
                     area_names = areas, 
                     N_sites = input_data$nim.constants$N_sites, 
@@ -175,7 +174,7 @@ plotModelComparison(modelPaths = c("rypeIDSM_dHN_multiArea_realData_Lierne_alt.r
                     minYear = minYear, 
                     max_years = input_data$nim.constants$max_years, 
                     survAreaIdx = input_data$nim.constants$SurvAreaIdx, 
-                    plotPath = "Plots/Comp_Lierne_R1", 
+                    plotPath = "Plots/Comp_LierneModels", 
                     returnData = FALSE)
 
 

### 2024-09-01 R/plotSimCheckMetrics_replicates.R
Restricted plotting to primary annual parameters
N and esw are derivatives of D and p, respectively. Displaying the model's ability to predict them adds no additional info as the fit is identical to the one for the primary parameters.

@@ -352,6 +352,14 @@ plotSimCheckMetrics_replicates <- function(plotColors = "customRainbow", thin = 
                                                Parameter == "D" ~ "D[t]")) %>%
     dplyr::mutate(ParamName = factor(ParamName, levels = c("R[t]", "esw[t]", "p[t]", "N[t]", "D[t]")))
   
+  
+  ## Drop derived parameters for simples representation
+  out.annual <- out.annual %>%
+    dplyr::filter(Parameter %in% c("R", "D", "p"))
+  
+  metrics.annual <- metrics.annual %>%
+    dplyr::filter(Parameter %in% c("R", "D", "p"))
+  
   ## Calculate metric means
   metricMeans.annual <- metrics.annual %>%
     dplyr::group_by(ParamName) %>%
@@ -455,7 +463,7 @@ plotSimCheckMetrics_replicates <- function(plotColors = "customRainbow", thin = 
   p_comb2 <- plot_grid(p_a2, p_b, p_c, nrow = 1, rel_widths = c(1.33, 1, 1))
   
   ## Plot to pdf
-  pdf(paste0("Plots/SimCheck_replicates/SimCheckMetrics_annualParams_", plotColors, ".pdf"), width = 9, height = 14)
+  pdf(paste0("Plots/SimCheck_replicates/SimCheckMetrics_annualParams_", plotColors, ".pdf"), width = 9, height = 7)
   suppressWarnings(
     print(p_comb)
   )
@@ -465,13 +473,13 @@ plotSimCheckMetrics_replicates <- function(plotColors = "customRainbow", thin = 
   dev.off()
   
   ## Plot to png
-  png(paste0("Plots/SimCheck_replicates/SimCheckMetrics_annualParams_", plotColors, ".png"), width = 9, height = 14, units = "in", res = 300)
+  png(paste0("Plots/SimCheck_replicates/SimCheckMetrics_annualParams_", plotColors, ".png"), width = 9, height = 7, units = "in", res = 300)
   suppressWarnings(
     print(p_comb)
   )
   dev.off()
 
-  png(paste0("Plots/SimCheck_replicates/SimCheckMetrics_annualParams_median_", plotColors, ".png"), width = 9, height = 14, units = "in", res = 300)
+  png(paste0("Plots/SimCheck_replicates/SimCheckMetrics_annualParams_median_", plotColors, ".png"), width = 9, height = 7, units = "in", res = 300)
   suppressWarnings(
     print(p_comb2)
   )

### 2024-09-01 Manuscript_Nilsen_Nater.qmd
Updated figures (manuscript repo & OSF)

@@
 
-![Detection parameters estimated from 10 distinct sets of simulated data (= colors).
-Upper panel depicts model parameters, and lower panels show derived estimated of effective strip width and mean detection probability.
-Dots represents true values (black for global values, colored for dataset-specific values), and density plot represent the posterior distributions for each model run.](Figures/SimCheck_Detects_Temps.png){#fig-sim-detection}
+![Vital rate and detection parameter averages estimated from 10 distinct sets of simulated data (= colors) using three model fits each.
+First row depicts posterior densities from each model run relative to the true value used to simulate data (black dashed line).
+Second and third rows visualize the distributions of Bayesian p-values (proportion of samples \> true value) and root-mean square deviations (RMSD) for all model runs, respectively.
+Purple lines and numbers mark the mean values across all model runs; dashed black line (second row only) marks the ideal Bayesian p-value of 0.5.](Figures/SimCheckMetrics_singleParams_Temps.png){#fig-sim-detection}
 
-![Demographic rate parameters estimated from 10 distinct sets of simulated data (= colors).
-Upper panel show annual and seasonal survival probabilities, hyperparameters for the random effects model for recruitment (mean and sd), and lower panel show annual estimates of the recruitment rate.
-Dots represent true values (black for global values, colored for dataset-specific values), and density curves represent the posterior distributions from each model run.](Figures/SimCheck_VRs_Temps.png){#fig-sim-demo}
+![Annual population density ($D_t$), detection probability ($p_t$), and recruitment rate ($R_t$) estimated from 10 distinct sets of simulated data (= colors) using three model fits each.
+First row depicts the relationship between posterior medians from each model run and the true value used to simulate data (purple solid line = relationship estimated from linear model with a = intercept and b = slope; black dashed line = perfect correlation).
+Second and third rows visualize the distributions of Bayesian p-values (proportion of samples \> true value) and root-mean square deviations (RMSD) for all model runs, respectively.
+Purple lines and numbers mark the mean values across all model runs; dashed black line (second row only) marks the ideal Bayesian p-value of 0.5.
+And equivalent figure showing posterior samples instead of posterior means is available in the supplementary material on OSF.](Figures/SimCheckMetrics_annualParams_median_Temps.png){#fig-sim-demo}
 

### 2024-09-01 Manuscript_Nilsen_Nater.qmd
Text edits

@@
 
-*Estimation of abundance and demographic rates for populations of wild species is a challenging but fundamental issue for both management and research into ecology and evolution.
-One set of approaches that has been used extensively to estimate abundance of wildlife populations is Distance Sampling (DS) methods for line or point transect survey data.
-The first versions of DS models were only available as closed population models, and did not allow for the estimation of changes in abundance through time.
+*Estimation of abundance and demographic rates for populations of wild species is a challenging but fundamental issue for both ecological research and wildlife management.
+One set of approaches that has been used extensively to estimate abundance of wildlife populations is Distance Sampling (DS) for line or point transect survey data.
+The first implementations of DS models were only available as closed population models, and did not allow for the estimation of changes in abundance through time.
 The advent of open population formulations based on the DS framework greatly extended the scope of the models, but much untapped potential remains in models that estimate temporal dynamics not only in abundance but also in the underlying demographic rates.
@@
 DS has been used for estimating animal abundance in a wide range of contexts and for a variety of taxa [@buckland2015].
-One reason for the method's popularity is that it does not require repeated visits to the same sites for estimating detection probability.
+One reason for the method's popularity is that it requires neither marking of individuals nor repeated visits to the same sites for estimating detection probability.
 This makes DS particularly useful for implementation in participatory monitoring programs, allowing stakeholders to take part in the data collection process.
 
-The most common implementations of DS models have long used closed-population formulations by only estimating population density or abundance for one point in time, and not including an explicit formulation of the process model that projects abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates @buckland2015.
+Classical implementations of DS models have used closed-population formulations, i.e.
+models that treat estimates of population density or abundance at different time points as independent and do not including an explicit formulation of the process model that links abundance across years based on estimates of population growth rate ($\lambda$) or underlying demographic rates [@buckland2015].
 In recent years, DS approaches have been extended in many ways, including applications that estimate changes in abundance over time in open populations via a hidden state model representing population dynamics [@moore2011; @sollmann2015; @schaub2021].
@@
 
-In this study, we present a new IPM (IDSM, integrated distance sampling model) which integrates data from line transect distance sampling survey data and survival data from marked birds.
-The main motivation behind the study is to showcase how distance sampling models could be used to not only estimate population density but also demographic rates in an IPM setting.
-We believe the model would serve as a good building block for later applications that would focus more on ecological dynamics.
+In this study, we present a new integrated distance sampling model (IDSM) which integrates data from line transect distance sampling survey data and survival data from marked animals.
 The model's core is a stage-structured matrix population model that projects population size from one time step to the next based on underlying survival and recruitment rates.
-Below, we first present the model and assess its robustness and performance through application to simulated data.
-We then apply the model to real data collected from a willow ptarmigan (*Lagopus lagopus*) study in Central Norway.
-Because demographic rates are often affected by environmental covariates (e.g.
-rodent abundance in the case of willow ptarmigans), we also illustrate how such covariates can be included in the modelling framework.
+We first present the model and assess its robustness and performance through applications to simulated data.
+By doing so, we showcase how distance sampling models can be used to not only estimate population density but also demographic rates in an IPM setting.
+Finally, we proceed to highlight the potential of this new modelling framework by applying it to a case study involving data collected on willow ptarmigan (*Lagopus lagopus*) in Central Norway.
 
@@
 
-### 2.1 An integrated distance sampling model
+### 2.1 Integrated distance sampling model
 
@@
 This structure commonly used for populations of passerine and game birds [@williams2002; @schaub2021].
-In the context of our willow ptarmigan case study (see below), the population is updated in late summer - which is when the annual distance-sampling survey takes place.
+In the context of our willow ptarmigan case study (see below), the population census is set in late summer - which is when the annual distance-sampling survey takes place.
 At this time, the juvenile class is about 1 - 2 months old.
@@
 Both juveniles and adults survive from year $t$ census to year $t+1$ census with survival probability $S_t$.
-As ptarmigan can reproduce already as 1-year old, all survivors may then produce offspring in late June which recruit into the population as juveniles just prior to the census in year $t+1$ according to a recruitment rate $R_{t+1}$.
-The changes in densities (numbers) of juveniles and adults in the population, $D_{juv}$ and $D_{ad}$, can thus be expressed as
+We assume that individuals can reproduce already as 1-year olds, meaning all survivors may produce offspring in late June which recruit into the population as juveniles just prior to the census in year $t+1$ according to a recruitment rate $R_{t+1}$.
+The changes in densities of juveniles and adults in the population, $D_{juv}$ and $D_{ad}$, can thus be expressed as
 
@@
 $$ where $\sigma$ is the half-normal detection parameter.
-We assumed $\sigma$ to vary among years (index $t$) but not between transect lines or animal group size.
+We assumed $\sigma$ to vary among years (index $t$) but not between transect lines or animal group sizes.
 Following @moore2011, the resulting $\sigma_t$ can be used to calculate effective strip width ($esw_{t}$) and, consequently, average detection probability per line transect with a truncation distance $W$ according to:
@@
 \end{equation}
-$$ Here, $Sk_t$ is the seasonal survival probability, and annual survival probability, $S_t$ is calculated as $S1_t * S2_t$.
-The motivation behind separating out the different seasons come from our ptarmigan case study (see below), but in cases where it is not relevant to distinguish between different seasons, $S_t$ could be estimated directly using the formulation above.
+$$ Here, $Sk_t$ is the survival probability over the relevant time interval $k$ in year $t$.
+The length and definition of $k$ will be specific to any given study.
+For the remainder of this article, we define $k$ as 6-month seasons to be consistent with our ptarmigan case study.
+Consequently, the annual survival probability, $S_t$ that appears in the population model above is calculated as the product of two seasonal survival probabilities, $S1_t$ and $S2_t$.
 
@@
 For both the tests with simulated data and the case study described below, we implemented log-normally distributed random year effects on recruitment rate and detection probability.
-Survival, on the other hand, was set to be constant.
-This was done because previous research has shown relatively modest interannual variation in survival [@Israelsen2020], but also to ensure that all parameters were estimable.
 In the case study, we additionally included a covariate effect (see details below) on log recruitment rates, resulting in the following model:
@@
 $\beta$ the slope of the covariate effect, and $\epsilon_t$ the normally distributed random effects.
+
+In both simulations and the case study, we treated survival as time-invariant.
+This was motivated by our case study: previous research has relatively low interannual variation in survival of our focal ptarmigan population [@Israelsen2020] and the  telemetry data used in this study has limited potential for accurately estimating time-variation as it is relatively sparse.
+We note, however, that also survival could be modelled as time-dependent if sufficient data is available.
 
@@
 Model implementation for simulated data tests was largely identical to that for real data and is described in detail below.
+For assessing model performance, fit, and bias, we 1) compared model estimates to the true values of parameters used for data simulations visually, 2) correlated estimated and true values, and 3) calculated two metrics to measure bias: the proportion of samples above the true value (corresponding to Bayesian p-values) and the root-mean square deviation (RMSD).
 
@@
 
-The willow ptarmigan has a circumpolar distribution (Fuglei et al., 2020), and lives year-round in heterogeneous alpine and artic ecosystems.
-In Norway, there has been a long-term decline in the willow ptarmigan abundance across more than a century [@Hjeljord2022], but in the last few decades abundance trends has fluctuated both in time and space.
-In Scandinavia, willow ptarmigan is a valued game species (see e.g.
+To demonstrate the applicability of the IDSM to real data, we applied it to a case study of willow ptarmigan, a small grouse species with a has a circumpolar distribution (Fuglei et al., 2020).
+In Norway, there has been a long-term decline in willow ptarmigan abundance across more than a century [@Hjeljord2022], but in the last few decades abundance trends have fluctuated substantially both across time and space.
+Willow ptarmigan is a valued game species (see e.g.
 @Andersen2014 ), and there have been several long-term research projects devoted to understanding how they respond to environmental variation and harvest management [@Israelsen2020; @Sandercock2011].
@@
 As the surveys are conducted in early August, juveniles can be distinguished from adults by their smaller body size.
-Males and females are manly distinguished by the sound (males often make a characteristic sound when being flushed).
-Nevertheless, mis-identification can occur, and in addition a proportion of the birds are registered as "unknown" age and/or sex.
-In the model version presented here, we have made the assumption that these birds were in fact juveniles.
+Males and females are mainly distinguished by sound (males often make a characteristic sound when being flushed).
+Observers are trained to distinguish age classes and sexes, but incomplete identification can occur.
+In this application we assumed that the resulting "unknown" age and/or sex class birds were in fact juveniles (see discussion for further considerations).
 Besides bird observations, field workers also record whether (1) or not (0) they encounter small rodents on any transect line, allowing the proportion of transect lines with small rodent detection to be used as measure of rodent occupancy (covariate ranging from 0 to 1).
-Since 2019, data has been collected through a mobile app tailormade for this project, which is available through App Store (for iOS phones) and Google Play (for Android phones).
-Before 2019, field workers reported their data through a dedicated web portal.
-After data are collected and reported, they undergo several steps of quality control: first by local contacts and subsequently by personnel at the Norwegian Institute for Nature Research (NINA).
-The data are then standardized based on the Darwin-Core standard [@Wieczorek2012], and made publicly available as a sampling-event data set published through GBIF [@Nilsen2023a].
+After data are collected they undergo quality control, get standardized based on the Darwin-Core standard [@Wieczorek2012], and made publicly available as a sampling-event data set published through GBIF [@Nilsen2023a].
 For additional description of the data collection procedures, see [@Bowler2020; @Kvasnes2018; @Nilsen2023a].
 
-The individual longitudinal study based on radio collared willow ptarmigan was conducted in 2015-2021.
+The radio-telemetry data is the result of an individual-based longitudinal study over the period 2015-2021.
 Each winter (in February-March), willow ptarmigan were located at night using snowmobiles and large hand nets with prolonged handles, as described in [@Israelsen2020].
-To prevent birds from flying off before the field personnel were close enough to capture them, a high-powered head lamp was used to dazzle the birds.
-After capture, birds were placed in an opaque bag to reduce stress.
-They were then fitted with a uniquely numbered leg ring (\~ 2.4g) and a Holohil RI-2BM or Holohil RI-2DM radio transmitter (\~ 14.1g) and subsequently released.
+High-powered head lamps were used to dazzle the birds and allow capture.
+Captured birds were fitted with a uniquely numbered leg ring (\~ 2.4g) and a Holohil RI-2BM or Holohil RI-2DM radio transmitter (\~ 14.1g) and subsequently released.
 The radio transmitters had an expected battery lifetime of 24 months (RI-2BM) or 30 months (RI-2DM), and included a mortality circuit that was activated if a bird had been immobile for 12 hours.
 We monitored the birds throughout most of the year by triangulation from the ground at least once a month for 10 months of the year (February -- November) by qualified field personnel.
-If a mortality signal was emitted from a transmitter, we attempted to recover it as soon as possible to determine cause of death.
 A number of birds dispersed out of the main study areas and was thus out of signal range for field personnel on the ground.
@@
 
-For the model fits to simulated and real data we ran 3 and 4 MCMC chains with NIMBLE's standard samples for 500k and 100k iterations, respectively.
-300k and 40k and thereof were discarded as burn-in prior to thinning with factors 5 and 20 , leaving us with 40k and 3k posterior samples per chain (total of 120k and 12k samples per run), respectively.
+For the model fits to simulated and real data we ran 3 (simulated) or 4 (real) MCMC chains with NIMBLE's standard samples for 100k iterations.
+40k thereof were discarded as burn-in prior to thinning with factor 20, leaving us with 3k posterior samples per chain (total of 12k samples per run).
 MCMC parameters were chosen to yield a representative number of samples from converged chains, and convergence was determined based on visual inspection of trace plots.

### 2024-09-02 Manuscript_Nilsen_Nater.qmd
Updated results with new sim check metrics

@@
 
-Posterior distributions for parameters estimated in three model fits to each of 10 simulated data sets are shown in @fig-sim-detection and @fig-sim-demo.
-Overall, the IDSM was able to correctly estimate both detection parameters and demographic rate parameters from all 10 simulated datasets without any systematic bias.
+```{r}
+#| echo: FALSE
+#| message: FALSE
+#| warning: FALSE
+
+simCheck <- readRDS("SimCheck_metrics/SimChecks_PerformanceMetrics.rds")
+
+```
+
+Models fit to simulated datasets reached MCMC convergence within the given number of iterations.
+Chain mixing was good for all parameters except average recruitment rate ($\mu_R$); for this parameter, an elevated degree of autocorrelation was visible in the MCMC chains in some of the replicate runs, but models still produced posterior distributions that well represented the true value used in simulations (@gig-sim-singles).
+
+Posterior estimates relative to true values, Bayesian p-values, and RMSD for parameters estimated in three model fits to each of 10 simulated data sets are shown in @fig-sim-singles and @fig-sim-annuals.
+Overall, the IDSM was able to correctly estimate the majority of parameters from all 10 simulated datasets without substantial systematic bias.
 The replicate runs for each dataset resulted in very similar posterior distributions, demonstrating that the models converged towards the same posterior distributions irrespective of starting values.
-Estimates of population sizes / densities were also adequate, and are presented in the supplementary materials (available at @Nilsen2024 in folder SimCheck_byDataSet).
+This may not seem to be the case for survival parameters (@fig-sim-singles), but this is largely due to the relatively low number of individuals in the simulated telemetry data; estimated posteriors match up well with simulated numbers of survivors in each datasets, and the averages of Bayesian p-values fell very close to 0.5 (= no bias).
+For time-variation in recruitment rate ($\sigma_R$), on the other hand, the average Bayesian p-value indicated a potential for overestimation (p = `r subset(simCheck$metricMeans.single, ParamName == "sigma[R]")$Bayes_p_mean`), and this is consistent with the relatively large spread of Bayesian p-values for year-specific recruitment rates (@fig-sim-annuals).
+Across all years, the correlation between predicted and true recruitment rates was very high (slope = `r subset(simCheck$metricMeans.annual, ParamName == "R[t])$Slope`), yet closer inspection showed that slight over- and under-estimation was present for certain years across all replicates (see supplementary figures in folder SimCheck_byDataSet at @Nilsen2024).
+This was also the case for year-specific estimates of population density and detection probability, but both were slightly more likely to be underestimated than overestimated (@fig-sim-annuals).
 
@@
 Second and third rows visualize the distributions of Bayesian p-values (proportion of samples \> true value) and root-mean square deviations (RMSD) for all model runs, respectively.
-Purple lines and numbers mark the mean values across all model runs; dashed black line (second row only) marks the ideal Bayesian p-value of 0.5.](Figures/SimCheckMetrics_singleParams_Temps.png){#fig-sim-detection}
+Purple lines and numbers mark the mean values across all model runs; dashed black line (second row only) marks the ideal Bayesian p-value of 0.5.](Figures/SimCheckMetrics_singleParams_Temps.png){#fig-sim-singles}
 
@@
 Purple lines and numbers mark the mean values across all model runs; dashed black line (second row only) marks the ideal Bayesian p-value of 0.5.
-And equivalent figure showing posterior samples instead of posterior means is available in the supplementary material on OSF.](Figures/SimCheckMetrics_annualParams_median_Temps.png){#fig-sim-demo}
+And equivalent figure showing posterior samples instead of posterior means is available in the supplementary material on OSF.](Figures/SimCheckMetrics_annualParams_median_Temps.png){#fig-sim-annuals}
 

### 2024-09-02 Manuscript_Nilsen_Nater.qmd
Text edits

@@
 Models fit to simulated datasets reached MCMC convergence within the given number of iterations.
-Chain mixing was good for all parameters except average recruitment rate ($\mu_R$); for this parameter, an elevated degree of autocorrelation was visible in the MCMC chains in some of the replicate runs, but models still produced posterior distributions that well represented the true value used in simulations (@gig-sim-singles).
+Chain mixing was good for all parameters except average recruitment rate ($\mu_R$); for this parameter, an elevated degree of autocorrelation was visible in the MCMC chains in some of the replicate runs, but models still produced posterior distributions that well represented the true value used in simulations (@fig-sim-singles).
 
@@
 For time-variation in recruitment rate ($\sigma_R$), on the other hand, the average Bayesian p-value indicated a potential for overestimation (p = `r subset(simCheck$metricMeans.single, ParamName == "sigma[R]")$Bayes_p_mean`), and this is consistent with the relatively large spread of Bayesian p-values for year-specific recruitment rates (@fig-sim-annuals).
-Across all years, the correlation between predicted and true recruitment rates was very high (slope = `r subset(simCheck$metricMeans.annual, ParamName == "R[t])$Slope`), yet closer inspection showed that slight over- and under-estimation was present for certain years across all replicates (see supplementary figures in folder SimCheck_byDataSet at @Nilsen2024).
+Across all years, the correlation between predicted and true recruitment rates was very high (slope = `r subset(simCheck$metricMeans.annual, ParamName == "R[t]")$Slope`), yet closer inspection showed that slight over- and under-estimation was present for certain years across all replicates (see supplementary figures in folder SimCheck_byDataSet in @Nilsen2024).
 This was also the case for year-specific estimates of population density and detection probability, but both were slightly more likely to be underestimated than overestimated (@fig-sim-annuals).
@@
 Second and third rows visualize the distributions of Bayesian p-values (proportion of samples \> true value) and root-mean square deviations (RMSD) for all model runs, respectively.
-Purple lines and numbers mark the mean values across all model runs; dashed black line (second row only) marks the ideal Bayesian p-value of 0.5.](Figures/SimCheckMetrics_singleParams_Temps.png){#fig-sim-singles}
+Purple lines and numbers mark the mean values across all model runs; dashed black line (second row only) marks the ideal Bayesian p-value of 0.5.](Figures/SimCheckMetrics_singleParams_Temps.png){#fig-sim-singles height=80%}
 
@@
 Purple lines and numbers mark the mean values across all model runs; dashed black line (second row only) marks the ideal Bayesian p-value of 0.5.
-And equivalent figure showing posterior samples instead of posterior means is available in the supplementary material on OSF.](Figures/SimCheckMetrics_annualParams_median_Temps.png){#fig-sim-annuals}
+And equivalent figure showing posterior samples instead of posterior means is available in the supplementary material on OSF.](Figures/SimCheckMetrics_annualParams_median_Temps.png){#fig-sim-annuals height=80%}
 
@@
 Having evaluated the overall performance of our model on simulated data, we used data from our case study in Lierne to estimate abundance, vital rates and detection probabilities from a real-world data set.
-Ptarmigan population density increased markedly across the study period, from \< 10 ptarmigan / $km^{2}$ in 2007 to \> 35 ptarmigan / $km^{2}$ in in 2021 (@fig-lierne-density).
+Like the model fits to simulated data, convergence was reached within the fiven amount of iterations and mixing was good, albeit with somewhat higher chain autocorrelation for the intercept in the recruitment model.
+Ptarmigan population density was estimated with a marked increase across the study period, from \< 10 ptarmigan / $km^{2}$ in 2007 to \> 35 ptarmigan / $km^{2}$ in in 2021 (@fig-lierne-density).
 The increase was most distinct from 2016 and onward.
@@
 
-In general, the model did a very good job at recreating the underlying parameters when fitted to simulated data where the true values are known.
-The simulated data sets were based on relatively wide ranges of parameter values, yet we did not find the model to have any particular problems or biases recovering the original parameters used in simulations.
-When the input data are unbiased (with respect to the underlying model formulation) we therefore consider the model to be able to extract unbiased and meaningful demographic parameters from age-structured distance sampling data.
+In general, the model did a good job at recreating the underlying parameters when fitted to simulated data with known underlying true parameter values.
+The simulated data sets were based relatively wide ranges of parameter values, yet model posteriors included the true values in almost all cases.
+While there seems to be potential for overestimating time variation in recruitment rates, this bias did not propagate into estimates of year-specific recruitment rates (@fig-sim-annuals), and may be related to the subpar mixing of the intercept of the recruitment model.
+The model was more likely to under- than overestimate detection probabilities and population densities, but bias in these estimates was generally small and spread out across time-series, i.e.
+bias did not seem to arise disproportionately at e.g.
+the starts or ends of time-series.
+Simulated data tests also revealed that telemetry data simulated with an average of 30 individuals may be too sparse to obtain robust and generalisable estimates of seasonal survival probabilites.
+This could be investigated further by repeating simulated data runs with different average numbers of individuals in the telemetry data, but we chose not to go down this path here as we were primarily interested in model performance given the amounts of data that are currently available for our ptarmigan case study.
+Based on the simulated data tests presented here, we conclude that the IDSM is able to provide meaningful and sufficiently accurate and robust estimates of demographic parameters from age-structured distance sampling data, provided that the input data are unbiased (with respect to the underlying model formulation).
 
@@
 In the model version presented in the results section, we made the assumption that these birds were in fact juveniles.
-Comparing estimates to an alternative scenario in which we discarded all birds classified as "unknown sex and age" (see supplementary figures in @Nilsen2024 ) we found that -- as expected -- estimated population density was virtually unaffected by the treatment of "unknown sex and age" observations, while the annual demographic parameters shifted proportional to the amount of "unknown sex and age" observations in the given year (towards higher recruitment and lower survival).
+Comparing estimates to an alternative scenario in which we discarded all birds classified as "unknown sex and age" (see supplementary figures in @Nilsen2024) we found that -- as expected -- estimated population density was virtually unaffected by the treatment of "unknown sex and age" observations, while the annual demographic parameters shifted proportional to the amount of "unknown sex and age" observations in the given year (towards higher recruitment and lower survival).
 Thus, biases in the reported age ratios may affect estimated of demographic rates, but not so much population density.
@@
 On a different note, we can also not exclude the possibility of a small degree of bias in survival estimates due to misclassification of age in the data (see above), especially seeing as the IDSM's recruitment rate estimates also appear somewhat high compared to other studies on willow ptarmigan [@Eriksen2023; @Kvasnes2014a; @Steen1988].
-Note however that the last period of the study period in Lierne, there was substantial increase in the density (ref Figure 2), which might indicate that the recruitment and/or survival rates were high in this period.
-
-Compared to previous studies on ptarmigan (see e.g.
-[@Israelsen2020; @Sandercock2011]), we could have expected somewhat higher estimates of survival probability.
-One potential reason for the overall lower estimates obtained here is that our IDSM analysis assumed constant survival over the period 2007-2021 while survival, in reality, must have varied over time.
-If survival in more recent years, when telemetry data was collected [and the study of e.g.
-\@ @Israelsen2020 was carried out] was higher than in earlier years -- something that seems likely given population increase over recent years -- an average over the entire time period is expected to be lower.
-On a different note, we can also not exclude the possibility of a small degree of bias in survival estimates due to misclassification of age in the data (see above), especially seeing as the IDSM's recruitment rate estimates also appear somewhat high compared to other studies on ptarmigan (+refs).
-<!-- If we have any "biological" thoughts about why somewhat high recruitment rates may not be that ridiculous here, then that could be added here -->
+As the ptarmigan case study served as an illustrative example for the IDSM framework in this article, we did not further investigate alternative models, such as implementations with time-varying survival or different treatment of uncertainty in age and/or sex.
+The model lends itself easily to such extensions, however, and methods such as posterior predictive checks and WAIC will be useful for assessing and optimizing the fit of the model to data from relevant case studies [@hooten2015; @conn2018].
 
@@
 Previous research has shown that ptarmigan recruitment is also sensitive to the weather in the late winter and spring (before and during the breeding season); as we did not fit any weather covariates to the model, there is a possibility that effects of spring conditions in certain years may have masked any remaining effects of small rodent abundance.
-Finally, the data set used in this analysis is relatively short (15 years), leaving us with somewhat limited statistical power to detect effects of temporal covariates .
+Finally, the data set used in this analysis is relatively short (15 years), leaving us with somewhat limited statistical power to detect effects of temporal covariates.
 Taken together we therefore do not consider this study as a particularly strong test of the underlying effect of small rodent fluctuations on ptarmigan recruitment rates.
@@
 We also thank James A.
-Martin for constructive discussions during the later stages of this work.
+Martin for constructive discussions during the later stages of this work and Simon Hansul, Tiago Marquez, and Sandrine Charles for constructive feedback on the manuscript during the review process.
 
@@
 
-The auxiliary radio-telemetry data and rodent occupancy data, and all code used for wrangling, analysing, and visualizing data and results can be found in the project's repository on GitHub: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.
-The results presented in this paper were created using version 1.3 of the code [@IDSMCode_2024].
+The auxiliary radio-telemetry data and rodent occupancy data, all code used for wrangling, analysing, and visualizing data and results, as well as vignettes providing rich text documentation of all steps in the workflows for simulated and real data analysis, can be found in the project's repository on GitHub: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.
+The results presented in this paper were created using version 1.4 of the code [@IDSMCode_2024].
 

### 2024-09-02 Analysis_RealData_LierneVest.R
Re-ran comparison with unknown unknowns removed

@@ -158,15 +158,16 @@ IDSM.out <- nimbleMCMC(code = model_setup$modelCode,
 Sys.time() - t.start
 
 saveRDS(IDSM.out, file = 'rypeIDSM_dHN_multiArea_realData_Lierne.rds')
+#saveRDS(IDSM.out, file = 'rypeIDSM_dHN_multiArea_realData_Lierne_ukukRremoved.rds')
 
 
 # COMPARE MODELS #
 #----------------#
 
 plotModelComparison(modelPaths = c("rypeIDSM_dHN_multiArea_realData_Lierne.rds",
-                                   "rypeIDSM_dHN_multiArea_realData_Lierne_alt.rds"), 
+                                   "rypeIDSM_dHN_multiArea_realData_Lierne_ukukRemoved.rds"), 
                     modelChars = c("Main model",
-                                   "Alternative model"), 
+                                   "Unknown removed"), 
                     N_areas = input_data$nim.constants$N_areas, 
                     area_names = areas, 
                     N_sites = input_data$nim.constants$N_sites, 
@@ -174,7 +175,7 @@ plotModelComparison(modelPaths = c("rypeIDSM_dHN_multiArea_realData_Lierne.rds",
                     minYear = minYear, 
                     max_years = input_data$nim.constants$max_years, 
                     survAreaIdx = input_data$nim.constants$SurvAreaIdx, 
-                    plotPath = "Plots/Comp_LierneModels", 
+                    plotPath = "Plots/Comp_ukukRemoved", 
                     returnData = FALSE)
 
 

### 2024-09-02 R/prepareInputData.R

@@ -185,6 +185,7 @@ prepareInputData <- function(d_trans, d_obs, d_cmr, d_rodent, localities = NULL,
     TaksObs <- d_obs_sub %>% 
       filter(between(DistanceToTransectLine, -0.1, W)) %>% # 
       dplyr::mutate(cs = unknownJuvenile+unknownunknown+FemaleAdult+MaleAdult) %>%
+      #dplyr::mutate(cs = unknownJuvenile+FemaleAdult+MaleAdult) %>% # Drop unknown unknowns
       reshape2::dcast(locationID~Year, value.var = "cs", sum) %>%
       dplyr::right_join(., temp, by = c("locationID" = "locationID")) %>%
       replace(., is.na(.), 0) %>%
@@ -209,6 +210,7 @@ prepareInputData <- function(d_trans, d_obs, d_cmr, d_rodent, localities = NULL,
     TaksObs_J <- d_obs_sub %>% 
       filter(between(DistanceToTransectLine, -0.1, W)) %>%
       dplyr::mutate(cs = unknownJuvenile + unknownunknown) %>%
+      #dplyr::mutate(cs = unknownJuvenile) %>% # Drop unknown unknowns
       reshape2::dcast(locationID~Year, value.var="cs", sum) %>%
       dplyr::right_join(., temp, by=c("locationID"="locationID")) %>%
       replace(., is.na(.), 0) %>%
@@ -261,6 +263,7 @@ prepareInputData <- function(d_trans, d_obs, d_cmr, d_rodent, localities = NULL,
     # Reformat data
     temp_Rec <- d_obs_sub %>% filter(between(DistanceToTransectLine, -0.1, W)) %>%
       dplyr::mutate(sumR = unknownJuvenile + unknownunknown,
+      #dplyr::mutate(sumR = unknownJuvenile, # Drop unknown unknowns
                     sumAd = MaleAdult + FemaleAdult,
                     sumAdF = FemaleAdult) %>%
       dplyr::mutate(Year2 = Year - (min(Year)) + 1)

### 2024-09-02 Analysis_RealData_LierneVest.R
Small fix in model comparison call

@@ -169,7 +169,7 @@ plotModelComparison(modelPaths = c("rypeIDSM_dHN_multiArea_realData_Lierne.rds",
                     modelChars = c("Main model",
                                    "Unknown removed"), 
                     N_areas = input_data$nim.constants$N_areas, 
-                    area_names = areas, 
+                    area_names = localities, 
                     N_sites = input_data$nim.constants$N_sites, 
                     N_years = input_data$nim.constants$N_years, 
                     minYear = minYear, 

### 2024-09-02 Manuscript_Nilsen_Nater.qmd
Updated remaining figures and rendered

@@
 To avoid loss of data, we conducted aerial triangulation using a helicopter or airplane three times a year (May, September and November) in the years 2016-2020.
+In the analysis here, we assume that the telemetry data is representative for the entire duration of study period (2007-2021), despite its collection only starting in 2015.
 
@@
 Second and third rows visualize the distributions of Bayesian p-values (proportion of samples \> true value) and root-mean square deviations (RMSD) for all model runs, respectively.
-Purple lines and numbers mark the mean values across all model runs; dashed black line (second row only) marks the ideal Bayesian p-value of 0.5.](Figures/SimCheckMetrics_singleParams_Temps.png){#fig-sim-singles height=80%}
+Purple lines and numbers mark the mean values across all model runs; dashed black line (second row only) marks the ideal Bayesian p-value of 0.5.](Figures/SimCheckMetrics_singleParams_Temps.png){#fig-sim-singles}
 
@@
 Purple lines and numbers mark the mean values across all model runs; dashed black line (second row only) marks the ideal Bayesian p-value of 0.5.
-And equivalent figure showing posterior samples instead of posterior means is available in the supplementary material on OSF.](Figures/SimCheckMetrics_annualParams_median_Temps.png){#fig-sim-annuals height=80%}
+And equivalent figure showing posterior samples instead of posterior means is available in the supplementary material on OSF.](Figures/SimCheckMetrics_annualParams_median_Temps.png){#fig-sim-annuals}
 
@@
 The auxiliary radio-telemetry data and rodent occupancy data, all code used for wrangling, analysing, and visualizing data and results, as well as vignettes providing rich text documentation of all steps in the workflows for simulated and real data analysis, can be found in the project's repository on GitHub: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.
-The results presented in this paper were created using version 1.4 of the code [@IDSMCode_2024].
+The results presented in this paper were created using version 1.5 of the code [@IDSMCode_2024].
 

### 2024-09-06 Analysis_RealData.R
Removed storing of toggles from main script

@@ -56,17 +56,6 @@ testRun <- TRUE
 # Run MCMC in parallel
 parallelMCMC <- FALSE
 
-# Store as .RDS
-storeToggles(downloadData = downloadData,
-             R_perF = R_perF,
-             R_parent_drop0 = R_parent_drop0,
-             sumR.Level = sumR.Level,
-             survVarT = survVarT,
-             fitRodentCov = fitRodentCov,
-             telemetryData = telemetryData,
-             testRun = testRun,
-             parallelMCMC = parallelMCMC)
-
 
 # DOWNLOAD/FETCH DATA #
 #---------------------#

### 2024-09-06 Analysis_RealData.R
Implemented definition of areaAggregation at start of script

@@ -87,7 +87,7 @@ LT_data <- wrangleData_LineTrans(DwC_archive_list = Rype_arkiv,
                                  duplTransects = duplTransects,
                                  #localities = localities,
                                  areas = areas,
-                                 areaAggregation = TRUE,
+                                 areaAggregation = areaAggregation,
                                  minYear = minYear, maxYear = maxYear)
 
 
@@ -105,7 +105,7 @@ d_cmr <- wrangleData_CMR(minYear = minYear)
 d_rodent <- wrangleData_Rodent(duplTransects = duplTransects,
                                #localities = localities,
                                areas = areas,
-                               areaAggregation = TRUE,
+                               areaAggregation = areaAggregation,
                                minYear = minYear, maxYear = maxYear)
 
 
@@ -119,7 +119,7 @@ input_data <- prepareInputData(d_trans = LT_data$d_trans,
                                d_rodent = d_rodent,
                                #localities = localities, 
                                areas = areas,
-                               areaAggregation = TRUE,
+                               areaAggregation = areaAggregation,
                                excl_neverObs = TRUE,
                                R_perF = R_perF,
                                R_parent_drop0 = R_parent_drop0,
@@ -149,10 +149,6 @@ model_setup <- setupModel(modelCode = modelCode,
                           testRun = testRun, 
                           nchains = nchains,
                           initVals.seed = MCMC.seeds)
-
-
-## Expand seed to get MCMC seeds
-MCMC.seeds <- expandSeed_MCMC(seed = mySeed, nchains = model_setup$mcmcParams$nchains)
 
 
 # MODEL (TEST) RUN #
@@ -322,7 +318,7 @@ plotVarDecomposition(mcmc.out = IDSM.out.tidy,
 ## Make map of Norwegian municipalities ("fylke")
 NorwayMunic.map <- setupMap_NorwayMunic(shp.path = "data/norway_municipalities/norway_municipalities.shp",
                                         d_trans = LT_data$d_trans,
-                                        areas = areas, areaAggregation = TRUE)
+                                        areas = areas, areaAggregation = areaAggregation)
 
 ## Plot population growth rate, density, and vital rates on map
 plotMaps(PostSum.list = PostSum.list, 

### 2024-09-06 Analysis_RealData.R
Added missing areaAggregation toggle

@@ -31,6 +31,9 @@ sourceDir('R')
 # downloadData <- FALSE
 downloadData <- TRUE
 
+# Aggregation to area level
+areaAggregation <- TRUE 
+
 # Recruitment per adult or per adult female
 R_perF <- FALSE
 

### 2024-09-06 Analysis_RealData_GNUparallel_PostProcessing.R
Added generation time calculations to GNU pipeline and walkthrough

@@ -177,21 +177,12 @@ plotLatitude(PostSum.list = PostSum.list,
              minYear = minYear, maxYear = maxYear,
              fitRodentCov = fitRodentCov)
 
+# OPTIONAL: GENERATION TIME #
+#---------------------------#
 
-# OPTIONAL: MODEL COMPARISON (PLOTS) #
-#------------------------------------#
-
-# plotModelComparison(modelPaths = c("rypeIDSM_dHN_multiArea_realData_allAreas_tidy.rds",
-#                                    "rypeIDSM_dHN_multiArea_realData_allAreas_tidy_noTelemetry.rds"), 
-#                     modelChars = c("Including telemetry",
-#                                    "Without telemetry"), 
-#                     N_areas = input_data$nim.constants$N_areas, 
-#                     area_names = areas, 
-#                     N_sites = input_data$nim.constants$N_sites, 
-#                     N_years = input_data$nim.constants$N_years, 
-#                     minYear = minYear, 
-#                     maxYear = maxYear, 
-#                     max_years = input_data$nim.constants$max_years, 
-#                     survAreaIdx = input_data$nim.constants$SurvAreaIdx, 
-#                     plotPath = "Plots/Comp_noTelemetry", 
-#                     returnData = FALSE)
+GT_estimates <- extract_GenTime(mcmc.out = IDSM.out.tidy, 
+                                N_areas = input_data$nim.constants$N_areas, 
+                                area_names = input_data$nim.constant$area_names, 
+                                area_coord = LT_data$d_coord,
+                                mapNM = NorwayMunic.map,
+                                save = TRUE)

### 2024-09-06 Analysis_RealData_GNUparallel_PostProcessing.R
Implemented saving/loading of LT_data for GNU pipeline
Elements from LT_data are needed for a handful of plotting function and this therefore needs to be available to read from file.

@@ -33,6 +33,7 @@ dir.create("Plots")
 
 ## Read in input data
 input_data <- readRDS("RypeData_forIM.rds")
+LT_data <- readRDS("LT_data.rds")
 
 ## Reinstate model toggles
 toggles <- readRDS("ModelRunToggles.rds")

### 2024-09-06 Analysis_RealData_GNUparallel_Setup.R

@@ -98,6 +98,8 @@ LT_data <- wrangleData_LineTrans(DwC_archive_list = Rype_arkiv,
                                  areaAggregation = TRUE,
                                  minYear = minYear, maxYear = maxYear)
 
+## Save line transect data for use in post-processing
+saveRDS(LT_data, file = "LT_data.rds")
 
 # WRANGLE KNOWN FATE CMR DATA #
 #-----------------------------#

### 2024-09-11 Analysis_RealData.R
Several small fixes to make entire workflow run as it should

@@ -370,18 +370,3 @@ plotModelComparison(modelPaths = c("rypeIDSM_dHN_multiArea_realData_allAreas_tid
                     survAreaIdx = input_data$nim.constants$SurvAreaIdx, 
                     plotPath = "Plots/Comp_noTelemetry", 
                     returnData = FALSE)
-
-plotModelComparison(modelPaths = c("rypeIDSM_dHN_multiArea_realData_allAreas_tidy.rds",
-                                   "rypeIDSM_dHN_multiArea_realData_allAreas_tidy_main.rds"), 
-                    modelChars = c("Using year 1 R",
-                                   "Original"), 
-                    N_areas = input_data$nim.constants$N_areas, 
-                    area_names = areas, 
-                    N_sites = input_data$nim.constants$N_sites, 
-                    N_years = input_data$nim.constants$N_years, 
-                    minYear = minYear, 
-                    #maxYear = maxYear, 
-                    max_years = input_data$nim.constants$max_years, 
-                    survAreaIdx = input_data$nim.constants$SurvAreaIdx, 
-                    plotPath = "Plots/Comp_year1R", 
-                    returnData = FALSE)

### 2024-09-11 Analysis_RealData_GNUparallel_PostProcessing.R

@@ -53,25 +53,26 @@ for(i in 1:nchains){
 # TIDY UP POSTERIOR SAMPLES #
 #---------------------------#
 
-IDSM.out.tidy <- tidySamples(IDSM.out = IDSM.out, 
+IDSM.out.tidy <- tidySamples(IDSM.out = IDSM.out,
                              save = TRUE,
-                             fileName = "rypeIDSM_dHN_multiArea_realData_allAreas_tidy_test.rds")
+                             fileName = "rypeIDSM_dHN_multiArea_realData_allAreas_tidy.rds")
 
-
+#IDSM.out.tidy <- readRDS("rypeIDSM_dHN_multiArea_realData_allAreas_tidy.rds")
 
 # MAKE POSTERIOR SUMMARIES PER AREA #
 #-----------------------------------#
 
-PostSum.list <- summarisePost_areas(mcmc.out = IDSM.out.tidy, 
-                                    N_areas = input_data$nim.constant$N_areas, 
-                                    area_names = input_data$nim.constant$area_names, 
-                                    N_sites = input_data$nim.constant$N_sites, 
-                                    min_years = input_data$nim.constant$min_years, 
-                                    max_years = input_data$nim.constant$max_years, 
+PostSum.list <- summarisePost_areas(mcmc.out = IDSM.out.tidy,
+                                    N_areas = input_data$nim.constant$N_areas,
+                                    area_names = input_data$nim.constant$area_names,
+                                    N_sites = input_data$nim.constant$N_sites,
+                                    min_years = input_data$nim.constant$min_years,
+                                    max_years = input_data$nim.constant$max_years,
                                     minYear = minYear, maxYear = maxYear,
                                     fitRodentCov = fitRodentCov,
                                     save = TRUE)
 
+#PostSum.list <- readRDS("PosteriorSummaries_byArea.rds")
 
 # OPTIONAL: MCMC TRACE PLOTS #
 #----------------------------#
@@ -84,12 +85,12 @@ plotMCMCTraces(mcmc.out = IDSM.out.tidy,
 # OPTIONAL: TIME SERIES PLOTS #
 #-----------------------------#
 
-plotTimeSeries(mcmc.out = IDSM.out.tidy, 
-               N_areas = input_data$nim.constant$N_areas, 
-               area_names = input_data$nim.constant$area_names, 
-               N_sites = input_data$nim.constant$N_sites, 
[…]

### 2024-09-11 R/extract_GenTime.R

@@ -101,7 +101,7 @@ extract_GenTime <- function(mcmc.out,
     theme_classic()
     
   pdf("Plots/GenerationTime/GenerationTime_Latitude.pdf", width = 8, height = 4)
-  p_latGT
+  print(p_latGT)
   dev.off()
   
 
@@ -110,7 +110,7 @@ extract_GenTime <- function(mcmc.out,
     dplyr::left_join(., subset(GT_data_all, GT_measure == "Inverse of fecundity elasticity"), by = "Area") 
   
   mapNM.GT_R0 <- mapNM %>%
-    dplyr::left_join(., subset(GT_data_all, GT_measure == "Per generation growth rate (R0)"), by = "Area") 
+    dplyr::left_join(., subset(GT_data_all, GT_measure == "Per generation growth rate"), by = "Area") 
   
   pdf("Plots/GenerationTime/GenerationTime_elasF_Map.pdf", width = 5, height = 6)
   print(

### 2024-09-11 R/plotDetectFunction.R

@@ -28,7 +28,7 @@ plotDetectFunction <- function(mcmc.out,
   for(i in 1:N_areas){
     
     ## Extract posterior samples of relevant parameters
-    sigma_avg <- mcmc.mat[, paste0("mu.dd[", i, "]")]
+    sigma_avg <- exp(mcmc.mat[, paste0("mu.dd[", i, "]")])
 
     ## Make, summarise, and store predictions for each covariate value
     for(x in 1:100){

### 2024-09-11 R/plotLatitude.R

@@ -119,7 +119,6 @@ plotLatitude <- function(PostSum.list,
   p_popDens1 <- ggplot(subset(popDens.sum, SummaryPeriod == paste0(minYear, "-", maxYear))) + 
     geom_pointrange(aes(x = Latitude, y = Median, ymin = lCI, ymax = uCI, colour = Longitude), size = 0, fatten = 4, alpha = 0.5) +
     geom_point(aes(x = Latitude, y = Median, fill = Longitude), shape = 21, color = "black", size = 2, alpha = 0.75) +
-    ylim(0, 500) + 
     ylab("Estimate") + 
     paletteer::scale_color_paletteer_c("grDevices::Temps") + 
     paletteer::scale_fill_paletteer_c("grDevices::Temps") + 
@@ -169,7 +168,6 @@ plotLatitude <- function(PostSum.list,
     geom_point(aes(x = dens.Median, y = Median, fill = Latitude), shape = 21, color = "black", size = 2, alpha = 0.75) +
     paletteer::scale_color_paletteer_c("grDevices::Temps") + 
     paletteer::scale_fill_paletteer_c("grDevices::Temps") + 
-    xlim(0, 500) + 
     xlab("Population density") +
     ylab("Population growth rate") + 
     theme_classic()

### 2024-09-11 R/plotModelComparison.R

@@ -20,7 +20,7 @@
 
 plotModelComparison <- function(modelPaths, modelChars, 
                                 N_areas, area_names, N_sites, N_years, 
-                                minYear, maxYera, max_years, survAreaIdx, 
+                                minYear, maxYear, max_years, survAreaIdx, 
                                 plotPath, returnData = FALSE){
 
   if(length(modelPaths) != length(modelChars)){
@@ -224,8 +224,8 @@ plotModelComparison <- function(modelPaths, modelChars,
         pDetect.sum <- rbind(pDetect.sum, pDetect_add)
         
         # Summarize annual average population densities
-        popDens_juv <- out.mat[, paste0("Density[",  i, ", 1, ", area_yearIdxs[t], "]")]
-        popDens_ad <- out.mat[, paste0("Density[",  i, ", 2, ", area_yearIdxs[t], "]")]
+        popDens_juv <- out.mat[, paste0("meanDens[",  i, ", 1, ", area_yearIdxs[t], "]")]
+        popDens_ad <- out.mat[, paste0("meanDens[",  i, ", 2, ", area_yearIdxs[t], "]")]
 
         popDens_mean <- popDens_juv + popDens_ad
         

### 2024-09-11 R/plotTimeSeries.R

@@ -98,11 +98,7 @@ plotTimeSeries <- function(mcmc.out,
       popDens.sum <- rbind(popDens.sum, popDens_add)
       
       # Summarize annual population age structure (proportion juveniles)
-      if(N_sites[i] > 1){
-        ageStr_mean <- rowMeans(popDens_juv/(popDens_juv + popDens_ad))
-      }else{
-        ageStr_mean <- popDens_juv/(popDens_juv + popDens_ad)
-      }
+      ageStr_mean <- popDens_juv/(popDens_juv + popDens_ad)
       
       ageStr_add <- data.frame(Area = area_names[i],
                                  Year = area_years[t], 

### 2024-09-11 generate_default_nix.R

@@ -18,8 +18,10 @@ rix(
         "ggplot2", 
         "ggpubr", 
         "MCMCvis", 
-        "paletteer",  
+        "paletteer", 
+        "popbio",
         "qs", 
+        "Rage",
         "reshape2",
         "RJSONIO", # https://github.com/LivingNorway/LivingNorwayR/issues/75
         "sf", 

### 2024-09-12 Analysis_RealData.R
Added assessments of VR correlations to nixGNUparallel workflow and fixed a couple of small typos

@@ -299,12 +299,12 @@ checkDD(mcmc.out = IDSM.out.tidy,
 # OPTIONAL: CHECK VITAL RATE SAMPLING CORRELATIONS #
 #--------------------------------------------------#
 
-checkVRcorr(mcmc.out = IDSM.out.tidy, 
-            N_areas = input_data$nim.constant$N_areas, 
-            area_names = input_data$nim.constant$area_names, 
-            area_coord = LT_data$d_coord,
-            min_years = input_data$nim.constant$min_years, 
-            max_years = input_data$nim.constant$max_years)
+checkVRcorrs(mcmc.out = IDSM.out.tidy, 
+             N_areas = input_data$nim.constant$N_areas, 
+             area_names = input_data$nim.constant$area_names, 
+             area_coord = LT_data$d_coord,
+             min_years = input_data$nim.constant$min_years, 
+             max_years = input_data$nim.constant$max_years)
 
 
 # OPTIONAL: CALCULATE AND PLOT VARIANCE DECOMPOSITION #

### 2024-09-12 Analysis_RealData_GNUparallel_PostProcessing.R

@@ -146,6 +146,17 @@ checkDD(mcmc.out = IDSM.out.tidy,
         max_years = input_data$nim.constant$max_years)
 
 
+# OPTIONAL: CHECK VITAL RATE SAMPLING CORRELATIONS #
+#--------------------------------------------------#
+
+checkVRcorrs(mcmc.out = IDSM.out.tidy, 
+             N_areas = input_data$nim.constant$N_areas, 
+             area_names = input_data$nim.constant$area_names, 
+             area_coord = LT_data$d_coord,
+             min_years = input_data$nim.constant$min_years, 
+             max_years = input_data$nim.constant$max_years)
+
+
 # OPTIONAL: CALCULATE AND PLOT VARIANCE DECOMPOSITION #
 #-----------------------------------------------------#
 

### 2024-09-12 R/checkVRcorrs.R

@@ -94,10 +94,12 @@ checkVRcorrs <- function(mcmc.out,
   ifelse(!dir.exists("Plots/VitalRate_corr"), dir.create("Plots/VitalRate_corr"), FALSE)
   
   pdf("Plots/VitalRate_corr/SurvRepCorr_Latitude.pdf", width = 8, height = 4)
-  ggpubr::ggarrange(p_VRcorr1 + ggtitle("Survival[t] vs. Recruitment[t+1]") + theme(axis.title.x = element_blank()),
-                    p_VRcorr2 + ggtitle("Survival[t] vs. Recruitment[t]") + theme(axis.title.x = element_blank()), 
-                    heights = c(1, 1), ncol = 2, 
-                    legend = "right", common.legend = TRUE)
+  print(
+    ggpubr::ggarrange(p_VRcorr1 + ggtitle("Survival[t] vs. Recruitment[t+1]") + theme(axis.title.x = element_blank()),
+                      p_VRcorr2 + ggtitle("Survival[t] vs. Recruitment[t]") + theme(axis.title.x = element_blank()), 
+                      heights = c(1, 1), ncol = 2, 
+                      legend = "right", common.legend = TRUE)
+  )
   dev.off()
   
   ## Return filepaths

### 2024-09-12 _targets.R

@@ -228,12 +228,12 @@ list(
   
   tar_target(
     VRcorr.checks,
-    checkVRcorr(mcmc.out = IDSM.out.tidy, 
-                N_areas = input_data$nim.constant$N_areas, 
-                area_names = input_data$nim.constant$area_names, 
-                area_coord = LT_data$d_coord,
-                min_years = input_data$nim.constant$min_years, 
-                max_years = input_data$nim.constant$max_years)
+    checkVRcorrs(mcmc.out = IDSM.out.tidy, 
+                 N_areas = input_data$nim.constant$N_areas, 
+                 area_names = input_data$nim.constant$area_names, 
+                 area_coord = LT_data$d_coord,
+                 min_years = input_data$nim.constant$min_years, 
+                 max_years = input_data$nim.constant$max_years)
   ),
   
   tar_target(
